---
title: "Lecture 14 - Comparison of Multiple Linear Regression Models"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: breezedark
    #highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: format.css
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)

# Ensure the plot options are set correctly
opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

# Load the dataset
climate <- read.csv("climate.csv")

```


## Learning outcomes



-- investigate the importance of a group of covariates simultaneously.


-- Define nested models.


-- Understand and implement the F statistics to compare nested models.



## Background

You have by now considered two types of testing problems in linear regression:
    
1.  Testing whether the response is related to at least one explanatory variable;

2.  Testing the effect of a given (single) explanatory variable having adjusted for other variables.


::: {.revealjs-slide data-vertical-align="middle"}

## Basic model

$$
\\[2em]
\Huge{
Y = X\boldsymbol{\beta} + \boldsymbol{\varepsilon}
}
\\[2em]
\text{Capital letters denote matrices.}
$$

:::

## Basic model


$$
\begin{align}
Y = \quad (The \quad Y \quad values) \quad \begin{bmatrix}
Y^{(1)} \\
Y^{(2)} \\
\vdots \\
Y^{(n)} \\
\end{bmatrix}
\end{align}
$$

## Basic model

$$
\begin{align}
X \quad (model\quad matrix) &= \begin{bmatrix}
1 & X_{2}^{(1)} & X_{6}^{(1)} & X_{1}^{(1)} \\
1 & X_{2}^{(2)} & X_{6}^{(2)} & X_{1}^{(2)} \\
\vdots & \vdots & \vdots & \vdots \\
1 & X_{2}^{(n)} & X_{6}^{(n)} & X_{1}^{(n)} \\
\end{bmatrix},\\
\\
\end{align}
$$


## Basic model

$$
\begin{align}
\boldsymbol{\beta \quad }\quad (regression \quad parameters) = \begin{bmatrix}
\beta_0 \\
\beta_2 \\
\beta_6 \\
\beta_1 \\
\end{bmatrix} 
\end{align}
$$

## Basic model

$$
\begin{align}
\boldsymbol{\varepsilon} \quad (vector \quad of \quad errors) \quad = \begin{bmatrix}
\varepsilon^{(1)} \\
\varepsilon^{(2)} \\
\vdots \\
\varepsilon^{(n)} \\
\end{bmatrix}
\end{align}
$$


#

We want to the estimates of the intercept and slopes: $\beta_0, \beta_1, …,\beta_p$

to make the sum of squared errors as small as possible. (i.e.  to minimize $\varepsilon_1^2 +\varepsilon_2^2+ ...+\varepsilon_n^2$)

#

Matrix algebra gives an immediate solution  that is easily programmed.  
$$ \boldsymbol{b} =(X^T X)^{-1} X^T \boldsymbol{y}  $$                             
This holds for all sizes of problem, in terms of both the number of rows (observations) and columns (variables) in the `X` model matrix.

The columns of `X` must be linearly independent, or more accurately, not have any linear dependencies.

## Nested Models

A linear model `M1` is nested within a more complex model `M2` if you can get `M1` just by removing some predictors from `M2`. Put in another way, by setting some of `M2`'s coefficients to zero.

## Nested Models

For the Climate data, we define model `M2` by

$$
\begin{align}
\\
\\
E[\mbox{MnJlyTemp}] =\\ 
\beta_0 + 
\beta_1 \mbox{Lat} + 
\beta_2 \mbox{Height} + 
\beta_3 \mbox{Sea} + 
\beta_4 \mbox{NorthIsland} + \\ 
\beta_5 \mbox{Rain} + 
\beta_6 \mbox{Sun} + 
\beta_7 \mbox{Long} + 
\beta_8 \mbox{MnJanTemp}
\end{align}
$$

## Nested Models

if we set $\beta_5 = \beta_6 = \beta_7 =\beta_8 =0$ in `M2` we obtain.

. . .


$$
\begin{align}
E[\mbox{MnJlyTemp}] =\\ 
\beta_0 + 
\beta_1 \mbox{Lat} + 
\beta_2 \mbox{Height} + 
\beta_3 \mbox{Sea} + 
\beta_4 \mbox{NorthIsland} + \\ 
\color{red}{
\beta_5 \mbox{Rain} + 
\beta_6 \mbox{Sun} + 
\beta_7 \mbox{Long} + 
\beta_8 \mbox{MnJanTemp}
}
\end{align}
$$

## Nested Models

We defined `M1` as

$$E[\mbox{MnJlyTemp}] = \\ \beta_0 + \beta_1 \mbox{Lat} + \beta_2 \mbox{Height} + \beta_3 \mbox{Sea} + \beta_4 \mbox{NorthIsland} $$

::: {.revealjs-slide data-vertical-align="middle"}
## Nested Models


--"accepted model": a simplified model with a set of variables  that we accept as being useful.


--"extended model": a bigger model (with more variables). We will refer to the additional variables as "candidate" variables.
:::


## The F Tests for Nested Models

The general ideas about model comparison  continue to  apply.    

-- We want a model with few parameters.

-- We want a model that fits well (i.e. one with small RSS).

-- The choice of model will be a question of whether the improvement in  goodness of fit of the extended model  `M2` over the simpler model `M1` is  worth the cost.


## The F Tests for Nested Models


-- Suppose that linear model `M1` has `q` explanatory variables (hence `q+1` regression parameters, including intercept).


. . .


-- Suppose that `M1` is nested within linear model `M2` which has  `p > q` explanatory variables (hence `p+1` regression  parameters, including intercept).

## The F Tests for Nested Models 

Comparison of the models can be achieved by testing

*H~0~* : $\beta_j = 0$ for all $j \in J$ (i.e. `M1` adequate); versus

*H~1~* : $\beta_j$ not zero for all $j \in J$ (i.e. `M2` better)
    
where `J` indexes the `p-q` variables that appear in `M2` but not `M1`.

## The F Tests for Nested Models

The `F`-statistic to test these hypotheses is

$$
F = \frac{[RSS_{M1} - RSS_{M2}]/(p-q)}{RSS_{M2}/(n-p-1)}
$$

As before, extreme, large values of `F` provide evidence against *H~0~* (hence evidence that we should prefer model `M2` to `M1`).


## The F Tests for Nested Models

If *H~0~* is correct then `F` has an `F` distribution with `(p-q),(n-p-1)` degrees of freedom.

Hence if *f~obs~* is the observed value of the test statistic, and `X` is a random variable from an *F~p-q,n-p-1~* distribution, then the `P`-value is given by 

$$P= P(X \ge f_{obs})$$


## Interpretation of Test Results

-- the importance of variables in a multiple regression is dependent upon context (i.e. what other variables are taken into account).

-- Retention of *H~0~* (i.e. acceptance that model `M1` is adequate) does not mean that the `p-q` variables indexed by `J` are unrelated to the response.

-- Rather, retention of *H~0~* indicates that the variables indexed by `J` do not provide additional information about the response having adjusted for all the other (`q`) variables.

## Model Comparison for the Climate data

Our `M2` and `M1` were given above. 

```{r firstModels, echo=-1}

Climate.lm1 <- lm(MnJlyTemp~ Lat + Height+ Sea+ NorthIsland, data=climate)
anova(Climate.lm1)
```
#

```{r}
Climate.lm2 <- lm(MnJlyTemp ~ Lat + Height+ Sea+ NorthIsland+ Rain+ Sun + Long+ MnJanTemp, data=climate)
anova(Climate.lm2)
```


## Model Comparison for the Climate data

-- For model `Climate.lm1`, calculations give `RSS M1` = `r round(anova(Climate.lm1)["Residuals", "Sum Sq"],3)` and `q=4`.

-- For model `Climate.lm2`, calculations give `RSS M2` =  `r round(anova(Climate.lm2)["Residuals", "Sum Sq"], 3)` and `p=8`.


## Model Comparison for the Climate data

$$F = \frac{[RSS_{M1} - RSS_{M2}]/(p-q)}{RSS_{M2}/(n-p-1)} = 
\\
\frac{[28.13 - 25.022]/(8-4)}{25.022/(27)} = 0.8384$$

$$
P(X \ge 0.8384) = 0.5129 \qquad \text{where } X \sim F_{4,27}
$$

## Model Comparison for the Climate data

- We conclude that *H~0~* can be retained (i.e. model `M1` is adequate). There is no evidence that Rain, Sun, Long and MnJanTemp (considered together)  provide further information about MnJlyTemp once we have adjusted for Lat, Height, Sea and NorthIsland.

*H~0~* : $\beta_j = 0$ for all $j \in J$ (i.e. `M1` adequate)

## Model Comparison for the Climate data

Can we simplify further? 

Suppose we want to know if we really need all four variables that are in our previous model `Climate.lm1`. 

Maybe we can get by with an even simpler model.

## Model Comparison for the Climate data

$$
E[\mbox{MnJlyTemp}] = \beta_0 + \beta_1 \mbox{Lat} + \beta_2 \mbox{Height}
$$ 
So in terms of M1 we are setting $\beta_3=\beta_4 =0$. 

## Model Comparison for the Climate data

```{r Climate.lm0}
Climate.lm0 =lm( MnJlyTemp ~ Lat+ Height, data = climate)
anova( Climate.lm0, Climate.lm1)
```

Clearly we reject the hypothesis that $\beta_3=\beta_4 =0$ (p < 0.01). There is something in these two variables that we need. 


## Connection with the Omnibus (for all variables) F Test

-- The omnibus `F` test for model utility discussed earlier is a particular case of the more general methodology for model comparison  presented in this lecture.

-- Specifically, in omnibus `F` test we always take `M1` to be the `null` model with `q=0` explanatory variables, and `M2` to be the full model with `p` explanatory variables.


## Connection with T Tests for Single Variables

-- `F` tests can be used to compare two models that differ by a single explanatory variable.

-- a `T` test can also be used to assess the importance of any given explanatory variable.

There is a connection between the two approaches.

## Connection with T Tests for Single Variables

Suppose we test *H~0~*: $$H_0:\beta_j = 0~~~~\mbox{versus}~~~~H_1:\beta_j \ne 0$$ having adjusted for certain other variables. 

If the observed `t`-test statistic is *t~obs~*, and the observed `F` test statistic is   *f~obs~*, then 

-- *f~obs~ = t~obs~^2^*.

-- The `P`-value from the two tests will be the same.


## More Testing for the Climate Data

-- Suppose that we wish to test for the importance of `NorthIsland` having adjusted for the other variables `Lat`, `Height` and `Sea`.

-- We want to test *H~0~*: $$H_0:\beta_4 = 0~~~~\mbox{versus}~~~~H_1:\beta_4 \ne 0$$   


## More Testing for the Climate Data

```{r secondModels}
Climate.lm00 <- lm(MnJlyTemp ~ Lat+ Height+ Sea, data=climate)
anova(Climate.lm0, Climate.lm00, Climate.lm1)
```

## More Testing for the Climate Data

-- The first `F`-statistic (from the anova() function output) is `f = 8.869` with corresponding `P`-value `P=0.00559`. So we need Sea in addition to Lat and Height.

-- The second  `F`-statistic is  `f = 5.797` with corresponding `P`-value `P=0.02219`. So we need NorthIsland  in addition to Lat, Height and Sea. 


## More Testing for the Climate Data

If we look at the summary() output for M1 we get
```{r summary M1}
summary(Climate.lm1)
```

## More Testing for the Climate Data

-- The `t`-statistic for NorthIsland  is `t=2.408` with corresponding `P`-value `P=0.02219`.

-- Note that `f` = 5.797 = 2.408^2^ = *t^2^*.
