---
title: "Lecture 23 - Orthogonal Factorial Models"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
library(kableExtra)
library(car)
library(DescTools)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## learning outcomes

-- Recognize and interpret an orthogonal factorial design.

-- Implement an orthogonal factorial design in R.

## What is an Orthogonal Factorial Design?

-- An `orthogonal factorial design` is a particular case of experimental designs, with interesting properties from a statistical point of view.

&nbsp;

-- A `factorial design` is when the number of observations in the different level groups of the different factors are chosen by an experimenter. 

## What is an Orthogonal Factorial Design?

-- A factorial design is `orthogonal` if the sum of squares attributable to one factor is the same whether or not the other factor has been included in the model.

&nbsp;

-- In terms of ANOVA tables, this means that *SSA* will be the same  whether *A* appears on the first or second line of the ANOVA table.

&nbsp;

--In terms of experimental design a `balanced` and `complete` factorial design is orthogonal.

## What is a balanced and complete factorial design?

- A factorial design is `complete` if observations are made at every possible combination of factor levels, or  treatment.

  - For example, if factor *A* has 3 levels, and factor *B* has 4 levels, then a complete design requires that we observe responses at each of the possible $3\times4 = 12$ treatments.

## What is a balanced and complete factorial design?

- A factorial design is `balanced` if the same number of  experimental units are observed at each treatment.

  - $n_{ij}$ = r `(number of replications is constant)`.
  
  
| Factor A \ Factor B | B1             | B2             | B3             |
|---------------------|----------------|----------------|----------------|
| A1                  | * * *          | * * *          | * * *          |
| A2                  | * * *          | * * *          | * * *          |


## What is a balanced and complete factorial design?


::: callout-important

**Balance and completeness need to be achieved by design (they will not usually be the case in an observational study or survey).**

:::

## Consequences of an Orthogonal Factorial Design

In a two-way orthogonal design:

-- The *P*-values for each factor in the ANOVA table remain precisely the same irrespective of the order in which the factors are listed.

&nbsp;

-- In considering the importance of factor *B* it does not matter whether or not we have adjusted for *A* (and vice versa).

&nbsp;

-- The idea of orthogonality can be extended to three or more factor models.


## Example: A Designed Experiment for Dairy Cattle

::: {.columns}
:::{.column width="65%"}

```{r Cows.lm.1, echo=-1}
Cows <- read.csv(file="./cows.csv", header=TRUE)
table(Cows$Breed, Cows$Age)
```

:::
:::{.column width="35%"}
**Design summary**
<div style="font-size:smaller;">
- Cells: \(5 * 2 = 10\)  
- Replicates per cell: \(r = 10\)
- Total observations: \(N = 10 * 10 = 100\)  
</div>
:::
:::

- The design is `complete` and `balanced`, therefore `orthogonal`.

## Analysis of Dairy Cattle Data

::: {.columns}
:::{.column width="50%"}

```{r, echo=-1}
Cows.lm.1 <- lm(Butterfat ~ Breed + Age, data=Cows)
anova(Cows.lm.1)
```

:::
:::{.column width="50%"}

```{r Cows.lm.2, echo = -1}
Cows.lm.2 <- lm(Butterfat ~ Age + Breed, data=Cows)
anova(Cows.lm.2)
```
:::
:::

-- There is overwhelming evidence of a breed effect (*P*-value smaller than $2 \times 10^{-16}$) on mean butterfat content.
    
-- There is no evidence of an age effect.

## Summary Table

```{r summariesCows}
summary(Cows.lm.1)$coefficients
summary(Cows.lm.2)$coefficients
```

Note that by default, level one of `Age` (2 years old) and level one of `Breed`, Ayrshire,
    are set as the reference levels for the treatment constraint.

The Jerseys seem to provide the highest butterfat concentration.

You should look at model diagnostics.

## summary factorial models so far

Remember the `one-way` model in multiple regression form

$$Y_i = \mu + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i\\(i=1,2,\ldots,n)$$
can be written as $$\boldsymbol{Y_{ij} = \mu + \alpha_i + \varepsilon_{ij}}$$

## summary factorial models so far

-- $Y_{ij}$ is the response of the $j$th unit at the $i$th level
    of the factor $(i=1,\ldots,K;~j=1,\ldots,n_i)$;

-- $K$ denotes the number of levels, and $n_i$ the number of
    observations at level $i$ of the factor.

## summary factorial models so far

- Assume treatment constraint $\alpha_1=0$.

- $\varepsilon_{11}, \varepsilon_{12}, \ldots, \varepsilon_{Kn_K}$ are random errors
    satisfying assumptions ***A1*** to ***A4***.

## summary factorial models so far

A `two-way` model with factors *A* and *B* in multiple regression form
$$Y_i = \mu + \alpha_2 z_{Ai2} + \ldots + \alpha_K z_{AiK} 
+ \beta_2 z_{Bi2} + \ldots + \beta_L z_{BiL} + \varepsilon_i~~~~~(i=1,2,\ldots,n)$$
becomes:
$$\boldsymbol{Y_{ijk} = \mu + \alpha_i + \beta_j + \varepsilon_{ijk}}$$

where $Y_{mji}$ is the response for the $k$th unit at level $m$ of factor *A* and level $j$ of factor *B* ($m=1,\ldots,K;~j=1,\ldots,L;~i=1,\ldots,n_{mj}$).

## summary factorial models so far

- $\alpha_1, \ldots, \alpha_K$ and $\beta_1, \ldots, \beta_L$ are parameters describing the `main effects` of *A* and *B* respectively.

&nbsp;

- Assume treatment constraints, $\alpha_1 = 0$ and $\beta_1 = 0$.

&nbsp;

- $\varepsilon_{111}, \varepsilon_{112}, \ldots, \varepsilon_{KLn_{KL}}$, are error
    terms satisfying assumptions ***A1*** to ***A4***.


