


We have previously considered Testing whether the response is related to at least one  explanatory variable.

We did this by comparing the model with all predictors in it, to a simpler model with just the intercept. 

In this topic, we investigate comparing  comparing a given model with a  simplified version of that model.  

To help our discussion later we may call the simplified model an "accepted model", that is, with a set of variables  that we think accept as being useful. The bigger model (with more variables) we may call an "extended model", and we may refer to the additional variables as "candidate" variables. 

## Nested Models

A linear model *M1* is said to be **nested** within another
model *M2* if *M1* can be recovered as a special case of *M2* by
setting the parameters of *M2* to the necessary values.

For the Climate data, define model *M2* by

$$E[\mbox{MnJlyTemp}] = \beta_0 + \beta_1 \mbox{Lat} + \beta_2 \mbox{Height} + \beta_3 \mbox{Sea} + \beta_4  \mbox{NorthIsland}      
+ \beta_5 \mbox{Rain}+ \beta_6 \mbox{Sun}+ \beta_7\mbox{Long}+ \beta_8 \mbox{MnJanTemp}$$

Then the model *M1* defined by

$$E[\mbox{MnJlyTemp}] = \beta_0 + \beta_1 \mbox{Lat} + \beta_2 \mbox{Height} + \beta_3 \mbox{Sea} + \beta_4 \mbox{NorthIsland} $$ 

This is nested within *M2* because we can obtain *M1* by setting
$\beta_5 = \beta_6 = \beta_7 =\beta_8 =0$ in *M2*.

## *F* Tests for Nested Models

  - The general ideas about model comparison   continue to  apply.
    
      - We want a "cheap" model (i.e. one with few parameters);
    
      - We want a model that fits well (i.e. one with small RSS.

  - The choice of model will be a question of whether the improvement in     goodness of fit of the extended model  *M2* over the simpler model *M1* is      worth the cost.

Selecting a model is equivalent to testing hypotheses about
    parameters of *M2* (the more complex model).


  - Suppose that linear model *M1* has *q* explanatory variables  (hence *q+1* regression parameters, including intercept).   

  - Suppose that *M1* is nested within linear model *M2* which has  *p > q* explanatory variables (hence *p+1* regression     parameters, including intercept).

Comparison of the models can be achieved by testing

*H~0~*: $\beta_j = 0$ for all $j \in J$ (i.e. *M1* adequate); versus

*H~1~*: $\beta_j$ not zero for all $j \in J$ (i.e. *M2* better)
    
where *J* indexes the *p-q* variables that appear in *M2* but not *M1*.


The *F*-statistic to test these hypotheses is

$$F = \frac{[RSS_{M1} - RSS_{M2}]/(p-q)}{RSS_{M2}/(n-p-1)}$$

As before, extreme, large values of *F* provide evidence against  *H~0~* (hence evidence that we should prefer model *M2* to  *M1*).

If *H~0~* is correct then *F* has an *F* distribution with
    *(p-q),(n-p-1)* degrees of freedom.

Hence if *f~obs~* is the observed value of the test statistic,    and *X* is a random variable from an *F~p-q,n-p-1~*       distribution, then the *P*-value is given by $$P= P(X \ge f_{obs})$$

## Interpretation of Test Results

  - As we have seen earlier, the importance of variables in a multiple      regression is dependent upon context (i.e. what other variables are       taken into account).

  - Retention of *H~0~* (i.e. acceptance that model *M1* is adequate)        does not mean that the *p-q* variables indexed by *J* are       unrelated to the response.

  - Rather, retention of *H~0~* indicates that the variables indexed by       *J* **do not provide additional information** about the response   having **adjusted** for all the other (*q*) variables.

## Model Comparison for the Climate data

Our M2 and M1 are given by the above. 

`r xfun::embed_file("../../data/Climate.csv")` 

```{r firstModels, echo=-1}
Climate <- read.csv(file="../../data/Climate.csv", header=TRUE, row.names=1) 
 
Climate.lm1 <- lm(MnJlyTemp~ Lat + Height+ Sea+ NorthIsland, data=Climate)
anova(Climate.lm1)
Climate.lm2 <- lm(MnJlyTemp ~ Lat + Height+ Sea+ NorthIsland+ Rain+ Sun + Long+ MnJanTemp, data=Climate)
anova(Climate.lm2)
```


  - For model *M1*, calculations give *RSS~M1~ = `r round(anova(Climate.lm1)["Residuals", "Sum Sq"],3)`* and
    *q=4*.

  - For model *M2*, calculations give *RSS~M2~ =  `r round(anova(Climate.lm2)["Residuals", "Sum Sq"], 3)`* and
    *p=8*.
 
  - Hence *F*-statistic is
    $$F = \frac{[RSS_{M1} - RSS_{M2}]/(p-q)}{RSS_{M2}/(n-p-1)} = \frac{[28.13 - 25.022]/(8-4)}{25.022/(27)}
    = 0.8384$$

  - Corresponding *P*-value is right hand tail probability:
    $$P(X \ge 0.8384) = 0.5129~~~~~~~~~~\mbox{where $X \sim F_{4,27}$}$$

  - We conclude that *H~0~* can be retained (i.e. model *M1* is      adequate). There is no evidence that Rain, Sun, Long and MnJanTemp (considered together)   provide further
    information about MnJlyTemp once we have adjusted for Lat, Height, Sea and NorthIsland.

## Model Comparison Using R

```{r anovaFirstModels}
anova(Climate.lm1, Climate.lm2)
```
** Can we simplify further? **

Suppose we want to know if we really need all four variables that are in our previous model M1. 

Maybe we can get by with just an even simpler model as the extended model  M0

$$E[\mbox{MnJlyTemp}] = \beta_0 + \beta_1 \mbox{Lat} + \beta_2 \mbox{Height} $$ 
So in terms of M1 we are setting $\beta_3=\beta_4 =0$. 

```{r M0}
Climate.lm0 =lm( MnJlyTemp ~ Lat+ Height, data = Climate)
anova( Climate.lm0, Climate.lm1)
```

Clearly we reject the hypothesis that $\beta_3=\beta_4 =0$. There is something in these two variables that we need. 


## Connection with the Omnibus *F* Test

  - The omnibus *F* test of model fit discussed in Topic 11 is a      particular case of the more general methodology for model comparison    presented in this lecture.

  - Specifically, in omnibus *F* test we always take *M1* to be the      *null* model with *q=0* explanatory variables, and       *M2* to be the full model with *p* explanatory variables.

## Connection with T Tests for Single Variables

  - *F* tests can be used to compare two models that differ by a single      explanatory variable.

  - T test can also be used to assess the importance of any given  explanatory variable.

  - There is a connection between the two approaches. Suppose we test
    *H~0~*: $$\beta_j = 0~~~~\mbox{versus}~~~~H_1:\beta_j \ne 0$$ having
    adjusted for certain other variables. If the observed *t*-test
    statistic is *t~obs~*, and the observed *F* test statistic is
    *f~obs~*, then *f~obs~ = t~obs~^2^*.

  - The *P*-value from the two tests will be the same.

## More Testing for the Climate  Data

  - Suppose that we wish to test for the importance of NorthIsland having
    adjusted for the other variables Lat, Height and Sea.

   - We want to test
    *H~0~*: $$\beta_4 = 0~~~~\mbox{versus}~~~~H_1:\beta_4 \ne 0$$   

  - Will perform *t* and *F* tests using R.


```{r secondModels}
Climate.lm00 <- lm(MnJlyTemp ~ Lat+ Height+ Sea, data=Climate)
anova(Climate.lm0, Climate.lm00, Climate.lm1)
```


  - The first *F*-statistic (from the anova() function output) is
    *f = 8.869* with corresponding *P*-value *P=0.00559*.
    So we need Sea in addition to Lat and Height.

- The second  *F*-statistic is 
 *f = 5.797* with corresponding *P*-value *P=0.02219*.
    So we need NorthIsland  in addition to Lat, Height and Sea. 
    
If we look at the summary() output for M1  we get
```{r summary M1}
summary(Climate.lm1)
```
  - The *t*-statistic for NorthIsland  is
    *t=2.408* with corresponding *P*-value *P=0.02219*.

  - Note that *f* = 5.797 = 2.408^2^ = *t^2^*.
