

*Collinearity and Orthogonality, Measuring multicollinearity, variance  inflation factors. *

##  Idea of Collinearity and Orthogonality  

Consider the model	$$Y  =  \beta_0 + \beta_1 X_1 + \beta_2X_2 + \varepsilon$$  .

We normally interpret the regression parameter  $\beta_1$  as the rate of change in  Y  per unit change in  $X_1$, where  $X_2$  is kept constant.

However, if  $X_1$  and  $X_2$  are strongly related, then this interpretation may not be valid (or even possible).  In particular if  $X_1$  increases,  $X_2$  may increase at the same time.  An example would be if $X_1$= personal income and $X_2$ =household income.    The effect of  $X_1$  is expressed both directly (through $\beta_1$) and indirectly (through $\beta_2$).  This complicates both the interpretation and the estimation of these parameters.

 
Now if $X_1$  and $X_2$ have a total lack of any linear relationship (uncorrelated) then they are said to be **orthogonal**.

In practice regressors are rarely orthogonal unless the regression arises from a carefully designed experiment.

Nevertheless a little non-orthogonality is usually not important to the analysis, especially in view of all the other variability involved.

However sometimes  $X_1$  and $X_2$ are so closely related that the regression results are very hard to interpret.  For example inclusion of one variable into a model already containing the other variable can radically alter the regression parameters.  The parameters may also be very sensitive to slight changes in the data and have large standard errors.

Severe non-orthogonality is referred to as **collinear** data or **multicollinearity**. This

*	can be difficult to detect

*	is not strictly a modeling error but a problem of inadequate data

*	requires caution in interpreting results.

So we address the following questions:
	How does multicollinearity affect inference and forecasting?
	How does one detect multicollinearity?
	How does one improve the analysis if multicollinearity is present?

##   A Numerical Example

We will consider the effects of multicollinearity in the TempA and TempB  data considered earlier.  Recall the model is  
$$Y = 50 + 5\times TempA  + 5\times TempB  + \varepsilon  ~~~~ (*) $$

Though you would not have noticed it, the TempA and TempB data were highly correlated.   

```{r read Simulation1}
Sim = read.csv("../../data/Simulation1.csv",header=TRUE)
Sim;    attach(Sim)
Y = 50 + 5*tempA +5*tempB+ Epsilon

library(scatterplot3d)
scatterplot3d(cbind(tempA,tempB,Y) ,pch=16, type="h")
```

This is what resulted in the   high  standard errors for each regression coefficient.  

```{r lm Simulation1}
lm.cor = lm( Y ~ tempA + tempB );  summary(lm.cor)$coef 
```

Now suppose we keep exactly the same numbers TempA, TempB and Epsilon, but break the correlation  between the TempA and TempB columns. 
As a simple way of spreading them out,  I randomized the order of numbers in column TempB. 

```{r Sim randomized}
tempBrandomized = tempB[order(rnorm(20))]
Yrand = 50 + 5*tempA +5*tempBrandomized+ Epsilon
scatterplot3d(cbind(tempA,tempBrandomized,Yrand) ,pch=16,  type="h")
lm.uncor = lm( Yrand ~ tempA + tempBrandomized );  summary(lm.uncor)$coef 
```

The standard errors of the coefficients are about 1/3 what they were.  

Note that  really the only thing we have done here is to break the correlation between TempA and TempB, but this has greatly improved the analysis. 

## Orthogonal Data

Consider the model	$$Y  =  \beta_0 + \beta_1 X  + \beta_2 Z + \varepsilon$$  

Recall that  $X$  and  $Z$ are uncorrelated, they are called orthogonal.  Then  

*	the least squares estimates $\hat\beta_1$ and $\hat\beta_2$  are independent of each other

*	the order of adding the variables does not affect the sums of squares in the ANOVA table.

In other words  it doesn’t matter whether we talk about the effect of  $X$ by itself or  “$X$ after $Z$".  The effect is the same.    

*	Interpretation of parameters a lot easier.

*	But it usually only happens in designed experiments with balanced numbers.

In summary, orthogonality of the explanatory variables  allows easier interpretation of the results of an experiment and is an important goal in the design of experiments.         

### Collinear data
Suppose we have three sets of numbers $X,Z,Y$ such that 
$$Y = \beta_0 +\beta_1 X  + \beta_2 Z +\varepsilon$$

but that, unknown to us,
$$Z = \alpha_0 + \alpha_1 X ~  $$
at least equal up to some level of rounding error.

Then the regression model could be rewritten in terms of any arbitrary combination of the X and Z figures, such as 

$$Y = \beta_0 +\beta_1 X  + \beta_2 [k(\alpha_0 + \alpha_1 X) + (1-k)Z]  +\varepsilon$$
$$= \{\beta_0+\alpha_0\beta_2k\}+ \{\beta_1+\alpha_1\beta_2k\} X  + \{1-k\} Z +\varepsilon$$
for any real number $k$.  In other words there are an infinite number of possible solutions, and if we do get a  numerical  answer for 
the regression coefficients, the particular answer may depend on  something as trivial as rounding error.  Different amounts of rounding could give very different answers.  
We describe this situation by saying the regression model is *not estimable*.  

A consequence of such high correlation between the predictor variables is that the regression coefficients may become totally uninterpretable  (for example something you  *know*
has a positive relationship with the response ends up with a negative coefficient)  and  the regression coefficients may have very  high standard errors.


## Measuring Multicollinearity

So we recognize a potential problem.  How do we assess whether it is a real problem in our data?

###  Correlations Among The Predictors

We could look at the correlations between the explanatory variables.  Since correlation is a measure of linear relationship, then if two variables are highly correlated then it doesn’t make sense to keep them both in the model.

Example.  Consider the data in the following  page.  This data, called Electric.csv
consists of household electricity bills, along with explanatory variables: 

* the income of the household; 
* the number of persons in the house; and
* the area of the house. 

Each of these explanatory variables would be expected to be related to the Bill, and yet they are also related to each other, so we may not be able to have them all in the same model.   


N.B. `r xfun::embed_file("../../data/Electric.csv")` if you wish to replicate the following examples.

```{r read electric}
electric = read.csv("../../data/Electric.csv", header=TRUE)
electric;   
cor(electric)
```

This indicates that the Area of the house is very strongly correlated to the Income of the occupants ($r= 0.961$)
, and so both may say much the same thing about the electricity bill.  We may not be able to have both these in the model.  

By contrast the number of persons in the house is almost unrelated (r  =  0.143) to the Income and only weakly related to the Area (r = 0.366) so persons provides more-or-less independent information.

 The following graph helps one visualise the relationships expressed by the correlations. 

```{r  electric pairs}
pairs(electric)
```

The question is: 

### How do we know this is causing  real problems? 

The aim of the analysis is to explain electricity consumption  (‘Bill’) in terms of household ‘Income’, ‘Persons’ and ‘Area’. 

```{r electric.lm0}
electric.lm0 = lm(Bill~ Area+ Persons + Income, data=electric)
summary(electric.lm0)
```


Note (fact 1) the regression as a whole is highly significant. 

However (fact 2) none of the regression coefficients in the model is significant.    

So these two facts together suggest multicollinearity. 

### Contradictory Regression Coefficients

Note that by themselves both variables are positively related to Bill and have small Std Errors, But put together,  one predictor has changed its sign and the Std Errors are much larger. 

This illustrates how multicollinearity can render the regression coefficients quite misleading. 

```{r single variables}
elec.lm1 = lm(Bill ~ Area, data=electric);  summary(elec.lm1)
elec.lm2= lm(Bill ~ Income, data=electric);  summary(elec.lm2)
elec.lm3 = lm(Bill ~ Income + Area, data=electric); summary(elec.lm3)
```

One solution to multicollinearity is to delete one of the variables.  However it can be difficult to decide which one to delete.   For example suppose we particularly want to quantify the effect of *Persons* on the electricity bill:

 

```{r pair of variables}
elec.lm4 = lm(Bill ~ Area+ Persons, data=electric); summary(elec.lm4)
elec.lm5= lm(Bill ~ Income+ Persons, data=electric); summary(elec.lm5)
```

Should we estimate that each extra person adds 42 to the electric bill, or 83 to  the bill?  Of course the interpretation of these regression models differs slightly (in the first model it is the effect of each additional person among families of the same income, while in the second it is the effect of person among families of with homes of the same area). But it still makes one uneasy to see such a big difference.  

## Regressing The Predictors Among Themselves

More generally (more than just looking at individual correlations) we could look at whether any of the variables are strongly predicted by a combination of the other regressors.

i.e.  Whether the $R^2$  for regressing variable $X_i$
on the other regressors  
$X_1, X_2,..., X_{i-1}, X_{i+1},...,X{p}$ is large.  We call this $R^2_i$.  

If it is large then we probably don’t need $X_i$  (or some other variables) in the model.

This is more general than just looking at simple correlations between variables, since the regression model picks up on combinations of variables.

A “rule of thumb” is that multicollinearity is a big problem if $R_i^2 > 0.9$,  i.e.  the $i$th variable is more than 90% 
explained by the other predictors.

We can get $R_i^2$ by doing the regressions ourselves, but there  is an easier way. 

## Variance Inflation Factors, VIFs

A major impact of multicollinearity is found in the variances (and hence standard errors) of the regression coefficients.

1.  If $X_i$   is alone  in the model	 $$Y  =\beta_0 +\beta_1 X_i +\varepsilon~,~~~~~\mbox{then}$$

$$\hat{\mbox{var}}(\hat\beta_1) = \frac{S^2}{\sum_j{(X_{ij}
- \bar X_i)^2}}$$
2. If If $X_i$   is in a model with other predictors then 
$$\hat{\mbox{var}}(\hat\beta_i) =  S^2   \left(X^T X\right)^{-1}_{ii}$$
where the last term is the corresponding diagonal entry of the $S^2 (X^T X )^{-1}$ matrix,  the *variance-covariance* matrix.  This is always $\ge$ the  $\hat{\mbox{var}}
(\hat\beta_1)$.

The ratio of these variances is called the Variance Inflation Factor (VIF).   It turns out that 
$$VIF_i = \frac{1}{1- R^2_i}$$    so we get $VIF_i \ge 10$ if 
$R_i^2 \ge 0.9$.

In conclusion  if the VIF is large  eg. $VIF_i> 10$ then
multicollinearity is a big problem. 
On the other hand if  the VIF is around 1, then the variable $X_i$ is pretty much orthogonal to the other variables.

To obtain VIFs we need to invoke   library(car).

```{r vifs1, include=FALSE}
library(car)
```

```{r vifs2, eval=FALSE}
library(car)
```

```{r}
vif( electric.lm0)
```

Clearly we need to do something to reduce the multicollinearity.   

Note: if a regression model includes a factor, then vif() defaults to a generalised VIF. We will interpret it the same way. 

### Example: SAP Data

The following example illustrates multicollinearity in a context where it is not immediately visible.  Chatterjee and Price  discuss a dataset relating  Sales to  A (advertising expenditure),  A1 (last year's advertising),  P (promotion expenditure), P1 (last year’s promotions expenditure) and other variables.  The analyst eventually found out that an approximate budgetary constraint 

$$ A + A1 + P + P1  =  5  $$
was hidden in the data.  The constraint was known to the advertising executives who had commissioned the study but it had not been mentioned to the statistical consultant who was given the data to analyze.  

Annual Data on Advertising, Promotion, and Sales Expenses,  and Sales:

A= advertising expenditure (at time t)  

A1 denotes lagged values, i.e. at time t-1.

P denotes promotion expenditures (at time t) 

P1  denotes the lagged values.

SE denotes sales expense and   

S the aggregate sales. 

N.B. `r xfun::embed_file("../../data/SAP.csv")` if you wish to replicate the following examples.


```{r  readSAP}
SAPdata = read.csv("../../data/SAP.csv", header=TRUE)
head(SAPdata)

SAP.lm = lm( S ~ A + P + SE + A1 + P1, data=SAPdata)
summary(SAP.lm)
```

We see that, apart from Sales Expenses, there is only weak
evidence of a relationship between advertising, promotions and sales. 

```{r  vif SAP}
vif(SAP.lm)
```

The VIF shows high multicollinearity for all advertising and promotions data. 
If we drop one of the variables the others change their coefficients unpredictably  depending on which variable we drop. 

```{r dropping}
SAP.lm2= lm( S ~  P + SE +A1 + P1, data=SAPdata)
summary(SAP.lm2)$coef
SAP.lm3= lm( S ~ A +P +SE  + P1, data=SAPdata)
summary(SAP.lm3)$coef
```

Suppose we look at the relationships among the variables. 

```{r looking in}
SAPA.lm = lm( A ~ P + SE + A1 +P1, data=SAPdata)
summary(SAPA.lm)$coef
```

This is approximately A = 5 - P -A1 - P1 as in the hidden constraint.

## What do we do about multicollinearity?

As far as 161.251 is concerned, we have few options. 

1. We can remove one (or more) of the correlated variables from the regression. Our aim is to fix the problem, but sometimes we have a bit of choice which variable to remove, in which case we would like the model to be as  interpretable as possible.    Sometimes the choice is between a variable with many missing values and one with few, and so we choose the latter if it allows us a bigger effective sample size $n$.

2.  We could construct another variable which is a combination of the two most highly correlated variables.  It might be an ordinary average $(X_1+X_2)/2$, or an average of standardized versions of the $X_i$s or some other linear combination. Sometimes subtracting or dividing one variable by another will help.    For example the incidence of a particular disease such as obesity may be higher in countries with high GDP and high Population.   But GDP depends on the size of the economy, which is correlated with the Population.  Dividing gives us  per capita GDP which is a better indicator of average wealth and hence may be more related to disease incidence. 

3. (Not examinable). Those who have studied multivariate statistics may find it useful to construct principal components  or do a factor analysis of a set of correlated variables.  Dropping less important components leads to a class of models called biased regression.  Another form of biased regression shrinks the regression coefficients towards zero - a method called shrinkage estimators. 

4. (Not examinable). It is possible to fit regression models that impose linear constraints on the regression coefficients (such as forcing them to add to 100%). Such restricted least squares is also beyond the scope of this course. 
Therefore for this course we focus on method 1, dropping variables. 



