
In this lecture, we will consider hypothesis tests for one-way models.

We also look at model diagnostics for these models.

## Similarities in Analysis for Factorial and Regression Models

Factorial models (using the treatment constraint) can be written as
multiple linear regression models, using dummy
variables:

$$Y_i = \mu + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i \quad(i=1,2,\ldots,n)$$

Why is there no $\alpha_1 z_{i1}$?

## Differences in Analysis for Factorial and Regression Models

Despite the two models being similar in how they are expressed,
    there are differences in the way these types of model should be
    analysed.

**Importantly**:
    
- In a multiple linear regression model it makes sense to add or
        remove a single explanatory variable.
- In a factorial model it (usually) only makes sense to add or
        remove a whole factor, not single levels of a factor.
- This will usually require *F* tests (not just t-tests for
    single parameters).

## Testing for Factor Effect

Typically the primary question of interest for a one-way model is
    this: is the mean response dependent on the factor?

In the factorial model (with treatment constraint) this question is
    addressed by testing
    *H~0~*: $$\alpha_2 = \alpha_3 = \cdots = \alpha_K = 0$$
    versus
    *H~1~*: $$\alpha_2, \alpha_3, \ldots, \alpha_K~\mbox{not all zero}$$

Because this is a test involving multiple parameters we will usually
    need to perform an *F* test.

### *F* tests for Factor Effect

Testing for the significance of the factor in a one-way model is the
    same as comparing the (nested) models:
    $$M0: Y_i = \mu + \varepsilon_i \quad (i=1,2,\ldots,n)$$ and
    $$M1: Y_i = \mu + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i  \quad (i=1,2,\ldots,n)$$

Retaining *H~0~* corresponds to $M0$ being adequate.

Accepting *H~1~* corresponds to $M1$ being better.


We can use the *F* tests for nested models developed earlier (since
    factorial models can be regarded as regression models).

The *F* test statistic is
    $$F = \frac{[RSS_{M0} - RSS_{M1}]/(K-1)}{RSS_{M1}/(n-K)}$$

Note that the model *M1* contains *K* ‘regression’ parameters,
    namely $\mu$ and $\alpha_2, \alpha_3, \ldots, \alpha_K$
    (assuming the treatment constraint).

If *f~obs~* is the observed value of the test statistic, and
    *X* is a random variable from an *F~K-1,n-K~* distribution,
    then the *p*-value is given by $P= P(X \ge f_{\mbox{obs}})$

### Omnibus F Test for the Caffeine Data


You can `r  xfun::embed_file("../../data/caffeine.csv")` to replicate the following  example.

```{r Caffeine.lm, echo=-1, eval=-2}
Caffeine <- read.csv(file="../../data/caffeine.csv", header=TRUE)
Caffeine <- read.csv(file="caffeine.csv",header=TRUE)
Caffeine$Dose <- factor(Caffeine$Dose, levels = c("Zero", "Low", "High"))
Caffeine.lm <- lm(Taps~Dose, data=Caffeine) 
summary(Caffeine.lm)
```

### Alternative Test for the Effect of Caffeine

In general, you conduct a test for factor effects using the `anova()`
    command.

For a one-way model, R returns the omnibus *F* test statistic and
    related quantities.

```{r anovaCaffeine.lm}
anova(Caffeine.lm)
```

#### Conclusions from the ANOVA table

The observed *F*-statistic for testing for the significance of
    caffeine dose is *f* = `r round(summary(Caffeine.lm)$fstatistic["value"], 3)` on *K-1* = `r summary(Caffeine.lm)$fstatistic["numdf"]` and *n-K* = `r summary(Caffeine.lm)$fstatistic["dendf"]`
    degrees of freedom.

The corresponding *P*-value is *P*=`r round(anova(Caffeine.lm)["Dose", "Pr(>F)"], 4)`; we therefore have very
    strong evidence against *H~0~*; i.e. finger tapping speed does seem
    to be related to caffeine dose.

The table of coefficients in the R output shows that the mean
    response difference between the high dose and control groups is a
    little more than twice as big as the difference between low dose and
    control group.

### Interpretation of the ANOVA Table

Use the ANOVA table produced by R to answer the following questions on
the caffeine data.

1.  What is the residual sum of squares for the full model *M1*?

2.  What is the mean sum of squares for the model *M1*?

3.  What is the residual sum of squares for the null model *M0*, given
    below? $$M0: Y_i = \mu + \varepsilon_i~~~~~~~(i=1,2,\ldots,30)$$

4.  What is the mean sum of squares for the model *M0*?

### Diagnostics for Caffeine Data



We have tested whether the speed of students’ finger tapping is related to the caffeine dose received.
As usual, the validity of our conclusions is contingent on the adequacy of the fitted model. We should therefore inspect standard model diagnostics.

```{r plotCaffeine.lm}
par(mfrow=c(2,2))
plot(Caffeine.lm)
```





### Comments on the Diagnostic Plots: Technical

For a factorial model, the fitted values are constant within each
    group. This explains the vertical lines in the left-hand plots.

A **balanced** factorial design exists when the same number
    of observations are taken at each factor level.

When the data are balanced the leverage is constant, so a plot of
    the leverage would be pointless.

As a result, R produces a plot of standardized residuals against
    factor levels for balanced factorial models in place of the usual
    residuals versus leverage plot.

### Comments on the Diagnostics: Application to the Caffeine Data

For factorial models, it is particularly important to scrutinize the
    plots for *heteroscedasticity* (non-constant variance),
    when the error variance changes markedly between the factor levels.

When heteroscedasticity is present, assumption A3 fails and
    conclusions based on *F* tests become unreliable.

The diagnostic plots do not provide any reason to doubt the
    appropriateness of the fitted model.

## Bartlett’s Test for Heteroscedasticity

In order to assess heteroscedasticity, we can use the Bartlett's test (or the Levene test).

The hypotheses for both tests are:

- *H~0~*: error variance constant across groups versus

- *H~1~*: error variance not constant across groups

Rejecting the null means that the equal variance assumption is
    violated.


### Heteroscedasticity in Caffeine Data


```{r Caffeine.bt}
Caffeine.bt = bartlett.test(Taps ~ Dose, data = Caffeine)
Caffeine.bt
```

The *p*-value for the Bartlett's test is *P*=`r round(Caffeine.bt$p.value, 4)` which confirms that, for the
caffeine data, we have no reason to doubt the constant variance
assumption.


```{r, include = FALSE}
library(car)
```


```{r Caffeine.lt}
Caffeine.lt = leveneTest(Taps ~ Dose, data = Caffeine)
Caffeine.lt
```

We reach the same conclusion with the Levene test.
