


As we have seen, simple linear regression seeks to model the relationship between the  mean of the response *y* and a single predictor *x*.

You will have met simple linear regression in earlier Statistics papers,  so some of the material in this topic should be revision.

This topic seeks to introduce some notation and remind you of some fundamental concepts of regression.

## Pulse and Height

The following R code

- reads the pulse data into R (from a text file); and then 
- prints out the data for simple viewing.

```{r getPulseData, echo=-1, eval=-2}
Pulse <- read.csv(file="../../data/pulse.csv", header=T)
Pulse <- read.csv(file="https://r-resources.massey.ac.nz/161221/data/pulse.csv", header=TRUE)
Pulse
```
`r xfun::embed_file("../../data/pulse.csv")` 

- We would normally then look at some plots. Only one scatter plot is possible. We could display it using either a simple plot, or a more fancy ggplot, but it is essentially the same graph.

```{r plot}
plot( Pulse~Height, data=Pulse)
```

```{r PulsePlot, fig.cap="Plot of resting pulse rate (beats per minute) versus height (in centimeters) for a sample of 50 hospital patients. Source: A Handbook of Small Data Sets by Hand, Daly, Lunn, McConway and Ostrowski."}
library(ggplot2)
ggplot(Pulse, mapping = aes(x=Height, y=Pulse)) + geom_point() +
labs(ylab="Resting pulse rate (beats per minute)", xlab="Height (in centimeters) ") 
```


It is of interest to know whether pulse rate is truly dependent on height.

View pulse as response (*y*) variable and height as predictor  (*x*) variable.

We can formally test whether this really is the case by using a simple linear regression model.

Note there is one observation (patient resting pulse 116) which is distant from the main bulk of data.

## Basics of Simple Linear Regression: Data

Quantitative data occur in pairs: *(x~1~,y~1~), ...,(x~n~,y~n~)*.

*y~i~* denotes the **response** variable and *x~i~* the explanatory variable for the *i*^th^ individual.

The aim is to estimate the **mean** response for any given value of the explanatory variable.

It therefore makes sense to think of the *x* values  *x~1~,...,x~n~* as fixed constants, while *y~1~, ..., y~n~* are **chance** values taken by random variables *Y~1~,..., Y~n~*.

## Basics of Simple Linear Regression: Model

Now recall the linear regression model is

$$Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i$$ 

where
- $\beta_0, \beta_1$ are **regression parameters**.
- $\varepsilon_1, \ldots ,\varepsilon_n$ are error terms, about  which we make the following **assumptions**.

-----------------------------------------------------
A1. The expected value of the errors $E[\varepsilon_i] = 0$ for *i=1,...,n*.

A2. The errors $\varepsilon_1, \ldots ,\varepsilon_n$ are independent.

A3.  The error variance $\mbox{var}(\varepsilon_i) = \sigma^2$ (an unknown constant) for *i=1,...,n*.

A4. The errors $\varepsilon_1, \ldots ,\varepsilon_n$ are normally distributed.

--------------------------------------------------------
N.B. A2 is a new assumption not mentioned before. Near the end of the course we will consider what happens if this assumption is wrong. 

We refer back to these four assumptions many times in this course; they will often be referred to using the shorthand  **A1** to **A4**.


The validity of the simple regression model given above  and assumption A1  implies that

$$E[Y_i] = \mu_i = \beta_0 + \beta_1 x_i$$

So we say that $E[Y_i] = \mu_i$ is the expected value of *Y~i~* given (or  **conditional upon**) *x~i~*.


## Parameter Estimation


How do we estimate the parameters $\beta_0$ and $\beta_1$? 

The simplest method, which is optimal if the assumptions A1-A4 are correct, is to use the method of least squares.


Least squares estimates (LSEs) $\hat{\beta_0}$ and $\hat{\beta_1}$, are those values that minimize the sum of squares

$$\begin{aligned} SS(\beta_0, \beta_1) &=& \sum_{i=1}^n (y_i - \mu_i)^2\\
&=& \sum_{i=1}^n (y_i - \beta_0 - \beta_1 x_i)^2.
\end{aligned}$$


Using calculus it can be shown that $\hat{\beta_0}$ and $\hat{\beta_1}$ are given by
$$\hat{\beta_1} = \frac{s_{xy}}{s_{xx}}$$ 
where

$s_{xy} = \sum_{i=1}^n (x_i - \bar x)(y_i - \bar y)$;

$s_{xx} = \sum_{i=1}^n (x_i - \bar x)^2$;

and $\hat{\beta_0} = \bar y - \hat{\beta_1} \bar x \; .$

The last formula implies that the least squares regression line goes through the centroid $\left(\bar x, \bar y \right)$ of the data. 


## Properties of LSEs

Some maths shows sampling distributions of $\hat{\beta_0}$ and $\hat{\beta_1}$ are:

$$\hat{\beta_1} \sim N \left(\beta_1,  \frac{\sigma^2}{s_{xx}}\right )$$
 and
$$\hat{\beta_0} \sim N \left(\beta_0 , \, \sigma^2\left[\frac{1}{n} + 
\frac{\bar x^2}{s_{xx}} \right] \right)\; .$$


The details of the  above formulas are not important.  The *practical* point of the formulas is that 

1. the estimates are unbiased (centred on the right thing). This implies that random sampling should give you about the right answer (at least if we have a big enough dataset),   and 
2. the estimates are normally distributed, which means we can calculate confidence intervals and hypothesis tests for $\beta_0$ and $\beta_1$.

The variance terms (second term within each of the N( , ) formulae) are given for completeness of your notes, but are not something we will need to calculate ourselves. 


## Estimating the Error Variance

In practice, we are almost certainly working with sample data, which means the error variance, $\sigma^2$ is also an unknown parameter.   

It is important to estimate it because its square root $\sigma$ gives us information about the natural variation in the data i.e.  about how far away the individual datapoints $Y_i$ typically are from the central line  $E[Y] =\beta_0 + \beta_1 x$.

It can be estimated using the **residual sum of squares (RSS)**, defined by 

$$RSS = \sum_{i=1}^n (y_i - \hat{\mu_i})^2 = \sum_{i=1}^n (y_i - \hat{\beta_0} - \hat{\beta_1} x_i )^2 \; .$$ 

where $\hat{\mu_i} = \hat{\beta_0} + \hat{\beta_1} x_i$ is the *i*^th^
 **fitted value**.
 
 We often use the notation $\hat y_i$  instead of $\hat{\mu_i}$.    
 
 The $e_i = y_i - \hat{y_i}$ is the *i*^th^  **residual**. 

Some maths give an estimate of $\sigma^2$ is given by 

$$s^2 = \frac{1}{n-2} RSS.$$

The $$s  = \sqrt{ \frac{1}{n-2} \sum{(y_i- \hat y_i)^2}}$$
is sometimes called the **residual standard error**.  i.e. it is the **standard deviation of the residuals**. 
The reason for the "n-2" will be explained later.

