
In a multiple regression model there are a variety of hypotheses
    that we might wish to test. For example:

1.  Is the mean response related to any of the predictors?
2.  Does a given predictor provide additional information about the
        response over and above that provided by the other predictors?

In this lecture we shall look at the methodology for testing question  1 above.

## The *F* Test for Overall Fit of a Multiple Regression Model

For any given model: $Y_i = \beta_0 + \beta_1 x_{i1} + \ldots + \beta_p x_{ip}
+ \varepsilon_i$ where *i=1,2,...,n*.

We ask: Is the mean response (linearly) related to any of the predictors?


We test:  

*H~0~*: $\beta_1 = \beta_2 = \ldots = \beta_p = 0$ i.e. mean response is
not linearly related to any of the predictors

against

*H~1~*: $\beta_1, \beta_2, \ldots, \beta_p~\mbox{not all zero}$ i.e. mean
response is linearly related to at least one of the predictors.

Testing *H~0~* versus *H~1~* is equivalent to comparing two different
models: 

*M0*: $Y_i = \beta_0 + \varepsilon_i$ and 

*M1*: $Y_i = \beta_0 + \beta_1 x_{i1} + \ldots + \beta_p x_{ip} +  \varepsilon_i$ where *i=1,2,...,n*.



Model *M0* corresponds to *H~0~*; Model *M1* corresponds to *H~1~*.

## General Thoughts on Choosing Between Models

In choosing between models, statisticians have two aims:
    
- to choose a simple (i.e. not too complex) model;
- to choose a model that fits the data well.

We can measure the complexity of a linear regression model by the
    number of regression parameters, *p+1*. The greater this value,
    the more complex the model.

We can measure the closeness of fit of the model to data using the
    residual sum of squares.

Think of model comparison like clothes shopping --- is it worth
    spending more (parameters) in order to get a better (model) fit?

## F-Tests for Overall Fit (again)

We want to compare:

Model *M0* --- cheap (just 1 regression parameter) but may
        fit badly (residual sum of squares, *RSS~M0~*, may be large).

Model *M1* --- more expensive (*p+1* regression parameters)
        but will fit better (residual sum of squares, *RSS~M1~*,
        smaller).

So we calculate residual sum of squares (RSS) for each model and
    compute the following ***F test statistic***:

$$F = \frac{[RSS_{M0} - RSS_{M1}]/p}{RSS_{M1}/(n-p-1)}$$

- Intuitively, this *F* statistic measures (a standardised version
    of) improvement in fit per "unit cost" (i.e. per extra parameter).

Large values of *F* suggests that we should prefer *M1* to *M0*. Intuitively, improvement in fit of model is worth the cost.

How large is "large"?

If model *M0* is correct (i.e. *H~0~* is correct) then the *F* test statistic has an *F* distribution with *p,(n-p-1)* degrees of freedom, often denoted *F~p,n-p-1~*.

We use these facts to calculate the *P*-value for the *F* statistic, and
    hence test *H~0~* versus *H~1~*.

## Aside: The *F* Distribution


An *F* distribution is defined by two degrees of freedom.

Random variables from the *F* distribution take only non-negative values.

Some examples of the density of various *F* distributions are displayed. The shape depends on the numerator and denominator degrees of freedom...


```{r FDistributions, echo=FALSE, fig.width=9, fig.height=7, fig.cap="Probability density functions for F distributions with numerator df of 3 (left) and 10 (right), and denominator df of (10 (upper) and 30 (lower)"}
par(mfrow=c(2,2))
x=(0:250)/10
y1 = df(x,3,10)
plot(x, y1, ylab="f(x)", ylim=c(0,0.8), type="l")
y2 = df(x,10,10)
plot(x, y2, ylab="f(x)", ylim=c(0,0.8), type="l")
y3 = df(x,3,30)
plot(x, y3, ylab="f(x)", ylim=c(0,0.8), type="l")
y4 = df(x,10,30)
plot(x, y4, ylab="f(x)", ylim=c(0,0.8), type="l")
```

... but large values of *x* are always unlikely to be observed by chance alone.


As a rough guide,  the median of an *F* distribution is always <1. So if you get an F value less than one,  then it is not going to be significant.  
If you get an F value greater than ten it definitely is very significant. 

## ANOVA table 

In most Statistical programs (but not R) it is common to represent the result of a regression with an Analysis of Variance table.   It is a useful way to keep track of things. (We'll soon see how R summarises this table.)



Source | Sum of squares | df | Mean Square | F 
--- | --- | --- | --- | ---
Regression | $SS_{regn}$| p | $MS_{regn} = \frac{SS_{regn}}{p}$| F=$\frac{MS_{regn}}{MS_{resid}}$
Residuals| $SS_{resid}$ | n-p-1  | $MS_{resid}=\frac{SS_{resid}
}{n-p-1}$ | 
Total |$SS_{Tot}$|n-1  | |


where  $MS_{total} = \sum( Y -\bar Y)^2$,  $~~MS_{error} =  \sum( Y -\hat Y)^2$  ,  p= number of regression parameters (predictors, not including the intercept)
the model formula and $n-p-1$ is the residual df. 

Note the First two SS add up to the total SS.  So this illustrates how regression separates the different sources of variation or "analyses the variance". 

The df add up as well:   p  + n-p-1= n-1.

The *F* is the ratio of the Mean Squared errors. 

### Analysis of Variance table for Climate data


N.B. `r xfun::embed_file("../data/climate.csv")` to be able to run the examples below.


For the climate data we have 

```{r climate}
Climate = read.csv(file= '../data/climate.csv', header=TRUE, row.names = 1)
Climate.lm0= lm(MnJlyTemp ~ 1, data= Climate)
Climate.lm1 = lm( MnJlyTemp~ ., data=Climate)
anova(Climate.lm0, Climate.lm1)
```

Note that we have explicitly created the null model `Climate.lm0` with no predictor variables.

The use of the `anova()` command has compared two models here. We show that this way of using `anova()` is unnecessary below, but we do make frequent use of the `anova()` command with reference to two (or more) models a lot when working with regression models.


## Omnibus *F* Test Statistics in R

The above test of whether the response is related to any of the
    explanatory variables is sometimes called an **omnibus** *F*    test, while others refer to it as a ***test of model utility***.

We do not have to do the test by hand, or by fitting the null model because R provides the *F* statistic
    and corresponding *P*-value for this test as a standard part of the
    `summary()` output for a linear model.




### Back to Climate example

```{r summary info}
Climate.lm1.sum = summary(Climate.lm1)
Climate.lm1.sum
names(Climate.lm1.sum)
Climate.lm1.sum$r.squared
Climate.lm1.sum$fstatistic
```

These figures are the same as were seen earlier when we used the `anova()` command. 


