
```{r extraPkgs, message=FALSE, warning=FALSE}
library(car)          # Type I vs Type III ANOVA
library(emmeans)      # Marginal means and trends
library(kableExtra)   # Table styling
library(patchwork)    # Plot composition
library(scales)
library(gridExtra)
library(grid)
```


## Learning objectives

By the end of this in-class session you will be able to:

1. Recall the five canonical GLM forms from Lecture 25 and how parameterization and contrasts affect interpretation.

2. Map natural orderings of GLMs and identify which pairs are nested.

3. Conduct partial F tests for nested GLMs and interpret RSS, df, F, and P.

4. Explain why p-values can change when multiple models are passed to `anova()`.

5. Compare three Samara models and select a defensible specification.

6. Transfer the workflow to a second dataset on body fat with one covariate and one factor.


## Quick recall of Lecture 25

**Five canonical model forms for one factor `A` and one covariate `x`:**

- Separate slopes: `Y ~ A + A:x` or equivalently `Y ~ A * x`
- Parallel slopes: `Y ~ A + x`
- Single regression line: `Y ~ x`
- Intercepts only by level: `Y ~ A`
- Null model: `Y ~ 1`

**Equivalence of parameterizations:** two formulas can span the same model space even if coefficients differ. Same fitted values implies same residuals and identical SSE.

**Centering covariates:** centering turns the intercept into the mean response at average `x`, which improves interpretability in interaction models.

## Comparison on Nested Linear Models

For any linear model $M0$ nested within $M1$, we can test:

* $H_0$: $M0$ adequate vs

* $H_1$: $M0$ not adequate.


using an *F* test.

The hypotheses can be stated in terms of parameters of model $M1$.

The *F* test statistic is

$$F = \frac{[RSS_{M0} - RSS_{M1}]/d}{RSS_{M1}/r}$$ where *d* is
    the difference in the number of (regression) parameters between the
    models, and *r* is the residual degrees of freedom for $M1$.


If $H_0$ is correct then *F* has an *F* distribution with $d,r$
    degrees of freedom.

Extreme large values of the test statistic provide evidence against
    $H_0$.

Hence if $f_{obs}$ is the observed value of the test statistic,
    and *X* is a random variable from an $F_{d,r}$ distribution,
    then the *P*-value is given by $P= P(X \ge f_{obs})$

This test can be carried out in R using `anova(M0, M1)` where `M0` and `M1` are respectively the $M0$ and $M1$ fitted models.

## The Samara Data revisited

Recall that we were modelling the mean speed of fall of samara
    (fruit from a maple tree) as a function of the disk loading (a
    covariate) and the tree from which it fell (a factor) for each
    fruit.

We fitted separate linear regressions for each of the three trees.

We shall now consider simpler models:
    
- Parallel regressions for each tree (i.e. equal slope for `Load` for all trees);
- A single regression model for every fruit (i.e. equal slope and intercept for each tree).
    
We shall compare these models with each other, and with the separate regressions model.


Grab the data here:`r xfun::embed_file("../data/samara.csv",text = " samara.csv")`

```{r SamaraReadData, echo=FALSE}
Samara <- read_csv("../data/samara.csv",show_col_types = FALSE) |>
  transform(Tree = factor(Tree))

```

```{r samara-plot, echo=FALSE}
p_samara <- Samara |>
  ggplot(aes(x = Load, y = Velocity, color = Tree, shape = Tree)) +
  geom_point(size = 2.6, alpha = 0.9) +
  labs(title = "Samara: Velocity vs Load by Tree",
       x = "Load (disk loading)",
       y = "Velocity") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
p_samara
```

## Natural ordering of GLMs and nesting

We focus on the nesting relations for one factor `Tree` and covariate `Load`:

- Null ⊂ Intercepts-only ⊂ Parallel slopes ⊂ Separate slopes

`Single line` vs `Intercepts-only` are not nested when groups differ in mean at `x = 0`. 
Valid partial F tests require true nesting and common error variance.


We fit:

- M2: Separate regressions: `Velocity ~ Tree + Load + Tree:Load`
- M1: Parallel regressions: `Velocity ~ Tree + Load`
- M0: Single regression: `Velocity ~ Load`

```{r samara-models}
M2_sep <- lm(Velocity ~ Tree + Load + Tree:Load, data = Samara)  # separate slopes
M1_par <- lm(Velocity ~ Tree + Load, data = Samara)              # parallel slopes
M0_one <- lm(Velocity ~ Load, data = Samara)                     # single line
```

```{r samara-summaries, echo=FALSE,  results='asis'}
summaries <- bind_rows(
  glance(M2_sep) |> mutate(model = "Separate (Tree + Load + Tree:Load)"),
  glance(M1_par) |> mutate(model = "Parallel (Tree + Load)"),
  glance(M0_one) |> mutate(model = "Single (Load)")
) |>
  mutate(rss = map_dbl(list(M2_sep, M1_par, M0_one), ~ sum(residuals(.x)^2))) |>
  transmute(model,
            df_resid = df.residual,
            rss,
            r.sq = r.squared,
            adj.r.sq = adj.r.squared,
            AIC = AIC,
            BIC = BIC) |>
  arrange(AIC)

```


```{r tbl-summaries, echo=FALSE, fig.cap="Model summaries with RSS, df, R^2, AIC, BIC",fig.width=12, fig.height=4, out.width="90%", fig.align='center'}


tbl <- summaries |>
  mutate(across(c(rss, r.sq, adj.r.sq, AIC, BIC), ~round(., 3)))

theme_tbl <- ttheme_minimal(
  base_size = 16,                                  # font size
  core    = list(padding = unit(c(6, 10), "pt")),  # cell padding
  colhead = list(fg_params = list(fontface = "bold"))
)

tg <- tableGrob(tbl, rows = NULL, theme = theme_tbl)
grid.newpage(); grid.draw(tg)
```

## Visualizing fitted lines

Let's look at these fits visually to remind ourselves what the data looks like and how the models differ.

```{r fittedlines, echo=FALSE, fig.width=14, fig.height=4, dpi=150, out.width="100%"}

# Make a version of plot_lines that maps color for the predicted lines
plot_lines_with_legend <- function(model, title) {
  newdat <- tidyr::crossing(
    Tree = factor(levels(Samara$Tree), levels = levels(Samara$Tree)),
    Load = seq(min(Samara$Load), max(Samara$Load), length.out = 100)
  )
  newdat$Velocity_hat <- predict(model, newdata = newdat)
  # Ensure both points and lines use the same color mapping (Tree)
  p <- p_samara +
    geom_line(data = newdat, aes(x = Load, y = Velocity_hat, color = Tree), linewidth = 1)
  p + labs(title = title)
}

p1 <- plot_lines_with_legend(M2_sep, "Separate slopes")
p2 <- plot_lines_with_legend(M1_par, "Parallel slopes")
p3 <- plot_lines_with_legend(M0_one, "Single regression line")

# Collect a single legend and put it at the bottom
(p1 + p2 + p3) + plot_layout(ncol = 3, guides = "collect") & theme(legend.position = "bottom")

```


## Pairwise nested comparisons with partial F tests

```{r samara-anovas}
# M1 vs M2 tests the interaction
anova(M1_par, M2_sep)
```

## Pairwise nested comparisons with partial F tests continued

```{r samara-anovas-2}
# M0 vs M1 tests Tree after Load
anova(M0_one, M1_par)
```


## Pairwise nested comparisons with partial F tests continued
```{r samara-anovas-3}
# M0 vs M2 tests Tree and interaction together
anova(M0_one, M2_sep)
```

## All models in one call and denominator effect

```{r samara-anova-3}
anova(M0_one, M1_par, M2_sep)
```

**Note:** The p-value for `M0` vs `M1` can change because the denominator uses the largest model in the call. Decide a maximal model ahead of time or clearly document the sequential plan.

## Manual check of the F test

We verify `M1` vs `M2` by hand.

Remember \[ F = \frac{(RSS_{M1} - RSS_{M2})/d}{RSS_{M2}/r} \]

```{r manual-F}
rss_M1 <- sum(residuals(M1_par)^2)
rss_M2 <- sum(residuals(M2_sep)^2)
d      <- attr(logLik(M2_sep),"df") - attr(logLik(M1_par),"df")  # number of added parameters
r      <- df.residual(M2_sep)                                    # residual df in larger model
Fval   <- ((rss_M1 - rss_M2)/d) / (rss_M2/r)
pval   <- pf(Fval, d, r, lower.tail = FALSE)

```

```{r manual-F-table, echo=FALSE}
tibble(comparison = "M1 vs M2", d = d, r = r, F = Fval, `Pr(>F)` = pval) |>
  kable(digits = 6, caption = "Manual partial F reproduces anova()") |>
  kable_styling(full_width = FALSE)
```



## Diagnostics checklist for the chosen model

```{r diagnostics}
chosen <- M1_par
p1 <- ggplot(data.frame(fitted = fitted(chosen), resid = resid(chosen)), aes(fitted, resid)) +
  geom_point() + geom_hline(yintercept = 0, linetype = 2) +
  labs(title = "Residuals vs Fitted") + theme_minimal(base_size = 13)
p2 <- ggplot(data.frame(stdres = rstandard(chosen)), aes(sample = stdres)) +
  stat_qq() + stat_qq_line() +
  labs(title = "Normal Q-Q") + theme_minimal(base_size = 13)
p1 | p2
```


## Body Fat case study

In this section, we are concerned with modelling a small data set on human body fat.

The response is percentage fat.  Age is a numerical covariate, gender is a factor.


Grab the data here: `r xfun::embed_file("../data/fat.csv",text = " fat.csv")`

```{r load-fat, echo=FALSE, message=FALSE}
Fat <- readr::read_csv("../data/fat.csv", show_col_types = FALSE)
```

Various model comparisons and model summaries are provided. Decide which is the best model, and write down the fitted model equation for males and females.

Columns: `Percent.Fat` response, `Age` numeric, `Gender` factor.

```{r fat-plot}
p_fat <- ggplot(Fat, aes(x = Age, y = Percent.Fat, color = Gender, shape = Gender)) +
  geom_point(size = 2.6, alpha = 0.9) +
  labs(title = "Percent Body Fat vs Age by Gender",
       x = "Age", y = "Percent Fat") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
p_fat
```

## Three models and comparisons

```{r fat-models}
F0 <- lm(Percent.Fat ~ Age, data = Fat)
F1 <- lm(Percent.Fat ~ Gender + Age, data = Fat)
F2 <- lm(Percent.Fat ~ Gender + Age + Gender:Age, data = Fat)

anova(F0, F1)
anova(F1, F2)

```
Since we are talking about a one degree of freedom difference, the p-value from 
anova(F1, F2) is the same as the p-value for the interaction term in F2, namely 0.0679. So we don't use the interaction model.  

Similarly the difference between the first two models is one df, so the p-value from anova(F0, F1) would be the same as for the coefficient of GenderM, namely 0.0182. 

So we need to use the second model.  

## Selected model terms

```{r fat-summaries}

summary(F1)

```


## Fitted equations for the selected model

If we select the parallel slopes `F1` with treatment contrasts and `Gender` coded with F as reference:

- Females: `E[Fat] = β0 + β_Age * Age`

- Males: `E[Fat] = (β0 + β_GenderM) + β_Age * Age`

```{r fat-equation}
co <- coef(F1)
beta0 <- unname(co["(Intercept)"])
beta_age <- unname(co["Age"])
beta_male <- unname(co["GenderM"])
tibble(
  model = c("Females", "Males"),
  equation = c(
    sprintf("Fat = %.2f + %.2f * Age", beta0, beta_age),
    sprintf("Fat = %.2f + %.2f * Age", beta0 + beta_male, beta_age)
  )
) |>
  kable(caption = "Fitted equations by Gender (parallel slopes model)") |>
  kable_styling(full_width = FALSE)
```




## Reporting template

- State the model forms, nesting, and the scientific contrast.
- Report RSS, df, F, and p with the denominator model identified.
- Add a sentence on diagnostics and effect interpretation.
- If two models are close, prefer the simpler one for clarity unless the complex model adds needed scientific nuance.
