
## From model comparison to model selection

In Lecture 26, the scientific question gave us a reduced model and a full model to compare. We knew in advance which additional terms we wanted to test.

Real modelling problems are often less tidy. We may have many plausible explanatory variables and many possible models. There may be no single reduced-versus-full comparison that answers the whole problem.

This changes the question from **model comparison** to **model selection**.

The central question is:

> How do we balance model fit against model complexity when several models are plausible?

### What are we trying to do?

Before choosing a model, identify the modelling goal:

- **Explanation or inference:** estimate and interpret effects after adjustment for other variables.
- **Prediction:** predict new observations accurately.
- **Description:** find a compact model that describes the observed relationships.

These goals can lead us to prefer different models. A variable can be scientifically important even if it contributes little to prediction, and a useful predictor does not necessarily have a causal interpretation.

## Overfitting and underfitting

Should all the plausible explanatory variables be used, or should some be omitted? The answer involves two risks.

```{r bias-variance-continuum, echo=FALSE, fig.width=9, fig.height=1.8, out.width="90%", fig.cap="The bias-variance trade-off as model flexibility changes."}
continuum <- tibble(
  x = c(1, 5, 9),
  label = c("Too simple\nunderfitting\nmore potential bias", "A useful balance", "Too flexible\noverfitting\nmore variance")
)

ggplot(continuum, aes(x = x, y = 0)) +
  geom_segment(aes(x = 1.4, xend = 8.6, y = 0, yend = 0),
               arrow = grid::arrow(ends = "both", length = grid::unit(0.12, "inches")),
               colour = "#325d88", linewidth = 1.2) +
  geom_point(aes(colour = label), size = 4, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.8, lineheight = 0.95, size = 3.7) +
  scale_colour_manual(values = c("#f47c3c", "#325d88", "#93c54b")) +
  coord_cartesian(xlim = c(0.5, 9.5), ylim = c(-0.2, 1.8), expand = FALSE) +
  theme_void()
```

This is the **bias-variance trade-off**. A model that is too simple can omit structure. A model that is too complicated can fit accidental features of the current data and produce less stable estimates.

### Dropping or Including Variables that have $\beta=0$

Suppose we have a regression model with several variables.  We might suspect one has regression coefficient $\beta=0$.  Should we drop it?

Some mathematical theory shows that if we omit a variable which (truly) has regression coefficient  $\beta=0$  then

- Variances of the remaining regression coefficients are reduced

- Variances of predictions from the model  at any point $x_0$  are reduced. 

Both these things are good.

Conversely if we include a variable whose coefficient $\beta=0$ this  increases these variances unnecessarily. This is called **overfitting**. We want to avoid this.

### Dropping or Including Variables that have $\beta \ne 0$

On the other hand suppose we make a mistake and omit a variable that has regression coefficient $\ne 0$.   

Then we have the opposite problem, called **underfitting**.  Mathematical theory shows the consequence of underfitting is that 

-	if the wrongly-omitted variables are correlated with the remaining variables then the expected values of the remaining regression coefficients $\beta_0,\beta_1,...\beta_{p-1}$ are **biased**. (i.e. may be a too big or too small).

- Whether or not the omitted variables are correlated with the existing ones, the predictions $\hat y_i$ are biased.

Both these things are bad!

**In practice we do not know if parameters are zero.** We could use a hypothesis test to help us decide, but this is not a definitive answer.

## A Numerical Illustration of Under and Overfitting

In this artificial example, we consider an industrial process whose Yield is thought  to depend on the process temperature at stage A of the process, and possibly also on the temperature at stage B.   

i.e.  We consider the model 
$$ Y=\beta_0 + \beta_1 A+ \beta_2 B +\varepsilon.$$
	
We will assume the values of $A$ and $B$ and $\varepsilon$ are given in the data below.  The errors are random $\varepsilon \sim \mbox{Normal}(0, 1)$.

`r xfun::embed_file("../data/simulation1.csv")` 

```{r read Simulation1, eval=-1, echo=-2, message=FALSE}
sim = read.csv("simulation1.csv",header=TRUE)
sim = read.csv("../data/simulation1.csv",header=TRUE)
```


What we want to know is, what are the implications of using an incorrect model, 	i.e. incorrect assumptions about the $\beta$s. To understand the implications, we simulate two scenarios. 
	
### Scenario 1: $\beta_0=50$, $\beta_1=10$ and $\beta_2=0$.

For this illustration, the correct model doesn't depend on $B$ at all:
$$Y  =  50 + 10 A + \varepsilon$$

The following shows the result of modelling $Y$ by (1) $A$ alone and  (2) by $A$ and $B$.

```{r overfitting}
dat <- sim |> mutate(Y = 50 + 10*A + epsilon)
lm1 = lm(Y ~ A, data=dat)
lm2 = lm(Y ~ A + B, data=dat)
lst(lm1, lm2) |> map_dfr(tidy, .id='model')
```

Adding the unnecessary variable doesn't introduce bias ($\hat\beta_1 \approx 10$), but the standard error of $\hat\beta_1$ is higher in the misspecified model.

### Scenario 2: $\beta_0=50$, $\beta_1=5$ and $\beta_2=5$.

We consider the model where $Y$ depends on both $A$ and $B$,
$$	Y  =  50 + 5 A + 5 B + \varepsilon $$

```{r underfitting}
dat <- sim |> mutate(Y = 50 + 5*A + 5*B + epsilon)
lm1 = lm(Y ~ A, data=dat)
lm2 = lm(Y ~ A + B, data=dat)
lst(lm1, lm2) |> map_dfr(tidy, .id='model')
```

In the first model, $B$ is wrongly ignored and $A$ is both biased and has a larger standard error. In the second model, the values of $\hat\beta_1$ and $\hat\beta_2$ are close to the true ones and the standard errors are the same as in the second analysis of Scenario 1.

### Summary: Bias vs Variance Trade-off

Omitting a needed variable can bias the remaining estimates and predictions, especially when the omitted predictor overlaps with variables retained in the model. Including an unnecessary variable can leave the estimates centred correctly while increasing their standard errors.

In practice we do not know which situation we are in. Model selection therefore requires a modelling goal and criteria that reflect the cost of complexity. A hypothesis test can contribute evidence, but it is not a complete decision rule.

## How should we judge extra complexity?

If adding variables always decreases RSS, how can we tell whether the extra complexity is worthwhile? The answer depends on the goal and on which candidate models are being compared.

### Fit measures

The coefficient of determination

$$R^2 = 1 - \frac{\sum(y_i-\hat y_i)^2}{\sum(y_i-\bar y)^2}$$

measures the proportion of sample response variation accounted for by the fitted model. It does not penalise complexity, so $R^2$ cannot decrease when variables are added.

Adjusted $R^2$ accounts for residual degrees of freedom:

$$R^2_{\mathrm{Adj}} = 1 - \frac{\sum(y_i-\hat y_i)^2/(n-p)}{\sum(y_i-\bar y)^2/(n-1)}.$$

It can decrease when an added term does not reduce RSS enough to justify its degrees of freedom. A decrease is evidence against that particular added structure, not proof that the variable should never be used.

The residual standard error is the residual scale estimate:

$$S = \sqrt{\frac{RSS}{df_{\mathrm{residual}}}}.$$

It measures remaining unexplained variation on the response scale. It is useful for comparing fitted models to the extent that they have the same response and observations.

### Partial *F* tests and AIC

Partial *F* tests are useful when we have a particular nested change in mind. AIC provides another criterion:

$$AIC = n\log(RSS/n) + 2p + \text{constant},$$

where $p$ is the number of fitted regression parameters. Smaller AIC is preferred when comparing candidate models fitted to the same response and observations.

Unlike the partial *F* test, AIC does not require one model to be nested inside another. It still does not identify a universally correct model. It gives one way to balance fitted error against model complexity.

Individual coefficient *P*-values answer a more specific question about one coefficient under one parameterisation. They should not be treated as a complete model-selection rule.

### Cross-validation: a prediction question

All of the criteria above are calculated from the data used to fit the model. If prediction is our main goal, the stronger question is:

> How well does the model predict observations it did not use to fit itself?

The basic cross-validation idea is:

```{r cross-validation-diagram, echo=FALSE, fig.width=8, fig.height=2.2, out.width="90%", fig.cap="The basic cross-validation cycle."}
cv_steps <- tibble(
  x = c(1, 3, 5, 7),
  label = c("Fit on\ntraining data", "Predict\nheld-out data", "Measure\nprediction error", "Repeat\nwith new folds")
)

ggplot(cv_steps, aes(x = x, y = 0)) +
  geom_segment(data = tibble(x = c(1.35, 3.35, 5.35), xend = c(2.65, 4.65, 6.65)),
               aes(x = x, xend = xend, y = 0, yend = 0),
               arrow = grid::arrow(length = grid::unit(0.12, "inches")),
               colour = "#325d88", linewidth = 1) +
  geom_label(aes(label = label), fill = "#f8f5f0", colour = "#3e3f3a", label.size = 0.4,
             size = 3.5, lineheight = 0.9) +
  coord_cartesian(xlim = c(0.4, 7.6), ylim = c(-0.7, 0.8), expand = FALSE) +
  theme_void()
```

We introduce that idea here but leave implementation, including `cv.glmnet()`, for Lecture 28.

## Select terms, not arbitrary coefficients

The scientific predictor and the individual regression coefficients are not always the same thing. A four-level factor creates three treatment coefficients, but it is still one scientific predictor.

Similarly, in

```
y ~ A * x
```

`A:x` is one interaction term, even though it can correspond to several interaction coefficients. If we retain an interaction, we normally retain the corresponding main effects as well. This is the hierarchy principle.

## Example:  Model Selection for Climate Data

We wish to model the mean temperature in July (winter) in Aotearoa New Zealand using data from 36 towns. We have information on location (latitude, longitude, height (elevation), whether they're by the sea or in the North island) as well as climate information (mean temperature in January, total sunshine hours and total rainfall).

`r xfun::embed_file("../data/climate.csv")` 

```{r read climate,eval=-1, echo=-2, message=FALSE}
climate = read_csv("climate.csv")
climate = read_csv("../data/climate.csv")
climate
```

### Climate Exploratory Data Analysis

```{r, fig.width=10, fig.height=4}
climate |>
  pivot_longer(-c(Place, MnJlyTemp)) |>
  ggplot() +
  geom_point(mapping=aes(x=value, y=MnJlyTemp)) +
  facet_wrap(vars(name), scales = 'free_x', ncol=4)
```

### Higher winter temperature associations

- Lower elevations (height)
- Lower latitudes (further north)
- Higher longitudes (further east - recall how Ao/NZ is oriented on a map - possibly correlation with latitude, will check next!)
- Higher summer temperatures
- North island (could just be latitude?)
- Increased rainfall - up to a point! (subtropical vs rainy westcoast - interaction with Island/latitude??)
- Closeness to the sea
- Increased sunshine hours

Many of these are likely correlated.

### Climate: Latitude/Longitude and North vs South

```{r}
ggplot(climate) +
  geom_point(mapping=aes(x=Long, y=-Lat, shape=as_factor(NorthIsland), col=MnJlyTemp)) +
  geom_text(mapping=aes(x=Long, y=-Lat, label=Place), size=2, hjust='right', nudge_x=0.1, nudge_y=0.2) +
  coord_equal()
```

#### Climate: Latitude/Longitude and North vs South

Ok, so we can see that the temperature is being driven by latitude (further north is warmer) - the longitude effect is possibly just due to orientation. It's clear there's an elevation (height) effect as well, though that is possibly confounded with whether the town is close to the sea (it can't really be both!)

So we'd expect some collinearity here and figuring out which are the best variables to be used might take a bit of playing!

We'll start with Latitude, then add in Height and Sea:

```{r}
lm1 = lm( MnJlyTemp ~ Lat, data=climate)
lm2 = lm( MnJlyTemp ~ Lat + Height, data=climate)
lm3 = lm( MnJlyTemp ~ Lat + Height + Sea, data=climate)
```

### Building the Climate Model

#### Model 1: Lat only

```{r}
lm1 |> summary()
```

#### Model 2: Lat + Height

```{r}
lm2 |> summary()
```

#### Model 3: Lat + Height + Sea

```{r}
lm3 |> summary()
```

#### First three predictors: assessment

The first three additions agree across the measures we have considered: the added-term *P*-values are small, $S$ decreases, and adjusted $R^2$ increases. This is consistent with retaining `Lat`, `Height` and `Sea` for this candidate model path.

The `Height` coefficient also changes noticeably when `Sea` is added. This tells us that the interpretation of `Height` depends on whether `Sea` is adjusted for. The predictors overlap in the information they carry, so the coefficient in a multiple regression is a conditional association, not an isolated effect of height.

#### Model 4: Adding NorthIsland

```{r fourth variable}
lm4 = lm( MnJlyTemp ~ Lat+ Height + Sea + NorthIsland, data=climate)
lm4 |> summary()
```

The added-term *P*-value is small, and the fit measures improve. The large change in the `Lat` coefficient tells us that the interpretation of latitude depends on whether `NorthIsland` is included. It does not, by itself, prove that the new variable is correcting bias.

**Why do you think adding North vs South island is useful over and above Latitude? Wouldn't Latitude do everything here??**

#### Model 5: Adding Rain

```{r fifth variable}
lm5 = lm( MnJlyTemp ~ Lat+ Height+ Sea + NorthIsland + Rain, data=climate)
lm5 |> summary()
```


The evidence for `Rain` is mixed. Its coefficient *P*-value is approximately 0.10, while $S$ decreases and adjusted $R^2$ increases. These are all in-sample measures, so they do not establish that `Rain` will improve prediction for new towns. If prediction were the primary goal, this would be a natural case for validation or cross-validation.

#### Model 6: Adding Longitude

```{r sixth variable, warning=FALSE}
lm6 = lm( MnJlyTemp ~ Lat+ Height+ Sea + NorthIsland+ Rain + Long, data=climate)
lm6 |> summary()
```

The evidence against adding `Long` is consistent: its *P*-value is greater than 0.5, adjusted $R^2$ decreases, and $S$ increases. This candidate addition does not improve the fitted model by these measures.

### Climate model-building progression

The progression is a sequence of candidate comparisons, not a proof that one final formula is universally correct. The table puts the main evidence side by side.

```{r climate-progression, echo=FALSE, results='asis'}
climate_progression <- tibble(
  Model = paste("lm", 1:6),
  `Added term` = c("Lat", "Height", "Sea", "NorthIsland", "Rain", "Long"),
  `Adjusted R2` = sapply(list(lm1, lm2, lm3, lm4, lm5, lm6), function(model) summary(model)$adj.r.squared),
  RSE = sapply(list(lm1, lm2, lm3, lm4, lm5, lm6), sigma),
  AIC = sapply(list(lm1, lm2, lm3, lm4, lm5, lm6), AIC),
  `Added-term evidence` = c("< 0.001", "< 0.001", "< 0.001", "< 0.001", "0.10", "> 0.5")
)

knitr::kable(climate_progression, digits = 3,
             caption = "Climate candidate models and in-sample comparison measures")
```

The table helps us describe what changes as terms are added. It does not replace the modelling goal, subject-matter reasoning, or validation of future prediction.

### A final model check

Choosing a plausible mean structure is not the end of the analysis. We still need to check whether the fitted model shows evidence of problems with the LINE assumptions.

`lm(MnJlyTemp ~ Lat + Height + Sea + NorthIsland + Rain, data = climate)`.

#### Final Model Diagnostic Plots
```{r diagnostic plots, echo=FALSE}
# Generate the 4 standard diagnostic plots
par(mfrow = c(2, 2))
plot(lm5)
```

#### Interpretation of diagnostic plots

- Residuals vs. Fitted: the points show no strong systematic pattern, so there is not strong evidence of a problem with the linear mean structure.
- Normal Q-Q: the points are reasonably close to the reference line, with one noticeable departure. This is not proof of normality, but it gives no strong warning at this scale.
- Scale-Location: the spread does not show a strong trend across fitted values, so there is not strong evidence of changing variance.
- Residuals vs. Leverage: no observation stands out as obviously influential from this plot alone.

These plots do not confirm the assumptions or certify the model. They give us evidence about whether the fitted model shows important departures that need further investigation.

## A more realistic running example

From this point in the course, we will keep coming back to one larger dataset to show how we can build up a more complex, and therefore more realistic, regression model. The data describe hourly bike rentals in Seoul, South Korea, together with weather and calendar information that might help explain or predict demand.

Rather than introducing a new small dataset for every method, we will use the Seoul bike data to see what happens as a model becomes more complicated. We will start with several plausible predictors here, then return to the same problem when we look at automated variable selection, penalised regression, multicollinearity and other complications later in the course.

`r xfun::embed_file("../data/seoul_bike_hourly.csv")`

```{r read seoul hourly, message=FALSE}
seoul_hourly = read_csv("../data/seoul_bike_hourly.csv")
seoul_hourly
```

The response is `rented_bike_count`. Potential explanatory variables include weather measurements such as temperature, humidity, wind speed and rainfall, as well as calendar information such as hour, season and whether the day is a holiday.

There is no obvious single model here. Several predictors look reasonable, some are likely to overlap in the information they contain, and the number of possible models grows quickly. That is much closer to what happens with real data.

```{r seoul candidate models, message=FALSE}
seoul_basic = lm(
  rented_bike_count ~ temperature + humidity + wind_speed,
  data = seoul_hourly
)

seoul_richer = lm(
  rented_bike_count ~ temperature + humidity + wind_speed +
    visibility + dew_point_temperature + solar_radiation +
    rainfall + snowfall + hour + season + holiday + functioning_day,
  data = seoul_hourly
)

tibble(
  model = c("basic weather model", "richer weather + calendar model"),
  adj_r_squared = c(
    summary(seoul_basic)$adj.r.squared,
    summary(seoul_richer)$adj.r.squared
  ),
  sigma = c(summary(seoul_basic)$sigma, summary(seoul_richer)$sigma),
  aic = c(AIC(seoul_basic), AIC(seoul_richer))
)
```

The richer model has higher adjusted $R^2$, lower residual standard error, and lower AIC in this sample. These are in-sample comparisons, so they do not establish that the richer model will predict future bike demand better.

We can still compare a handful of candidate models by hand. Once there are many plausible predictors, that becomes tedious very quickly. In the next lecture we look at ways to automate that search, and at a different strategy where we keep a larger model but shrink its coefficients rather than repeatedly adding and removing variables.
