
 

## Automated Variable Selection Procedures

Automated variable selection procedures usually work as follows:

1. Choose a starting model, often either the model with no covariates or the model with all covariates included.
2. Check whether there is an advantage in adding or removing covariates.
3. Repeat this process until there is no clear advantage in changing the model.

## Forward Model Selection

In this technique:

1. Fit all one-variable models, and choose the one with the smallest p-value, provided at least one of the simple linear regression models is significant at the chosen significance level.
2. Next, fit all two-variable models *that include the first chosen X-variable*. Choose the variable whose regression coefficient is most significant, provided its p-value is below the chosen cutoff.
3. Then fit all three-variable models *that include the two variables already chosen*, and continue in the same way until there are no more significant variables to add.
4. Stop when no further variable meets the entry criterion.

The criterion for "significance" can be chosen by the user. For example, one may decide to only include variables with p-value less than the usual significance level \(\alpha = 0.05\).

### When to stop

Sometimes it is useful to allow a higher p-value for entry into the model in forward selection, such as p-value < \(\alpha = 0.25\). This can help us avoid overlooking potentially useful variables. In some settings it is better to cast the net a little wider at first, and then remove weak predictors later.

To decide whether a variable should enter the model, we can look at two equivalent quantities. One option is to check whether the t-value for that variable's regression coefficient is significant. For example, when `df = 60`, the t-value will be significant at \(\alpha = 0.05\) if \(|t| > 2\). Because we are comparing two nested models, we can also use the `anova()` function, for example `anova(model_1, model_2)`, and check whether the F statistic is significant with p-value < 0.05. The two approaches are equivalent when we add just one variable, because \(F = t^2\) where \(F \sim F_{1,df}\).

## Backwards Variable Selection

1. Start with a model containing all possible explanatory variables, and possibly interactions as well.
2. For each variable in turn, investigate the effect of removing that variable, or interaction if any, from the current model.
3. Remove the least informative variable or interaction, unless the term is supplying significant information about the response.
4. Go to step 2. Stop only if all variables or interactions in the current model are important.

## Stepwise (direction ="both")

The phrase *stepwise variable selection* is usually taken to mean that we start with forward selection, but after each variable is added we check whether any variables already in the model have fallen below the significance criterion \(\alpha\). If so, we remove them. One approach is to use *F* tests, or equivalently *t* tests.

### Example: Hours of sunshine (backwards)

Use backwards elimination on a full model of `Sun` from the `climate` data.

`r xfun::embed_file("../data/climate.csv")` 

```{r read Climate data 30, eval=-1, echo=-2}
climate = read.csv("climate.csv",header=TRUE,row.names=1)
climate = read.csv("../data/climate.csv",header=TRUE,row.names=1)
climate.lm1= lm(Sun ~ . , data=climate)
summary(climate.lm1)
```

#### Drop Lat

We would drop the variable with the highest p-value, in this case `Lat`. Only drop one variable at a time. We can use the `update()` function to save typing.

```{r update}
climate.lm2 = update(climate.lm1, . ~ .-Lat)
summary(climate.lm2)
```

#### Drop MnJlyTemp

The highest p-value is now for `MnJlyTemp`, so we update the model to drop that variable as well. Instead of printing the whole summary, we can use the `drop1()` function to help decide which variable to drop next.

```{r update 2}
climate.lm3 = update(climate.lm2, . ~ .-MnJlyTemp)
drop1(climate.lm3, test="F")
```

The F test indicates that `Height` has p-value = 0.08, so we can drop that one as well.

#### Drop Height

```{r update 3}
climate.lm4 = update(climate.lm3, . ~ .-Height)
drop1(climate.lm4, test="F")
```

Now `Sea` has p-value = 0.15, so we drop that as well.

#### Drop Sea

```{r update 4}
climate.lm5 = update(climate.lm4, . ~ .-Sea)
drop1(climate.lm5, test="F")
```

Finally we drop `Rain`, which has p-value = 0.10.

#### Drop Rain

```{r update 5 }
climate.lm6 = update(climate.lm5, . ~ .-Rain)
summary(climate.lm6)
```

We see that annual hours of sunshine are greater in places with high longitude, that is, on the eastern side of New Zealand, and in places with high mean January temperature. The model also suggests, perhaps surprisingly, that hours of sunshine are lower in the North Island. However, `NorthIsland` and `Longitude` are correlated, so that apparent effect should be interpreted with care.

## Forward selection

Now try forward selection instead. Do we finish with the same model?

```{r stepwise}
climate.null= lm(Sun ~ 1, data = climate)
add1(climate.null,  scope = ~. -Lat+Long+MnJanTemp+MnJlyTemp +Height+Rain+Sea+NorthIsland, test="F")
```

#### Forward Step 1: Add MnJanTemp

```{r}
climate.s1 = update(climate.null, .~.+MnJanTemp)
add1(climate.s1,  scope = ~Lat+Long+MnJanTemp+MnJlyTemp +Height+Rain+Sea+NorthIsland, test="F")
```

#### Forward Step 2: Add Rain

```{r}
climate.s2 = update(climate.s1, .~.+Rain)
add1(climate.s2, scope = ~Lat+Long+MnJanTemp+MnJlyTemp +Height+Rain+Sea+NorthIsland, test="F")
```

There are no additional variables to add.

#### Forward Selection Final Model

```{r}
summary(climate.s2)
```

The model chosen by forward selection is different from the one chosen by backwards elimination.

#### Forwards vs Backwards Comparison

Sometimes the two approaches lead to the same model, but in this example they do not. The reason is that two variables can be significant when they appear **together** in a regression, even if neither is quite strong enough to enter one at a time during forward selection. This is one reason why backwards elimination can sometimes work better.

## Model Selection using AIC

An alternative approach to model selection is to define a numerical measure of model quality and then choose the model that scores best by that measure. The so-called information criteria are examples of such measures. We will look at information criteria in general, the Akaike information criterion (AIC) in particular, and how they can be used for model selection.

### Information Criteria

A good regression model should do two things:

1. fit the data well;
2. remain reasonably simple, with as few predictors as possible.

So we can measure model quality using a quantity of the form
$$\mbox{Measure of badness of fit} + k \times \mbox{Number of predictors} + \mbox{constant}$$
where *k* controls the balance between fit and simplicity. Any additive constant is unimportant for model comparison, because it cancels out when we compare different models.

### Akaike Information Criterion

Several information criteria in statistics have this general form. One of the most widely used is the Akaike Information Criterion (AIC). For a linear model with unknown error variance, it is defined by
$$AIC = n \log (RSS/n) + 2p + \mbox{constant}.$$
When comparing models, we prefer the one with the **smaller** AIC. This is not the only criterion in common use, but it is perhaps the most common.

### Back to the Climate data

We have found two different models, `climate.lm6` and `climate.s2`. They have different numbers of variables, and they are not even nested. Even so, we can compare them using the adjusted \(R^2\) or the `AIC()` function. Both criteria suggest that `climate.lm6` is preferred.

```{r comparing}
summary(climate.lm6)$adj.r.square
summary(climate.s2)$adj.r.square
AIC(climate.lm6)
AIC(climate.s2)
```

#### Comparing Models with extractAIC()

An alternative to `AIC()` is the `extractAIC()` command. Its output gives the number of regression parameters, \((p+1)\), and the model's AIC.

```{r comparing 2}
extractAIC(climate.lm6)
extractAIC(climate.s2)
```

### AIC in Stepwise Variable Selection

We have seen that sequential variable selection can be carried out using a sequence of *F* tests. Another option is to compare models using AIC. At each step of the algorithm, we move to the model with the smallest AIC among those formed by adding or dropping a single variable. We stop only when no single addition or deletion reduces AIC further. This procedure is implemented in R by the `step()` command.

```{r step}
climate.null <- lm(Sun ~ 1, data=climate)
climate.full <- lm(Sun ~ Lat + Long + MnJanTemp + MnJlyTemp + Rain + Height + Sea + NorthIsland, data=climate)
scope <- list(lower = formula(climate.null), upper=formula(climate.full))
```

#### Stepwise Search with step()

```{r step1}
climate.step <- step(climate.null, scope = scope, direction = "both")
```

#### Stepwise Model Summary

```{r stepcont}
summary(climate.step)
```

Notice that AIC is more liberal than the F test here, and it has allowed a non-significant variable into the model.

### Backwards elimination via step() and AIC

The syntax below shows how to carry out backwards elimination. This also leaves a non-significant variable in the model, although it gives a higher adjusted R-square.

```{r step2}
climate.step2 <- step(climate.full, scope = scope, direction="backward")
```

#### Backwards AIC Model Summary

```{r}
summary(climate.step2)
```

**Question:  Which of these models is preferred?**

## Running Example Bridge: Stepwise Search on Seoul Bike Data

The Seoul bike data give us a more modern example of the same problem. Suppose we want to predict hourly demand from weather and calendar information, but we do not want to hand-build every candidate model.

```{r seoul stepwise read, message=FALSE}
seoul_hourly = read_csv("../data/seoul_bike_hourly.csv")
```

```{r seoul stepwise setup}
seoul_subset = seoul_hourly |>
  select(
    rented_bike_count,
    hour,
    season,
    holiday,
    functioning_day,
    temperature,
    humidity,
    wind_speed,
    visibility,
    dew_point_temperature,
    solar_radiation,
    rainfall,
    snowfall
  )

seoul_null = lm(rented_bike_count ~ 1, data = seoul_subset)
seoul_full = lm(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)
seoul_scope = list(lower = formula(seoul_null), upper = formula(seoul_full))
```

```{r seoul add1 preview}
add1(seoul_null, scope = formula(seoul_full), test = "F")
```

```{r seoul step preview}
seoul_step = step(seoul_null, scope = seoul_scope, direction = "both", trace = 0)
formula(seoul_step)
```

This is exactly the kind of setting where automated search looks tempting:

- there are many plausible predictors,
- some are related to each other,
- and the design can quickly become too large to compare by hand.

The next question is whether we should search by adding and dropping whole terms, or whether we should keep a richer model and shrink the coefficients.

## Penalised Regression: What Actually Happens to the Coefficients?

Stepwise regression makes discrete decisions. A term is either in the model or out of the model. Penalised regression uses a different strategy. We start with a larger model, but discourage the model from using large coefficients. For ordinary least squares, we minimise \[RSS = \sum_{i=1}^{n} (y_i - \hat y_i)^2.\] For penalised regression, we minimise \[RSS + \text{penalty}.\] The penalty is controlled by a tuning parameter \(\lambda\). When \(\lambda = 0\), there is no penalty and we get ordinary least squares. When \(\lambda\) is large, the coefficients are pulled more strongly toward zero. The intercept is usually not penalised. The penalty is applied to the slope coefficients.

## A Clean Case: Orthogonal Predictors

The clearest way to see ridge and lasso is to consider a simple case where the predictors are orthogonal. Orthogonal predictors means that the predictors are uncorrelated with each other in the model matrix. In matrix terms, the columns of \(X\) are perpendicular. Suppose also that the predictors have been centred and scaled so that \[\sum_{i=1}^{n} x_{ij}^{2} = n.\] This is not always true in real data, but it gives a useful clean result.

## Ridge Regression in the Orthogonal Case

Ridge regression minimises \[\sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij} \right)^2 + \lambda \sum_{j=1}^{p}\beta_j^2.\] The penalty uses squared coefficients. This is called an \(L_2\) penalty. In the orthogonal case, the ridge estimate for each slope coefficient is \[\hat\beta^{\text{ridge}}_j = \frac{n}{n+\lambda} \hat\beta^{\text{OLS}}_j.\] This is the main ridge intuition. Each ordinary least squares coefficient is multiplied by a number between 0 and 1. If \(\lambda = 0\), then \(\frac{n}{n+\lambda} = 1\), so ridge gives the ordinary least squares coefficient. If \(\lambda\) is large, then \(\frac{n}{n+\lambda}\) is closer to 0, so the coefficient is shrunk toward zero. Ridge regression therefore shrinks coefficients, but it usually does not make them exactly zero.

## Lasso Regression in the Orthogonal Case

Lasso regression minimises \[\sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p}\beta_j x_{ij} \right)^2 + \lambda \sum_{j=1}^{p}|\beta_j|.\] The penalty uses absolute coefficient values. This is called an \(L_1\) penalty. In the orthogonal case, the lasso estimate is a soft-thresholded version of the ordinary least squares estimate: \[\hat\beta^{\text{lasso}}_j = \text{sign}(\hat\beta^{\text{OLS}}_j) \left( |\hat\beta^{\text{OLS}}_j| - \frac{\lambda}{2n} \right)_+,\] where \((a)_+ = \max(a, 0)\). This formula says: 1. take the ordinary least squares coefficient; 2. reduce its magnitude; 3. if it becomes small enough, set it exactly to zero. That is why lasso can do variable selection. Ridge multiplies coefficients by a shrinkage factor. Lasso subtracts from the absolute size of the coefficient and truncates at zero.

## Why the Geometry Matters

For ridge regression, the constraint has the form \(\beta_1^2 + \beta_2^2 \leq c\). In two dimensions, this is a circle. For lasso regression, the constraint has the form \(|\beta_1| + |\beta_2| \leq c\). In two dimensions, this is a diamond, or a square rotated by 45 degrees. The lasso diamond has corners on the axes. Those corners occur where one coefficient is exactly zero. Because the lasso constraint has corners, the best solution often lands on an axis. When that happens, one coefficient is exactly zero. The ridge constraint is smooth, so the solution is usually shrunk toward zero but not exactly onto an axis. This is the geometric reason that ridge shrinks, while lasso can shrink and select.

```{r ridge lasso geometry, fig.width=8, fig.height=4, message=FALSE, warning=FALSE}
library(tidyverse)

theta = seq(0, 2 * pi, length.out = 500)

ridge_boundary = tibble(
  beta_1 = cos(theta),
  beta_2 = sin(theta),
  method = "Ridge: circular L2 constraint"
)

lasso_boundary = tibble(
  beta_1 = c(seq(0, 1, length.out = 100),
             seq(1, 0, length.out = 100),
             seq(0, -1, length.out = 100),
             seq(-1, 0, length.out = 100)),
  beta_2 = c(seq(1, 0, length.out = 100),
             seq(0, -1, length.out = 100),
             seq(-1, 0, length.out = 100),
             seq(0, 1, length.out = 100)),
  method = "Lasso: diamond L1 constraint"
)

constraint_plot = bind_rows(ridge_boundary, lasso_boundary)

ggplot(constraint_plot, aes(x = beta_1, y = beta_2)) +
  geom_path(linewidth = 1) +
  facet_wrap(~ method) +
  coord_equal() +
  labs(
    x = expression(beta[1]),
    y = expression(beta[2]),
    title = "Ridge and lasso constraints"
  ) +
  theme_minimal()
```

## Why Standardisation Matters

Before ridge or lasso, predictors should usually be standardised. The reason is that the penalty is applied to the coefficient values. Suppose one predictor is measured in metres and another is measured in millimetres. The numerical sizes of their coefficients will be different just because of the measurement scale. Without standardisation, the penalty would not treat the predictors fairly. The `glmnet` package standardises predictors by default using `standardize = TRUE`. This is usually what we want for ridge and lasso.

## Using `glmnet`

The `glmnet` package fits penalised regression models. It expects the predictors as a numeric matrix and the response as a vector. From this point on, we are treating the problem mainly as prediction, not formal hypothesis testing.

```{r seoul penalised setup, message=FALSE, warning=FALSE}
library(glmnet)
library(tidyverse)

X = model.matrix(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)[, -1]

y = seoul_subset$rented_bike_count
```

The `model.matrix()` function converts the regression formula into the numeric matrix needed by `glmnet`. It also converts factors into dummy variables. The argument `alpha` controls the penalty. `alpha = 0` gives ridge regression. `alpha = 1` gives lasso regression. Values between 0 and 1 give elastic net regression.

## Ridge Regression in `glmnet`

```{r seoul ridge cv, message=FALSE, warning=FALSE}
set.seed(42)

seoul_ridge = cv.glmnet(
  x = X,
  y = y,
  alpha = 0,
  standardize = TRUE
)

plot(seoul_ridge, main = "Ridge regression")
```

The cross-validation chooses the amount of shrinkage. The value `seoul_ridge$lambda.min` gives the \(\lambda\) with the smallest cross-validated prediction error. The value `seoul_ridge$lambda.1se` gives a more strongly penalised model whose error is still within one standard error of the minimum. We can extract the ridge coefficients.

```{r seoul ridge coefs, include=FALSE}
ridge_coef_min = coef(seoul_ridge, s = "lambda.min")
ridge_coef_1se = coef(seoul_ridge, s = "lambda.1se")

ridge_coef_table = tibble(
  coefficient = rownames(ridge_coef_min),
  ridge_lambda_min = as.numeric(ridge_coef_min),
  ridge_lambda_1se = as.numeric(ridge_coef_1se)
)
```

```{r seoul ridge coefs display, echo=FALSE}
ridge_coef_table
```

Ridge usually keeps all predictors in the model, but reduces the size of the coefficients.

## Lasso Regression in `glmnet`

```{r seoul lasso cv, message=FALSE, warning=FALSE}
set.seed(42)

seoul_lasso = cv.glmnet(
  x = X,
  y = y,
  alpha = 1,
  standardize = TRUE
)

plot(seoul_lasso, main = "Lasso regression")
```

The lasso can set some coefficients exactly to zero.

```{r seoul lasso coefs, include=FALSE}
lasso_coef_min = coef(seoul_lasso, s = "lambda.min")
lasso_coef_1se = coef(seoul_lasso, s = "lambda.1se")

lasso_coef_table = tibble(
  coefficient = rownames(lasso_coef_min),
  lasso_lambda_min = as.numeric(lasso_coef_min),
  lasso_lambda_1se = as.numeric(lasso_coef_1se),
  selected_min = lasso_lambda_min != 0,
  selected_1se = lasso_lambda_1se != 0
)
```

```{r seoul lasso coefs display, echo=FALSE}
lasso_coef_table
```

Count the number of selected predictors, excluding the intercept.

```{r seoul lasso counts, include=FALSE}
nonzero_min = sum(lasso_coef_min[-1] != 0)
nonzero_1se = sum(lasso_coef_1se[-1] != 0)
```

```{r seoul lasso counts display, echo=FALSE}
cat("Number of non-zero coefficients at lambda.min:", nonzero_min, "\n")
cat("Number of non-zero coefficients at lambda.1se:", nonzero_1se, "\n")
```

List the selected predictors for each choice.

```{r seoul lasso selected, include=FALSE}
selected_min = rownames(lasso_coef_min)[as.vector(lasso_coef_min != 0)]
selected_min = selected_min[selected_min != "(Intercept)"]

selected_1se = rownames(lasso_coef_1se)[as.vector(lasso_coef_1se != 0)]
selected_1se = selected_1se[selected_1se != "(Intercept)"]
```

```{r seoul lasso selected display, echo=FALSE}
cat("Selected at lambda.min:\n")
selected_min
cat("\nSelected at lambda.1se:\n")
selected_1se
```

The `lambda.min` model is chosen to minimise cross-validated prediction error. The `lambda.1se` model is chosen to be more regularised while still having similar prediction error. If `lambda.min` keeps nearly every variable, that is not a failure of lasso. It means that, for this dataset and this set of predictors, cross-validation found that the less sparse model predicted better.

## Coefficient Path Plot

A coefficient path plot shows how the lasso coefficients change as \(\lambda\) varies.

```{r seoul lasso path, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
plot(seoul_lasso$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(seoul_lasso$lambda.min), lty = 2, col = "red")
abline(v = log(seoul_lasso$lambda.1se), lty = 3, col = "blue")
legend("topright",
       legend = c("lambda.min", "lambda.1se"),
       lty = c(2, 3), col = c("red", "blue"))
```

Each line in the plot shows one coefficient. The vertical lines show the two common choices of \(\lambda\). At large \(\lambda\) (left side of the plot), many coefficients are zero. As \(\lambda\) decreases, more coefficients become non-zero.

```{r seoul lasso counts table, include=FALSE}
lasso_counts = tibble(
  lambda_choice = c("lambda.min", "lambda.1se"),
  lambda = c(seoul_lasso$lambda.min, seoul_lasso$lambda.1se),
  nonzero_predictors = c(
    sum(coef(seoul_lasso, s = "lambda.min")[-1] != 0),
    sum(coef(seoul_lasso, s = "lambda.1se")[-1] != 0)
  )
)
```

```{r seoul lasso counts table display, echo=FALSE}
lasso_counts
```

## Comparing Stepwise and Lasso

Stepwise regression and lasso are both automated procedures, but they answer slightly different questions. Stepwise regression asks: which terms should be added or removed to improve the model selection criterion? Lasso asks: how much should the coefficients be shrunk, and which coefficients become zero under that amount of shrinkage? For the Seoul bike data, the lasso at `lambda.min` may keep many or all predictors. That makes it a poor example if our only goal is to show variable selection. But it is a good example for a more realistic point: lasso does not guarantee a tiny model. It produces sparse models only when the cross-validated penalty is strong enough to push some coefficients to zero. In some datasets, many predictors really do contain predictive information. In that case, ridge, lasso, and stepwise selection may give different-looking models because they are controlling complexity in different ways. Let us compare the stepwise formula with the lasso-selected predictors.

```{r seoul stepwise vs lasso, include=FALSE}
step_terms = attr(terms(seoul_step), "term.labels")

lasso_coef_1se = coef(seoul_lasso, s = "lambda.1se")

lasso_selected_coefficients = rownames(lasso_coef_1se)[
  as.vector(lasso_coef_1se != 0)
]

lasso_selected_coefficients = lasso_selected_coefficients[
  lasso_selected_coefficients != "(Intercept)"
]

# Map dummy-coded lasso columns back to model terms for comparison
# season -> season, functioning_day -> functioning_day, etc.
lasso_selected_terms = unique(
  gsub("^(season|holiday|functioning_day).*$", "\\1",
       lasso_selected_coefficients)
)

in_both = intersect(step_terms, lasso_selected_terms)
only_stepwise = setdiff(step_terms, lasso_selected_terms)
only_lasso = setdiff(lasso_selected_terms, step_terms)
```

```{r seoul stepwise vs lasso display, echo=FALSE}
cat("Stepwise model terms:\n")
print(step_terms)
cat("\nLasso selected coefficients at lambda.1se:\n")
print(lasso_selected_coefficients)
cat("\n--- Comparison (mapping lasso dummies back to model terms) ---\n")
cat("\nIn both stepwise and lasso:\n")
print(in_both)
cat("\nOnly in stepwise (not selected by lasso):\n")
print(only_stepwise)
cat("\nOnly in lasso (not in stepwise formula):\n")
print(only_lasso)
```

Be careful when comparing stepwise and lasso here. Stepwise works with model terms. For example, `season` is one model term. `glmnet` works with columns of the model matrix. A factor such as `season` becomes several dummy-variable columns. So the lasso count is a count of selected columns, not necessarily a count of selected scientific variables.

## MSE Comparison Across Methods

We can compare the fitted performance of each method by computing the mean squared error on the training data.

```{r seoul mse comparison, include=FALSE}
# Stepwise model predictions
seoul_step_pred = predict(seoul_step, seoul_subset)

# Ridge predictions
ridge_pred_min = predict(seoul_ridge, newx = X, s = "lambda.min")
ridge_pred_1se = predict(seoul_ridge, newx = X, s = "lambda.1se")

# Lasso predictions
lasso_pred_min = predict(seoul_lasso, newx = X, s = "lambda.min")
lasso_pred_1se = predict(seoul_lasso, newx = X, s = "lambda.1se")

# Compute MSE and RMSE
mse_comparison = tibble(
  Method = c("Stepwise (AIC)",
             "Ridge (lambda.min)",
             "Ridge (lambda.1se)",
             "Lasso (lambda.min)",
             "Lasso (lambda.1se)"),
  MSE = c(
    mean((seoul_subset$rented_bike_count - seoul_step_pred)^2),
    mean((seoul_subset$rented_bike_count - ridge_pred_min)^2),
    mean((seoul_subset$rented_bike_count - ridge_pred_1se)^2),
    mean((seoul_subset$rented_bike_count - lasso_pred_min)^2),
    mean((seoul_subset$rented_bike_count - lasso_pred_1se)^2)
  ),
  RMSE = sqrt(MSE),
  `Non-zero predictors` = c(
    length(coef(seoul_step)) - 1,
    sum(coef(seoul_ridge, s = "lambda.min")[-1] != 0),
    sum(coef(seoul_ridge, s = "lambda.1se")[-1] != 0),
    sum(coef(seoul_lasso, s = "lambda.min")[-1] != 0),
    sum(coef(seoul_lasso, s = "lambda.1se")[-1] != 0)
  )
)
```

```{r seoul mse comparison display, echo=FALSE}
mse_comparison = mse_comparison |>
  mutate(
    RMSE = sqrt(MSE),
    delta_MSE = MSE - min(MSE),
    delta_RMSE = RMSE - min(RMSE)
  )

mse_comparison
```

The first thing to notice is that the training MSE values are very similar. The RMSE column puts the errors in the original units: bikes per hour. All five models have an RMSE around 432 bikes per hour. The differences between them are only a few bikes per hour. The lowest training RMSE is for lasso at `lambda.min` (432.3 bikes/hour), but it is almost identical to the stepwise AIC model (432.4 bikes/hour). The difference of 0.1 bikes per hour is practically meaningless. The ridge models have slightly higher RMSE (433.6 and 437.3 bikes/hour). This is expected, because ridge deliberately shrinks coefficients. It is not trying to minimise the training RSS as aggressively as ordinary least squares. Ridge accepts a small loss in training fit in exchange for potentially more stable predictions on new data. The lasso at `lambda.min` keeps all 14 model-matrix predictors. In this example, lasso at `lambda.min` is therefore not giving a sparse model. It is mainly acting as a shrinkage method. The lasso at `lambda.1se` is more interesting for variable selection. It keeps 11 predictors and has an RMSE of 436.1 bikes/hour, only about 4 bikes/hour higher than the best model. This illustrates the usual trade-off: a simpler model may have slightly worse fitted performance, but may be easier to interpret and may generalise better. The number of non-zero predictors should be interpreted carefully. Ridge regression usually does not set coefficients exactly to zero, so the ridge models keep all predictors. Also, `glmnet` works with columns of the model matrix, whereas stepwise regression works with model terms. A categorical variable can therefore count as one term in stepwise regression but several dummy-variable columns in `glmnet`. The main conclusion is not that one method clearly wins. Instead, the table shows that several different modelling strategies can give very similar fitted accuracy, while producing different levels of shrinkage and different apparent model complexity. In practice, we would prefer to compare these methods using a held-out test set or cross-validated prediction error, rather than training MSE. Training MSE is useful for illustration, but it is not a fair estimate of future predictive performance. The Seoul bike example shows that penalised regression does not automatically give a dramatically better or much smaller model. Its value is that it gives a principled way to trade off fit, shrinkage, stability, and sparsity.

## Key Takeaway

Penalised regression gives us another way to control model complexity. Instead of searching through many candidate models, we fit a richer model and penalise large coefficients. Ridge regression uses an \(L_2\) penalty and shrinks coefficients toward zero. Lasso regression uses an \(L_1\) penalty and can shrink coefficients exactly to zero. In `glmnet`, ridge is fitted with `alpha = 0`, lasso with `alpha = 1`, and the amount of shrinkage is controlled by \(\lambda\), usually chosen by cross-validation.

