```{r setup, include=FALSE}
library(tidyverse)
```

## From manual model building to automated search

In Lecture 27, we built candidate models by hand. That works when there are only a few plausible predictors. It becomes awkward very quickly when the set of possible models grows.

At that point, we need a rule for how to move through a larger model space. There are two useful ways to respond.

1. Search through candidate models by adding or dropping terms.
2. Keep a larger model and shrink its coefficients.

This lecture is about those two strategies.

## Part I: automated model search

### Forward, backward, and stepwise

The basic idea is simple.

- Forward selection starts small and adds one term at a time.
- Backward elimination starts large and removes one term at a time.
- Stepwise selection allows moves in either direction.

This matters because the model-building unit is the term, not an arbitrary coefficient. A four-level factor is one model term even though it corresponds to several coefficients. An interaction is also one term in the model formula. That is why the criterion should be phrased in terms of adding or dropping terms, not hunting for the coefficient with the smallest p-value.

That is the same idea we stressed in Lecture 27: select terms, not arbitrary coefficients.

### What criterion do we use?

For a nested comparison, a partial F test is still a good choice. For a broader search, AIC is often more convenient. Lecture 27 introduced AIC as a fit-versus-complexity criterion, so we can use it directly. If two candidates differ only slightly in AIC, the lower value alone does not settle which model we should use.

In practice, a stepwise algorithm evaluates models one term away from the current model and moves to the model with the lower AIC if that reduces the criterion. This is exactly what `step()` does in R.

### Climate example: path dependence matters

The `climate` data are small enough that we can watch the search move.

```{r read-Climate-data-30, message=FALSE}
climate <- read.csv("../data/climate.csv", header = TRUE, row.names = 1)
climate <- climate |>
  drop_na(Sun, Lat, Long, MnJanTemp, MnJlyTemp, Rain, Height, Sea, NorthIsland)
```

A full model for annual sunshine is

```{r climate-full}
climate_null <- lm(Sun ~ 1, data = climate)
climate_full <- lm(
  Sun ~ Lat + Long + MnJanTemp + MnJlyTemp + Rain + Height + Sea + NorthIsland,
  data = climate
)
climate_scope <- list(
  lower = formula(climate_null),
  upper = formula(climate_full)
)
```

Let us run the search from both ends of the candidate set. Forward selection starts with an intercept and asks which term to add next. Backward elimination starts with the full model and asks which term to remove. Each step is local: the algorithm compares models it can reach from its current position.

```{r climate-backward-check}
climate_forward <- step(
  climate_null,
  scope = climate_scope,
  direction = "forward",
  trace = 0
)

climate_backward <- step(
  climate_full,
  scope = climate_scope,
  direction = "backward",
  trace = 0
)

climate_forward$anova
climate_backward$anova
formula(climate_forward)
formula(climate_backward)
c(
  forward_AIC = climate_forward$anova$AIC[nrow(climate_forward$anova)],
  backward_AIC = climate_backward$anova$AIC[nrow(climate_backward$anova)]
)
```

Compare the paths and final formulas before looking at any P-values. If they differ, the starting point has changed the answer. If they agree, that is reassuring about this particular search, but it does not prove that the resulting formula is the best model among every possible combination. We still have to decide whether the predictors and assumptions make sense for the scientific question.

Here the forward search retains `MnJanTemp`, `Rain`, `Sea` and `Height` (AIC 371.040), while the backward search retains `Long`, `MnJanTemp`, `Rain`, `Height`, `Sea` and `NorthIsland` (AIC 368.958). The searches end at different formulas because each follows a local path through the candidate set.

## Stepwise search on the Seoul data

The Seoul bike data give us a more realistic version of the same problem. We have many weather and calendar variables, and a large number of plausible models.

```{r seoul-stepwise-read, message=FALSE}
seoul_hourly <- read_csv("../data/seoul_bike_hourly.csv", show_col_types = FALSE) |>
  mutate(hour = factor(hour))
```

`hour` records a position in the day. We will treat it as a factor so each hour can have its own mean shift, rather than forcing demand to change in one straight line from midnight to 11 pm. A CSV does not preserve R's factor class, so we convert it after import. The same coding must be used when we build the model matrix for ridge and lasso.

```{r seoul-stepwise-setup}
seoul_subset <- seoul_hourly |>
  select(
    rented_bike_count,
    hour,
    season,
    holiday,
    functioning_day,
    temperature,
    humidity,
    wind_speed,
    visibility,
    dew_point_temperature,
    solar_radiation,
    rainfall,
    snowfall
  ) |>
  drop_na()

seoul_null <- lm(rented_bike_count ~ 1, data = seoul_subset)
seoul_full <- lm(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)
seoul_scope <- list(lower = formula(seoul_null), upper = formula(seoul_full))
```

```{r seoul-step-preview}
seoul_step <- step(seoul_null, scope = seoul_scope, direction = "both", trace = 0)
formula(seoul_step)
```

This is the kind of setting where automated search is useful. We do not want to fit and compare every plausible model by hand. But the search still makes a hard decision: a term is either in or out.

The next question is whether we want to search through many smaller models at all. Another approach is to keep a richer model and shrink the coefficients.

## Part II: shrinkage instead of searching

Stepwise regression makes discrete decisions. Penalised regression takes a different approach. We keep the larger model, but we add a penalty that discourages large coefficient values.

For ordinary least squares we minimise

$$RSS = \sum_{i=1}^{n} (y_i - \hat y_i)^2.$$

For penalised regression we minimise

$$RSS + \lambda \times \text{penalty}.$$

The tuning parameter \(\lambda\) controls the strength of the penalty. When \(\lambda = 0\), we get ordinary least squares. As \(\lambda\) increases, the coefficients are pulled more strongly toward zero. The intercept is usually not penalised.

The two main choices are:

- Ridge: penalty uses squared coefficients, \(\lambda \sum \beta_j^2\)
- Lasso: penalty uses absolute coefficients, \(\lambda \sum |\beta_j|\)

The practical distinction is straightforward:

- ridge shrinks coefficients toward zero, usually retaining all predictors;
- lasso shrinks coefficients and can set some exactly to zero.

That gives us a second way to control complexity: not by selecting a model, but by shrinking the fitted coefficients.

### Why standardisation matters

The penalty is applied to coefficient size, so predictors measured on different scales are treated differently unless we standardise them first. This is one reason why ridge and lasso are usually fitted after standardising the predictors.

The `glmnet` package does this by default when `standardize = TRUE`.

## Using `glmnet` on the Seoul data

```{r seoul-penalised-setup, message=FALSE, warning=FALSE}
library(glmnet)

X = model.matrix(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)[, -1]

y = seoul_subset$rented_bike_count

set.seed(42)
foldid <- sample(rep(seq_len(10), length.out = nrow(X)))
```

The argument `alpha` controls the penalty:

- `alpha = 0` gives ridge regression;
- `alpha = 1` gives lasso regression;
- values between 0 and 1 give elastic net.

The amount of shrinkage is chosen by cross-validation rather than by hand.

### Ridge regression

```{r seoul-ridge-cv, message=FALSE, warning=FALSE}
seoul_ridge = cv.glmnet(
  x = X,
  y = y,
  alpha = 0,
  foldid = foldid,
  standardize = TRUE
)

plot(seoul_ridge, main = "Ridge regression")
```

`lambda.min` is the penalty with the lowest average error across these held-out folds. `lambda.1se` is the largest penalty whose estimated error is within one standard error of that minimum. The latter accepts an estimated error close to the minimum in exchange for a more strongly regularised model. The CV curve selects a penalty from this dataset; it is not an untouched final test of the selected model's performance.

```{r seoul-ridge-summary}
coef(seoul_ridge, s = "lambda.min")
coef(seoul_ridge, s = "lambda.1se")
```

Ridge usually keeps all predictors in the model, but makes the coefficients smaller and therefore more stable.

### Lasso regression

```{r seoul-lasso-cv, message=FALSE, warning=FALSE}
seoul_lasso = cv.glmnet(
  x = X,
  y = y,
  alpha = 1,
  foldid = foldid,
  standardize = TRUE
)

plot(seoul_lasso, main = "Lasso regression")
```

The lasso can set some coefficients exactly to zero. That is the main difference from ridge.

```{r seoul-lasso-summary}
coef(seoul_lasso, s = "lambda.min")
coef(seoul_lasso, s = "lambda.1se")
```

### A coefficient path plot

```{r seoul-lasso-path, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
plot(seoul_lasso$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(seoul_lasso$lambda.min), lty = 2, col = "red")
abline(v = log(seoul_lasso$lambda.1se), lty = 3, col = "blue")
legend("topright",
       legend = c("lambda.min", "lambda.1se"),
       lty = c(2, 3), col = c("red", "blue"))
```

As \(\lambda\) grows, more coefficients get pushed toward zero. For lasso, some coefficients can actually become exactly zero. That is why lasso is often described as doing both shrinkage and selection.

## Stepwise search versus shrinkage

These methods solve different problems.

| Method | Main idea | What it does |
|---|---|---|
| Stepwise search | Add or remove terms | Produces a small model by searching through candidate formulas |
| Ridge | Penalise large coefficients | Shrinks all coefficients, usually keeping every predictor |
| Lasso | Penalise absolute coefficients | Shrinks coefficients and can set some to zero |

This is more than a technical distinction. It reflects different modelling goals.

- Stepwise selection is natural when we want a smaller, interpretable model.
- Ridge is often attractive when prediction is the goal and predictors overlap.
- Lasso sits between the two: it can select predictors while also shrinking the remaining coefficients.

The important conceptual point is that these are not just different algorithms for the same job. They are different ways of controlling model complexity.

We also need to be careful when comparing them. Stepwise search works with model terms. `glmnet` works with the columns of the model matrix. A factor such as `season` may correspond to several dummy-variable columns. So a lasso coefficient count is not always a count of scientific variables in the same sense as a stepwise term count.

Neither a stepwise coefficient P-value nor the chosen lasso set has the same inferential interpretation it would have had if that model had been specified before looking at the outcome.

There is one limitation to the Seoul example. Hours close together share weather, calendar conditions and demand patterns. Randomly splitting rows across folds allows the training and validation sets to contain near neighbours. That is enough to illustrate how `cv.glmnet()` chooses a penalty, but it is not a credible test of forecasting future days. Once time becomes the modelling question, validation must keep later observations out of the training set.

## Why this leads naturally to Lecture 29

Ridge is especially useful when predictors carry overlapping information. Ordinary least squares coefficients can then become unstable, hard to interpret, and sensitive to small changes in the data.

That is the multicollinearity problem we turn to next.

## Summary

By the end of this lecture, the message is simple.

There are two broad ways to handle a large collection of plausible predictors:

1. Search through nested or near-nested models by adding and dropping terms.
2. Keep a richer model and shrink the coefficients instead.

Forward, backward, and stepwise methods are examples of the first idea. Ridge and lasso are examples of the second. The choice depends on the scientific goal, the prediction problem, and the structure of the predictors themselves.
