```{r setup, include=FALSE}
library(tidyverse)
```

## From manual model building to automated search

In Lecture 27, we built candidate models by hand. That works when there are only a few plausible predictors. It becomes awkward very quickly when the set of possible models grows.

At that point, we need a rule for how to move through a larger model space. There are two useful ways to respond.

1. Search through candidate models by adding or dropping terms.
2. Keep a larger model and shrink its coefficients.

This lecture is about those two strategies.

## Part I: automated model search

### Forward, backward, and stepwise

The basic idea is simple.

- Forward selection starts small and adds one term at a time.
- Backward elimination starts large and removes one term at a time.
- Stepwise selection allows moves in either direction.

This matters because the model-building unit is the term, not an arbitrary coefficient. A four-level factor is one model term even though it corresponds to several coefficients. An interaction is also one term in the model formula. That is why the criterion should be phrased in terms of adding or dropping terms, not hunting for the coefficient with the smallest p-value.

That is the same idea we stressed in Lecture 27: select terms, not arbitrary coefficients.

### What criterion do we use?

For a nested comparison, a partial F test is still a good choice. For a broader search, AIC is often more convenient. We do not need to re-derive AIC here. Lecture 27 already introduced it as a fit-versus-complexity criterion. We can use it directly.

In practice, a stepwise algorithm evaluates models one term away from the current model and moves to the model with the lower AIC if that reduces the criterion. This is exactly what `step()` does in R.

### Climate example: path dependence matters

The `climate` data are small enough that we can watch the search move.

```{r read Climate data 30, eval=-1, echo=-2, message=FALSE}
climate = read.csv("climate.csv", header = TRUE, row.names = 1)
climate = read.csv("../data/climate.csv", header = TRUE, row.names = 1)
```

A full model for annual sunshine is

```{r climate full}
climate_full = lm(Sun ~ Lat + Long + MnJanTemp + MnJlyTemp + Rain + Height + Sea + NorthIsland, data = climate)
```

Backwards elimination asks which term is least useful to remove. Forward selection asks which term is most useful to add.

```{r climate backward check}
drop1(climate_full, test = "F")
```

We could do the whole sequence by hand, but the important lesson is not the exact list of updates. It is that the final model depends on where the search starts and which direction it follows.

Forward search and backward search can end at different models because they are not exploring the same path through the model space. This is a problem of path dependence: the search is constrained by the route it takes, not by some unique best model that everyone would discover eventually.

That is the key point. A stepwise model is not the uniquely best model. It is a model generated by a particular search strategy.

## Stepwise search on the Seoul data

The Seoul bike data give us a more realistic version of the same problem. We have many weather and calendar variables, and a large number of plausible models.

```{r seoul stepwise read, message=FALSE}
seoul_hourly = read_csv("../data/seoul_bike_hourly.csv")
```

```{r seoul stepwise setup}
seoul_subset = seoul_hourly |>
  select(
    rented_bike_count,
    hour,
    season,
    holiday,
    functioning_day,
    temperature,
    humidity,
    wind_speed,
    visibility,
    dew_point_temperature,
    solar_radiation,
    rainfall,
    snowfall
  )

seoul_null = lm(rented_bike_count ~ 1, data = seoul_subset)
seoul_full = lm(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)
seoul_scope = list(lower = formula(seoul_null), upper = formula(seoul_full))
```

```{r seoul step preview}
seoul_step = step(seoul_null, scope = seoul_scope, direction = "both", trace = 0)
formula(seoul_step)
```

This is the kind of setting where automated search is useful. We do not want to fit and compare every plausible model by hand. But the search still makes a hard decision: a term is either in or out.

The next question is whether we want to search through many smaller models at all. Another approach is to keep a richer model and shrink the coefficients.

## Part II: shrinkage instead of searching

Stepwise regression makes discrete decisions. Penalised regression takes a different approach. We keep the larger model, but we add a penalty that discourages large coefficient values.

For ordinary least squares we minimise

$$RSS = \sum_{i=1}^{n} (y_i - \hat y_i)^2.$$

For penalised regression we minimise

$$RSS + \lambda \times \text{penalty}.$$

The tuning parameter \(\lambda\) controls the strength of the penalty. When \(\lambda = 0\), we get ordinary least squares. As \(\lambda\) increases, the coefficients are pulled more strongly toward zero. The intercept is usually not penalised.

The two main choices are:

- Ridge: penalty uses squared coefficients, \(\lambda \sum \beta_j^2\)
- Lasso: penalty uses absolute coefficients, \(\lambda \sum |\beta_j|\)

The practical distinction is straightforward:

- ridge shrinks coefficients toward zero, usually retaining all predictors;
- lasso shrinks coefficients and can set some exactly to zero.

That gives us a second way to control complexity: not by selecting a model, but by shrinking the fitted coefficients.

### Why standardisation matters

The penalty is applied to coefficient size, so predictors measured on different scales are treated differently unless we standardise them first. This is one reason why ridge and lasso are usually fitted after standardising the predictors.

The `glmnet` package does this by default when `standardize = TRUE`.

## Using `glmnet` on the Seoul data

```{r seoul penalised setup, message=FALSE, warning=FALSE}
library(glmnet)

X = model.matrix(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    temperature + humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_subset
)[, -1]

y = seoul_subset$rented_bike_count
```

The argument `alpha` controls the penalty:

- `alpha = 0` gives ridge regression;
- `alpha = 1` gives lasso regression;
- values between 0 and 1 give elastic net.

The amount of shrinkage is chosen by cross-validation rather than by hand.

### Ridge regression

```{r seoul ridge cv, message=FALSE, warning=FALSE}
set.seed(42)

seoul_ridge = cv.glmnet(
  x = X,
  y = y,
  alpha = 0,
  standardize = TRUE
)

plot(seoul_ridge, main = "Ridge regression")
```

`lambda.min` gives the penalty that minimises cross-validated prediction error. `lambda.1se` gives a more regularised model whose error is still close to the minimum.

```{r seoul ridge summary}
coef(seoul_ridge, s = "lambda.min")
coef(seoul_ridge, s = "lambda.1se")
```

Ridge usually keeps all predictors in the model, but makes the coefficients smaller and therefore more stable.

### Lasso regression

```{r seoul lasso cv, message=FALSE, warning=FALSE}
set.seed(42)

seoul_lasso = cv.glmnet(
  x = X,
  y = y,
  alpha = 1,
  standardize = TRUE
)

plot(seoul_lasso, main = "Lasso regression")
```

The lasso can set some coefficients exactly to zero. That is the main difference from ridge.

```{r seoul lasso summary}
coef(seoul_lasso, s = "lambda.min")
coef(seoul_lasso, s = "lambda.1se")
```

### A coefficient path plot

```{r seoul lasso path, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
plot(seoul_lasso$glmnet.fit, xvar = "lambda", label = TRUE)
abline(v = log(seoul_lasso$lambda.min), lty = 2, col = "red")
abline(v = log(seoul_lasso$lambda.1se), lty = 3, col = "blue")
legend("topright",
       legend = c("lambda.min", "lambda.1se"),
       lty = c(2, 3), col = c("red", "blue"))
```

As \(\lambda\) grows, more coefficients get pushed toward zero. For lasso, some coefficients can actually become exactly zero. That is why lasso is often described as doing both shrinkage and selection.

## Stepwise search versus shrinkage

These methods solve different problems.

| Method | Main idea | What it does |
|---|---|---|
| Stepwise search | Add or remove terms | Produces a small model by searching through candidate formulas |
| Ridge | Penalise large coefficients | Shrinks all coefficients, usually keeping every predictor |
| Lasso | Penalise absolute coefficients | Shrinks coefficients and can set some to zero |

This is more than a technical distinction. It reflects different modelling goals.

- Stepwise selection is natural when we want a smaller, interpretable model.
- Ridge is often attractive when prediction is the goal and predictors overlap.
- Lasso sits between the two: it can select predictors while also shrinking the remaining coefficients.

The important conceptual point is that these are not just different algorithms for the same job. They are different ways of controlling model complexity.

We also need to be careful when comparing them. Stepwise search works with model terms. `glmnet` works with the columns of the model matrix. A factor such as `season` may correspond to several dummy-variable columns. So a lasso coefficient count is not always a count of scientific variables in the same sense as a stepwise term count.

## Why this leads naturally to Lecture 29

The reason ridge matters so much in practice is that predictors often carry overlapping information. When predictors are strongly related, ordinary least squares coefficients become unstable, hard to interpret, and sensitive to small changes in the data.

That is the core problem of multicollinearity. We will look at it next, because it is exactly the setting where shrinkage methods are most useful.

## Summary

By the end of this lecture, the message is simple.

There are two broad ways to handle a large collection of plausible predictors:

1. Search through nested or near-nested models by adding and dropping terms.
2. Keep a richer model and shrink the coefficients instead.

Forward, backward, and stepwise methods are examples of the first idea. Ridge and lasso are examples of the second. The choice depends on the scientific goal, the prediction problem, and the structure of the predictors themselves.
