
```{r ExtraPkgs, include=FALSE}
library(car)
```


## Collinearity and orthogonality

Before worrying about diagnostics, look at the geometry of the predictors.

```{r collinearity-geometry-29, echo=FALSE, fig.width=10, fig.height=4.5}
set.seed(42)
n_demo <- 80
x_demo <- scale(rnorm(n_demo), scale = FALSE)[, 1]
z_seed <- rnorm(n_demo)
z_orth <- residuals(lm(z_seed ~ x_demo))
z_col <- x_demo + rnorm(n_demo, sd = 0.18)

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(x_demo, z_orth, pch = 19,
     xlab = "Predictor X", ylab = "Predictor Z",
     main = "Different information")
abline(lm(z_orth ~ x_demo), lwd = 2)

plot(x_demo, z_col, pch = 19,
     xlab = "Predictor X", ylab = "Predictor Z",
     main = "Overlapping information")
abline(lm(z_col ~ x_demo), lwd = 2)
par(old_par)
```

::: incremental
### Orthogonal predictors

When the centred predictor columns are orthogonal, one predictor contains no linear information about the other. Their coefficient estimates can be separated cleanly.

### Collinear predictors

When predictors move together, they contain overlapping information. The regression can still fit the response well, but it becomes harder to separate their individual conditional coefficients.
:::

A little overlap is normal. The problem becomes important when predictors are so strongly related that the individual coefficient estimates become unstable.

## What does collinearity do to a model?

Consider

$$
Y = \beta_0 + \beta_1 A + \beta_2 B + \varepsilon,
$$

where both $A$ and $B$ genuinely contribute to the response.

```{r read-Simulation1, eval=-1, echo=-2, message=FALSE}
sim = read_csv("simulation1.csv")
sim = read_csv("../data/simulation1.csv")

sim_final <- sim |>
  mutate(Y = 50 + 5*A + 5*B + epsilon)
```

```{r correlated-vs-orthogonal-29, echo=FALSE, fig.width=10, fig.height=4.5}
set.seed(42)
sim_rand <- sim_final |>
  mutate(
    B_rand = sample(B),
    Y_rand = 50 + 5*A + 5*B_rand + epsilon
  )

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
plot(sim_final$A, sim_final$B, pch = 19,
     xlab = "A", ylab = "B", main = "Correlated predictors")
abline(lm(B ~ A, data = sim_final), lwd = 2)

plot(sim_rand$A, sim_rand$B_rand, pch = 19,
     xlab = "A", ylab = "B", main = "Correlation broken")
abline(lm(B_rand ~ A, data = sim_rand), lwd = 2)
par(old_par)
```

## Models with correlated predictors

```{r compare-correlated-models-29}
lm_cor <- lm(Y ~ A + B, data = sim_final)
lm_uncor <- lm(Y_rand ~ A + B_rand, data = sim_rand)

bind_rows(
  tidy(lm_cor) |> mutate(Model = "Correlated predictors"),
  tidy(lm_uncor) |> mutate(Model = "Correlation broken")
) |>
  select(Model, term, estimate, std.error)
```

The response noise has not become smaller. What changed is the relationship between the predictors.

With correlated predictors, the model has less independent information for deciding how much of the response belongs to $A$ and how much belongs to $B$. The individual standard errors increase and the coefficient estimates become more sensitive to small changes in the data.

## Exact dependence: coefficients are not estimable

Strong correlation is inconvenient. **Exact** linear dependence is more fundamental.

Suppose

$$
Z = \alpha_0 + \alpha_1 X.
$$

Then $Z$ contains no information that is not already determined by the intercept and $X$. There are infinitely many coefficient combinations that give exactly the same fitted values.

R shows us this directly.

```{r exact-collinearity-29}
exact_dat <- tibble(
  x = 1:8,
  z = 2 + 3 * x,
  y = c(4.1, 5.0, 6.2, 7.0, 8.1, 9.2, 10.0, 11.1)
)

exact_fit <- lm(y ~ x + z, data = exact_dat)
summary(exact_fit)
```

One coefficient is reported as `NA` because the design matrix is rank deficient. R can still calculate fitted values, but it cannot identify separate coefficients for all linearly dependent columns.

Near dependence is less obvious: R returns coefficients for all terms, but those coefficients can have large standard errors and change substantially when the sample or formula changes.
## Electricity data: correlated predictors in practice

Now move from the artificial example to household electricity bills. We will model `Bill` using household `Income`, the number of `Persons`, and floor `Area`.

```{r read-electric, eval=-1, echo=-2, message=FALSE}
electric = read_csv("electric.csv")
electric = read_csv("../data/electric.csv")
```

## Correlations among the predictors

A correlation matrix is a useful first check.

```{r electric-cor}
round(cor(electric), 3)
```

`Area` and `Income` are very strongly correlated. `Persons` overlaps much less with either one.

```{r electric-pairs-29, fig.width=9, fig.height=7, message=FALSE, warning=FALSE}
library(GGally)

ggpairs(
  electric,
  upper = list(continuous = wrap("cor", size = 4)),
  lower = list(continuous = wrap("points", alpha = 0.55, size = 1.4)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.35))
) +
  theme_bw(base_size = 12)
```

Pairwise correlation is a warning, not a decision rule. A predictor can also be well explained by several other predictors together even if no individual pairwise correlation looks extreme.

## Spotting multicollinearity in the fitted model

```{r electric-full-29}
electric_lm0 <- lm(Bill ~ Area + Persons + Income, data = electric)
summary(electric_lm0)
```

The model can explain substantial variation in electricity bills while the individual coefficients are much less stable.

What should make us suspicious?

::: incremental
- The overall regression can look useful while individual terms have large standard errors.
- Coefficient signs or magnitudes can change sharply when correlated predictors enter together.
- Small changes to the formula can produce large changes in the conditional coefficients.
:::

## Contradictory coefficient behaviour

First fit `Area` and `Income` separately, then together.

```{r electric-contradictory-29}
area_only <- lm(Bill ~ Area, data = electric)
income_only <- lm(Bill ~ Income, data = electric)
area_income <- lm(Bill ~ Area + Income, data = electric)

bind_rows(
  tidy(area_only) |> mutate(Model = "Area only"),
  tidy(income_only) |> mutate(Model = "Income only"),
  tidy(area_income) |> mutate(Model = "Area + Income")
) |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, std.error, p.value)
```

By themselves, both `Area` and `Income` are positively associated with `Bill`. When they enter together, the model is asking a much harder conditional question: what is the association with income among houses of the same area, and vice versa?

Because area and income move together so strongly, the data contain little independent information for those comparisons. A sign change should not be read as evidence that higher income causes lower electricity use.

Now compare two models that adjust the `Persons` coefficient in different ways.

```{r electric-persons-29}
area_persons <- lm(Bill ~ Area + Persons, data = electric)
income_persons <- lm(Bill ~ Income + Persons, data = electric)

bind_rows(
  tidy(area_persons) |> mutate(Model = "Area + Persons"),
  tidy(income_persons) |> mutate(Model = "Income + Persons")
) |>
  filter(term == "Persons") |>
  select(Model, term, estimate, std.error, p.value)
```

The coefficient for `Persons` changes because the two models condition on different household characteristics. This is not merely a numerical nuisance. With correlated predictors, the scientific interpretation of a coefficient depends strongly on what else is held fixed.

## Can we detect this systematically?

Pairwise plots are useful, but we want a diagnostic that asks a broader question:

> How well can each predictor be explained by all of the other predictors together?

That leads directly to variance inflation factors.
## Regressing predictors among themselves

For predictor $X_i$, regress it on all of the other predictors and call the resulting coefficient of determination $R_i^2$.

$$
X_i \sim X_1 + \cdots + X_{i-1} + X_{i+1} + \cdots + X_p.
$$

If $R_i^2$ is close to 1, most of the variation in $X_i$ is already explained by the other predictors. Very little independent variation remains for estimating its separate conditional coefficient.

The corresponding variance inflation factor is

$$
VIF_i = \frac{1}{1-R_i^2}.
$$

So if $R_i^2 = 0$, then $VIF_i = 1$: there is no inflation from linear overlap with the other predictors. As $R_i^2$ approaches 1, the variance inflation increases rapidly.

For a numerical predictor, we can write the coefficient variance as

$$
\widehat{\operatorname{var}}(\hat\beta_i)
=
\frac{s^2}{(n-1)\operatorname{var}(X_i)}
\times
\frac{1}{1-R_i^2}.
$$

The final factor is the VIF. It tells us how much the variance of that coefficient has been enlarged by linear overlap with the other predictors, holding the residual variance and spread of $X_i$ fixed.

## Computing VIFs

The `car` package provides `vif()`.

```{r vifs1-29}
car::vif(electric_lm0)
```

The large values for `Area` and `Income` match what we already saw in the pairwise plots and coefficient behaviour.

A VIF is a diagnostic, not an instruction to delete a variable. Whether the overlap matters depends on whether we need the separate coefficient, prediction is the main goal, and whether there is a scientifically defensible alternative representation.

## SAP example: a hidden constraint

The SAP data relate sales (`S`) to current and previous advertising (`A`, `A1`), current and previous promotion spending (`P`, `P1`), and sales expenses (`SE`).

The analyst did not know that four of the predictors were approximately constrained by a fixed budget:

$$
A + A1 + P + P1 \approx 5.
$$

```{r read-SAP, eval=-1, echo=-2, message=FALSE}
SAP = read_csv("SAP.csv")
SAP = read_csv("../data/SAP.csv")
```

## SAP: full model

```{r sap-full}
SAP_lm <- lm(S ~ A + P + SE + A1 + P1, data = SAP)
summary(SAP_lm)
```

The model is trying to estimate separate advertising and promotion coefficients from predictors that are close to a linear constraint.

## SAP: VIFs

```{r vif-SAP}
car::vif(SAP_lm)
```

The VIFs make the problem obvious. Apart from `SE`, the advertising and promotion variables contain very little independent information for separating their individual coefficients.

## SAP: drop one term and the coefficients move

```{r dropping-29}
SAP_lm2 <- lm(S ~ P + SE + A1 + P1, data = SAP)
SAP_lm3 <- lm(S ~ A + P + SE + P1, data = SAP)

bind_rows(
  tidy(SAP_lm2) |> mutate(Model = "Drop A"),
  tidy(SAP_lm3) |> mutate(Model = "Drop A1")
) |>
  filter(term != "(Intercept)") |>
  select(Model, term, estimate, std.error)
```

The coefficient pattern changes depending on which member of the constrained set is removed. That is exactly the instability we expect when the predictors are nearly linearly dependent.

```{r looking-in-29}
SAPA_lm <- lm(A ~ P + SE + A1 + P1, data = SAP)
summary(SAPA_lm)
```

The predictor regression reconstructs the hidden constraint: `A` is almost determined by the remaining advertising and promotion variables.

## Practical responses to multicollinearity

There is no single automatic fix. The response should match the modelling goal.

::: incremental
1. **Remove or redefine one of the overlapping predictors.** Prefer the variable with the stronger scientific interpretation, better measurement, or less missingness when the choice is defensible.

2. **Construct a meaningful combined quantity.** This might be an average, a standardised composite, a difference, or a scientifically meaningful ratio such as GDP per capita rather than GDP and population separately.

3. **Use penalised regression.** As we saw in Lecture 28, ridge uses an **L2** penalty and lasso uses an **L1** penalty. Ridge is particularly useful when we want to retain overlapping predictors but stabilise the fitted coefficient vector. Lasso can instead set some model-matrix coefficients to zero.
:::

Penalisation does not make correlated predictors independent. It changes the estimation problem so that unstable coefficient combinations are discouraged.

## Seoul bike data: the same problem in a realistic model

The Seoul weather predictors give us a useful example because temperature, humidity and dew point overlap strongly.

```{r seoul-multi-read-29, message=FALSE}
seoul_hourly <- read_csv("../data/seoul_bike_hourly.csv", show_col_types = FALSE)

seoul_weather <- seoul_hourly |>
  select(
    rented_bike_count, temperature, humidity, dew_point_temperature,
    solar_radiation, rainfall, snowfall
  ) |>
  drop_na()
```

```{r seoul-pairs-29, fig.height=7, fig.width=8, message=FALSE, warning=FALSE}
ggpairs(
  seoul_weather |>
    select(temperature, humidity, dew_point_temperature, solar_radiation),
  upper = list(continuous = wrap("cor", size = 4)),
  lower = list(continuous = wrap("points", alpha = 0.12, size = 0.4)),
  diag = list(continuous = wrap("densityDiag", alpha = 0.35))
) +
  theme_bw(base_size = 12)
```

## Seoul: OLS and VIFs

```{r seoul-multi-model-29}
seoul_weather_lm <- lm(
  rented_bike_count ~ temperature + humidity + dew_point_temperature +
    solar_radiation + rainfall + snowfall,
  data = seoul_weather
)

summary(seoul_weather_lm)
car::vif(seoul_weather_lm)
```

The model can still be useful as a prediction model, but the separate coefficients for highly overlapping weather measurements are conditional on holding almost redundant quantities fixed.

## Seoul: four ways to handle the same predictor set

We will use `lambda.1se`, following the convention from Lecture 28.

```{r seoul-penalised-compare-29, include=FALSE}
library(glmnet)

X_weather <- model.matrix(
  rented_bike_count ~ temperature + humidity + dew_point_temperature +
    solar_radiation + rainfall + snowfall,
  data = seoul_weather
)[, -1]

y_weather <- seoul_weather$rented_bike_count

set.seed(42)
foldid_weather <- sample(rep(seq_len(10), length.out = nrow(X_weather)))

ridge_weather <- cv.glmnet(
  X_weather, y_weather, alpha = 0, foldid = foldid_weather, standardize = TRUE
)

lasso_weather <- cv.glmnet(
  X_weather, y_weather, alpha = 1, foldid = foldid_weather, standardize = TRUE
)

ols_drop_dew <- lm(
  rented_bike_count ~ temperature + humidity + solar_radiation + rainfall + snowfall,
  data = seoul_weather
)

coef_to_tibble <- function(x, label) {
  tibble(term = rownames(x), value = as.numeric(x)) |>
    rename(!!label := value)
}

ols_tbl <- tibble(
  term = names(coef(seoul_weather_lm)),
  OLS = as.numeric(coef(seoul_weather_lm))
)

ridge_tbl <- coef_to_tibble(
  as.matrix(coef(ridge_weather, s = "lambda.1se")),
  "Ridge L2"
)

lasso_tbl <- coef_to_tibble(
  as.matrix(coef(lasso_weather, s = "lambda.1se")),
  "Lasso L1"
)

drop_tbl <- tibble(
  term = names(coef(ols_drop_dew)),
  `OLS, drop dew point` = as.numeric(coef(ols_drop_dew))
)

seoul_coef_compare <- ols_tbl |>
  full_join(ridge_tbl, by = "term") |>
  full_join(lasso_tbl, by = "term") |>
  full_join(drop_tbl, by = "term")
```

```{r seoul-penalised-compare-display-29, echo=FALSE}
knitr::kable(seoul_coef_compare, digits = 3)
```

These columns answer related but different modelling questions.

::: incremental
- **OLS** estimates the unpenalised conditional coefficients with all six weather predictors.
- **Ridge, L2** keeps the whole predictor set but regularises unstable coefficient combinations.
- **Lasso, L1** can shrink some coefficients exactly to zero.
- **Dropping dew point** changes the model itself and asks the remaining predictors to absorb information previously shared with dew point.
:::

The important distinction is between **prediction** and **coefficient interpretation**. Several models can predict similarly while assigning very different numerical coefficients to overlapping predictors.

## Looking ahead

We can now build models with numerical and categorical predictors, compare and search through candidate models, and deal more deliberately with overlapping predictors.

One limitation remains: our mean structure has still mostly been built from straight-line effects of numerical predictors.

Lecture 30 first deals with a different issue, observations that do not all carry the same precision. After that, we move to **LOESS and splines**, where the fitted mean itself can bend rather than being forced to follow a straight line.
