
```{r ExtraPkgs, include=FALSE}
library(car)
```


## Idea of Collinearity and Orthogonality

In multiple regression, we often use several explanatory variables at the same time. Sometimes these explanatory variables give us different kinds of information. For example, in a model for electricity bills, the number of people in a house and the area of the house may both be useful predictors. But sometimes explanatory variables overlap strongly. For example, household income and house area may be strongly related. Higher-income households may tend to live in larger houses. If both variables are included in the same model, the regression has to separate two predictors that are partly measuring the same underlying thing.

This problem is called **multicollinearity**. Multicollinearity occurs when two or more explanatory variables are strongly related to each other. It does not necessarily mean that the model predicts badly. The main problem is that the individual regression coefficients can become hard to interpret.

Consider the model \[Y  =  \beta_0 + \beta_1 X_1 + \beta_2X_2 + \varepsilon.\] We usually interpret \(\beta_1\) as the rate of change in \(Y\) for a one-unit increase in \(X_1\), holding \(X_2\) constant. That interpretation becomes difficult if \(X_1\) and \(X_2\) usually move together.

If \(X_1\) and \(X_2\) have no linear relationship, they are called **orthogonal**. Orthogonality makes regression easier to interpret because each predictor is contributing separate information. In practice, explanatory variables are rarely perfectly orthogonal unless the data come from a carefully designed experiment. A little non-orthogonality is usually not a problem.

When predictors overlap strongly, small changes to the model can shift their separate coefficients. We may see large standard errors or unexpected signs, even when the model as a whole is useful. We will use these examples to see when that instability matters, how to detect it, and what we can do about it.

Two cases need to be kept separate. With exact linear dependence, one predictor is an exact combination of the others. The model cannot identify every coefficient separately, although it may still identify fitted values or particular combinations of coefficients. With near dependence, the coefficients are identifiable, but their separate estimates can be very imprecise and can change markedly when the sample or formula changes. Neither case says, by itself, that the predictors are scientifically unnecessary or that the model cannot predict.

## A Numerical Example

We will consider the effects of multicollinearity using a simple artificial example. Suppose the true model is \[Y = 50 + 5A + 5B + \epsilon.\] Both \(A\) and \(B\) genuinely affect \(Y\). The issue is that \(A\) and \(B\) are highly correlated.

`r xfun::embed_file("../data/simulation1.csv")` 

```{r read-Simulation1, eval=-1, echo=-2, message=FALSE}
sim = read_csv("simulation1.csv")
sim = read_csv("../data/simulation1.csv")

sim_final <- sim |> mutate(Y = 50 + 5*A +5*B+ epsilon)
```

### Correlated A and B scatterplot

```{r, fig.height=5}
ggplot(sim_final) + 
  geom_point(mapping=aes(x=A, y=B, size=Y))
```

### Model with correlated predictors

```{r lm-Simulation1}
lm.cor = lm( Y ~ A + B, data=sim_final)
summary(lm.cor)
```

The model fits very well overall, but the standard errors for the individual coefficients are larger than we might expect. The reason is that \(A\) and \(B\) overlap. The model has difficulty separating the contribution of \(A\) from the contribution of \(B\).

### Breaking the correlation

Now suppose we keep the same values of \(A\), \(B\), and \(\epsilon\), but break the correlation between \(A\) and \(B\). One simple way to do this is to randomly reorder the values in column \(B\).

```{r Sim-randomized}
set.seed(42)

sim_rand <- sim_final |>
  mutate(B_rand = sample(B)) |>
  mutate(Y_rand = 50 + 5*A + 5*B_rand + epsilon)
```

Now \(A\) and `B_rand` contain the same numerical values as before, but the strong relationship between them has been broken.

```{r}
ggplot(sim_rand) + 
  geom_point(mapping=aes(x=A, y=B_rand, size=Y_rand))
```

### Model with uncorrelated predictors

```{r}
lm.uncor = lm( Y_rand ~ A + B_rand, data=sim_rand)
summary(lm.uncor)
```

The standard errors of the coefficients are now much smaller. The important point is that we did not make the response less noisy. We mainly changed the relationship between the predictors. When the predictors are less correlated, the model can estimate their separate effects more precisely.

### Orthogonal Data

Consider the model	$$Y  =  \beta_0 + \beta_1 X  + \beta_2 Z + \varepsilon$$  

If $X$ and $Z$ are orthogonal in the observed data, adding $Z$ does not change the estimated slope for $X$ or the sum of squares attributed to $X$. Under independent, constant-variance errors, the two slope estimates are uncorrelated; under normal errors they are independent. That makes the separate coefficients easier to interpret. We often see this in balanced designed experiments, but should not expect it automatically in observational data.

### Collinear data

Suppose we have three sets of numbers $X,Z,Y$ such that 
$$Y = \beta_0 +\beta_1 X  + \beta_2 Z +\varepsilon$$

but that, unknown to us,
$$Z = \alpha_0 + \alpha_1 X$$
exactly. We will consider near dependence separately.

We can rewrite the model with different coefficients: add $k$ copies of $\alpha_0 + \alpha_1 X$ and subtract $kZ$. Since the two are equal, this does not change the fitted value:

$$\begin{aligned}Y &= \beta_0 +\beta_1 X + \beta_2 Z + k(\alpha_0 + \alpha_1 X) - k Z  +\varepsilon\\ & = (\beta_0 + k\alpha_0) + (\beta_1 + k\alpha_1)X + (\beta_2 - k)Z + \varepsilon\end{aligned}$$

for any real number $k$.

There are infinitely many sets of coefficients with the same fitted values. With near dependence, a unique solution exists, but even a small change in the data can move the separate coefficients substantially.

#### Not estimable

Here the dependence is exact. Many sets of intercept and slope coefficients produce the same fitted values, so we cannot estimate the individual coefficients uniquely from these data. In less extreme examples the dependence is only approximate: the model can be fitted, but separating the coefficients requires information that the observed predictors contain very little of.

How do we assess whether it is a real problem in our data?

## Correlations among the predictors

Pairwise correlations are a useful first check. A strong correlation tells us that two predictors tend to move together, so their separate conditional coefficients may be hard to estimate. It does not tell us to discard either variable. The scientific question may require both, and a predictor can also be nearly explained by several others even when no single pairwise correlation is striking. We will look at both the correlation matrix and what the fitted model says about coefficient precision.

Consider household electricity bills with income, number of people and floor area as predictors. These variables can all be included when the question calls for them, but income and area move closely together in these data. We will check whether that overlap makes their separate conditional coefficients difficult to estimate.

`r xfun::embed_file("../data/electric.csv")` 

```{r read-electric, eval=-1, echo=-2, message=FALSE}
electric = read_csv("electric.csv")
electric = read_csv("../data/electric.csv")
electric
```

### Correlation matrix

We could check all pair-wise correlations with the `cor()` function:

```{r electric-cor}
cor(electric)
```

This indicates that the `Area` of the house is very strongly correlated with the `Income` of the occupants ($r=0.961$). By contrast, the number of persons in the house is less strongly related to `Income` ($r=0.143$) and `Area` ($r=0.366$), so it may provide additional information. A pairs plot can help us see these relationships.

### Pairs plot

```{r electric-pairs}
pairs(electric)
```

But how do we know this is causing real problems? 

### Spotting multicollinearity

```{r lm}
electric.lm0 = lm(Bill~ Area+ Persons + Income, data=electric)
summary(electric.lm0)
```

The overall F test gives evidence that the predictors jointly explain variation in bills, while the individual coefficient tests are much less precise. With the strong `Area`–`Income` correlation we saw above, overlap is a plausible reason. Let's look at what happens when we fit these predictors separately and together.

### Contradictory regression coefficients

Let's look at the variable we know are highly correlated (i.e. `Area` and `Income`) separately as predictors:

By themselves both variables are positively related to Bill and have small Std Errors, as we would expect given the pairs chart.

But let's see what happens if we put them together:

#### Area and Income together

```{r correlated-variables}
elec.lm3 <- lm(Bill ~ Area + Income, data=electric)
summary(elec.lm3)
```

Once `Area` and `Income` enter together, the income coefficient changes sign and both coefficients have larger standard errors. We should not read that sign change as evidence that a higher income reduces a household's bill. These data provide little independent variation in income after holding area fixed.

#### Area + Persons model

We could fit a smaller model, but the choice of predictor changes the question. Suppose we want to estimate the association between household size and the bill after adjusting for either area or income:

```{r pair-of-variables}
elec.lm4 = lm(Bill ~ Area + Persons, data=electric)
elec.lm5= lm(Bill ~ Income + Persons, data=electric)
summary(elec.lm4)
```

#### Income + Persons model

```{r}
summary(elec.lm5)
```

The estimated increase per additional person is about \$42 when we hold house area fixed, but about \$85 when we hold income fixed. These models compare different households: the first holds area constant, and the second holds income constant. That is a sizeable change, so we need to be clear about which comparison answers our question.

## Regressing the predictors among themselves

More generally, we can ask how well each predictor is explained by a combination of the other regressors. Let $R_i^2$ be the $R^2$ from regressing $X_i$ on $X_1, X_2, \ldots, X_{i-1}, X_{i+1}, \ldots, X_p$. A large value means that little variation in $X_i$ is left to estimate its separate conditional coefficient. It is more general than checking pairwise correlations, but no single cutoff decides whether that matters for the question at hand. This leads to the concept of variance inflation.

## Variance inflation factors, VIFs

A major impact of multicollinearity is found in the variances (and hence standard errors) of the regression coefficients. If $X_i$ is alone in the model $Y =\beta_0 +\beta_1 X_i +\varepsilon$, then $\hat{\mbox{var}}(\hat\beta_1) = \frac{s^2}{(n-1)\mbox{var}(X_i)}$. If $X_i$ is in a model with other predictors then $\hat{\mbox{var}}(\hat\beta_i) =  s^2   \left(X^T X\right)^{-1}_{ii}$ where the last term is the corresponding diagonal entry of the $(X^T X )^{-1}$ matrix, the *variance-covariance* matrix.

It turns out that we can rewrite this as: $\hat{\mbox{var}}(\hat\beta_i) = \frac{s^2}{(n-1)\mbox{var}(X_i)}\cdot \frac{1}{1 - R_i^2}$ where $R_i^2$ is the multiple $R^2$ of the regression of $X_i$ on the other predictors. Thus, when going from a model with just $X_i$ to a full model, the variance of the coefficient of $X_i$ is multiplied by $\mbox{VIF}_i = \frac{1}{1 - R_i^2}$.

VIF describes how much the variance of one coefficient is enlarged by linear overlap with the other predictors, with the error variance and the spread of that predictor held fixed. A high VIF flags limited information for that *separate conditional effect*. It is a diagnostic, not an instruction to delete the predictor. Whether we change the model depends on whether we need that coefficient, the available data and the modelling goal.

### Computing VIFs

To obtain VIFs we need to invoke `library(car)`.

```{r vifs1}
library(car)
```

```{r}
vif( electric.lm0)
```

The VIFs flag overlap in the information carried by `Area` and `Income`. Whether to change the formula depends on the question: we may need both predictors, or we may decide that their separate coefficients are not the target of this analysis.

What we do depends on the question. We might retain both predictors, choose one with a stronger scientific justification, or construct a meaningful combined measure.

## Example: SAP data

The SAP example from Chatterjee and Price relates sales (`S`) to current and previous advertising (`A`, `A1`), current and previous promotion spending (`P`, `P1`), and sales expenses (`SE`). The four advertising and promotion variables were subject to an approximate budget constraint:

$$ A + A1 + P + P1  =  5  $$

The advertising executives knew this, but the analyst did not. The constraint makes it difficult to separate the conditional coefficients for those four predictors.

`r xfun::embed_file("../data/SAP.csv")` 

```{r read-SAP, eval=-1, echo=-2, message=FALSE}
SAP = read_csv("SAP.csv")
SAP = read_csv("../data/SAP.csv")
SAP
```

### SAP full model

We start with a linear model for sales in terms of all other variables:

```{r sap-full}
SAP.lm = lm( S ~ A + P + SE + A1 + P1, data=SAP)
summary(SAP.lm)
```

### SAP VIFs

We see that, apart from Sales Expenses, there is only weak
evidence of a relationship between advertising, promotions and sales. 

```{r vif-SAP}
vif(SAP.lm)
```

The VIF shows high multicollinearity for all advertising and promotions data. 

### Dropping variables from SAP

If we drop one of the variables the others change their coefficients unpredictably  depending on which variable we drop. 

```{r dropping}
SAP.lm2= lm( S ~  P + SE +A1 + P1, data=SAP)
SAP.lm3= lm( S ~ A +P +SE  + P1, data=SAP)
SAP.lm2 |> tidy()
SAP.lm3 |> tidy()
```

### Hidden constraint revealed

Suppose we look at the relationships among the predictor variables: 

```{r looking-in}
SAPA.lm = lm( A ~ P + SE + A1 + P1, data=SAP)
SAPA.lm |> tidy()
```

This is approximately A = 5 - P -A1 - P1 as in the hidden constraint.

## What do we do about multicollinearity?

In this course, we will focus on three practical responses to multicollinearity.

1. We can remove one (or more) of the correlated variables from the regression. Our aim is to fix the problem, but sometimes we have a bit of choice which variable to remove, in which case we would like the model to be as interpretable as possible. Sometimes the choice is between a variable with many missing values and one with few, and so we choose the latter if it allows us a bigger effective sample size $n$.

2. We could construct another variable which is a combination of the two most highly correlated variables. It might be an ordinary average $(X_1+X_2)/2$, or an average of standardized versions of the $X_i$s or some other linear combination. Sometimes subtracting or dividing one variable by another will help. For example the incidence of a particular disease such as obesity may be higher in countries with high GDP and high Population. But GDP depends on the size of the economy, which is correlated with the Population. Dividing gives us per capita GDP which is a better indicator of average wealth and hence may be more related to disease incidence. 

3. We can use penalised regression methods. This links back to the previous lecture.

## Link back to penalised regression

In the previous lecture, we introduced ridge and lasso regression. Both methods add a penalty to the regression fit. Ridge uses an \(L_2\) penalty: \(\lambda \sum_{j=1}^{p}\beta_j^2\). Lasso uses an \(L_1\) penalty: \(\lambda \sum_{j=1}^{p}|\beta_j|\). Both can be useful when predictors overlap, but they behave differently. Ridge keeps the predictors in the model and shrinks their coefficients toward zero. This can help when several predictors carry related information and we do not want the fitted coefficients to become too unstable. Lasso can shrink some coefficients exactly to zero. This can help when we want a smaller selected set of predictors, but it may choose one variable from a group of correlated variables and discard another similar variable. So, when the problem is mainly coefficient instability from overlapping predictors, ridge is often the more natural first example. When the goal is variable selection, lasso is often more natural. Neither method makes correlated observational predictors independent. They are modelling tools, not substitutes for better study design.

## Running example revisited: Seoul bike predictors

Last lecture we used ridge and lasso to control model complexity. We now return to the Seoul weather predictors to ask a different question: how much information do they provide about each separate coefficient? Temperature, humidity and dew point overlap, so the model may describe demand reasonably well while being uncertain about which one carries a particular conditional association.

```{r seoul-multi-read, message=FALSE}
seoul_hourly = read_csv("../data/seoul_bike_hourly.csv")
```

```{r seoul-selected-correlations, include=FALSE}
seoul_weather = seoul_hourly |>
  select(
    temperature,
    humidity,
    wind_speed,
    visibility,
    dew_point_temperature,
    solar_radiation,
    rainfall,
    snowfall
  )
```

```{r seoul-cor-display, echo=FALSE}
round(cor(seoul_weather), 2)
```

The correlation matrix shows that some weather predictors are strongly related. For example, temperature and dew point are strongly correlated. Humidity and visibility are also related. This matters because the regression model tries to estimate the separate effect of each predictor while holding the others constant. That is difficult when the predictors tend to move together.

A pairs plot gives the same idea visually.

```{r seoul-pairs, fig.height=7, fig.width=8, message=FALSE, warning=FALSE}
library(GGally)

seoul_pairs_data = seoul_hourly |>
  transmute(
    `Bike count` = rented_bike_count,
    `Temperature` = temperature,
    `Humidity` = humidity,
    `Dew point` = dew_point_temperature,
    `Solar radiation` = solar_radiation
  )

ggpairs(
  seoul_pairs_data,
  upper = list(
    continuous = wrap(
      "cor",
      size = 4
    )
  ),
  lower = list(
    continuous = wrap(
      "points",
      alpha = 0.12,
      size = 0.4
    )
  ),
  diag = list(
    continuous = wrap(
      "densityDiag",
      alpha = 0.35
    )
  )
) +
  theme_bw(base_size = 12) +
  labs(
    title = "Seoul bike demand and overlapping weather predictors",
    subtitle = "Strong correlations indicate predictors that may cause unstable regression coefficients"
  )
```

Now fit an ordinary linear model using only the weather predictors.

```{r seoul-multi-model}
seoul_weather.lm = lm(
  rented_bike_count ~ temperature + humidity + dew_point_temperature +
    solar_radiation + rainfall + snowfall,
  data = seoul_hourly
)
summary(seoul_weather.lm)
```

This model may still be useful for prediction, but the individual weather coefficients should be interpreted cautiously. Weather is still relevant, but the predictors overlap so strongly that their separate coefficients are hard to interpret.

We can check this more formally using variance inflation factors.

```{r seoul-vifs}
vif(seoul_weather.lm)
```

Large VIF values tell us that some predictors are strongly explained by the other predictors.

The same Seoul bike model from the previous lecture is not just a model-selection example. It is also a **multicollinearity example**. In the previous lecture, ridge and lasso were introduced as penalised regression methods. Here, we can now see another reason they are useful. Ridge regression is helpful when we want to keep overlapping predictors in the model but reduce coefficient instability. Lasso regression is helpful when we want to shrink coefficients and possibly remove some predictors. These methods respond to the same problem in different ways. Ridge keeps the overlapping predictors and shrinks their coefficients, whereas lasso may shrink some coefficients all the way to zero. This is why ridge is often discussed in connection with multicollinearity. It is not mainly a variable-selection method. It is a *coefficient-stabilisation method*.

The important interpretation is: the Seoul bike predictors can give similar prediction accuracy under several methods, but the ordinary regression coefficients may still be unstable because the predictors overlap. Prediction and coefficient interpretation are not the same problem.

### Recall from Lecture 28

Last lecture we used `alpha = 0` for ridge and `alpha = 1` for lasso, and compared `lambda.min` with `lambda.1se`. Here we fit ridge and ordinary least squares to the same six weather predictors. Let's compare their coefficients at `lambda.min`.

```{r seoul-ols-ridge-compare, include=FALSE}
library(glmnet)

X_weather = model.matrix(
  rented_bike_count ~ temperature + humidity + dew_point_temperature +
    solar_radiation + rainfall + snowfall,
  data = seoul_hourly
)[, -1]

y_weather = seoul_hourly$rented_bike_count

set.seed(42)
seoul_ridge_weather = cv.glmnet(X_weather, y_weather, alpha = 0, standardize = TRUE)

ols_weather = lm(rented_bike_count ~ temperature + humidity + dew_point_temperature +
                   solar_radiation + rainfall + snowfall, data = seoul_hourly)

ridge_coef = coef(seoul_ridge_weather, s = "lambda.min")

coef_compare = tibble(
  Coefficient = rownames(ridge_coef),
  `OLS estimate` = as.numeric(coef(ols_weather)),
  `Ridge estimate (lambda.min)` = as.numeric(ridge_coef)
)
```

```{r seoul-ols-ridge-compare-display, echo=FALSE}
coef_compare
```

`glmnet` standardises predictors internally while fitting, then reports coefficients on their original scales. The two columns are therefore on compatible scales.

Ridge penalises the *whole coefficient vector*. With orthogonal predictors, each coefficient simply moves towards zero as the penalty grows. With correlated predictors, the coefficients are estimated jointly and can rebalance. An individual coefficient may even move away from zero relative to its OLS estimate. Judge the change in the collection of coefficients and the fitted predictions; do not expect every entry in this table to be a smaller copy of its OLS counterpart. Ridge can reduce instability, but it does not create independent information about highly overlapping predictors.

Ridge and lasso both penalise large coefficients. That penalty can reduce the instability caused by multicollinearity. Ridge usually keeps all predictors and shrinks them. Lasso can shrink some predictors exactly to zero. So the same fitted models from the previous lecture can be interpreted in a second way: as tools for controlling model complexity; and as tools for reducing coefficient instability when predictors overlap.

So far we have asked how much each *predictor* contributes. Next we change a different part of the problem: some observed responses are measured more precisely than others.
