
## Splines and LOESS: Flexible Regression Curves

So far we have focused on parametric regression models, where the mean response is described by a small number of coefficients. In this lecture we use the Seoul Bike Sharing data to study a different problem: the relationship between demand and weather is clearly nonlinear.

By the end of this lecture, students should be able to:

1. Explain why a straight-line mean function can be too restrictive.
2. Use LOESS to explore nonlinear structure.
3. Fit regression splines using basis functions.
4. Explain the difference between ordinary and natural splines.
5. Explain how degrees of freedom control the flexibility of a regression spline.

### Running Example: Hourly Bike Demand

`r xfun::embed_file("../data/seoul_bike_hourly.csv")`

```{r read-seoul-hourly, message=FALSE}
seoul_hourly = read_csv("../data/seoul_bike_hourly.csv")
seoul_hourly
```

The core question for this lecture is:

> How does hourly bike demand change with temperature?

### Why Go Beyond Straight Lines?

A straight-line model can underfit. A highly flexible curve can overfit. Our aim is to describe the mean relationship without losing the regression framing.

```{r seoul-temperature-scatter, warning=FALSE, message=FALSE}
seoul_hourly |>
  ggplot(aes(x = temperature, y = rented_bike_count)) +
  geom_point(alpha = 0.06) +
  labs(
    x = "Temperature (C)",
    y = "Hourly rented bike count",
    title = "Bike demand changes nonlinearly with temperature"
  )
```

## Example 1: LOESS

### The LOESS Idea

LOESS fits many small local regressions instead of one global line.

1. Choose a target point.
2. Borrow nearby observations.
3. Fit a weighted local regression.

### LOESS with the Seoul Bike Data

We have already fitted polynomials and piecewise regression models. A polynomial uses one global curve, so changing its coefficients can alter the fit across the entire range. A piecewise model lets the shape change at selected predictor values. LOESS takes a more local approach: at each temperature, it fits a small regression using nearby observations, with closer observations contributing more. It is a way to inspect the shape before committing to a particular regression formula.

```{r loess-seoul, warning=FALSE, message=FALSE}
seoul_hourly |>
  ggplot(aes(x = temperature, y = rented_bike_count)) +
  geom_point(alpha = 0.05) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    span = 0.15,
    se = FALSE,
    linewidth = 1,
    aes(colour = "Span = 0.15", alpha = "Span = 0.15")
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    span = 0.45,
    se = FALSE,
    linewidth = 1,
    aes(colour = "Span = 0.45", alpha = "Span = 0.45")
  ) +
  scale_colour_manual(values = c("Span = 0.15" = "green4", "Span = 0.45" = "blue")) +
  scale_alpha_manual(values = c("Span = 0.15" = 0.8, "Span = 0.45" = 0.8)) +
  labs(
    x = "Temperature (C)",
    y = "Hourly rented bike count",
    colour = "LOESS span",
    alpha = "LOESS span",
    title = "LOESS span controls the smoothness of the fitted curve"
  )
```

### The Span Parameter

- `span` controls how local the fit is.
- Small span gives a more flexible but noisier curve.
- Large span gives a smoother and more stable curve.

The smaller span follows local changes more closely, including random fluctuations. The larger span smooths those fluctuations but may miss a sharp change in the mean. Let's check which features persist across both curves before choosing a regression formula.

## Example 2: Splines

### The Key Conceptual Point

A spline model is still linear in its coefficients:

$$y = \beta_0 + \beta_1 b_1(x) + \beta_2 b_2(x) + \cdots + \beta_k b_k(x) + \varepsilon.$$

It is nonlinear in the original predictor, but linear in the unknown coefficients. That means we can still use standard regression tools.

### Linear, Spline, and Natural Spline Fits

A regression spline joins polynomial pieces at *knots*, with constraints that keep the fitted curve smooth. R constructs basis functions for those pieces and `lm()` estimates their coefficients. The result is curved as a function of temperature, but still linear in the unknown coefficients. The `df = 5` argument controls the size of the basis and therefore the permitted flexibility; it does not mean a fifth-degree polynomial. We should compare curves at the same predictor values before judging whether that flexibility helps.

```{r spline-seoul-models, warning=FALSE, message=FALSE}
library(splines)

seoul_temp = seoul_hourly |>
  select(temperature, rented_bike_count) |>
  filter(is.finite(temperature), is.finite(rented_bike_count))

lm_linear = lm(rented_bike_count ~ temperature, data = seoul_temp)
lm_bs = lm(rented_bike_count ~ bs(temperature, df = 5), data = seoul_temp)
lm_ns = lm(rented_bike_count ~ ns(temperature, df = 5), data = seoul_temp)
spline_knots <- attr(bs(seoul_temp$temperature, df = 5), "knots")

temp_grid = tibble(
  temperature = seq(min(seoul_temp$temperature), max(seoul_temp$temperature), length.out = 250)
)

predictions = temp_grid |>
  mutate(
    Linear = predict(lm_linear, newdata = temp_grid),
    `B-spline` = predict(lm_bs, newdata = temp_grid),
    `Natural spline` = predict(lm_ns, newdata = temp_grid)
  ) |>
  pivot_longer(
    cols = c(Linear, `B-spline`, `Natural spline`),
    names_to = "model",
    values_to = "fitted_bikes"
  )

seoul_temp |>
  ggplot(aes(x = temperature, y = rented_bike_count)) +
  geom_point(alpha = 0.05) +
  geom_vline(
    xintercept = spline_knots,
    colour = "grey45",
    linetype = "dashed",
    linewidth = 0.5
  ) +
  geom_line(
    data = predictions,
    aes(x = temperature, y = fitted_bikes,
        colour = model, linetype = model, alpha = model),
    linewidth = 1
  ) +
  scale_colour_manual(
    values = c("Linear" = "red", "B-spline" = "blue", "Natural spline" = "forestgreen")
  ) +
  scale_linetype_manual(
    values = c("Linear" = "solid", "B-spline" = "dotted", "Natural spline" = "dotdash")
  ) +
  scale_alpha_manual(
    values = c("Linear" = 0.5, "B-spline" = 1, "Natural spline" = 1)
  ) +
  labs(
    x = "Temperature (C)",
    y = "Hourly rented bike count",
    colour = "Model",
    linetype = "Model",
    alpha = "Model",
    title = "Spline pieces join smoothly at the marked knots",
    caption = "Dashed guides mark the internal B-spline knots. All fitted curves use the same observations and temperature range."
  )
```

### Natural Splines vs Ordinary Splines

```{r spline-boundary, warning=FALSE, message=FALSE}
predictions |>
  filter(model %in% c("B-spline", "Natural spline")) |>
  ggplot(aes(x = temperature, y = fitted_bikes, colour = model)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Temperature (C)",
    y = "Fitted hourly bike count",
    colour = "Model",
    title = "Natural splines constrain the fit near the boundaries"
  )
```

An ordinary cubic regression spline can continue curving at the edges. A natural cubic spline imposes a boundary restriction: its fitted function is linear beyond the boundary knots. This limits one form of unstable tail behaviour, but it does not guarantee reliable extrapolation. We have few observations at extreme temperatures, so uncertainty there still matters.

A flexible curve will usually fit the observations used to build it more closely. That alone is not a reason to prefer it. We should ask whether the departure from a line is systematic, whether the shape is stable where there are fewer observations, and whether the extra flexibility serves the description or prediction we are trying to make.

## A Second Nonlinear Signal

Temperature is not the only candidate for nonlinear modelling. Solar radiation may also have a curved relationship with demand.

```{r seoul-solar-loess, warning=FALSE, message=FALSE}
seoul_hourly |>
  ggplot(aes(x = solar_radiation, y = rented_bike_count)) +
  geom_point(alpha = 0.05) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 1) +
  labs(
    x = "Solar radiation",
    y = "Hourly rented bike count",
    title = "Other weather predictors may also need flexible mean functions"
  )
```

## Lecture Summary

- LOESS is a flexible local smoother that is very useful for exploring shape.
- Splines create flexible mean functions inside a regression framework.
- Natural splines help control unstable boundary behaviour.

## Bridge to the Next Lecture

The Seoul bike data now present a new problem. After modelling nonlinear weather effects, we switch from hourly demand to daily demand and ask how time itself can enter the regression model through trend and seasonal predictors.
