
```{r time-indexed-regression-extra, include=FALSE}
library(patchwork)
```

## Time-Indexed Regression: From Flexible Models to Trend

The previous lectures focused on choosing predictors and flexible mean functions. We now turn to a different complication: observations collected through time.

In ordinary regression we usually write
$$y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i,$$
and assume that the errors $\varepsilon_i$ are independent. For the moment we keep that assumption and focus on how regular time patterns can be represented in the **mean** of the model.

By the end of this lecture, students should be able to:

1. Explain why time-indexed data often need explicit time predictors.
2. Use indicator variables to represent simple seasonal effects.
3. Distinguish mean structure in time from residual dependence over time.
4. Recognise the same modelling idea in both a classical and a modern dataset.

## Ski Sales: A First Seasonal Example

Often time series show a regular seasonal pattern. One simple way to handle this in a regression model is to create indicator variables for the most obvious seasonal effect.

We begin with quarterly ski equipment sales and personal disposable income (PDI). Sales increase over time as income increases, but there also appears to be a seasonal pattern.

`r xfun::embed_file("../data/SkiSales.csv")`

```{r read-ski-sales}
SkiSales <- read.csv("../data/SkiSales.csv", header = TRUE)
SkiSales |> head() |> kable()
```

### Fitting Sales on Income Alone

```{r ski-sales-and-residuals}
ski_plot <- SkiSales |>
  mutate(Quarter = factor(Quarter)) |>
  ggplot(aes(x = PDI, y = Sales)) +
  geom_point(aes(shape = Quarter, colour = Quarter), size = 2.5) +
  labs(title = "Ski sales rise with disposable income")

SkiSales_lm1 <- lm(Sales ~ PDI, data = SkiSales)

ski_resid_plot <- SkiSales |>
  mutate(
    Quarter = factor(Quarter),
    Residuals = resid(SkiSales_lm1)
  ) |>
  ggplot(aes(x = PDI, y = Residuals)) +
  geom_point(aes(shape = Quarter, colour = Quarter), size = 2.5) +
  geom_hline(yintercept = 0, linetype = 2, colour = "blue") +
  labs(title = "Residuals still show a seasonal pattern")

(ski_plot | ski_resid_plot) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The residuals tend to be positive in quarters 1 and 4, and negative in quarters 2 and 3. That suggests winter sales are systematically higher than summer sales, even after allowing for income.

### A Winter Indicator

```{r ski-winter-indicator}
SkiSales <- SkiSales |>
  mutate(Winter = as.numeric(Quarter == 1 | Quarter == 4))

SkiSales_lm2 <- lm(Sales ~ PDI + Winter, data = SkiSales)
summary(SkiSales_lm2)

plot(
  SkiSales_lm2$residuals ~ PDI,
  data = SkiSales,
  pch = SkiSales$Quarter,
  col = SkiSales$Quarter,
  xlab = "PDI",
  ylab = "Residuals"
)
abline(h = 0, lty = 2)
```

The winter indicator is strongly associated with sales, and the residuals look much less structured by quarter.

### Is Winter Enough?

We can still ask whether there is any extra difference between the first and second half of winter, or between the two summer quarters.

```{r ski-quarter-check}
SkiSales <- SkiSales |>
  mutate(
    Q4 = as.numeric(Quarter == 4),
    Q2 = as.numeric(Quarter == 2)
  )

SkiSales_lm3 <- lm(Sales ~ PDI + Winter + Q4 + Q2, data = SkiSales)
summary(SkiSales_lm3)
anova(SkiSales_lm2, SkiSales_lm3)
```

For these data, the simple winter indicator captures the main seasonal effect well enough. This is our first example of time-indexed regression: the mean changes with both income and time-of-year.

## A Modern Comparison: Daily Seoul Bike Demand

The same idea appears in a richer modern dataset. Here the response is daily bike demand, and time enters the model through a progression of increasingly flexible terms.

`r xfun::embed_file("../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-time-indexed, message=FALSE}
seoul_daily <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date)

seoul_outage_days <- sum(seoul_daily$functioning_day_all != "yes")
seoul_outage_days
seoul_daily |> head() |> kable()
```

### Plot 1: Raw Data with Non-Functioning Days Highlighted

Let's check whether every observation measures ordinary demand. On days when the bike system was not functioning all day, the count also reflects the lack of access to bikes.

```{r seoul-plot1-raw-highlight}
seoul_daily |>
  mutate(non_functioning = functioning_day_all != "yes") |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(colour = "grey50") +
  geom_point(
    data = seoul_daily |> filter(functioning_day_all != "yes"),
    colour = "red", size = 2
  ) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    title = "Non-functioning days (red) are not ordinary demand"
  )
```

We therefore exclude the red days before modelling ordinary bike demand.

### Plot 2: Clean Time Series After Filtering

```{r seoul-plot2-clean-series}
seoul_daily_clean <- seoul_daily |>
  filter(functioning_day_all == "yes")

ggplot(seoul_daily_clean, aes(x = date, y = rented_bike_count)) +
  geom_line() +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    title = "Daily bike demand on functioning days only"
  )
```

We removed 13 outage days. The remaining rows retain their original dates and `day_index` values, so consecutive rows are not always consecutive days. Keep that distinction in mind when we examine residual dependence.

This dataset covers one year. Its rise and fall within that year are visible, but one year cannot establish that the same annual cycle will recur or that a fitted linear trend persists over several years.

### Plot 3: Trend-Only Model

A linear trend captures the broad direction but ignores the seasonal pattern entirely.

```{r seoul-plot3-trend-only}
seoul_trend <- lm(rented_bike_count ~ day_index, data = seoul_daily_clean)

seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(colour = "grey50") +
  geom_line(aes(y = fitted(seoul_trend)), colour = "blue", linewidth = 1) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    title = "Trend-only model misses the seasonal pattern"
  )
```

The fitted line describes a broad change across the observed year, but it misses the rise and fall within that year. It underpredicts in some parts of the year and overpredicts in others, as we also saw in the ski sales residuals.

### Plot 4: Adding a Fourier Seasonal Term

A sine and cosine pair with a one-year period describes a smooth within-year pattern without needing a separate indicator for each date.

One sine term completes an annual cycle, and one cosine term completes the same cycle with a different starting position. Fitting them together lets the data determine where the seasonal peak falls. This adds two coefficients to the trend-only model. Next lecture we will unpack why the pair represents a shifted seasonal wave.

```{r seoul-plot4-fourier-prep}
seoul_daily_clean <- seoul_daily_clean |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365)
  )

seoul_fourier <- lm(
  rented_bike_count ~ day_index + S1 + C1,
  data = seoul_daily_clean
)
summary(seoul_fourier)
```

```{r seoul-plot4-fourier-plot}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(colour = "grey50") +
  geom_line(aes(y = fitted(seoul_fourier)), colour = "red", linewidth = 1) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    title = "Trend plus Fourier terms captures much of the within-year pattern"
  )
```

The Fourier pair adds only two parameters and captures much of the observed rise and fall. We have a one-year record, so this fitted cycle describes the pattern within these data; it does not establish that the same pattern will recur in later years.

### Plot 5: Comparing Trend Only and Trend + Fourier

Overlaying both fitted lines on the same axes makes the improvement clear.

```{r seoul-plot5-comparison}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(colour = "grey50") +
  geom_line(aes(y = fitted(seoul_trend), colour = "Trend only"), linewidth = 1) +
  geom_line(aes(y = fitted(seoul_fourier), colour = "Trend + Fourier"), linewidth = 1) +
  scale_colour_manual(
    values = c("Trend only" = "blue", "Trend + Fourier" = "red")
  ) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Model",
    title = "Trend and Fourier terms capture more of the within-year pattern"
  )
```

The logic is the same as in the ski example. We are still using ordinary regression, but one predictor records where the observation sits in time and a pair of trigonometric terms describes the within-year pattern.

## Mean Structure First, Error Dependence Later

In both examples we used time predictors to describe the mean response:
$$E(Y_t) = \beta_0 + \beta_1 t + \beta_2 z_t + \cdots.$$

That is different from saying the **errors** are autocorrelated. A model can capture clear time structure in the mean and still leave dependent residuals behind. The next lecture develops richer seasonal terms, and later lectures return to the question of autocorrelation in the errors.
