
```{r time-indexed-regression-extra, include=FALSE, cache=FALSE}
library(patchwork)
library(splines)
```

## Time-Indexed Regression: From Flexible Models to Trend

The previous lectures focused on choosing predictors and flexible mean functions. We now turn to a different complication: observations collected through time.

In ordinary regression we usually write
$$y_i = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i,$$
and assume that the errors $\varepsilon_i$ are independent. For the moment we keep that assumption and focus on how regular time patterns can be represented in the **mean** of the model.

By the end of this lecture, students should be able to:

1. Explain why time-indexed data often need explicit time predictors.
2. Use indicator variables to represent simple seasonal effects.
3. Distinguish mean structure in time from residual dependence over time.
4. Recognise the same modelling idea in both a classical and a modern dataset.

## Ski Sales: A First Seasonal Example

Often time series show a regular seasonal pattern. One simple way to handle this in a regression model is to create indicator variables for the most obvious seasonal effect.

We begin with quarterly ski equipment sales and personal disposable income (PDI). Sales increase over time as income increases, but there also appears to be a seasonal pattern.

`r xfun::embed_file("../data/SkiSales.csv")`

```{r read-ski-sales}
SkiSales <- read.csv("../data/SkiSales.csv", header = TRUE)
SkiSales |> head() |> kable()
```

### Fitting Sales on Income Alone

```{r ski-income-model}
SkiSales_lm1 <- lm(Sales ~ PDI, data = SkiSales)
```

```{r ski-sales-and-residuals, echo=FALSE}
ski_plot <- SkiSales |>
  mutate(Quarter = factor(Quarter)) |>
  ggplot(aes(x = PDI, y = Sales)) +
  geom_point(aes(shape = Quarter, colour = Quarter), size = 2.5) +
  labs(
    x = "PDI",
    y = "Sales",
    colour = "Quarter",
    shape = "Quarter",
    title = "Ski sales rise with disposable income"
  )

ski_resid_plot <- SkiSales |>
  mutate(
    Quarter = factor(Quarter),
    Residuals = resid(SkiSales_lm1)
  ) |>
  ggplot(aes(x = PDI, y = Residuals)) +
  geom_point(aes(shape = Quarter, colour = Quarter), size = 2.5) +
  geom_hline(yintercept = 0, linetype = 2, colour = "blue") +
  labs(
    x = "PDI",
    y = "Residuals",
    colour = "Quarter",
    shape = "Quarter",
    title = "Residuals still show a seasonal pattern"
  )

(ski_plot | ski_resid_plot) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

The residuals tend to be positive in quarters 1 and 4, and negative in quarters 2 and 3. That suggests winter sales are systematically higher than summer sales, even after allowing for income.

## A Winter Indicator

```{r ski-winter-indicator}
SkiSales <- SkiSales |>
  mutate(Winter = as.numeric(Quarter == 1 | Quarter == 4))

SkiSales_lm2 <- lm(Sales ~ PDI + Winter, data = SkiSales)
summary(SkiSales_lm2)
```

```{r ski-winter-residuals, echo=FALSE}
SkiSales |>
  mutate(
    Quarter = factor(Quarter),
    Residuals = resid(SkiSales_lm2)
  ) |>
  ggplot(aes(x = PDI, y = Residuals)) +
  geom_point(aes(shape = Quarter, colour = Quarter), size = 2.5) +
  geom_hline(yintercept = 0, linetype = 2, colour = "blue") +
  labs(
    x = "PDI",
    y = "Residuals",
    colour = "Quarter",
    shape = "Quarter",
    title = "Residuals after allowing for winter sales"
  ) +
  theme(legend.position = "bottom")
```

The winter indicator is strongly associated with sales, and the residuals look much less structured by quarter.

### Is Winter Enough?

We can still ask whether there is any extra difference between the first and second half of winter, or between the two summer quarters.

```{r ski-quarter-check}
SkiSales <- SkiSales |>
  mutate(
    Q4 = as.numeric(Quarter == 4),
    Q2 = as.numeric(Quarter == 2)
  )

SkiSales_lm3 <- lm(Sales ~ PDI + Winter + Q4 + Q2, data = SkiSales)
summary(SkiSales_lm3)
anova(SkiSales_lm2, SkiSales_lm3)
```

For these data, the simple winter indicator captures the main seasonal effect well enough. This is our first example of time-indexed regression: the mean changes with both income and time-of-year.

## A Modern Comparison: Daily Seoul Bike Demand

The same idea appears in a richer modern dataset. Here the response is daily bike demand, and time enters the model through a progression of increasingly flexible terms.

`r xfun::embed_file("../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-time-indexed, message=FALSE}
seoul_daily <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date)

seoul_daily |> head() |> kable()
```

## Bikes: Daily Demand Before Filtering

```{r seoul-plot1-raw-series, echo=FALSE}
ggplot(seoul_daily, aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  scale_colour_manual(values = c("Observed rentals" = "grey50")) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Daily bike rentals before filtering"
  ) +
  theme(legend.position = "bottom")
```

Several days drop right down to zero. Before fitting a model, let's go back to the data and check what happened on those days.

## Bikes: Checking the Non-Functioning Days

The `functioning_day_all` variable records whether the bike system was functioning throughout the day. Let's count the days that are not marked `"yes"`.

```{r seoul-count-non-functioning-days}
seoul_outage_days <- sum(seoul_daily$functioning_day_all != "yes")
seoul_outage_days
```

There are 12 days with zero rentals, but 13 days marked as not functioning all day. One of those days still recorded some rentals. We use the functioning-day flag to identify the days to exclude.

## Bikes: Demand on Functioning Days

```{r seoul-clean-data}
seoul_daily_clean <- seoul_daily |>
  filter(functioning_day_all == "yes")
```

```{r seoul-plot2-clean-series, echo=FALSE}
ggplot(seoul_daily_clean, aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  scale_colour_manual(values = c("Observed rentals" = "grey50")) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Daily bike demand on functioning days only"
  ) +
  theme(legend.position = "bottom")
```

On the excluded days, the rental count also reflects restricted access to bikes, so those days do not measure ordinary demand.

We removed 13 outage days. The remaining rows retain their original dates and `day_index` values, so consecutive rows are not always consecutive days. Keep that distinction in mind when we examine residual dependence.

This dataset covers one year. Its rise and fall within that year are visible, but one year cannot establish that the same annual cycle will recur or that a fitted linear trend persists over several years.

## Bikes: Trend-Only Model

A linear trend captures the broad direction but ignores the seasonal pattern entirely.

```{r seoul-trend-model}
seoul_trend <- lm(rented_bike_count ~ day_index, data = seoul_daily_clean)
```

```{r seoul-plot3-trend-only, echo=FALSE}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  geom_line(aes(y = fitted(seoul_trend), colour = "Trend only"), linewidth = 1) +
  scale_colour_manual(values = c("Observed rentals" = "grey50", "Trend only" = "blue")) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Trend-only model misses the seasonal pattern"
  ) +
  theme(legend.position = "bottom")
```

The fitted line describes a broad change across the observed year, but it misses the rise and fall within that year. It underpredicts in some parts of the year and overpredicts in others, as we also saw in the ski sales residuals.

## Bikes: Fitting a Natural Spline Through Time

Last lecture we used natural splines to fit a curved relationship with temperature. We can do the same thing with time. We keep `df = 5` here and let the fitted mean bend across the year.

```{r seoul-spline-model}
seoul_spline <- lm(
  rented_bike_count ~ ns(day_index, df = 5),
  data = seoul_daily_clean
)
```

```{r seoul-plot3-spline, echo=FALSE}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  geom_line(aes(y = fitted(seoul_spline), colour = "Natural spline"), linewidth = 1) +
  scale_colour_manual(values = c("Observed rentals" = "grey50", "Natural spline" = "forestgreen")) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Natural spline follows the pattern across the observed year"
  ) +
  theme(legend.position = "bottom")
```

## Bikes: Adding a Fourier Seasonal Term

A sine and cosine pair with a one-year period describes a smooth within-year pattern without needing a separate indicator for each date.

One sine term completes an annual cycle, and one cosine term completes the same cycle with a different starting position. Fitting them together lets the data determine where the seasonal peak falls. This adds two coefficients to the trend-only model. Next lecture we will unpack why the pair represents a shifted seasonal wave.

The spline lets the mean bend across the dates we observed, but it does not require that shape to repeat. A sine and cosine pair defines a wave that repeats every 365 days, so it is a natural choice when we expect a seasonal pattern to recur. The linear trend can still move the overall level up or down from year to year; it is the seasonal component that repeats.

```{r seoul-plot4-fourier-prep}
seoul_daily_clean <- seoul_daily_clean |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365)
  )

seoul_fourier <- lm(
  rented_bike_count ~ day_index + S1 + C1,
  data = seoul_daily_clean
)
summary(seoul_fourier)
```

```{r seoul-plot4-fourier-plot, echo=FALSE}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  geom_line(aes(y = fitted(seoul_fourier), colour = "Trend + Fourier"), linewidth = 1) +
  scale_colour_manual(values = c("Observed rentals" = "grey50", "Trend + Fourier" = "red")) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Trend plus Fourier terms captures much of the within-year pattern"
  ) +
  theme(legend.position = "bottom")
```

The Fourier pair adds only two parameters and captures much of the observed rise and fall. We have a one-year record, so this fitted cycle describes the pattern within these data; it does not establish that the same pattern will recur in later years.

## Bikes: Comparing Trend, Natural Spline and Fourier Fits

We can now compare three ways of describing the mean through time: a straight line, a natural spline, and a trend with an annual sine and cosine pair.

:::: {.columns style="display: flex; gap: 1em; align-items: center;"}
::: {.column width="70%"}

```{r seoul-plot5-comparison, echo=FALSE, out.width="100%"}
seoul_daily_clean |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals")) +
  geom_line(aes(y = fitted(seoul_trend), colour = "Trend only"), linewidth = 1) +
  geom_line(aes(y = fitted(seoul_spline), colour = "Natural spline"), linewidth = 1) +
  geom_line(aes(y = fitted(seoul_fourier), colour = "Trend + Fourier"), linewidth = 1) +
  scale_colour_manual(
    values = c(
      "Observed rentals" = "grey50",
      "Trend only" = "blue",
      "Natural spline" = "forestgreen",
      "Trend + Fourier" = "red"
    )
  ) +
  labs(
    x = "Date",
    y = "Daily rented bike count",
    colour = "Series",
    title = "Three regression fits to the observed year"
  ) +
  guides(colour = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom")
```

:::

::: {.column width="30%"}

```{r seoul-model-r-squared, echo=FALSE}
tibble(
  Model = c("Trend only", "Natural spline", "Trend + Fourier"),
  R_squared = c(
    summary(seoul_trend)$r.squared,
    summary(seoul_spline)$r.squared,
    summary(seoul_fourier)$r.squared
  )
) |>
  kable(col.names = c("Model", "$R^2$"), digits = 3)
```

:::
::::

All three fits use ordinary regression on the same functioning days. The straight line gives a constant rate of change, the natural spline lets the mean bend through time, and the Fourier pair adds an annual wave to the trend. A closer fit to these observations alone does not tell us which model will predict future demand best.

## Mean Structure First, Error Dependence Later

In both examples we used time predictors to describe the mean response:
$$E(Y_t) = \beta_0 + \beta_1 t + \beta_2 z_t + \cdots.$$

That is different from saying the **errors** are autocorrelated. A model can capture clear time structure in the mean and still leave dependent residuals behind. The next lecture develops richer seasonal terms, and later lectures return to the question of autocorrelation in the errors.

## Lecture Summary

We are still using ordinary regression. We have simply added predictors that tell the model when an observation was collected.

For ski sales, income got us part of the way, but the residuals still differed by quarter. Adding a winter indicator dealt with most of that seasonal pattern. We checked whether separating the quarters helped further, and the simpler model was adequate for these data.

With the Seoul bike data, a straight-line trend missed much of the rise and fall across the year. The natural spline let the mean bend across the year without requiring an annual cycle. The sine and cosine pair gave the fitted mean enough flexibility to follow that pattern with just two extra coefficients. We first removed the outage days so that we were modelling demand when bikes were available. Even then, we only have one year: we cannot yet say whether that cycle repeats or whether the trend continues.

As we move on, keep the mean and the errors separate. We have improved how the mean changes through time, but the errors may still be related from one day to the next. Next lecture we look more closely at how the seasonal terms work. Then we return to the residuals and ask whether the independence assumption is reasonable.
