
## Regression Models for Trend and Seasonality

In the previous lecture, a simple seasonal indicator was enough for the ski-sales data. In other situations, the pattern across the year is smoother and more complicated. We want a model that is more flexible than a single winter indicator, but more parsimonious than fitting a separate effect for every month.

By the end of this lecture, students should be able to:

1. Use factor indicators to represent seasonality.
2. Build sine and cosine predictors for periodic behaviour.
3. Explain why sine-plus-cosine allows an unknown phase shift.
4. Recognise Fourier terms as a systematic version of the same idea.

## Electricity Bills: Month Effects and Seasonal Shape

We consider monthly electricity bills for a household from May 1994 to May 1998. During some months there were paying guests staying in part of the house, so the householder wanted to estimate the extra business-related electricity cost.

We first allow for bill length, whether the bill was estimated, guest weeks, and a time trend. Then we examine the seasonal pattern that remains.

`r xfun::embed_file("../data/Electricity.csv")`

```{r read-electricity}
Electricity <- read.csv("../data/Electricity.csv", header = TRUE)
Electricity |> head() |> kable()
```

### Residual Seasonality After Basic Adjustment

```{r electricity-initial-model}
Electricity_lm1 <- lm(Bill ~ Days + Estimate + Guestwks + Time, data = Electricity)
resids <- resid(Electricity_lm1)

plot(resids ~ Time, data = Electricity, ylab = "Residuals")
lines(resids ~ Time, data = Electricity, col = 4)
abline(h = 0, lty = 2)
```

After this adjustment, residuals tend to be positive in some months and negative in others. That suggests a seasonal pattern remains.

We use the residual plot to choose seasonal terms for the *mean* model. To test their contribution, we add them to the original `Bill` regression while keeping `Days`, `Estimate`, `Guestwks` and `Time` in the model.

## Sinusoidal Terms

A sine and cosine pair with the same period can represent a seasonal wave whose peak occurs at any point in the cycle. The origin used to define the pair is a choice of coordinates, not a claim about which month is the average.

```{r electricity-phase-invariance}
Electricity <- Electricity |>
  mutate(
    SinTime4 = sin((Time - 4) * 2 * pi / 12),
    CosTime4 = cos((Time - 4) * 2 * pi / 12),
    SinTime5 = sin((Time - 5) * 2 * pi / 12),
    CosTime5 = cos((Time - 5) * 2 * pi / 12),
    Sin2Time = sin((Time - 5) * 4 * pi / 12),
    Cos2Time = cos((Time - 5) * 4 * pi / 12)
  )

Electricity_first4 <- lm(
  Bill ~ Days + Estimate + Guestwks + Time + SinTime4 + CosTime4,
  data = Electricity
)
Electricity_first5 <- lm(
  Bill ~ Days + Estimate + Guestwks + Time + SinTime5 + CosTime5,
  data = Electricity
)

c(
  maximum_fitted_value_difference = max(abs(fitted(Electricity_first4) - fitted(Electricity_first5))),
  cosine_coefficient_shift_4 = coef(Electricity_first4)["CosTime4"],
  cosine_coefficient_shift_5 = coef(Electricity_first5)["CosTime5"]
)

Electricity <- Electricity |>
  mutate(
    SinTime = SinTime5,
    CosTime = CosTime5
  )
```

We can choose any convenient month as the origin of the sine and cosine pair. With a 12-month period, shifting the origin changes the two coefficients, but it does not change the family of curves the pair can fit. To see this, fit the pair using `Time - 4`, then repeat using `Time - 5`. The fitted values agree to numerical precision. A smaller cosine coefficient in the second version therefore says something about how we labelled the two basis functions, not that we found a better seasonal model. We will use `Time - 5` from here on because its coefficients are easier to describe.

### A Sharper Seasonal Shape: Add Second Harmonics

One sine/cosine pair produces a smooth cycle with one peak and one trough per year. The residual pattern suggests a sharper rise or fall than this pair can produce. Adding a second pair, at twice the frequency, allows that shape to change while keeping the original adjustment variables in the model. Compare the model with the first pair against the model with both pairs. The partial F test asks whether those *two additional coefficients* improve the fit after the other terms have already been included.

```{r electricity-second-harmonic}
Electricity_full1 <- lm(
  Bill ~ Days + Estimate + Guestwks + Time + SinTime + CosTime,
  data = Electricity
)
Electricity_full2 <- lm(
  Bill ~ Days + Estimate + Guestwks + Time + SinTime + CosTime + Sin2Time + Cos2Time,
  data = Electricity
)

anova(Electricity_full1, Electricity_full2)
```

```{r electricity-second-harmonic-plot}
Electricity |>
  mutate(
    first_harmonic = fitted(Electricity_full1),
    second_harmonic = fitted(Electricity_full2)
  ) |>
  ggplot(aes(x = Time, y = Bill)) +
  geom_point() +
  geom_line(aes(y = first_harmonic, colour = "First harmonic")) +
  geom_line(aes(y = second_harmonic, colour = "Second harmonic")) +
  scale_colour_manual(values = c("First harmonic" = "blue", "Second harmonic" = "red")) +
  labs(x = "Month index", y = "Electricity bill ($)", colour = "Model")
```

Under the usual OLS error assumptions, the second harmonic provides evidence of additional seasonal shape in these data: $F(2, 40) = 3.83$, $p = 0.0301$. This two-degree-of-freedom test compares models with the same original adjustment variables. Later we will check whether the residual errors are dependent, which matters for that P-value. The fitted pattern describes this household after adjustment for the recorded covariates; it does not establish that the same pattern applies elsewhere.

### Putting the Model Together

```{r electricity-full-model}
summary(Electricity_full2)
```

The final model is still a regression model. It adjusts for the bill length, estimated status, guest weeks and time trend, then adds smooth seasonal waves rather than a separate effect for each month.

## Fourier Terms in a Modern Example

The Seoul bike data use the same idea, but on a daily time scale. The sine and cosine terms are usually written more systematically as **Fourier terms**.

`r xfun::embed_file("../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-trend-seasonality, message=FALSE}
seoul_daily <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date) |>
  filter(functioning_day_all == "yes") |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )

seoul_daily |> head() |> kable()
```

We keep the same 352 functioning days used in Lecture 32 and retain their original dates and `day_index` values. The gaps remain in the time index after the 13 outage days are removed.

```{r seoul-fourier-model}
seoul_fourier <- lm(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2 +
    mean_temperature + total_rainfall + total_solar_radiation,
  data = seoul_daily
)
summary(seoul_fourier)
```

```{r seoul-fourier-plot}
seoul_daily |>
  mutate(fitted = fitted(seoul_fourier)) |>
  ggplot(aes(x = day_index, y = rented_bike_count)) +
  geom_line(alpha = 0.6) +
  geom_line(aes(y = fitted), colour = "red", linewidth = 1) +
  labs(
    x = "Day index",
    y = "Daily rented bike count",
    title = "Fourier terms capture the observed within-year pattern"
  )
```

The electricity example introduced the ideas in a small, interpretable setting. The Seoul bike model shows the same method with daily data. The sine and cosine pairs compactly represent the rise and fall observed within this one-year record; they do not establish a recurring annual pattern.

## Looking Ahead

So far we have assumed that once the mean structure is well chosen, the residuals behave independently. The next lecture asks whether that is actually true.
