
```{r diagnosing-autocorrelation-setup, include=FALSE}
library(DescTools)
library(lmtest)
```

## Diagnosing Autocorrelated Errors

In the last two lectures we concentrated on fitting the mean structure of a time series with regression. Now we ask whether the residuals behave independently after that mean structure has been removed.

By the end of this lecture, students should be able to:

1. Explain what autocorrelation means in regression residuals.
2. Describe why autocorrelation makes ordinary standard errors unreliable.
3. Use residual plots, runs tests, Durbin-Watson, ACF plots, and Ljung-Box tests.
4. Recognise when dependence remains after a reasonable mean model has been fitted.

## What Is Autocorrelation?

A time predictor can explain a pattern in the *mean* while nearby deviations from that mean remain related. Positive residual autocorrelation means a positive residual tends to be followed by another positive residual, and likewise for negative residuals. A negative correlation at lag one tends to produce alternation. The usual independent-error calculations then give unreliable conventional standard errors. Depending on the regressors and the dependence pattern, they can be too small or too large. If the conditional mean is correctly specified and the errors have mean zero given the predictors, autocorrelation alone need not bias the OLS coefficients. It can, however, make the apparent precision and tests wrong.

## A Classical Example: Consumer Expenditure and Money Stock

This example studies consumer expenditure (`ConsExp`) and the stock of money (`MoneyStk`) in billions of US dollars.

`r xfun::embed_file("../data/MoneyStock.csv")`

```{r read-money-stock}
MoneyStock <- read.csv("../data/MoneyStock.csv", header = TRUE)
MoneyStock |> head() |> kable()
```

```{r money-stock-pairs}
pairs(MoneyStock[3:5])
```

### Fit the Regression Model

```{r money-stock-regression}
MoneyStock_lm1 <- lm(ConsExp ~ MoneyStk, data = MoneyStock)
summary(MoneyStock_lm1)
```

`MoneyStk` and `ConsExp` have a strong linear association, and the regression output gives a precise-looking estimate of the slope. Before trusting that precision, we need to check whether the residuals are independent.

### Residual Plot Over Time

```{r money-stock-residual-plot}
plot(residuals(MoneyStock_lm1), type = "b", ylab = "Residuals", xlab = "Quarter")
abline(h = 0, lty = 2)
```

The residuals show the long, slow wandering pattern typical of positive autocorrelation. Extended sequences of positive or negative residuals should not be common if the errors are independent.

## Diagnostic Tools

### Runs Test

The default `RunsTest()` for a numerical vector divides residuals above and below their median and counts how often the sequence switches sides. A long run is a reason to look more closely at the time plot; the test does not tell us which time process produced it.

```{r money-stock-runs-test}
RunsTest(residuals(MoneyStock_lm1))
```

A small P-value suggests the sequence of observations above and below its median does not look random under this test. It does not identify the source of any pattern.

### Durbin-Watson Test

The Durbin-Watson test is a classical test for first-order autocorrelation in regression residuals.

```{r money-stock-durbin-watson}
dwtest(MoneyStock_lm1, alternative = "greater")
```

Here the one-sided alternative reflects the suspicion of **positive** autocorrelation.

### ACF Plot

The autocorrelation function (ACF) shows the correlation between residuals separated by different time lags.

```{r money-stock-acf, fig.width=8, fig.height=4}
acf(residuals(MoneyStock_lm1), main = "ACF of MoneyStock residuals")
```

Let's check how long the pattern lasts. A large lag-1 value means neighbouring residuals are strongly correlated. If the ACF decays slowly, the relationship extends over several time lags. We should also check whether a missing time trend could explain part of this pattern.

### Ljung-Box Test

The Ljung-Box test checks several autocorrelations at once rather than focusing only on lag 1.

```{r money-stock-ljung-box}
Box.test(residuals(MoneyStock_lm1), lag = 10, type = "Ljung-Box")
```

If this test rejects, we have evidence of residual autocorrelation across the lags tested. It does not tell us which change to the model will address it.

## A Modern Comparison: Seoul Bike Residuals

The same diagnostic logic applies to the daily bike model from the previous lecture.

`r xfun::embed_file("../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-diagnostics, message=FALSE}
seoul_daily <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date) |>
  filter(functioning_day_all == "yes") |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )

seoul_mean <- lm(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2 +
    mean_temperature + total_rainfall + total_solar_radiation,
  data = seoul_daily
)

seoul_diagnostics <- seoul_daily |>
  mutate(residual = residuals(seoul_mean))
```

```{r seoul-residual-plot}
ggplot(seoul_diagnostics, aes(x = date, y = residual)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Actual date", y = "Residual", title = "Seoul bike residuals on functioning days")
```

We fitted the mean model to all 352 functioning days. The dates remain visible here, including the gaps left by 13 outage days. For the sequence-based summaries below, use only the longest uninterrupted run, from 1 December 2017 through 10 April 2018. Otherwise, an ordinary ACF would treat consecutive retained rows as though they were always one day apart.

```{r seoul-diagnostic-tests}
seoul_run_residuals <- seoul_diagnostics |>
  filter(date >= as.Date("2017-12-01"), date <= as.Date("2018-04-10")) |>
  pull(residual)

RunsTest(seoul_run_residuals)
Box.test(seoul_run_residuals, lag = 14, type = "Ljung-Box")
```

```{r seoul-acf, fig.width=8, fig.height=4}
acf(seoul_run_residuals, main = "ACF of Seoul bike residuals: consecutive days")
```

The diagnostics describe the residual sequence on those 131 consecutive days. They do not establish that the mean model is complete or that the errors are independent over the entire gappy year.

## Diagnostic Summary

We start with the residuals plotted against actual time, then inspect their ACF. A test can support what we see, but several small P-values do not identify the correct error model. First ask whether a missing trend, seasonal term or outage explains the pattern. If a reasonable mean model still leaves dependence, the next lecture changes the error model itself.

## Looking Ahead

Next we will fit a regression with AR(1) errors using generalised least squares, then check whether it accounts for the dependence we found.
