
```{r autoregressive-errors-setup, include=FALSE}
library(nlme)
library(broom.mixed)
```

## Regression with Autoregressive Errors

The previous lecture showed how to diagnose autocorrelation in regression residuals. This lecture keeps the same mean structure ideas, but changes the model for the errors.

By the end of this lecture, students should be able to:

1. Write down a regression model with AR(1) errors.
2. Explain why autocorrelated errors distort ordinary standard errors.
3. Fit a regression model with `gls()` and `corAR1()`.
4. Check whether normalised residuals look closer to white noise.

## Why Ordinary Least Squares Is Not Enough

If the mean model is correctly specified and the errors average to zero given the predictors, correlated errors do not by themselves force the OLS slope to be biased. They do make the usual independent-error standard errors unreliable; the direction of the error is not fixed. We now keep the trend and seasonal predictors and model the dependence among the remaining errors.

## The AR(1) Error Model

A first-order autoregressive error model is
$$\varepsilon_t = \phi \varepsilon_{t-1} + \eta_t,$$
where $\eta_t$ is independent white noise with mean zero and constant variance.

- If $\phi > 0$, nearby errors tend to move in the same direction.
- If $\phi < 0$, nearby errors tend to alternate.
- If $\phi = 0$, we are back to independent errors.

For the process to be stationary, we require $|\phi| < 1$. Its lag-$k$ correlation is $\phi^k$, so dependence weakens as the time gap grows. These correlations describe the errors after we have accounted for the mean predictors.

## Main Worked Example: Tourism in Victoria

These data are monthly room nights occupied in hotels, motels, and guesthouses in Victoria from January 1980 to December 1994.

`r xfun::embed_file("../data/motel.csv")`

```{r read-tourism}
tourism <- read_csv("../data/motel.csv", show_col_types = FALSE) |>
  mutate(
    Date = as.Date(Date),
    Time = round(time_length(Date - min(Date), unit = "month")),
    Month = factor(month(Date, label = TRUE), ordered = FALSE),
    Year = year(Date)
  )

stopifnot(all(diff(tourism$Time) == 1))

tourism |> head() |> kable()
```

### Ordinary Regression First

```{r tourism-lm}
tourism_lm <- lm(RoomNights ~ Time + Month, data = tourism)
anova(tourism_lm)
summary(tourism_lm)
```

```{r tourism-lm-fit-plot}
augment(tourism_lm, tourism) |>
  ggplot(aes(x = Date)) +
  geom_point(aes(y = RoomNights)) +
  geom_line(aes(y = .fitted), colour = "blue") +
  labs(title = "Trend and seasonality explain much of the tourism series")
```

This is a sensible mean model, but the residuals may still be dependent.

```{r tourism-lm-residual-acf, fig.height=4.5}
acf(resid(tourism_lm), main = "ACF of tourism residuals")
```

The residual ACF suggests positive autocorrelation, so AR(1) errors are a reasonable first correction.

## Regression Model with AR(1) Errors

For the tourism data we now write
$$\begin{aligned}
Y_t &= \beta_0 + \beta_1 \mathrm{Time} + \beta_2 z_{t,2} + \cdots + \beta_{12} z_{t,12} + \varepsilon_t, \\
\varepsilon_t &= \phi \varepsilon_{t-1} + \eta_t.
\end{aligned}$$

The mean still depends on trend and month. The new ingredient is the correlation structure among the errors.

## Why GLS Works

Suppose
$$\begin{aligned}
y_t &= \beta_0 + \beta_1 x_t + \varepsilon_t, \\
\varepsilon_t &= \phi \varepsilon_{t-1} + \eta_t.
\end{aligned}$$

Subtracting $\phi$ times the previous equation from the current one gives
$$y_t - \phi y_{t-1}
= \beta_0(1 - \phi) + \beta_1(x_t - \phi x_{t-1}) + \eta_t.$$

The transformation removes the AR(1) dependence, leaving the independent errors $\eta_t$. **Generalised least squares** uses this covariance structure when estimating the regression coefficients.

## Fitting the AR(1) Model in R

The `gls()` function from `nlme` uses the usual model formula plus an explicit correlation structure.

```{r tourism-gls}
tourism_gls <- gls(
  RoomNights ~ Time + Month,
  correlation = corAR1(form = ~ Time),
  data = tourism
)
anova(tourism_gls)
summary(tourism_gls)
```

`Time` is the integer covariate that defines correlation distance in `correlation = corAR1(form = ~ Time)`.

`Time` counts months, so observations one month apart have correlation $\phi$ and observations two months apart have correlation $\phi^2$ under this model. The regression formula still describes mean room nights. The `corAR1()` term describes the pattern left in the errors.

### Comparing OLS and GLS Standard Errors

```{r tourism-standard-error-comparison}
list(
  gls = tidy(tourism_gls),
  lm = tidy(tourism_lm)
) |>
  bind_rows(.id = "model") |>
  filter(term == "Time") |>
  select(model, term, estimate, std.error)
```

Compare the trend estimates *and* their standard errors. GLS may change the estimated trend as well as its precision because observations are no longer treated as providing independent information. The direction and size of the change are properties of these data and this fitted correlation model, not a universal correction factor.

### Checking the Normalized Residuals

```{r tourism-normalized-residual-acf}
acf(
  residuals(tourism_gls, type = "normalized"),
  main = "ACF of normalized residuals after GLS"
)
```

We want much less remaining structure in the normalised residuals than in the OLS residuals. A quieter ACF makes AR(1) a more plausible working error model; it does not prove independence or rule out a missing part of the mean.

## A Short Comparison: Seoul Bike with AR(1) Errors

The same modelling pattern can be used for the daily Seoul bike data once we have already chosen a reasonable mean structure.

`r xfun::embed_file("../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-gls, message=FALSE}
seoul_daily <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date) |>
  filter(functioning_day_all == "yes") |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )

seoul_lm <- lm(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2 +
    mean_temperature + total_rainfall + total_solar_radiation,
  data = seoul_daily
)

seoul_gls <- gls(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2 +
    mean_temperature + total_rainfall + total_solar_radiation,
  correlation = corAR1(form = ~ day_index),
  data = seoul_daily
)

seoul_normalized_residuals <- residuals(seoul_gls, type = "normalized")
seoul_diagnostics <- seoul_daily |>
  mutate(normalized_residual = seoul_normalized_residuals)
```

The Seoul models use all 352 functioning days, and `day_index` retains its original values after the 13 outage days are removed. `corAR1(form = ~ day_index)` uses the integer distance between days, so a gap of several days is not treated as one step.

```{r seoul-full-series-plot}
augment(seoul_lm, seoul_daily) |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = rented_bike_count), colour = "grey55") +
  geom_line(aes(y = .fitted), colour = "red") +
  labs(x = "Actual date", y = "Daily rented bike count", title = "Seoul demand and fitted mean on functioning days")
```

```{r seoul-gls-comparison}
list(
  gls = tidy(seoul_gls),
  lm = tidy(seoul_lm)
) |>
  bind_rows(.id = "model") |>
  filter(term == "day_index") |>
  select(model, term, estimate, std.error)
```

```{r seoul-normalized-residual-acf}
seoul_run_normalized <- seoul_diagnostics |>
  filter(date >= as.Date("2017-12-01"), date <= as.Date("2018-04-10")) |>
  pull(normalized_residual)

acf(seoul_run_normalized, main = "ACF of normalised Seoul residuals: consecutive days")
```

We want much less remaining structure in the normalised residuals than in the OLS residuals. A quieter ACF makes AR(1) a more plausible working error model; it does not prove independence or rule out a missing part of the mean.

The daily Seoul example uses the same distinction between a model for expected demand and a model for residual dependence. We retain the actual day index after removing system outages so a gap of several days is not treated as one AR(1) step. We should not read an ordinary ACF of the shortened residual vector as though every retained row were exactly one day apart.

## Final Comment

AR(1) gives us a first model for the dependence. If the normalised residuals still show a pattern, we need to reconsider the mean or the error model. Fitting a trend and seasonal terms does not finish the job; we also need to check what remains in the residuals.
