---
title: "Lecture 21 - Factor or Numerical Covariate?"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
library(kableExtra)
library(car)
library(DescTools)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## learning outcomes

-- Differentiate between a categorical variable used as factor or covariate.

-- implement and compare both models in R.

## Factor or Covariate?

When we have a categorical variable expressed as measured levels we can consider the question:

&nbsp;

"Is it better to use it as a factor or a numerical covariate?"

&nbsp;

For example, should the effect of a drug be modelled using a four level factor (zero, low, medium and high doses) or as a numeric variable (0, 10, 20, 30 ml doses)?

## Nesting of Linear Effects

Consider a one factor experiment in which all individuals at factor level *j* take the same value of *x* for this variable ($j=1,2,\ldots,K$).

We could fit a simple linear regression model to the data:

$$M0: Y_i = \beta_0 + \beta_1 x_i + \varepsilon_i~~~~~~(i=1,2,\ldots,n)$$

We could also fit a standard one-way factorial model which can be written as a multiple regression model:

$$M1: Y_i = \mu + \alpha_2 z_{i2} + \cdots + \alpha_K z_{iK}+\varepsilon_i~~~~~~(i=1,2,\ldots,n)$$

where $z_{ij}$ is the indicator variable for unit *i* being at  factor level *j* as usual.

## Nesting of Linear Effects

-- $M0$ assumes that the expected values (means) of the response for each level of the factor follow a straight line.

&nbsp;

-- $M1$ allows the mean response to vary arbitrarily across the levels of the factor.

&nbsp;

Therefore, the $M0$ model will be preferred if the relationship between the factor and the response is linear. The $M1$ model will be preferred if their relationship is not linear.

## Nesting of Linear Effects

-- We can show that model $M0$ `is nested within` model $M1$.

-- We can compare models $M0$ (the simpler one) and $M1$ (the more complex one) using an *F* test:

  -- *H~0~*: model $M0$ adequate.
  
  VS

  -- *H~1~*: model $M0$ not adequate.
 
## Nesting of Linear Effects
 
 using the *F* test statistic:

$$F = \frac{[RSS_{M0} - RSS_{M1}]/(K-2)}{RSS_{M1}/(n-K)}.$$


-- The *F*-statistic has *K-2, n-K* degrees of freedom.

## Nesting of Linear Effects

-- Retention of *H~0~* indicates that a simple linear regression model is adequate; i.e. the effect of the variable *x* is adequately modelled as linear.

-- Rejection of *H~0~* indicates that any association between *y* and *x* is non-linear.


## Vitamin C and Tooth Growth experiment

-- 30 guinea pigs were divided (at random) into three groups of ten and dosed with different ammounts of vitamin C in mg (administered in orange juice).

  -- Group 1: low (0.5mg).
  
  -- Group 2: medium (1mg).
  
  -- Group 3: high (2mg).

-- Length of odontoblasts (teeth) was measured as response variable.

## Vitamin C and Tooth Growth experiment

```{r boxplotLengthDose}
data(ToothGrowth)
TG = ToothGrowth |>  filter(supp == "VC")
plot(len ~ as.factor(dose), data=TG, col="grey")
```

## Vitamin C and Tooth Growth experiment


The box plot is slightly suggestive of a change of dispersion in the response between the groups.

```{r}
bartlett.test(len ~ as.factor(dose), data=TG)
```

The test does not provide clear evidence of heteroscedasticity (*P*-value is *P=0.1048*).

## Vitamin C and Tooth Growth experiment

Could the dependence of tooth growth on vitamin C be adequately modelled by:

-- a linear regression of `len` on `dose`

&nbsp;

or

&nbsp;

-- a factorial model with `len` and `as.factor(dose)` to be preferred?
 
## Vitamin C and Tooth Growth experiment

::::{.columns}


:::{.column with="45%"}

```{r Tooth.lm.0}
Tooth.lm.0 <- lm(len ~ dose, data=TG)
summary(Tooth.lm.0)$coefficients
summary(Tooth.lm.0)$fstatistic
```

<span style="font-size:smaller;"> *p*-value: 1.509e-11  </span>  

:::

:::{.column with="45%"}

```{r Tooth.lm.1}
Tooth.lm.1 <- lm(len ~  as.factor(dose), data=TG)
summary(Tooth.lm.1)$coefficients
summary(Tooth.lm.1)$fstatistic
```

<span style="font-size:smaller;"> *p*-value: 3.357e-11  </span> 

:::

::::

## Vitamin C and Tooth Growth experiment

```{r anova1}
anova(Tooth.lm.0, Tooth.lm.1)
```

-- We would retain *H~0~* at the 5% significance level (*p*-value:0.058).

-- but there is certainly some evidence to doubt that the relationship should be modelled as a linear effect.
 


