---
title: "Lecture 19 - One-way (factor) Models"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
library(kableExtra)
#library(car)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## learning outcomes

-- Interpret and build a one-way (or facotr) model.

-- Implement a one-way model in R.


## The One-Way Model

-- The one-way (or one factor) model is used when a continuous numerical response **Y** is dependent on a single factor (categorical explanatory variable).

&nbsp;

-- In such models, the categories defined by a factor are called the **levels** of the factor.


## Mathematical Formulation of the One-Way Model

It is common to express the One-way model as

$$y_{ij} = \mu_j + \varepsilon_{ij}~~~~~~~~(j=1,\ldots,K~~~~~i=1,\ldots,n_j)$$ 

-- $y_{ij}$: is the response of the *i*^th^ unit at the *j*^th^ level of the factor; *K* denotes the number of levels.

-- $\mu_j$: is the (unknown) population mean for level $j$.

-- $n_{j}$: the number of observations (replications) at level *j* for the factor.

-- $\varepsilon_{11},..,\varepsilon_{Kn_K}$: random errors satisfying assumptions (A1)-(A4).


## Mathematical Formulation of the One-Way Model

The null hypothesis is then expressed as 

$$H_0:  ~~~ \mu_1 = \mu_2 = ... =\mu_K $$
and the alternative hypothesis is 

$$H_1: ~~~\mbox{ means not all equal.}$$


## Mathematical Formulation of the One-Way Model

Another way of writing it 
$$y_{ij} = \mu + \alpha_j + \varepsilon_{ij}~~~~~~~~(j=1,\ldots,K,~~~~~i=1,\ldots,n_i)$$ 

where $\alpha_j = \mu_j - \mu$ (mean difference to the baseline)  for  $j=1,...,K$.  

The mean response is $$E[Y_{ij}] = \mu + \alpha_j$$

## Parameterisation of the One-Way Model

We have $K$ means and we are trying to estimate *K+1* parameters $\mu,\alpha_1,...,\alpha_K$

We cannot estimate *K+1* numbers from *k* bits of information. 

As it stands, the model is  **overparameterised**.

## Solution to Overparameterisation

We impose a constraint on the parameters  $\alpha_1, \alpha_2, \ldots, \alpha_K$.

-- Sum constraints: $\sum_{j=1}^K \alpha_j = 0$:

In this case $\mu$ can be interpreted as a kind of "grand mean", and $\alpha_1, \alpha_2, \ldots, \alpha_K$ measure deviations from this grand mean.
    
## Solution to Overparameterisation

-- Treatment constraint: $\alpha_1 = 0$:

Level 1 of the factor is regarded as the baseline or reference level, and $\alpha_2, \alpha_3 \ldots, \alpha_K$ measure deviations from this baseline. This is highly appropriate if level 1 corresponds to a
control group

&nbsp;

We will work with the treatment constraint. It is also used by  default for factors in R.

## Expressing Factor Models as Regression Models

We can express a one-way factor model as a regression model by using dummy (indicator) variables.

Define dummy variables $z_{11}, z_{12}, \ldots, z_{n K}$ by
    $$z_{ij} = 
    \left \{ 
    \begin{array}{ll}
    1 & \mbox{unit i observed at factor level j}\\
    0 & \mbox{otherwise.}
    \end{array}
    \right .$$



## Regression form of one-way factor model

A one-way factor model can be expressed as a regregression as follows:
$$y_i = \mu + \alpha_1 z_{i1} + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i~~~~~(i=1,2,\ldots,n)$$

The parameter $\mu$ is the regression intercept and, under the treatment constraint ($\alpha_1 = 0$), the mean baseline (factor level  1) response.

## Expressing Factorial Models using Matrices

For a one-way factorial model with treatment constraint:
$${\boldsymbol{y}} = X {\boldsymbol{\beta}} + \varepsilon$$ 
    
$$
\small
\boldsymbol{y} =
\begin{bmatrix}
Y_1 \\
Y_2 \\
\vdots \\
Y_n
\end{bmatrix},
\quad
X =
\begin{bmatrix}
1 & z_{12} & \ldots & z_{1K} \\
1 & z_{22} & \ldots & z_{2K} \\
\vdots & \vdots & \ddots & \vdots \\
1 & z_{n2} & \ldots & z_{nK}
\end{bmatrix},
\small
\boldsymbol{\beta} =
\begin{bmatrix}
\mu \\
\alpha_2 \\
\vdots \\
\alpha_K
\end{bmatrix},
\quad
\boldsymbol{\varepsilon} =
\begin{bmatrix}
\varepsilon_1 \\
\varepsilon_2 \\
\vdots \\
\varepsilon_n
\end{bmatrix}.
$$

<span style="font-size:smaller;"> No $\alpha_1$ in the parameter vector and row for $z_{i1}$ in the design matrix, due to the treatment constraint. </span>   

## Example: stimulating Effects of Caffeine

Does caffeine stimulation affect the rate at which individuals can tap their fingers?

Thirty male students randomly allocated to three treatment groups of 10 students each. Groups treated as follows:

-- Group 1: zero caffeine dose
    
-- Group 2: low caffeine dose
    
-- Group 3: high caffeine dose

Blind treatment allocation (subjects did not know their caffeine dosage). Two hours after treatment, each subject tapped fingers as quickly as possible for a minute.

## Example: stimulating Effects of Caffeine

```{r getCaffeineData, echo=-1}
Caffeine <- read.csv(file="./caffeine.csv", header=TRUE)
boxplot(Taps~Dose, data=Caffeine) 
```

## Modelling the Caffeine Data

Our one-way factor model, where the factor has three levels, can be expressed as follows:

$$y_i = \mu + \alpha_2 z_{i2} + \alpha_{3} z_{i3} + \varepsilon_i$$

where:

-- *y~i~* is the number of taps recorded for the *i*th subject;

-- *z~i2~ = 1* if subject *i* is on low dose; *z~i2~ = 0* otherwise;

-- and *z~i3~ = 1* if subject *i* is on high dose; *z~i3~ = 0* otherwise.

-- $\mu$ is mean response for a subject on zero dose;

-- $\alpha_2$ is the effect of low dose in contrast to zero dose;

-- and, $\alpha_3$ is the effect of high dose in contrast to zero dose.

## Modelling the Caffeine Data

This are first lines of the **design matrix** for the one-way model for the caffeine data.

```{r getModelMatrix, echo=FALSE}
Caffeine.lm <- lm(Taps~Dose, data=Caffeine) 
Caffeine.lm |> model.matrix() |> head() |> knitr::kable()
```


## R Code considerations

We fix up the order of the factor levels (default is alphabetical).

```{r getCaffeineData 2 }
levels(factor(Caffeine$Dose))
Caffeine$Dose <- factor(Caffeine$Dose, levels=c("Zero",
"Low", "High"))
levels(Caffeine$Dose)
```
The second line reorders the factor levels so that zero, low and high dose are interpreted as levels 1, 2 and 3 respectively.


## Fitting the Model

We can use the `lm()` function in R to fit model in the same manner as for regression models.

```{r Caffeine.lm}
Caffeine.lm <- lm(Taps~Dose, data=Caffeine) 
summary(Caffeine.lm)
```

## Fitting the Model

Using the notation introduced earlier, the parameter estimates are
    $\hat \mu = 244.8$, $\hat \alpha_2 = 1.6$ and
    $\hat \alpha_3 = 3.5$.


Dose (level) | mean response      | fitted value 
--- | --- | --- 
Zero (1)   | $\mu$            | 244.8        
Low (2)      | $\mu + \alpha_2$ | 246.4        
High (3)     | $\mu + \alpha_3$ | 248.3    



## Hypothesis tests

There are three ways we can think of the hypothesis test. 

**First** the model (with $H_0:  ~~~ \mu_1 = \mu_2 = ... =\mu_K$) can be tested with the $F$ test at the bottom of the summary(lm)  output, or applying the anova() command to the output.

```{r anova version}
anova(Caffeine.lm)
```

## Hypothesis tests

**Second** using the treatment constraint the model can be written as multiple linear regression models, using dummy variables:

$$Y_i = \mu + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i \quad(i=1,2,\ldots,n)$$

## Hypothesis tests

The F test for the hypothesis that all slopes are zero gives the same omnibus F test. 
$$H_0: \alpha_2 = ...=\alpha_K = 0$$

```{r regression version}
z2 =   Caffeine$Dose == "Low"
z3 =   Caffeine$Dose == "High" 
Caffeine.lm2 = lm( Caffeine$Taps ~ z2 + z3 )
summary(Caffeine.lm2)$coefficients
```


## Hypothesis tests

**Finally** we can consider the problem as a comparison of two models  

   $$M0:~~~~ Y_i = \mu + \varepsilon_i \quad (i=1,2,\ldots,n)$$
   
   and
    $$M1:~~~~ Y_i = \mu + \alpha_2 z_{i2} + \ldots + \alpha_K z_{iK} + \varepsilon_i  \quad (i=1,2,\ldots,n)$$

## Hypothesis tests

```{r comparison}
Caffeine.M0 = lm(Caffeine$Taps ~ 1)
Caffeine.M1 = lm(Caffeine$Taps ~ z2 + z3)
anova(Caffeine.M0, Caffeine.M1)
```

## Hypothesis tests

The observed *F*-statistic for testing for the significance of caffeine dose is *f* = `r round(summary(Caffeine.lm)$fstatistic["value"], 3)` on *K-1* = `r summary(Caffeine.lm)$fstatistic["numdf"]` and *n-K* = `r summary(Caffeine.lm)$fstatistic["dendf"]` degrees of freedom.

&nbsp;

The corresponding *P*-value is *P*=`r round(anova(Caffeine.lm)["Dose", "Pr(>F)"], 4)`; There is strong evidence against *H~0~*; i.e. Finger tapping speed does seem to be related to caffeine dose. 

## Conclusions from the ANOVA table

```{r}
coef(Caffeine.lm)
```

The coefficients shows that the mean response difference between the high dose and control groups is a little more than twice as big as the difference between low dose and control group.

## Hypothesis tests

**Important**:
    
-- In a multiple linear regression model it makes sense to add or remove a single explanatory variable.

-- In a factor model it (usually) only makes sense to add or remove a whole factor, not single levels of a factor.

-- This will usually require *F* tests (not just t-tests for single parameters).

