---
title: "Lecture 24 - Factors with Interactions"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
library(kableExtra)
library(car)
library(DescTools)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## learning outcomes

-- Design and interpret a factor model with interactions.

&nbsp;

-- Implement a factor model with interactions in R.

## An Intuitive Introduction to Interactions

-- So far, we have considered that the overall treatment effect is purely the addition of the main effects of the factors.

&nbsp;

-- Sometimes this will not be the case because the factors `interact`.

## An Intuitive Introduction to Interactions

Suppose

- Drug 1 on its own reduces blood pressure by 3 units.

  - Factor *A*: drug 1 absent, and drug 1 present.

- Drug 2 on its own reduces blood pressure by 5 units.

  - Factor *B* has two levels: drug 2 absent, and drug 2 present.



## An Intuitive Introduction to Interactions

The `main effects` model for two factors is

$$Y_{ijk} = \mu + \alpha_i + \beta_j + \varepsilon_{ijk}$$

For the two drug example we would have $\alpha_2 = -3$ and $\beta_2 = -5$.

According to the main effects model, the treatment effect with both drugs (relative to the baseline level) is $\alpha_2 + \beta_2 = -8$.

## An Intuitive Introduction to Interactions

To represent an interactive effect between the drugs we need to  introduce an extra term in the model:

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

Here $(\alpha\beta)_{ij}$ is the so-called `interaction term` in the model.

The notation $(\alpha\beta)_{ij}$ represents new parameters  (`NOT` a multiplication of the $\alpha$ and $\beta$  parameters).

## An Intuitive Introduction to Interactions

-- It is necessary to impose constraints on the interaction parameters to avoid overparameterization.

-- The treatment constraint sets all $(\alpha\beta)_{ij}=0$ when either *i* or *j* is one.

-- Hence, in the two drug example, only $(\alpha\beta)_{22}$ is non zero.

## An Intuitive Introduction to Interactions

For the model with interaction, the treatment effect in the presence of both drugs is

$$\alpha_2 + \beta_2 + (\alpha\beta)_{22} = -8 + (\alpha\beta)_{22}$$

so that $(\alpha\beta)_{22}$ here represents the change from the  main effects model that is caused by the interaction of the drugs.

## Two-Way Model with Interaction

For a general model with two factors, the model with interaction is

$$Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

## Two-Way Model with Interaction

-- Both the main effects parameters ($\alpha_i$, $\beta_j$) and the interaction parameters $(\alpha\beta)_{ij}$ must be constrained.

&nbsp;

-- We assume use of the treatment constrained (all  parameters indexed by level 1 of either factor are zero).

## Two-Way Model with Interaction

-- If factor *A* has *K* levels and factor *B* has *L* levels  then there are *(K-1)(L-1)* unconstrained interaction parameters.

&nbsp;

-- Modelling with an interaction term allows any pattern of means between the different factor levels (no particular structure).

## Inference for the Two-Way Model with Interaction

-- Parameter estimation is done by the method of least squares. Fitted values and residuals continue to be defined in the usual way.

-- The first test that is commonly done with the interaction model is to assess whether the interaction term in the model is statistically  significant.

-- In the *F* test, if the interaction term is important, then no more testing is necessary.

## Inference for the Two-Way Model with Interaction

::: callout-important

If a model contains an interaction term then it must also contain the corresponding main effects (even if these appear to be non-significant).

:::

## Testing for an Interaction Between Factors

Testing for the importance of the interaction is equivalent to  comparing models

$$M0:~~~Y_{ijk} = \mu + \alpha_i + \beta_j + \varepsilon_{ijk}$$

$$M1:~~~Y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \varepsilon_{ijk}$$

## Testing for an Interaction Between Factors

Hypothesis tests for parameters

*H~0~*: $$(\alpha\beta)_{ij} = 0~\mbox{for all}~i, j$$

*H~1~*: $$(\alpha\beta)_{ij}~\mbox{not zero for all}~i, j$$

## Testing for an Interaction Between Factors

The *F* test statistic is

$$F = \frac{[RSS_{M0} - RSS_{M1}]/[(K-1)(L-1)]}{RSS_{M1}/(n-KL)}$$

with the corresponding *P*-value computed in the usual manner.

## ANOVA Table for the Interaction Model

<div style="font-size:smaller;"> 

|                           |       Df       |    Sum Sq |   Mean Sq |     F value |     P value |
| :------------------------ | :------------: | --------: | --------: | ----------: | ----------: |
| Factor *A*              | $K-1$     | $SSA$ | $MSA$ | $f_a$ | $P_A$ |
| Factor *B* (adj.*A*) | $L-1$     | $SSB|A$ | $MSB|A$ | $f_{B|A}$ | $P_{B|A}$ |
| Interaction               | $(K-1)(L-1)$ | $SSI$ | $MSI$ | $f_I$ | $P_I$ |
| Residual                  | $n-KL$    | $RSS$ | $RMS$ |             |             |
| Total                     | $n-1$     |   $TSS$ |           |             |             |

</div>

## ANOVA Table for the Interaction Model

The sum of squares for the interaction ($SSI$ in the table) can be computed by  subtraction (difference in RSS for models *M0* and *M1*).

&nbsp;

The *F* test statistic for the interaction is $f_I = MSI/RMS$

## Example: rat diet

This task concerns a complete and balanced experiment into rat weight gain.

Two factors:

1. Protein: either beef or cereal;
2. Amount: either low or high.

We have ten replicates at each treatment.

## Example: rat diet

```{r, echo=-1}
RatDiet <- read.csv(file="./ratdiet.csv", header=TRUE)
table(RatDiet$Protein, RatDiet$Amount)
```

## Example: rat diet

Additive model

```{r, echo=FALSE}
RatDiet$Amount <- factor(RatDiet$Amount, levels = c("Low", "High"))
levels(factor(RatDiet$Amount))
```


```{r RatDiet.add.lm}
RatDiet.add.lm <- lm(Gain ~ Amount + Protein, data=RatDiet)
anova(RatDiet.add.lm)
```

What can you conclude (if anything) about the effect of `Protein` ignoring the effects of `Amount`?

## Example: rat diet

Interaction model

```{r RatDiet.lm}
RatDiet.lm <- lm(Gain ~ Amount * Protein, data=RatDiet)
```

The syntax `A*B` for factors `A` and `B` indicates the main effects of the factors themselves and their  interaction, which is written `A:B`.

  Gain ~ Amount `*` Protein
  
  and  
  
  Gain ~ Amount + Protein `+ Amount:Protein`

are the same formula.

## Example: rat diet

```{r anovaRatDiet.lm}
anova(RatDiet.lm)
```


 *P*-value for interaction between `Protein` and `Amount` is *P=0.055* (marginal), from an *F*-statistic of *3.95* on *1, 36*  degrees of freedom.

Weak evidence of the existence of an interaction.

## Example: rat diet

Summary Table:

```{r summaryRatDiet.lm}
summary(RatDiet.lm)$coefficients
```

## Example: rat diet

```{r}

interaction.plot(RatDiet$Amount, RatDiet$Protein, RatDiet$Gain,
                 fun = mean,
                 col = c("darkgreen", "orange"), xlab = "Feed Amount", ylab = "Mean Gain",
                 trace.label = "Protein")
```

## Fitted Values 

In the previous example, write down the fitted values for each of the four possible treatments.

Obtain these using R code.

```{r}
cbind(RatDiet, RatDiet.lm$fitted.values)
```
