---
title: "Lecture 16 - Polynomial regression models"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
#library(car)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## Learning outcomes

-- Understand the role and interpretation of polynomial terms in regression models.

-- Implement polynomial regression models in R.


## Introduction to Polynomial regression

Simple and multiple regression models are types of `linear models`.

Remember that the word `linear` refers to the fact that the response is linearly related to the `regression parameters`, not to the predictors.

 $$Y_i = \color{cyan}{\beta_0} + \color{cyan}{\beta_1} x_{i}$$

## Introduction to Polynomial regression

-- In most regression models, the relationship between the response and each of the explanatory variables is linear.

&nbsp;

-- We can represent more general relationships using polynomial functions of the explanatory variables.

&nbsp;

-- The resulting models are often referred to as `polynomial regression models`.

## The Polynomial Regression Model

$$Y_i = \beta_0 + \beta_1 x_{i} + \beta_2 x_i^2 + \ldots + \beta_p x_{i}^p
+ \varepsilon_i 
\\
(i=1,\ldots,n)$$ 


where $Y_i$ and $x_i$ are the response and explanatory variable
    observed on the $i$^th^ individual.

## The Polynomial Regression Model

-- $\varepsilon_1, \ldots \varepsilon_n$ are random errors satisfying
    the usual assumptions:
    
    
  * [A1.]{.highlight} The expected value of the errors $E[\varepsilon_i] = 0$ for i=1,...,n.

  * [A2.]{.highlight} The errors $\varepsilon_1, \ldots ,\varepsilon_n$ are independent.
  
  * [A3.]{.highlight}  The error variance $\mbox{var}(\varepsilon_i) = \sigma^2$ (an unknown constant) for i=1,...,n.

  * [A4.]{.highlight} The errors $\varepsilon_1, \ldots ,\varepsilon_n$ are normally distributed.

&nbsp;

-- $\beta_0, \beta_1, \ldots, \beta_p$ are the regression parameters.

## The Polynomial Regression Model

The polynomial regression equation and [A1]{.highlight} imply that
$$E[Y_i] = \mu_i = \beta_0 + \beta_1 x_{i} + \beta_2 x_i^2 + \ldots + \beta_p x_{i}^p$$

&nbsp;

With the powers of the predictor ($x_i^2, x_i^3$, etc) in the model the relationship between the predictor and the response variables is `no longer a line but a curve`.

&nbsp;

Adding higher power terms results in more complex shapes.

## Notes on The Polynomial Regression Model

We have defined a $p^{th}$ order polynomial regression model.

&nbsp;

-- Typically statisticians aim for low order polynomial models.

&nbsp;

-- having $p$ larger than 5 or 6 is unusual.

 

## Notes on The Polynomial Regression Model

 The hierarchical principle. As an example, a cubic polynomial should include linear and quadratic terms:
        $$E[Y_i] = \beta_0 + \beta_1 x_{i} + \beta_2 x_i^2 + \beta_3 x_{i}^3$$
    is correct; we should `not` omit (e.g.) the quadratic term:
        $$E[Y_i] = \beta_0 + \beta_1 x_{i} + \beta_3 x_{i}^3$$

## Child Lung Function Data


-- Response variable: FEV (forced expiratory volume) is a measure of lung function.

-- The data include determinations of FEV on 318 female children who were seen in a childhood respiratory disease study in Massachusetts in the U.S.

-- Child's age (explanatory variable, in years) was also recorded to model FEV as function of it.

&nbsp;

<span style="font-size: small;">
Tager, I. B., Weiss, S. T., Rosner, B., and Speizer,
F. E. (1979). Effect of parental cigarette smoking on pulmonary function in children. American Journal of Epidemiology, 110, 15-26.
</span>




## Scatter Plot

```{r getFevData, echo=-1, eval=TRUE}
# Load the dataset
Fev <- read.csv(file="./fev.csv", header=TRUE)
plot(FEV~Age, data=Fev)
```

The relationship between FEV and age is not linear. Will try to model FEV as a polynomial in age.


## Traditional model Fitting in R

A sixth order polynomial model can be fitted with

&nbsp;

```r
lm(FEV~Age+I(Age^2)+I(Age^3)+I(Age^4)+I(Age^5)+I(Age^6),data=Fev)
```
&nbsp;

Notice the function `I()`nhibit. This function "protects" its contents, in the sense that R reads the contents as a normal mathematical expression.

Modern practice is to avoid all that typing though!

## Sixth Order Model: modern form

```{r Fev.lm.6}
Fev.lm.6 <- lm(FEV~poly(Age,6, raw=TRUE), data=Fev)
summary(Fev.lm.6)
```
Thinking in the  hierarchical principle, The coefficient of the $6^{th}$ order is not significant. We can remove the term.

## Fifth Order Model

```{r Fev.lm.5}
Fev.lm.5 <- lm(FEV~poly(Age,5, raw=TRUE), data=Fev)
summary(Fev.lm.5)
```

The coefficient of the  $5^{th}$ order is not significant, we can remove the term from the model as well.

## Fourth Order Model

```{r Fev.lm.4}
Fev.lm.4 <- lm(FEV~poly(Age,4, raw=TRUE), data=Fev)
summary(Fev.lm.4)
```

`The fourth order term is statistically significant`. This is our preferred model.

## Fourth Order Model

Our fitted model equation is:
    $$
    E[\mbox{FEV}] = 3.57 - 1.39\text{ Age} + 0.27\text{ Age}^2  
    \\
    -0.018\text{ Age}^3 + 0.00041\text{ Age}^4
    $$

## Fourth Order Model

```{r Fev.lm.4.plot, echo = TRUE}
newdat <- data.frame(Age = seq(min(Fev$Age), max(Fev$Age), length.out = 100))
newdat$pred <- predict(Fev.lm.4, newdata = newdat)
plot(FEV~Age, data=Fev)
with(newdat, lines(x = Age, y = pred, col = "red", lwd = 2))
```


## Fourth Order Model

```{r Fev.lm.4.plot2, echo = FALSE}
newdat <- data.frame(Age = seq(min(Fev$Age), max(Fev$Age), length.out = 100))
newdat$pred <- predict(Fev.lm.4, newdata = newdat)
plot(FEV~Age, data=Fev)
with(newdat, lines(x = Age, y = pred, col = "red", lwd = 2))
```

-- The fitted model is adequate for ages from 5 to 15 years.

-- The ends of the plot exhibits spurious upward curvature. This is an artifact of the form of the polynomial rather than a characteristic of the data.



## Collinearity in the Fitted Model

-- For the 6th and 5th order models none of the regression coefficients are significant.

&nbsp;

-- This is due to `collinearity` between the powers of `Age`.

&nbsp;

-- `Collinearity` in polynomial regression can lead to `numerical instability` in software packages.


## Numeric instability example

Let's use a predictor taking on values from ten to one-hundred in steps of ten.

```{r getNewX}
x=(1:10)*10; x
```

and take it out to the sixth power...

```{r sixthPower}
x^6
```

When these two vectors end up in the same model matrix, the magnitude of the largest value swamps the little numbers in the matrix when finding the matrix $(X^t X)^{-1}$ fitting the linear model.

