---
title: "Lecture 15 - Models with binary predictors"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: breezedark
    #highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: format.css
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
#library(car)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

# Load the dataset
Climate <- read.csv("climate.csv")

```


## Learning outcomes



-- Implement a binary predictor in model design.

-- Implement a binary predictor with a numeric predictor



## Comparing groups using lm()

Suppose we denote the $i$th data value (response) from group $j$ by $Y_{ij}$, where $j= 0,1$ and $i=1,2,…,n_0$ or $1,2,…,n_1$   respectively.  Thus the group means are  $\mu_0$  and $\mu_1$ .

One way to analyse the data would be to use a two-sample t-test 

-- $H_0:~~\mu_0= \mu_1$ vs

-- $H_1:~~\mu_0 \ne\mu_1$.


## Comparing groups using lm()

Another way of writing down the hypothesis:

Represent one group as a `baseline group`  with mean $\mu$

and the other group as a  `shifted` or `treatment` with mean  $\mu+\delta$.

$\mu+\delta$ represents how much the mean of the treatment is away from the baseline group.

Then the hypotheses become

$H_0:~\delta=0$  vs

$H_1:~\delta\ne 0$.


## Comparing groups using lm()

To write  it this way we will define an `indicator variable`  as $z_i = 0$ if  the datum is in group 0, and $z_i=1$ if it is in group 1. 
   
Then we can write a formula for the response as    
$$ E[Y_i] = \mu + \delta z_i   = \beta_0 + \beta_1 z_i$$
where $\beta_0 = \mu$ and $\beta_1 = \delta$.


## Comparing groups using lm()

In other words the two-group comparison can be put in the same form as a linear regression. So  all we would need to do is:

1. set up the group indicator variable `z`, 

2. regress `Y` on `z` and 

3. test the hypothesis that the slope = 0. Therefore, there is no change in mean from the baseline group.

## An example with the climate data

The Climate data example has two binary predictors `NorthIsland` and `Sea`. In this lecture we will focus on the predictor `NorthIsland` only.


-- where 0 is `Belongs to the South Island`

-- and 1 is `Belong to the North island`.

## An example with the climate data

Using boxplot `NorthIsland` seems to have an effect on temperature.


```{r Climate boxplots, eval = -2, echo = -1}
boxplot(MnJlyTemp ~ NorthIsland, data = Climate)
```
is the effect statistically significant?



## Testing using lm()

Testing the significance of the relationship between `NorthIsland` and `MnJlyTemp`.

```{r lmTesting}
Climate.lm1 <- lm(MnJlyTemp ~ NorthIsland, data = Climate)
summary(Climate.lm1) 
```
## Testing using lm()

The t-test for the slope of the model is significant (`p`-value = `1.66e-05`), 

The mean difference of 3.58°C in July is significantly  higher in the North Island compared to the South Island.

## Testing using t.test()

```{r t.test Testing}
t.test(MnJlyTemp ~ NorthIsland, data = Climate)
``` 
 
Looking at the result the conclusion is the same but the p-value and df are not.

## Testing using t.test()

Remember `lm()`  assumes the variance of residuals does not depend on the values of the predictor. From the boxplot it looks like this assumption is justified.

The reason is that the default settings for `t.test()` do not assume equal variances, and so uses a method called Welch's method. Essentially it does a fudge on the df  to make the t-distribution work.   

## Testing using t.test()

If we want to get exactly the same results in the `t.test()` as in the `lm()`, then we need to set the var.equal=TRUE option.

```{r t.testVar.equal}
t.test(MnJlyTemp~NorthIsland, var.equal=TRUE, data=Climate)
``` 

In this case the t statistics, df and `p`-value are equal to those from `lm`.

## How do we know whether to assume equal variances?

A simple rule of thumb we can use is that if the standard deviation of residuals for `the two groups does not differ by more than a factor of 2`, then we can assume equal variances.

```{r Compare sd}
with(Climate, sd( MnJlyTemp[NorthIsland==0] )/ sd( MnJlyTemp[NorthIsland==1] ))
```
Clearly the ratio is much less than 2.


## How do we know whether to assume equal variances?

There are two formal tests commonly used to compare variances:

-- $H_0: \sigma_0^2 = \sigma_1^2$ (variances are equal)  vs 

-- $H_1:\sigma_0^2 \neq \sigma_1^2$ (variances are not equal). 


## How do we know whether to assume equal variances?
 
Bartlett's test (which depends more on the errors being normally distributed)

```{r Climate.bt}
bartlett.test(MnJlyTemp ~ factor(NorthIsland), data=Climate)
```

## How do we know whether to assume equal variances?

Levene's test (which works even if the data are not normal).  

We will normally use the Levene's test. The corresponding R function is available through the `car` package.


```{r}
library(car)
leveneTest(MnJlyTemp ~ factor(NorthIsland), data=Climate)
```
The `p`-value for both these tests is large,  which confirms that, for the Climate data, we have no reason to doubt the equal variance  assumption.


## single numeric predictor and a binary group variable

Binary predictors can also be used to deal with changes in slope between two groups.

For example, in their book "Regression Analysis by Example", Chatterjee and Price present a dataset on English exam scores, revision time, and gender.. Let:

## single numeric predictor and a binary group variable

- $Y_i$	= score in English exam,
-	$x_i$	= number of hours spent in revision,
- $z_i$ = sex (0 for boys, 1 for girls).

```{r getEnglish, eval = -2, echo = 0}
English <- read.csv(file="./EnglishExam.csv", header=TRUE)
#head(English)
```

## single numeric predictor and a binary group variable

Looking at the data graphically we see:

```{r EnglishEDA, echo = FALSE}
English |> 
  ggplot(aes(y=resulty, x= revision, group= sex, colour = as.factor(sex))) +  
  geom_smooth(formula = y ~ x, method = "loess", alpha = 0.3) +
  geom_point() +
  labs(x = "Revision (hours)",
       y = "Result",
       colour = "Sex") + 
  scale_colour_brewer(palette = "Set1")
```

## single numeric predictor and a binary group variable

The question we are asking is: should the lines for boys and girls have a different slope?

Test for a difference in slope by setting a `baseline group` (say, `the boys`) with slope $\beta_1$, and expressing the slope for `the girls` group as a `shifted` slope $\beta_1 + \delta$.

$$E[Y_i] = \beta_0 + \beta_1 x_i$$ `for boys`, and

## single numeric predictor and a binary group variable

for `the girls`

$$E[Y_i] = \beta_0 + (\beta_1 + \delta) x_i ~\mbox{ for girls}$$

The group indicator variable $z_i$ is very convenient to construct the appropriate model:

$$E[Y_i] = \beta_0 + (\beta_1 + \delta z_i) x_i = \beta_0 + \beta_1 x_i + \delta x_i z_i$$

Here, $x_i z_i$ corresponds to a new variable that is equal to 0 for boys and to $x_i$ for girls.

## single numeric predictor and a binary group variable

 In order to fit the model with `lm()` in the usual way, we can construct this new variable.

```{r constructNewVariable}
English$revisionGirls <- English$revision * English$sex
head(English)
```


## single numeric predictor and a binary group variable

We can now fit the model:

```{r DiffSlopesModels}
English.lm1 <- lm(resulty ~ revision, data = English)
English.lm2 <- lm(resulty ~ revision + revisionGirls, data = English)
summary(English.lm2)
```

## single numeric predictor and a binary group variable
```{r}
anova(English.lm1, English.lm2)
```

The coefficient associated with our constructed variable `revisionGirls` is significant (`p`-value $\approx 0.04$), which indicates that there is a difference in the slope between boys and girls.




