---
title: "Lab 11 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
library(car)
library(glmnet)
```

# Computer Laboratory 11: Search, Shrinkage, and Weighting

Last week we built and compared several Auto MPG models by hand. That worked because the candidate set was still manageable.

This week we keep the same data but make the problem harder. We first automate part of the model search, then look at what happens when several predictors carry almost the same information. In the second part, the problem changes: the magazine data contain group means with very different levels of precision.

The lab uses [auto_mpg.csv](https://knowlton.co.nz/161251/data/auto_mpg.csv) and [AveMags.csv](https://knowlton.co.nz/161251/data/AveMags.csv). The downloadable R Markdown file reads both datasets directly from the course site.

# Part A: Auto MPG

```{r import-auto, message=FALSE}
auto <- read_csv("https://knowlton.co.nz/161251/data/auto_mpg.csv") |>
  mutate(origin = factor(origin))

levels(auto$origin)
```

Europe is the reference level for `origin`. As in Lab 10, we will model `mpg` and exclude `car_name` from the candidate predictors.

## Search from both ends

Let's start with the same response and candidate predictors from Lab 10. The null model contains only an intercept; the full model contains the seven predictors we are willing to consider. We will run the search from both ends and see whether the two paths finish in the same place.

```{r search-models}
null_model <- lm(mpg ~ 1, data = auto)

full_model <- lm(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)
```

Fit the null and full Auto MPG models shown here. Run forward search from the null model and backward search from the full model using the same AIC scope. Print each search path and final formula.

```{r run-searches}
search_scope <- list(
  lower = formula(null_model),
  upper = formula(full_model)
)

forward_search <- step(
  null_model,
  scope = search_scope,
  direction = "forward",
  trace = 0
)

backward_search <- step(
  full_model,
  scope = search_scope,
  direction = "backward",
  trace = 0
)

```

```{r search-paths}
forward_search$anova
backward_search$anova

formula(forward_search)
formula(backward_search)
```

Discuss: Did the two routes end at the same model? What would agreement or disagreement tell us about this particular search, and what would it *not* establish?

**Solution.** Forward search adds `weight`, `model_year`, `origin`, `displacement`, `horsepower`, and then `cylinders`. Backward search starts from the full model and removes only `acceleration`.

Both searches therefore finish at

```text
mpg ~ cylinders + displacement + horsepower + weight +
      model_year + origin
```

They agree for these data and this candidate set. Stepwise procedures can still be path-dependent: at each stage they make a local decision from the models currently available. Agreement does not establish that this is the uniquely best model, and it does not account for models or terms outside the specified scope.

Compare the AIC values for the selected and full models.

```{r compare-search-aic}
AIC(full_model, forward_search, backward_search)
```

**Solution.** The two searches agree and end about 1.3 AIC units below the full model. That is a small separation, so the lower AIC by itself is a weak reason to prefer the selected formula. The full model keeps `acceleration`; we should consider whether that term matters for the modelling question, and how we would assess prediction on new cars.

## Keep `origin` in every candidate model

The first search allowed every candidate term to be removed. Now repeat the same idea after making one modelling decision in advance: `origin` must stay in the model.

Fit a lower-bound model containing `origin`, then repeat backward search with that model as the lower scope limit. Print the selected formula.

**Solution.** Put `origin` in the lower model. It then becomes part of the lower boundary of the search.

```{r forced-origin}
origin_model <- lm(mpg ~ origin, data = auto)

forced_origin_search <- step(
  full_model,
  scope = list(
    lower = formula(origin_model),
    upper = formula(full_model)
  ),
  direction = "backward",
  trace = 0
)

formula(forced_origin_search)
```

Discuss: What decision did we make before the search? Why could an automatic AIC step not remove `origin` this time?

This controls which models the algorithm is allowed to visit. It does not make the other statistical limitations of stepwise search disappear.

## How much do the numerical predictors overlap?

The search has given us a candidate formula, but several of these predictors describe related features of a car. Start with the pairwise correlations among `cylinders`, `displacement`, `horsepower`, `weight`, `acceleration` and `model_year`. Then use VIFs to ask how strongly each term overlaps with the remaining predictors in the full model.

```{r correlation-matrix}
auto |>
  select(
    cylinders,
    displacement,
    horsepower,
    weight,
    acceleration,
    model_year
  ) |>
  cor() |>
  round(3)
```

**Solution.** `cylinders`, `displacement`, `horsepower`, and `weight` are strongly correlated. For example, the correlations of cylinders with displacement and displacement with weight are about 0.95 and 0.93. These variables are all carrying information about related aspects of vehicle size and power.

Now inspect variance inflation factors for the full model.

```{r vif-full}
car::vif(full_model)
```

Discuss: Which conditional coefficients might be hard to estimate separately? Why would a high VIF alone be an inadequate reason to delete one of them?

**Solution.** For the one-degree-of-freedom numerical terms, the `GVIF` column is the usual VIF. The values are about 10.7 for `cylinders`, 22.9 for `displacement`, 10.0 for `horsepower`, and 11.1 for `weight`. These values match what we saw in the correlation matrix: the model has difficulty separating the conditional contributions of these overlapping predictors.

`origin` has two degrees of freedom, so its raw GVIF is not directly comparable with the one-degree-of-freedom terms. The adjusted `GVIF^(1/(2*Df))` column is provided for that comparison.

## What happens to the coefficients?

The VIFs tell us that some coefficients are difficult to separate. Now look at what that does to the estimates themselves.

Fit separate simple regressions for `cylinders`, `displacement`, `horsepower` and `weight`. Compare each simple slope with the corresponding coefficient in the full model.

```{r coefficient-comparison}
overlapping_predictors <- c(
  "cylinders",
  "displacement",
  "horsepower",
  "weight"
)

simple_models <- map(
  overlapping_predictors,
  ~ lm(reformulate(.x, response = "mpg"), data = auto)
)

coefficient_comparison <- tibble(
  Predictor = overlapping_predictors,
  `Simple coefficient` = map2_dbl(
    simple_models,
    overlapping_predictors,
    ~ unname(coef(.x)[.y])
  ),
  `Full-model coefficient` = unname(
    coef(full_model)[overlapping_predictors]
  ),
  `Full-model SE` = unname(
    coef(summary(full_model))[overlapping_predictors, "Std. Error"]
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

knitr::kable(coefficient_comparison)
```

Discuss: Which signs or magnitudes changed? Describe the full-model coefficient as an association conditional on the other included predictors; do not interpret a sign reversal as a causal benefit.

**Solution.** Every simple-regression coefficient is negative: cars with more cylinders, greater displacement, more horsepower, or greater weight tend to have lower `mpg`. In the combined model, the coefficient magnitudes change substantially and the displacement coefficient becomes positive.

That sign change does not mean greater displacement causes fuel economy to improve. The full-model displacement coefficient compares cars with the other included predictors held fixed, including weight, cylinders, and horsepower. The data contain relatively little independent variation with which to make that comparison. The fitted model can still describe or predict the response reasonably while its individual conditional coefficients are difficult to estimate precisely.

## Ridge and lasso on the same folds

We have now seen the problem with trying to separate strongly overlapping predictors using ordinary least squares. Ridge and lasso take a different approach: keep the larger predictor set, then control the coefficient sizes with a penalty.

Build the model matrix below and use one fold assignment for both `cv.glmnet()` fits. Keeping the folds fixed matters here because we want the difference between ridge and lasso to come from the penalty, not from a different random split.

```{r glmnet-data}
X <- model.matrix(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)[, -1]

y <- auto$mpg
```

Last week we introduced cross-validation as a way to evaluate performance on held-out observations. Here `cv.glmnet()` uses the same idea to choose the amount of shrinkage. We use the same folds for ridge and lasso so that their cross-validation results are directly comparable.

```{r cross-validated-penalties}
set.seed(161251)
fold_id <- sample(rep(1:10, length.out = nrow(auto)))

ridge_cv <- cv.glmnet(
  X,
  y,
  alpha = 0,
  foldid = fold_id
)

lasso_cv <- cv.glmnet(
  X,
  y,
  alpha = 1,
  foldid = fold_id
)
```

### Compare the coefficient paths

Plot the ridge and lasso coefficient paths. As the penalty becomes stronger, what happens to the coefficients under each method?

```{r coefficient-paths, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge coefficient paths")
plot(lasso_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso coefficient paths")
par(mfrow = c(1, 1))
```

**Solution.** Both methods shrink coefficients towards zero as the penalty increases. Ridge shrinks the correlated predictors together and does not set them exactly to zero. The lasso paths can reach zero, so lasso can produce a smaller active set of predictors.

The predictors use different units, so the vertical sizes of the returned coefficients should not be read as a direct ranking of predictor importance. `glmnet` standardises internally while fitting, then reports coefficients on the original scales.

### Choose the penalty by cross-validation

Now compare the cross-validation curves. The two vertical reference values answer slightly different questions: `lambda.min` minimises the estimated cross-validation error, while `lambda.1se` deliberately accepts stronger regularisation while staying within one standard error of that minimum.

```{r cv-curves, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv)
title("Ridge cross-validation")
plot(lasso_cv)
title("Lasso cross-validation")
par(mfrow = c(1, 1))

penalty_choices <- tibble(
  Method = c("Ridge", "Lasso"),
  `lambda.min` = c(ridge_cv$lambda.min, lasso_cv$lambda.min),
  `lambda.1se` = c(ridge_cv$lambda.1se, lasso_cv$lambda.1se)
)

knitr::kable(penalty_choices, digits = 4)
```

Discuss: What trade-off is made by choosing `lambda.1se` rather than `lambda.min`?

**Solution.** `lambda.min` gives the smallest mean cross-validated error among the fitted penalty values. `lambda.1se` chooses the most strongly regularised model whose estimated error is within one standard error of that minimum. It accepts a small amount of uncertainty in predictive performance in exchange for stronger shrinkage and, for lasso, usually a simpler active set.

Cross-validation estimates performance on held-out folds from this dataset. It is more relevant to prediction than training RSS or adjusted R-squared, but it is still an estimate with sampling and fold-to-fold uncertainty.

### Which coefficients remain?

```{r penalised-coefficients}
round(as.matrix(coef(ridge_cv, s = "lambda.min")), 4)
round(as.matrix(coef(ridge_cv, s = "lambda.1se")), 4)

round(as.matrix(coef(lasso_cv, s = "lambda.min")), 4)
round(as.matrix(coef(lasso_cv, s = "lambda.1se")), 4)
```

Discuss: Which coefficients are exactly zero under lasso? What trade-off does `lambda.1se` make? Why is a zero dummy coefficient for one `origin` level different from removing the `origin` term as a whole?

**Solution.** Ridge retains every coefficient at both penalty choices, but shrinks them. At `lambda.min`, lasso sets `cylinders`, `displacement`, and `acceleration` to zero. At the more strongly regularised `lambda.1se`, it also sets the Japan-versus-Europe origin contrast to zero.

For this particular fold assignment, lasso retains weight and horsepower from the strongly correlated size-and-power group while dropping cylinders and displacement. This is one way lasso can behave with correlated predictors: it may retain some members of an overlapping group and remove others. It is not a general promise about which member will be chosen, and a different sample or fold assignment could change the active set.

# Part B: Weighted regression for magazine readership

So far we have been changing how predictors enter the mean model. Weighted regression solves a different problem. Here each row is already a group mean, and some of those means are estimated much more precisely than others.

Each row in the magazine data is the mean number of magazines read for one age and sex group. The columns `N` and `StDev` give the group size and within-group standard deviation.

```{r import-magazines, message=FALSE}
mags <- read_csv("https://knowlton.co.nz/161251/data/AveMags.csv") |>
  mutate(
    Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
    Older = as.integer(Age >= 50)
  )

glimpse(mags)
```

## Start with the mean model

Plot `Avemags` against `Age`, marking `Sex`, group size `N` and within-group standard deviation `StDev`.

```{r magazine-data-plot}
ggplot(mags, aes(x = Age, y = Avemags)) +
  geom_point(aes(size = N, colour = StDev, shape = Sex)) +
  labs(
    x = "Age",
    y = "Mean number of magazines",
    size = "Group size",
    colour = "Within-group SD"
  ) +
  theme_minimal()
```

Before changing the weights, keep the mean structure fixed. The supplied variable `Adult` separates the teenage rows from the adult rows, and `Older` marks ages 50 and above. We will allow the age slope to change across those ranges while keeping the sex difference additive.

```{r magazine-mean-model}
magazine_unweighted <- lm(
  Avemags ~ Age * Adult + Age * Older + Sex,
  data = mags
)
```

The teenage slope is the coefficient of `Age`. The adult slope adds the `Age:Adult` coefficient, and the older-adult slope also adds `Age:Older`. The main effects for `Adult` and `Older` allow level changes at the boundaries, so the fitted pieces are not forced to meet exactly.

```{r magazine-unweighted-fit}
mags |>
  mutate(fitted = fitted(magazine_unweighted)) |>
  ggplot(aes(x = Age, y = Avemags, colour = Sex)) +
  geom_point() +
  geom_line(aes(y = fitted)) +
  labs(
    x = "Age",
    y = "Mean number of magazines"
  ) +
  theme_minimal()
```

Discuss: What broad age pattern does this unweighted model capture? What does the absence of an age-by-sex interaction assume about the fitted sex difference?

**Solution.** The unweighted fit increases through the teenage ages, is approximately flat through adulthood, and decreases for the older group. Females have a higher fitted mean than males at each age because the model contains an additive sex effect but no age-by-sex interaction.

## Which magazine means are most precise?

Compute each group's precision weight as `N/StDev^2` and list the groups from highest to lowest weight.

For a mean based on $N_i$ independent people with within-group standard deviation $SD_i$, its sampling variance is proportional to

\[
\frac{SD_i^2}{N_i}.
\]

We therefore use inverse-variance weights proportional to

\[
w_i = \frac{N_i}{SD_i^2}.
\]

```{r magazine-weights}
mags <- mags |>
  mutate(precision_weight = N / StDev^2)
```

```{r inspect-magazine-weights}
mags |>
  select(Age, Sex, N, StDev, precision_weight) |>
  arrange(desc(precision_weight))
```

Discuss: Why do both group size and within-group variation enter the weight?

**Solution.** The age-67 groups receive by far the largest weights because they combine large sample sizes with relatively modest standard deviations. The teenage groups generally receive much less weight. The 16-year-old male group has the smallest weight because its estimated mean is based on only 18 people and has a large within-group standard deviation.

## Hold the mean formula fixed and change the weights

Now change one thing: fit the same mean formula with `weights = precision_weight`. Compare its coefficients and fitted values with the unweighted model. Then compare ordinary residuals from the unweighted fit with $\sqrt{w_i}(y_i-\hat y_i)$ from the weighted fit.

```{r fit-weighted-model}
magazine_weighted <- lm(
  Avemags ~ Age * Adult + Age * Older + Sex,
  data = mags,
  weights = precision_weight
)

coefficient_comparison <- tibble(
  Term = names(coef(magazine_unweighted)),
  Unweighted = unname(coef(magazine_unweighted)),
  Weighted = unname(coef(magazine_weighted))
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(coefficient_comparison)
```





**Solution.** Weighting changes the fit because the highly weighted adult and older groups now have more influence than the relatively imprecise teenage groups. The fitted female-minus-male difference decreases from about 1.13 magazines in the unweighted model to about 0.72 in the weighted model. The weighted teenage fitted values, especially for females, move down, while several fitted values for older males move up slightly.

The weighting has not changed the response or the mean-model formula. It has changed the criterion being minimised: discrepancies from more precise group means count more heavily.

```{r compare-magazine-fits}
magazine_fits <- bind_rows(
  mags |>
    mutate(
      fitted = fitted(magazine_unweighted),
      fit = "Unweighted"
    ),
  mags |>
    mutate(
      fitted = fitted(magazine_weighted),
      fit = "Weighted"
    )
)

ggplot(magazine_fits, aes(x = Age, colour = Sex)) +
  geom_point(aes(y = Avemags, size = precision_weight), alpha = 0.5) +
  geom_line(aes(y = fitted, linetype = fit), linewidth = 0.9) +
  labs(
    x = "Age",
    y = "Mean number of magazines",
    size = "Precision weight",
    linetype = "Fit"
  ) +
  theme_minimal()
```

### Ordinary and weighted residuals

For the unweighted model, inspect the ordinary residuals. For the weighted model, inspect

\[
\sqrt{w_i}(y_i - \hat{y}_i),
\]

which places the residuals on the scale used by weighted least squares.

```{r compare-magazine-residuals}
residual_comparison <- bind_rows(
  mags |>
    transmute(
      Age,
      Sex,
      residual = resid(magazine_unweighted),
      model = "Unweighted: ordinary residuals"
    ),
  mags |>
    transmute(
      Age,
      Sex,
      residual = resid(magazine_weighted) * sqrt(precision_weight),
      model = "Weighted: weighted residuals"
    )
)

ggplot(residual_comparison, aes(x = Age, y = residual, colour = Sex)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point() +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  labs(x = "Age", y = "Residual") +
  theme_minimal()
```

Discuss: Which fitted contrasts change? What assumption about the precision of a group mean justifies the weights? Why do the two residual panels have different scales?

**Solution.** The ordinary residual plot treats every group mean as equally precise even though the sample sizes and standard deviations differ considerably. The weighted residuals rescale each discrepancy by the square root of its precision weight, so the two panels are on different scales. A raw discrepancy from a precise group mean matters more; the same discrepancy from a noisy group mean matters less.

We should not compare the vertical magnitudes across the two panels as if they were in identical units. The useful comparison is whether the residual structure is reasonable on the error scale assumed by each fit. Some variation and individual large residuals remain, but the weighted analysis no longer asks the low-information teenage rows to influence the fit as much as the large, more precise older groups.

## Final decision

In one short paragraph, state when you would use term-based search, ridge or lasso, and weighted regression in these two examples. Explain what you would need to assess before claiming the selected Auto MPG model predicts new cars well.

The two parts of this lab solve different problems:

- Stepwise search and penalised regression respond to a large or redundant predictor set.
- VIFs and coefficient comparisons diagnose why individual conditional effects can become unstable when predictors overlap.
- Cross-validation chooses the amount of shrinkage using held-out predictions.
- Weighted regression changes how much each observation contributes when the observations have different precisions.

None of these methods chooses a scientifically meaningful model on its own. The model still depends on the question, the available variables, and the assumptions connecting the fitted calculation to that question.

Everything in this lab still assumes that we have chosen a sensible shape for the mean. In Lab 12 we relax that too: what do we do when a straight-line effect is not enough?
