---
title: "Lab 10 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

Let's work through each complete analysis with the wage data first, then repeat the whole workflow with Auto MPG.

The wage data contain 3,000 male workers recorded from 2003 to 2009. Each row is one worker. Wage is recorded in thousands of US dollars. The Auto MPG data contain 392 cars from model years 1970 to 1982. Each row is one car.

Both datasets are observational, so the fitted relationships describe associations rather than causal effects.

| Wage variable | Meaning |
|:--|:--|
| `wage` | recorded wage, thousands of US dollars |
| `age` | age, years |
| `education` | education level |
| `year` | year the wage was recorded |
| `jobclass` | industrial or information work |

| Auto MPG variable | Meaning |
|:--|:--|
| `mpg` | fuel economy, miles per gallon |
| `weight` | vehicle weight, pounds |
| `origin` | USA, Europe or Japan |
| `model_year` | model year, 70--82 |
| `horsepower` | engine power, horsepower |

Let's leave `car_name` out of the modelling.

```{r import-data, message=FALSE}
wage <- readr::read_csv("../data/wage.csv") |>
  mutate(education = factor(education))

auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(wage)
glimpse(auto)
```

## What changes across groups?

### Wage example

Let's work through the complete three-model comparison using wage and age.

First plot wage against age, with education shown by colour.

```{r wage-raw-plot}
ggplot(wage, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Discuss: How does wage vary with age? Do the education groups appear to have similar vertical positions and slopes?

Now fit the three models.

```{r wage-fit-lines}
w1 <- lm(wage ~ age, data = wage)
w2 <- lm(wage ~ age + education, data = wage)
w3 <- lm(wage ~ age * education, data = wage)
```

#### One common line

Plot the fitted values from `w1`.

```{r wage-one-line}
wage |>
  mutate(fitted = predict(w1)) |>
  arrange(age) |>
  ggplot(aes(x = age, y = wage)) +
  geom_point(aes(colour = education), alpha = 0.2) +
  geom_line(aes(y = fitted), colour = "black", linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What is this model assuming about the education groups?

It assumes one common age-wage relationship for all education groups. Education is not in the model.

#### Parallel lines

Now plot the fitted values from `w2`.

```{r wage-parallel-lines}
wage |>
  mutate(fitted = predict(w2)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What changed from the one-line model? What is still being held the same across education groups?

Education can shift the fitted lines vertically, but the age slope is common across groups.

#### Separate lines

Finally, plot the fitted values from `w3`.

```{r wage-separate-lines}
wage |>
  mutate(fitted = predict(w3)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What changed from the parallel-lines model? Which term in the formula allows that change?

The interaction allows the age slopes to differ across education groups.

```text
education * age = education + age + education:age
```

#### Read the separate-lines coefficients

Run the usual model summary.

```{r wage-read-interaction}
summary(w3)
```

The `age` coefficient is the age slope for the reference education group. The education coefficients change the intercept, and the age-by-education coefficients change the slope. Age zero is outside these data, so the intercept differences are not useful overall wage comparisons.

### Auto MPG: repeat the complete analysis

Now repeat the whole comparison using fuel economy, weight and origin.

First plot `mpg` against `weight`, with `origin` shown by colour.

```{r auto-raw-plot}
ggplot(auto, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  )
```

Discuss: How does fuel economy vary with weight? Do the origin groups appear to have similar vertical positions and slopes?

Fuel economy falls as weight increases. The origins occupy different parts of the weight range, and the fitted relationship may differ in both position and slope.

Fit the three models below.

```text
a1: mpg ~ weight
a2: mpg ~ weight + origin
a3: mpg ~ weight * origin
```

```{r auto-fit-lines}
a1 <- lm(mpg ~ weight, data = auto)
a2 <- lm(mpg ~ weight + origin, data = auto)
a3 <- lm(mpg ~ weight * origin, data = auto)
```

#### One common line

Plot the fitted values from `a1`.

```{r auto-one-line}
auto |>
  mutate(fitted = predict(a1)) |>
  arrange(weight) |>
  ggplot(aes(x = weight, y = mpg)) +
  geom_point(aes(colour = origin), alpha = 0.3) +
  geom_line(aes(y = fitted), colour = "black", linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

Discuss: What is this model assuming about the origin groups?

It assumes one common weight-MPG relationship across all origins.

#### Parallel lines

Plot the fitted values from `a2`.

```{r auto-parallel-lines}
auto |>
  mutate(fitted = predict(a2)) |>
  arrange(origin, weight) |>
  ggplot(aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fitted, group = origin), linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

Discuss: What changed from `a1`? What is still being held the same across origins?

Origin can shift the fitted lines vertically, but the weight slope is common.

#### Separate lines

Plot the fitted values from `a3`.

```{r auto-separate-lines}
auto |>
  mutate(fitted = predict(a3)) |>
  arrange(origin, weight) |>
  ggplot(aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fitted, group = origin), linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

Discuss: What changed from `a2`? Which term allows the weight slopes to differ?

The `weight:origin` interaction allows the slopes to differ.

#### Read the separate-lines coefficients

Run `summary(a3)`.

```{r auto-read-interaction}
summary(a3)
```

USA is the reference origin, so `weight` is the USA slope. `originEurope` and `originJapan` change the intercepts, while `weight:originEurope` and `weight:originJapan` change the slopes. Because the interaction is present, the main-effect origin coefficients describe differences at zero pounds, which is outside the data.

## Do the groups improve the model?

A partial F test compares a reduced model with a larger model containing all of the reduced model's terms. It tests the added terms together.

### Wage example

Let's complete the set of nested comparisons before changing datasets.

#### Does education add information after age?

```{r compare-wage-education}
anova(w1, w2)
```

Four education parameters are added. The P-value is below 0.001, so there is evidence that education improves the model after age is already included.

#### Do the age slopes differ by education?

```{r compare-wage-slopes}
anova(w2, w3)
```

Four interaction coefficients are added. The P-value is about 0.013, giving evidence that at least some age slopes differ by education.

#### What does the full comparison test?

```{r compare-full-wage}
anova(w1, w3)
```

The wage comparison adds eight parameters and strongly favours the separate-lines model. This is a joint test and does not say that every added coefficient is individually useful.

#### Does term order matter?

A sequential ANOVA table gives each term credit for variation it explains when it enters the model.

```{r wage-sequential-anova}
anova(lm(wage ~ education + age, data = wage))
anova(lm(wage ~ age + education, data = wage))
```

Education and age share some information. The first term gets credit for the shared part before the second term is added.

### Auto MPG: repeat the complete comparison set

Now answer the same four questions using `a1`, `a2` and `a3`.

#### Does origin add information after weight?

```{r compare-auto-origin}
anova(a1, a2)
```

Two origin contrasts are added. The P-value is about 0.0024, so there is evidence that origin improves the model after weight is included.

#### Do the weight slopes differ by origin?

```{r compare-auto-slopes}
anova(a2, a3)
```

Two interaction coefficients are added. The P-value is about 0.044, giving modest evidence that the weight slopes differ among origins.

#### What does the full comparison test?

```{r compare-full-auto}
anova(a1, a3)
```

The Auto MPG comparison adds four parameters and favours separate lines over one common line, with P about 0.0011.

#### Does term order matter?

```{r auto-sequential-anova}
anova(lm(mpg ~ origin + weight, data = auto))
anova(lm(mpg ~ weight + origin, data = auto))
```

The direct comparison `anova(a1, a2)` writes the question into the models: `a1` already contains weight, and `a2` adds origin. The answer therefore does not depend on whichever term happened to appear first in a sequential table.

## When does another predictor earn its place?

A smaller RSS is guaranteed when another predictor is added, so RSS alone cannot decide whether a larger model is worth keeping.

Let's use adjusted R-squared, residual standard error and AIC alongside the partial F test. `broom::glance()` puts these model summaries into one tidy row.

### Wage example: build a compact model

Start from `w3` and add year.

```{r wage-add-year}
w4 <- lm(wage ~ age * education + year, data = wage)

bind_rows(
  w3 = broom::glance(w3),
  w4 = broom::glance(w4),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w3, w4)
```

Discuss: Do adjusted R-squared, residual standard error, AIC and the partial F test all favour adding year? How large is the improvement?

All four favour `w4`, although the changes in the fit measures are modest. Adjusted R-squared rises from about 0.260 to 0.263, residual standard error falls slightly, AIC falls by about 9.5, and the partial F test gives P about 0.0007.

Now add `jobclass`.

```{r wage-add-jobclass}
w5 <- lm(wage ~ age * education + year + jobclass, data = wage)

bind_rows(
  w4 = broom::glance(w4),
  w5 = broom::glance(w5),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w4, w5)
```

Discuss: Does job class still earn its place after year has been added?

The changes are small but point in the same direction. Adjusted R-squared rises slightly, residual standard error falls slightly, AIC falls by about 8.8, and the partial F test gives P about 0.001. Within this candidate set, keeping job class is reasonable.

So `w5` is the wage model to carry forward to diagnostics.

### Auto MPG: repeat the model-building sequence

Now repeat the whole process with Auto MPG.

Start from `a3`. Add `model_year` and call the model `a4`.

```{r auto-add-model-year}
a4 <- lm(mpg ~ weight * origin + model_year, data = auto)

bind_rows(
  a3 = broom::glance(a3),
  a4 = broom::glance(a4),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a3, a4)
```

Discuss: Does model year earn its place?

Yes. Adjusted R-squared rises from about 0.703 to 0.830, residual standard error falls from about 4.25 to 3.22 MPG, AIC falls by more than 200, and the partial F test gives P below 0.001.

Now add `horsepower` to `a4` and call the model `a5`.

```{r auto-add-horsepower}
a5 <- lm(mpg ~ weight * origin + model_year + horsepower, data = auto)

bind_rows(
  a4 = broom::glance(a4),
  a5 = broom::glance(a5),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a4, a5)
```

Discuss: Does horsepower earn its place after model year has been added?

No. Adding horsepower slightly reduces adjusted R-squared, barely changes residual standard error, raises AIC, and gives a partial F-test P-value of about 0.38.

So `a4` is the compact Auto MPG model to carry forward to diagnostics.

## What if prediction is the goal?

### 1. Separate model fit from prediction

Everything so far has been calculated on the same observations used to fit the models.

Discuss: What could be done to assess prediction for cars that were not used to fit the model? Why are adjusted R-squared, residual standard error and AIC not direct estimates of that prediction error?

Use validation data or cross-validation so predictions are assessed on observations that did not fit the model. The in-sample measures above are not direct estimates of prediction error for new cars.

## Does the working model fit the pattern?

### Wage example: diagnose and revise the model

Let's check `w5` before treating it as finished.

Run the standard diagnostic plots.

```{r wage-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w5)
par(mfrow = c(1, 1))
```

Now plot the residuals against age.

```{r wage-residuals-by-age}
wage |>
  mutate(residual = residuals(w5)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: What pattern should be visible if the straight-line age effect is adequate? What pattern is actually present?

The residuals should have no systematic relationship with age. Instead, there is a broad curve, so the straight-line age term is missing structure in the mean.

Quadratic regression was introduced earlier in the course, so let's add `age^2` while keeping the linear age term.

```{r wage-revise-mean}
w6 <- lm(
  wage ~ age * education + year + jobclass + I(age^2),
  data = wage
)

bind_rows(
  w5 = broom::glance(w5),
  w6 = broom::glance(w6),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w5, w6)
```

Repeat the standard diagnostic plots and the residual-versus-age plot for `w6`.

```{r wage-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w6)
par(mfrow = c(1, 1))

wage |>
  mutate(residual = residuals(w6)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: Has the broad age pattern been reduced? What concerns remain?

The broad age curve is reduced, although other departures remain. The plots do not establish independence.

### Auto MPG: repeat the diagnose-and-revise workflow

Now repeat the complete workflow for `a4`.

First run the standard diagnostic plots and plot the residuals against `weight`.

```{r auto-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(a4)
par(mfrow = c(1, 1))

auto |>
  mutate(residual = residuals(a4)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

Discuss: Is there a systematic pattern left in the mean?

There is clear curvature in the residuals against weight, so the mean structure still needs work.

Fit:

```text
a6: mpg ~ weight * origin + I(weight^2) + model_year
```

```{r auto-revise-mean}
a6 <- lm(
  mpg ~ weight * origin + I(weight^2) + model_year,
  data = auto
)

bind_rows(
  a4 = broom::glance(a4),
  a6 = broom::glance(a6),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a4, a6)

par(mfrow = c(2, 2))
plot(a6)
par(mfrow = c(1, 1))

auto |>
  mutate(residual = residuals(a6)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

Discuss: What improved? What still needs attention?

The broad U-shaped pattern is substantially reduced. Some tail and spread concerns remain. The standard diagnostic plots do not establish independence; that has to be considered from the way the observations were collected and whether groups of observations might share unmodelled structure.

## Pulling it together

A reasonable Auto MPG conclusion is:

Fuel economy decreases as weight increases, and allowing origin to change the fitted lines improves on one common line. There is modest evidence that the weight slopes differ among origins. Adding model year produces a large improvement, while horsepower adds little after weight, origin and model year are already included, so `a4` is a reasonable compact model before diagnostics. The residual plot for `a4` shows clear curvature, and adding `weight^2` improves the mean structure. The revised model still has tail and spread concerns, so the diagnostics do not establish that it is fully adequate. Finally, these are in-sample comparisons; validation or cross-validation is needed to assess prediction for new cars.

With a small candidate set, these decisions can still be made by hand. Once the predictor list gets longer, that becomes tedious very quickly. Lab 11 returns to Auto MPG and looks at automated search and shrinkage.
