---
title: "Lab 10 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

I will demonstrate each modelling step using wage data. When you reach a **Your turn** section, repeat that step using Auto MPG before moving on. The two datasets have different variables and different answers, but the modelling questions are the same.

The wage data contain 3,000 male workers, recorded from 2003 to 2009. Each row is one worker. Wage is recorded in thousands of US dollars. The Auto MPG data contain 392 cars from model years 1970 to 1982. Each row is one car. Both datasets are observational, so the fitted relationships describe associations, not causal effects.

| Wage variable | Meaning |
|:--|:--|
| `wage` | recorded wage, thousands of US dollars |
| `age` | worker's age, in years |
| `education` | education level, from less than high school to advanced degree |
| `year` | year the wage was recorded, 2003--2009 |
| `jobclass` | industrial or information work |

| Auto MPG variable | Meaning |
|:--|:--|
| `mpg` | fuel economy, miles per gallon |
| `weight` | vehicle weight, pounds |
| `origin` | USA, Europe or Japan |
| `model_year` | vehicle model year, 70--82 |
| `horsepower` | engine power, horsepower |

We will not use `car_name` as a predictor.

By the end of the lab, you should be able to read factor-by-covariate models as fitted lines, test nested models, compare a small set of candidate models, and use residual plots to identify a missing feature of the mean structure.

```{r import-data, message=FALSE}
wage <- readr::read_csv("../data/wage.csv") |>
  mutate(education = factor(education))

auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(wage)
glimpse(auto)
```

## What changes across groups?

### Instructor demonstration: education and age

Start with the wage data. Does the relationship between age and wage look similar at every education level?

```{r wage-raw-plot}
ggplot(wage, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Fit three models. The first has one line for everyone, the second allows education groups to have different intercepts but a common age slope, and the third allows both intercepts and slopes to differ.

```{r wage-fit-lines}
w1 <- lm(wage ~ age, data = wage)
w2 <- lm(wage ~ age + education, data = wage)
w3 <- lm(wage ~ age * education, data = wage)
```

```{.text}
education * age = education + age + education:age
```

The interaction lets the age slope vary with education. The next plot shows the fitted lines for all three models. Lines for each education group stop at that group's observed age range.

```{r wage-model-geometry}
age_values <- seq(min(wage$age), max(wage$age), length.out = 100)

education_grid <- wage |>
  group_by(education) |>
  summarise(
    age = list(seq(min(age), max(age), length.out = 100)),
    .groups = "drop"
  ) |>
  unnest(age)

common_lines <- tibble(age = age_values) |>
  mutate(
    fitted = predict(w1, newdata = tibble(age = age_values)),
    education = factor(NA_character_, levels = levels(wage$education)),
    model = "One common line"
  )

parallel_lines <- education_grid |>
  mutate(
    fitted = predict(w2, newdata = education_grid),
    model = "Parallel lines"
  )

separate_lines <- education_grid |>
  mutate(
    fitted = predict(w3, newdata = education_grid),
    model = "Separate lines"
  )

model_levels <- c("One common line", "Parallel lines", "Separate lines")
line_data <- bind_rows(common_lines, parallel_lines, separate_lines) |>
  mutate(model = factor(model, levels = model_levels))

point_data <- crossing(
  wage,
  model = factor(model_levels, levels = model_levels)
)

ggplot(point_data, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.18) +
  geom_line(
    data = filter(line_data, model != "One common line"),
    aes(y = fitted, group = interaction(education, model))
  ) +
  geom_line(
    data = filter(line_data, model == "One common line"),
    aes(x = age, y = fitted),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  facet_wrap(~ model) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Read the coefficient table for the separate-lines model.

```{r wage-read-interaction}
summary(w3)
```

The reference group is workers with less than a high-school education. Its age slope is the `age` coefficient. The education coefficients give intercept differences at age zero, and the interaction coefficients give slope differences from the reference group. Age zero is outside the data, so those intercept differences are not useful overall wage comparisons.

### Your turn: origin and vehicle weight

Now repeat the same steps for Auto MPG.

1. Plot `mpg` against `weight`, using `origin` for colour.
2. Fit `a1`, `a2` and `a3` with the formulas below.
3. Draw the one-line, parallel-lines and separate-lines fits. Use the wage geometry code as your template and keep each origin's fitted lines within its observed weight range.
4. Run `summary(a3)`. Identify the USA weight slope, the Europe and Japan intercept adjustments, and the slope adjustments. Explain why the intercept adjustments are not overall origin differences.

```{.text}
a1: mpg ~ weight
a2: mpg ~ weight + origin
a3: mpg ~ weight * origin
```

```{r auto-plot-and-models}
ggplot(auto, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  )

a1 <- lm(mpg ~ weight, data = auto)
a2 <- lm(mpg ~ weight + origin, data = auto)
a3 <- lm(mpg ~ weight * origin, data = auto)

weight_values <- seq(min(auto$weight), max(auto$weight), length.out = 100)

origin_grid <- auto |>
  group_by(origin) |>
  summarise(
    weight = list(seq(min(weight), max(weight), length.out = 100)),
    .groups = "drop"
  ) |>
  unnest(weight)

common_lines <- tibble(weight = weight_values) |>
  mutate(
    fitted = predict(a1, newdata = tibble(weight = weight_values)),
    origin = factor(NA_character_, levels = levels(auto$origin)),
    model = "One common line"
  )

parallel_lines <- origin_grid |>
  mutate(
    fitted = predict(a2, newdata = origin_grid),
    model = "Parallel lines"
  )

separate_lines <- origin_grid |>
  mutate(
    fitted = predict(a3, newdata = origin_grid),
    model = "Separate lines"
  )

model_levels <- c("One common line", "Parallel lines", "Separate lines")
line_data <- bind_rows(common_lines, parallel_lines, separate_lines) |>
  mutate(model = factor(model, levels = model_levels))

point_data <- crossing(
  auto,
  model = factor(model_levels, levels = model_levels)
)

ggplot(point_data, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.3) +
  geom_line(
    data = filter(line_data, model != "One common line"),
    aes(y = fitted, group = interaction(origin, model))
  ) +
  geom_line(
    data = filter(line_data, model == "One common line"),
    aes(x = weight, y = fitted),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  facet_wrap(~ model) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  )
```

```{r auto-read-interaction}
summary(a3)
```

USA is the reference origin. The `weight` coefficient is its slope. The `originEurope` and `originJapan` terms shift the intercepts, while `weight:originEurope` and `weight:originJapan` adjust the slopes. Because the interactions are in the model, the origin coefficients describe differences at zero pounds, not overall origin differences. Zero pounds is outside the data.

## Do the groups improve the model?

A partial F test compares a reduced model with a larger model that contains all the reduced model's terms. It tests the added terms together, not one coefficient at a time.

### Instructor demonstration: does education add information after age?

The question is whether education improves the model after age is already included. Compare `w1` with `w2`.

```{r compare-wage-education}
anova(w1, w2)
```

The test adds four education parameters. The P-value is below 0.001, so the parallel-lines model improves on one common line. This is evidence that at least one education group differs after age is accounted for.

### Your turn: does origin add information after weight?

Compare `a1` with `a2`. State how many parameters are added, report the P-value, and describe what changes in the fitted lines.

```{r compare-auto-origin}
anova(a1, a2)
```

The test adds two origin contrasts. The P-value is 0.0024, so there is evidence that the parallel-lines model improves on one common line. At least one origin differs after weight is accounted for.

### Instructor demonstration: do the age slopes differ?

Compare `w2` with `w3`. The question is whether the age slope varies across education groups.

```{r compare-wage-slopes}
anova(w2, w3)
```

The test adds four age-by-education interaction coefficients. The P-value is about 0.013, which is evidence that at least some age slopes differ by education.

### Your turn: do the weight slopes differ?

Compare `a2` with `a3`. Report the added terms and the P-value. What does the test say about the weight slopes across origins?

```{r compare-auto-slopes}
anova(a2, a3)
```

The added terms are the two origin-by-weight interactions. The P-value is about 0.044. There is modest evidence that the weight slopes differ among origins.

### Compare the simplest and fullest models

The earlier tests changed one part of the model at a time. Now compare `w1` directly with `w3`, then make the same comparison for Auto MPG. State the number of added parameters and interpret the joint test.

```{r compare-full-lines}
anova(w1, w3)
anova(a1, a3)
```

The wage comparison adds eight parameters and strongly favours the separate-lines model (P < 0.001). The Auto MPG comparison adds four parameters and also favours separate lines over one common line (F = 4.68, P = 0.0011). Neither joint test says that every individual coefficient is useful.

### Why does sequential ANOVA depend on order?

A sequential ANOVA table gives each term credit for the variation it explains when it enters the model. When predictors overlap, changing the order can change the sums of squares.

```{r wage-sequential-anova}
anova(lm(wage ~ education + age, data = wage))
anova(lm(wage ~ age + education, data = wage))
```

When education enters first, its sum of squares is about 1,226,364; when age enters first, age gets about 199,870 and education gets about 1,154,224. The full model and its residual sum of squares are unchanged. The terms share information, and the first term gets credit for the part they share.

Repeat this check for `mpg ~ origin + weight` and `mpg ~ weight + origin`. Why is comparing `a1` with `a2` a clearer answer to “does origin add information after weight?”

```{r auto-sequential-anova}
anova(lm(mpg ~ origin + weight, data = auto))
anova(lm(mpg ~ weight + origin, data = auto))
```

The same issue appears for the cars. If origin enters first, it gets a sum of squares of about 7,904; if weight enters first, origin gets about 224. The sequential table answers a different question depending on the order. `anova(a1, a2)` directly compares models that differ only by adding origin after weight.

## When does another predictor earn its place?

Adjusted R-squared, residual standard error and AIC compare fit while accounting for model complexity in different ways. Larger adjusted R-squared and smaller residual standard error or AIC are preferred when the models use the same response and observations. A partial F test answers a specific nested-model question.

### Instructor demonstration: add year

Add the year the wage was recorded to `w3`, and call the result `w4`.

```{r wage-add-year}
w4 <- lm(wage ~ age * education + year, data = wage)

tibble(
  model = c("w3", "w4"),
  adjusted_R2 = c(summary(w3)$adj.r.squared, summary(w4)$adj.r.squared),
  residual_SE = c(sigma(w3), sigma(w4)),
  AIC = c(AIC(w3), AIC(w4))
)

anova(w3, w4)
```

Year improves all three measures, although the changes are modest: adjusted R-squared rises from 0.260 to 0.263, residual standard error falls from 35.89 to 35.83, and AIC falls by about 9.5. The partial F test gives F = 11.50 and P = 0.0007. There is evidence that recorded wage differs by year after accounting for age and education.

### Your turn: add model year

Fit `a4` by adding `model_year` to `a3`. Recreate the same comparison table for `a3` and `a4`, then run `anova(a3, a4)`. State which measures improve and what the partial F test says.

```{r auto-add-model-year}
a4 <- lm(mpg ~ weight * origin + model_year, data = auto)

tibble(
  model = c("a3", "a4"),
  adjusted_R2 = c(summary(a3)$adj.r.squared, summary(a4)$adj.r.squared),
  residual_SE = c(sigma(a3), sigma(a4)),
  AIC = c(AIC(a3), AIC(a4))
)

anova(a3, a4)
```

All three measures improve substantially: adjusted R-squared rises from 0.703 to 0.830, residual standard error falls from 4.25 to 3.22 MPG, and AIC falls by about 217.5. The partial F test gives F = 289.0 and P < 0.001. Model year adds substantial information after weight, origin and their interaction are included.

### Instructor demonstration: add job class

Add `jobclass` to `w4` and call the result `w5`.

```{r wage-add-jobclass}
w5 <- lm(wage ~ age * education + year + jobclass, data = wage)

tibble(
  model = c("w4", "w5"),
  adjusted_R2 = c(summary(w4)$adj.r.squared, summary(w5)$adj.r.squared),
  residual_SE = c(sigma(w4), sigma(w5)),
  AIC = c(AIC(w4), AIC(w5))
)

anova(w4, w5)
```

The changes are small but all three fit measures favour `w5`: adjusted R-squared rises from 0.263 to 0.265, residual standard error falls from 35.83 to 35.77, and AIC falls by about 8.8. The partial F test gives F = 10.80 and P = 0.0010. Within this small candidate set, keeping job class is reasonable for a descriptive model.

### Your turn: add horsepower

Fit `a5` by adding `horsepower` to `a4`. Recreate the same comparison table and run `anova(a4, a5)`. Do the four measures tell a consistent story?

```{r auto-add-horsepower}
a5 <- lm(
  mpg ~ weight * origin + model_year + horsepower,
  data = auto
)

tibble(
  model = c("a4", "a5"),
  adjusted_R2 = c(summary(a4)$adj.r.squared, summary(a5)$adj.r.squared),
  residual_SE = c(sigma(a4), sigma(a5)),
  AIC = c(AIC(a4), AIC(a5))
)

anova(a4, a5)
```

They tell a consistent story against keeping horsepower in this candidate model. Adjusted R-squared changes from 0.830 to 0.830 (a slight decrease), residual standard error changes from 3.219 to 3.220 MPG, and AIC rises from 2037.9 to 2039.1. The partial F test gives F = 0.79 and P = 0.376. The fit does not improve enough to justify the extra term here.

Choose between `a4` and `a5` as a compact descriptive model. Support your choice with at least three pieces of evidence. Do not conclude that an excluded predictor has no association in every possible model.

`a4` is a reasonable choice. Adding model year produced a large improvement, but adding horsepower after that slightly lowers adjusted R-squared, barely changes residual standard error, raises AIC and gives a partial F-test P-value of 0.376. This conclusion applies to these candidate models and these data.

## What if prediction is the goal?

The measures above are calculated on the observations used to fit each model. They are not direct estimates of prediction error for new observations.

For the Auto MPG analysis, what could you do to assess how well a model predicts cars that were not used to fit it? Why would the answer to that question matter if prediction, rather than describing these data, were the goal?

Use a validation set or cross-validation so that predictions are assessed on observations that did not fit the model. The in-sample comparisons above do not estimate that prediction error.

## Does the working model fit the pattern?

Use the standard diagnostic plots for `w5`. For each plot, describe what you see and explain why it matters. Also plot residuals against age: that makes it easier to see whether a straight-line age effect leaves a pattern in the mean.

### Instructor demonstration: diagnose and revise the wage model

```{r wage-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w5)
par(mfrow = c(1, 1))
```

```{r wage-residuals-by-age}
wage |>
  mutate(residual = residuals(w5)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

The residuals against age show a broad curve, so the straight-line age term misses structure in the mean. Add a quadratic age term to `w5`, giving `w6`. In an R formula, write the squared term as `I(age^2)`.

```{r wage-revise-mean}
w6 <- lm(
  wage ~ age * education + year + jobclass + I(age^2),
  data = wage
)

anova(w5, w6)

tibble(
  model = c("w5", "w6"),
  adjusted_R2 = c(summary(w5)$adj.r.squared, summary(w6)$adj.r.squared),
  residual_SE = c(sigma(w5), sigma(w6)),
  AIC = c(AIC(w5), AIC(w6))
)
```

The quadratic term improves the model: adjusted R-squared rises from 0.265 to 0.292, residual standard error falls from 35.77 to 35.11, and AIC falls from 29,990.3 to 29,879.7. The partial F test gives F = 114.2 and P < 0.001.

Repeat both diagnostic checks for `w6`. Has the broad age pattern in the residuals changed? What concerns remain?

```{r wage-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w6)
par(mfrow = c(1, 1))
```

```{r wage-revised-residuals-by-age}
wage |>
  mutate(residual = residuals(w6)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

The broad age curve is reduced, though the diagnostic plots still show departures that deserve attention. The plots do not establish independence. Consider how the workers entered the data and whether observations may share unmodelled structure.

### Your turn: diagnose and revise the Auto MPG model

Use the same checks on your chosen model, `a4`. First inspect the standard diagnostic panels.

```{r auto-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(a4)
par(mfrow = c(1, 1))
```

Now plot residuals against `weight`.

```{r auto-residuals-by-weight}
auto |>
  mutate(residual = residuals(a4)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

The residuals against weight show strong curvature. Now fit `a6`:

```{.text}
mpg ~ weight * origin + I(weight^2) + model_year
```

Compare `a4` and `a6` with a partial F test and the same three fit measures. Repeat both diagnostic checks. State what improved and what still needs attention.

```{r auto-revise-mean}
a6 <- lm(
  mpg ~ weight * origin + I(weight^2) + model_year,
  data = auto
)

anova(a4, a6)

tibble(
  model = c("a4", "a6"),
  adjusted_R2 = c(summary(a4)$adj.r.squared, summary(a6)$adj.r.squared),
  residual_SE = c(sigma(a4), sigma(a6)),
  AIC = c(AIC(a4), AIC(a6))
)
```

The partial F test gives F = 57.68 and P < 0.001. Adjusted R-squared rises from 0.830 to 0.852, residual standard error falls from 3.219 to 3.005 MPG, and AIC falls from 2037.9 to 1985.0.

```{r auto-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(a6)
par(mfrow = c(1, 1))

auto |>
  mutate(residual = residuals(a6)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

The broad U-shaped residual pattern is substantially reduced. Some tail and spread concerns remain. We cannot diagnose independence from these plots alone. Consider how the cars entered the data and whether groups of observations might share unmodelled structure. Do not treat an influential-point label as an automatic reason to delete a car.

## Independent synthesis

A reasonable Auto MPG conclusion is:

Fuel economy decreases as weight increases, and allowing origin to change the fitted lines improves on one common line. There is modest evidence that the weight slopes differ among origins. Adding model year produces a large improvement, while horsepower adds little after weight, origin and model year are already included, so `a4` is a reasonable compact model before diagnostics. The residual plot for `a4` shows clear curvature, so adding `weight^2` improves the mean structure. The revised model still has tail and spread concerns, so the diagnostics do not establish that it is fully adequate. Finally, these are in-sample comparisons; validation or cross-validation is needed to assess prediction for new cars.

Lab 11 returns to Auto MPG with a larger candidate set.
