---
title: "Lab 10 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

This week we bring together Lectures 25--27 using a different dataset. The aim is not to repeat the Samara or Climate examples. I want you to use the same modelling logic on a new problem.

> How does vehicle weight relate to fuel economy, does that relationship differ by origin, and when is a more complicated model worth keeping?

The Auto MPG data contain 392 cars from model years 1970--1982. The course copy was prepared from the UCI Auto MPG data, with rows missing horsepower removed. Each row is one car. These are observational data, so the fitted relationships are associations rather than causal effects.

| Variable | Meaning | Units / type |
|:--|:--|:--|
| `mpg` | fuel economy | miles per gallon |
| `weight` | vehicle weight | pounds |
| `horsepower` | engine power | horsepower |
| `model_year` | model year | 70--82 |
| `origin` | region of origin | USA, Europe, Japan |

We will not use `car_name` as a predictor.

```{r import, message=FALSE}
auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(auto)
```

By the end of the lab you should be able to translate factor-by-covariate formulas into fitted lines, compare nested models with a partial F test, compare a small set of candidate models by hand, and use diagnostics to decide whether the mean structure needs more work.

## What does origin change?

### Demonstration: one line, parallel lines, separate lines

```{r raw-plot}
ggplot(auto, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.7) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  ) +
  theme_minimal()
```

Heavier cars tend to have lower fuel economy. The origins also occupy rather different parts of the weight range, so any apparent origin difference needs to be interpreted after accounting for weight. The relationship is not perfectly straight across the full range.

```{r define-models}
m1 <- lm(mpg ~ weight, data = auto)
m2 <- lm(mpg ~ weight + origin, data = auto)
m3 <- lm(mpg ~ weight * origin, data = auto)
```

The three formulas correspond to one common line, parallel lines, and separate lines. The interaction expands as

```text
origin * weight = origin + weight + origin:weight
```

The code below makes the geometry explicit.

```{r fitted-lines}
weight_values <- seq(min(auto$weight), max(auto$weight), length.out = 200)
model_levels <- c("One common line", "Parallel lines", "Separate lines")

common_grid <- tibble(weight = weight_values)
common_fit <- common_grid |>
  mutate(
    fitted = predict(m1, newdata = common_grid),
    model = factor("One common line", levels = model_levels)
  )

origin_grid <- crossing(
  weight = weight_values,
  origin = levels(auto$origin)
)

origin_fits <- bind_rows(
  origin_grid |>
    mutate(
      fitted = predict(m2, newdata = origin_grid),
      model = "Parallel lines"
    ),
  origin_grid |>
    mutate(
      fitted = predict(m3, newdata = origin_grid),
      model = "Separate lines"
    )
) |>
  mutate(model = factor(model, levels = model_levels))

plot_data <- crossing(
  auto,
  model = factor(model_levels, levels = model_levels)
)

ggplot(plot_data, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.35) +
  geom_line(
    data = origin_fits,
    aes(y = fitted),
    linewidth = 1
  ) +
  geom_line(
    data = common_fit,
    aes(x = weight, y = fitted),
    inherit.aes = FALSE,
    colour = "black",
    linewidth = 1
  ) +
  facet_wrap(~ model) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  ) +
  theme_minimal()
```

Adding `origin` moves the fitted lines vertically while keeping one common slope. Adding `origin:weight` then allows the lines to rotate away from that common slope.

### Your turn: read the separate-lines model

```{r separate-lines-summary}
summary(m3)
```

USA is the reference origin. The `weight` coefficient is therefore the USA weight slope. `originEurope` and `originJapan` shift the intercepts relative to USA, while `weight:originEurope` and `weight:originJapan` adjust the slopes.

The main-effect origin coefficients are not overall Europe-versus-USA or Japan-versus-USA differences. In an interaction model they describe intercept differences at `weight = 0`, while the interaction terms allow those origin differences to change with weight. Zero pounds is also far outside the data, so those intercept contrasts have little direct practical meaning.

## Which extra structure is supported?

A partial F test compares a reduced model with a larger model containing all of the reduced model's terms.

### Demonstration: does origin add information after weight?

```{r compare-common-parallel}
anova(m1, m2)
```

The P-value is about 0.0024. There is evidence that the parallel-lines model improves on one common line. At least one of the two origin contrasts contributes after weight is already in the model. This is a joint test of the `origin` term.

### Your turn: do the weight slopes differ?

```{r compare-parallel-separate}
anova(m2, m3)
```

The added terms are the two origin-by-weight interaction coefficients. The P-value is about 0.044. There is modest evidence that allowing different weight slopes improves on the parallel-lines model.

### Independent comparison: does origin affect either intercepts or slopes?

```{r compare-common-separate}
anova(m1, m3)
```

This comparison adds four parameters: two origin contrasts and two origin-by-weight interactions. The P-value is about 0.0011. Taken together, allowing origin to change intercepts or slopes improves on one common regression line. The test does not imply that every added coefficient is individually important.

### One wrinkle: sequential ANOVA depends on order

```{r sequential-anova}
anova(lm(mpg ~ origin + weight, data = auto))
anova(lm(mpg ~ weight + origin, data = auto))
```

These are sequential sums of squares. Origin and weight overlap in the information they carry, so the term entered first receives credit for the shared variation it can explain before the second term is added.

That is why `anova(m1, m2)` is the cleaner answer when our question is specifically whether origin adds information after weight. We write down the reduced and full models that match the question directly.

## When does another predictor earn its place?

Adjusted R-squared, residual standard error and AIC give us different views of the fit-complexity trade-off. Higher adjusted R-squared is better, lower residual standard error means less remaining variation in MPG, and lower AIC is preferred among models fitted to the same response and observations.

### Demonstration: add model year

```{r add-model-year}
m4 <- lm(mpg ~ weight * origin + model_year, data = auto)

tibble(
  model = c("m3", "m4"),
  adjusted_r_squared = c(
    summary(m3)$adj.r.squared,
    summary(m4)$adj.r.squared
  ),
  residual_standard_error = c(sigma(m3), sigma(m4)),
  AIC = c(AIC(m3), AIC(m4))
)

anova(m3, m4)
```

Model year produces a substantial improvement. Adjusted R-squared increases from about 0.703 to 0.830, residual standard error drops from about 4.25 to 3.22 MPG, and AIC drops by more than 200. The partial F test also gives very strong evidence that model year adds information.

### Your turn: does horsepower add anything further?

```{r add-horsepower}
m5 <- lm(
  mpg ~ weight * origin + model_year + horsepower,
  data = auto
)

tibble(
  model = c("m4", "m5"),
  adjusted_r_squared = c(
    summary(m4)$adj.r.squared,
    summary(m5)$adj.r.squared
  ),
  residual_standard_error = c(sigma(m4), sigma(m5)),
  AIC = c(AIC(m4), AIC(m5))
)

anova(m4, m5)
```

Horsepower makes only a small reduction in RSS. Adjusted R-squared decreases slightly, residual standard error barely changes, AIC increases from about 2037.9 to 2039.1, and the partial F test gives a P-value of about 0.38.

For this deliberately small candidate set, `m4` is the more economical working model. That does not mean horsepower has no association with MPG in every possible model. It means that after the terms already in `m4` are present, horsepower is not buying us enough here to justify keeping it.

### Independent decision: choose a working model

A reasonable answer is `m4`. The evidence is consistent across the measures we used: adding model year produces a large improvement, while adding horsepower after that produces little change and slightly worsens the complexity-adjusted measures.

## What if prediction is the goal?

Adjusted R-squared, residual standard error and AIC are all calculated from the same observations used to fit the model. They are not observed prediction errors for new cars.

For prediction, we would use validation data or cross-validation so that each prediction is judged on an observation that did not fit that version of the model. We identify that distinction here and implement cross-validation next week.

## Does the working model actually fit the pattern?

For each diagnostic we use the same routine: what pattern would be reassuring, what do we actually see, and why does it matter?

### Demonstration: diagnose the working model

```{r diagnostics-initial, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(m4)
par(mfrow = c(1, 1))
```

The residuals-versus-fitted plot has pronounced curvature. The current linear mean structure has missed systematic structure. The Q-Q plot is close to its reference line through much of the distribution but departs in the tails. The scale-location plot also suggests some change in residual spread. A few observations have relatively large residuals or leverage, but a labelled point is not automatically an error and should not be deleted mechanically.

### Your turn: revise and recheck

Polynomial regression was introduced earlier in the course. We keep the linear weight term and add the quadratic term with `I(weight^2)`.

```{r revise-mean-structure}
m6 <- lm(
  mpg ~ weight * origin + I(weight^2) + model_year,
  data = auto
)

anova(m4, m6)

tibble(
  model = c("m4", "m6"),
  adjusted_r_squared = c(
    summary(m4)$adj.r.squared,
    summary(m6)$adj.r.squared
  ),
  residual_standard_error = c(sigma(m4), sigma(m6)),
  AIC = c(AIC(m4), AIC(m6))
)
```

Adding the quadratic term makes a large reduction in RSS. Adjusted R-squared increases from about 0.830 to 0.852, residual standard error falls from about 3.22 to 3.01 MPG, and AIC falls from about 2037.9 to 1985.0.

```{r diagnostics-revised, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(m6)
par(mfrow = c(1, 1))
```

The strong U-shaped residual pattern is substantially reduced, although some mild structure remains. The tail and changing-spread concerns also remain. The standard diagnostic plots do not directly establish independence here. Row order has no scientific meaning, and independence has to be considered from the way the cars entered the dataset and whether groups of cars may share unmodelled structure.

## Independent synthesis

One reasonable conclusion is:

Fuel economy decreases as vehicle weight increases, and allowing origin to change the fitted lines improves substantially on one common line. There is modest evidence that the weight slope itself differs among origins. Adding model year produces a large improvement, while horsepower adds little once weight, origin and model year are already present, so `m4` is a reasonable compact model before diagnostics. The residual plot for `m4` shows clear curvature, so retaining the earlier polynomial-regression tool and adding `weight^2` gives a better mean structure. The revised model still has some tail and variance concerns, so the diagnostics do not certify the model as correct. Finally, all of these comparisons are in-sample; prediction on new cars would need validation or cross-validation.

With a small candidate set we can still work through these choices by hand. Once the predictor list becomes longer, that gets tedious very quickly. Next week we look at two different responses: automate the search through candidate models, or keep a larger model and shrink its coefficients.
