---
title: "Computer Laboratory 12: Curves, trend and seasonality : solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated 08 October 2026"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
library(broom)
```

# Computer Laboratory 12: Curves, trend and seasonality

This week we relax the shape of the mean model. We will use daily Seoul bike demand to compare LOESS, cubic splines and natural cubic splines, then let time itself enter the model through a trend and smooth seasonal terms.

The lab uses [seoul_bike_daily.csv](https://knowlton.co.nz/161251/data/seoul_bike_daily.csv). The downloadable R Markdown file reads the dataset directly from the course site.

## How much curve do we need for temperature?

Each row in `seoul_bike_daily.csv` is one calendar day. `rented_bike_count` is the number of rentals and `mean_temperature` is in °C. The 13 days when the system was not fully functioning should not represent ordinary demand. Keep the original `date` and `day_index` after filtering so that a gap in service remains a gap in time.

```{r seoul-data, message=FALSE}
seoul_all <- read_csv("https://knowlton.co.nz/161251/data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date)
seoul <- seoul_all |>
  filter(functioning_day_all == "yes")
c(all_days = nrow(seoul_all), outage_days = nrow(seoul_all) - nrow(seoul),
  analysed_days = nrow(seoul))
```

### Seoul: compare the smooths and spline models

Start with the points and two LOESS curves. `span` is the fraction of nearby observations used in each local fit. A small span can follow local changes but can also chase noise; a large span smooths more strongly.

```{r seoul-loess}
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.3,
              se = FALSE, aes(colour = "span 0.3")) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.9,
              se = FALSE, aes(colour = "span 0.9")) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "LOESS")
```

Now fit a straight line, a cubic spline and a natural cubic spline, using `df = 5` for both splines. The cubic spline joins cubic pieces smoothly. With the default settings in `bs()`, `df = 5` gives two internal knots, not a fifth-degree polynomial.

The natural cubic spline adds the restriction that its tails are linear beyond the boundary knots. The same `df` does not imply the same internal knots in the two fits. Both models are still linear in their fitted coefficients, even though their curves bend with temperature. Predict on the same temperature grid to compare the curves directly.

```{r seoul-splines}
seoul_line <- lm(rented_bike_count ~ mean_temperature, data = seoul)
seoul_bs <- lm(rented_bike_count ~ splines::bs(mean_temperature, df = 5), data = seoul)
seoul_ns <- lm(rented_bike_count ~ splines::ns(mean_temperature, df = 5), data = seoul)
temperature_grid <- tibble(mean_temperature = seq(
  min(seoul$mean_temperature), max(seoul$mean_temperature), length.out = 250
))
curve_fits <- temperature_grid |>
  mutate(
    Line = predict(seoul_line, newdata = temperature_grid),
    `Cubic spline` = predict(seoul_bs, newdata = temperature_grid),
    `Natural cubic spline` = predict(seoul_ns, newdata = temperature_grid)
  ) |>
  pivot_longer(-mean_temperature, names_to = "model", values_to = "fitted")
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.2) +
  geom_line(data = curve_fits,
            aes(y = fitted, colour = model), linewidth = 1) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "Model")
```

Plot the natural cubic spline residuals against temperature to check for any remaining shape.

```{r seoul-spline-residuals}
seoul |>
  mutate(residual = residuals(seoul_ns)) |>
  ggplot(aes(mean_temperature, residual)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Mean temperature (°C)", y = "Natural cubic spline residual")
```

Discuss: Where does a line miss the shape in the points? How does the LOESS span change the curve? Compare the cubic spline and natural cubic spline near the observed temperature limits. A natural cubic spline becomes linear beyond its boundary knots, but does that make extrapolation safe? Would a more flexible fit to these observations by itself justify choosing that model?

The straight line misses the broad curve in the temperature relationship. With span 0.3, LOESS follows more of the local ups and downs than it does with span 0.9. Notice where the two splines differ near the temperature limits: we have fewer observations there to guide the fit. The natural cubic spline continues as a straight line beyond its boundary knots, but we still have no data out there to check it. We use the residual plot to look for shape that remains; a closer fit to these observations alone does not tell us which curve will predict new days better.

## Can we separate trend from seasonality?

Lectures 32/33 used sine and cosine together to describe an annual wave. Let's first leave cosine out and see why the pair matters. We choose a period of 365 days and keep the linear trend in every model.

### Seoul: start with one sine term

The coefficient of sine changes its height, and a negative coefficient flips it. On its own, sine restricts where the seasonal peak can fall. Fit the unshifted sine term alongside the trend and inspect the fitted mean over the actual dates. We use `tidy()` for the coefficient estimates and their uncertainty. `glance()` gives one row for the whole model, including R².

```{r seoul-sine-model}
seoul_time <- seoul |>
  mutate(S1 = sin(2 * pi * day_index / 365))

seoul_trend <- lm(rented_bike_count ~ day_index, data = seoul_time)
seoul_sine <- lm(rented_bike_count ~ day_index + S1, data = seoul_time)

tidy(seoul_sine)

bind_rows(
  Trend = glance(seoul_trend),
  Sine = glance(seoul_sine),
  .id = "Model"
) |>
  select(Model, r.squared) |>
  knitr::kable(col.names = c("Model", "$R^2$"), digits = 3)
```

```{r seoul-sine-plot}
seoul_time |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals"), alpha = 0.6) +
  geom_line(aes(y = fitted(seoul_trend), colour = "Trend only"), linewidth = 0.9) +
  geom_line(aes(y = fitted(seoul_sine), colour = "Trend + sine"), linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Observed rentals" = "grey60", "Trend only" = "forestgreen", "Trend + sine" = "blue"
  )) +
  labs(
    x = "Date", y = "Daily rentals", colour = "Series",
    title = "A trend with one unshifted annual sine term"
  ) +
  theme(legend.position = "bottom")
```

Discuss: What does the trend alone miss? How much of the observed rise and fall does this sine term capture?

The trend can only give us a straight change across the year. Adding sine gives us a curve, but it still misses much of the rise and fall. Here R² increases from 0.377 to 0.416. We can change the height of that wave or flip it, but we cannot freely move its peak while the sine origin is fixed.

### Seoul: move the origin by 90 days

Now create `S1_shift = sin(2*pi*(day_index-90)/365)` and fit `seoul_sine_shift` with `day_index + S1_shift`. Keep the same observations, the original dates and the unshifted `day_index` trend. Move only the origin inside sine; the period is still 365 days.

```{r seoul-sine-shift-model}
seoul_time <- seoul_time |>
  mutate(S1_shift = sin(2 * pi * (day_index - 90) / 365))

seoul_sine_shift <- lm(
  rented_bike_count ~ day_index + S1_shift,
  data = seoul_time
)
```

Using `seoul-sine-plot` as a template, overlay the unshifted and shifted sine fits on the observed rentals. Show the fits by colour and give the graph a title and a titled legend. Use `tidy()` to compare the coefficients and `glance()` to build an R² table for both models, and calculate the largest absolute difference between their fitted values.

```{r seoul-sine-shift-comparison}
bind_rows(
  Unshifted = tidy(seoul_sine),
  Shifted = tidy(seoul_sine_shift),
  .id = "Model"
) |>
  select(Model, term, estimate, std.error, p.value)

bind_rows(
  Unshifted = glance(seoul_sine),
  Shifted = glance(seoul_sine_shift),
  .id = "Model"
) |>
  select(Model, r.squared) |>
  knitr::kable(col.names = c("Model", "$R^2$"), digits = 3)
max(abs(fitted(seoul_sine) - fitted(seoul_sine_shift)))

seoul_time |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals"), alpha = 0.6) +
  geom_line(aes(y = fitted(seoul_sine), colour = "Unshifted sine"), linewidth = 0.9) +
  geom_line(aes(y = fitted(seoul_sine_shift), colour = "Shifted sine"), linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Observed rentals" = "grey60", "Unshifted sine" = "blue", "Shifted sine" = "red"
  )) +
  labs(
    x = "Date", y = "Daily rentals", colour = "Series",
    title = "Moving the origin changes a sine-only fit"
  ) +
  theme(legend.position = "bottom")
```

Discuss: What changed in the coefficients, R² and predicted rentals? Did moving the origin change the annual period, or where this single wave is allowed to peak?

R² changes from 0.416 to 0.601, and the largest difference between the two fitted counts is about 10,733 rentals. We have kept the same days and the same 365-day period. What changed is where sine is allowed to peak, so the model now fits a different curve. That curve follows this year more closely, but we have not checked whether it predicts next year better. These two sine-only models are not nested, so we cannot use a partial F test to compare them.

### Seoul: keep sine and cosine together

Let's add cosine to the unshifted model. The two coefficients now let the annual wave have its peak anywhere in the year.

```{r seoul-first-pair-model}
seoul_time <- seoul_time |>
  mutate(C1 = cos(2 * pi * day_index / 365))

seoul_first <- lm(
  rented_bike_count ~ day_index + S1 + C1,
  data = seoul_time
)
```

Now repeat the 90-day shift with both terms. Add `C1_shift = cos(2*pi*(day_index-90)/365)`, reuse `S1_shift`, and fit `seoul_first_shift` with `day_index + S1_shift + C1_shift`. Both trigonometric predictors must use the same shift; leave the trend unchanged.

Use `tidy()` to compare coefficients and `glance()` to compare R², overlay the two paired fits against the actual dates, and calculate `max(abs(fitted(seoul_first) - fitted(seoul_first_shift)))`.

```{r seoul-pair-shift-comparison}
seoul_time <- seoul_time |>
  mutate(C1_shift = cos(2 * pi * (day_index - 90) / 365))

seoul_first_shift <- lm(
  rented_bike_count ~ day_index + S1_shift + C1_shift,
  data = seoul_time
)

bind_rows(
  Unshifted = tidy(seoul_first),
  Shifted = tidy(seoul_first_shift),
  .id = "Model"
) |>
  select(Model, term, estimate, std.error, p.value)

bind_rows(
  Unshifted = glance(seoul_first),
  Shifted = glance(seoul_first_shift),
  .id = "Model"
) |>
  select(Model, r.squared) |>
  knitr::kable(col.names = c("Model", "$R^2$"), digits = 3)
max(abs(fitted(seoul_first) - fitted(seoul_first_shift)))

seoul_time |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals"), alpha = 0.6) +
  geom_line(aes(y = fitted(seoul_first), colour = "Unshifted pair"), linewidth = 0.9) +
  geom_line(aes(y = fitted(seoul_first_shift), colour = "Shifted pair"), linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Observed rentals" = "grey60", "Unshifted pair" = "blue", "Shifted pair" = "red"
  )) +
  labs(
    x = "Date", y = "Daily rentals", colour = "Series",
    title = "Both origins give the same fit when we retain the pair"
  ) +
  theme(legend.position = "bottom")
```

Discuss: Do the coefficients still change? Do the predicted rentals change? Why do we get a different result from shifting sine alone?

Now both fits have $R^2 = 0.641$. The largest fitted-value difference is only about $6.55 \times 10^{-11}$ rentals, so the lines sit on top of each other. The coefficients change because sine and cosine share the wave differently after we move the origin. Together they can put the peak wherever the data suggest, so we do not need to search for a better origin. For the same reason, we keep the pair when the peak position is unknown and test it jointly; one large individual P-value does not justify dropping that component.

### Seoul: add a second seasonal pair

Keep the unshifted first pair and add `S2 = sin(4*pi*day_index/365)` and `C2 = cos(4*pi*day_index/365)`. This second pair allows a sharper within-year shape. Fit `seoul_second` using `day_index + S1 + C1 + S2 + C2`, then plot the trend-only, first-pair and second-pair fits over the observed rentals.

Compare `seoul_first` and `seoul_second` with `anova()`. These models are nested: the partial F test jointly assesses the two additional coefficients under the usual independent-error assumptions.

```{r seoul-fourier}
seoul_time <- seoul_time |>
  mutate(
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )
seoul_second <- lm(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2,
  data = seoul_time
)

seoul_time |>
  ggplot(aes(x = date, y = rented_bike_count)) +
  geom_line(aes(colour = "Observed rentals"), alpha = 0.6) +
  geom_line(aes(y = fitted(seoul_trend), colour = "Trend only"), linewidth = 0.9) +
  geom_line(aes(y = fitted(seoul_first), colour = "First pair"), linewidth = 0.9) +
  geom_line(aes(y = fitted(seoul_second), colour = "Second pair"), linewidth = 0.9) +
  scale_colour_manual(values = c(
    "Observed rentals" = "grey60", "Trend only" = "forestgreen",
    "First pair" = "blue", "Second pair" = "red"
  )) +
  labs(
    x = "Date", y = "Daily rentals", colour = "Series",
    title = "A trend with one and two seasonal pairs"
  ) +
  theme(legend.position = "bottom")

anova(seoul_first, seoul_second)
```

Discuss: What does the trend alone miss? Does the second pair change the fitted seasonal shape, and what does the partial F test say about its extra terms? We have only one year of Seoul data: what can we say about a pattern recurring next year?

The first pair lets us move the seasonal peak; the second pair lets the shape bend further by adding a wave with twice the frequency. For these 352 days, the partial F test gives $F(2,346) = 21.79$ and $p < 0.001$ under the usual independent-error assumptions. The two extra terms improve the fit here. We still need to check the residual assumptions, and one year cannot tell us whether this pattern repeats next year.

A factor for calendar season or month gives each category its own offset. Sine and cosine give a smooth within-year pattern. The terms answer different questions; neither makes a one-year record into evidence about future years.

## Wrapping up

We have changed the mean in two different ways. Splines let the relationship with a numerical predictor bend, while trend and sine/cosine terms let the expected response change systematically through time.

In one short paragraph, explain when you would use LOESS, a regression spline, and a trend plus Fourier terms. Then state what these mean-model choices still do **not** tell us about observations collected close together in time.

We can use LOESS to explore the shape before choosing a regression model. A spline then lets us put a curved relationship into that model and control how much it bends through its basis and degrees of freedom. For time, a trend describes an overall change, while Fourier terms add a repeating wave. All of these choices change the fitted mean. The residuals from nearby days may still be related, even after we have fitted that mean.

That remaining question is the starting point for Lab 13: after fitting a reasonable mean, are the residuals still dependent?
