---
title: "Computer Laboratory 12: Curves, seasons and dependent errors : solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
library(nlme)
```

# Computer Laboratory 12: Curves, seasons and dependent errors

We have been changing the *mean* in a regression model: a curve for temperature, a trend over time and a seasonal pattern. Now we need to look at what remains. We will work through daily Seoul bike demand, repeat the time-series analysis on monthly tourism and finish with quarterly Perth house prices.

## How much curve do we need for temperature?

Each row in `seoul_bike_daily.csv` is one calendar day. `rented_bike_count` is the number of rentals and `mean_temperature` is in °C. The 13 days when the system was not fully functioning should not represent ordinary demand. Keep the original `date` and `day_index` after filtering so that a gap in service remains a gap in time.

```{r seoul-data, message=FALSE}
seoul_all <- read_csv("../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date)
seoul <- seoul_all |>
  filter(functioning_day_all == "yes")
c(all_days = nrow(seoul_all), outage_days = nrow(seoul_all) - nrow(seoul),
  analysed_days = nrow(seoul))
```

### Seoul: compare the smooths and spline models

Start with the points and two LOESS curves. `span` is the fraction of nearby observations used in each local fit. A small span can follow local changes but can also chase noise; a large span smooths more strongly.

```{r seoul-loess}
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.3,
              se = FALSE, aes(colour = "span 0.3")) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.9,
              se = FALSE, aes(colour = "span 0.9")) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "LOESS")
```

Now fit a line, a cubic B-spline and a natural cubic spline, using `df = 5` for both splines. Here `df` controls the size of the spline basis, not the degree of a single polynomial. The spline model is linear in its fitted coefficients, even though its curve bends with temperature. Predict on the same temperature grid to compare the curves directly.

```{r seoul-splines}
seoul_line <- lm(rented_bike_count ~ mean_temperature, data = seoul)
seoul_bs <- lm(rented_bike_count ~ splines::bs(mean_temperature, df = 5), data = seoul)
seoul_ns <- lm(rented_bike_count ~ splines::ns(mean_temperature, df = 5), data = seoul)
temperature_grid <- tibble(mean_temperature = seq(
  min(seoul$mean_temperature), max(seoul$mean_temperature), length.out = 250
))
curve_fits <- temperature_grid |>
  mutate(
    Line = predict(seoul_line, newdata = temperature_grid),
    `B-spline` = predict(seoul_bs, newdata = temperature_grid),
    `Natural spline` = predict(seoul_ns, newdata = temperature_grid)
  ) |>
  pivot_longer(-mean_temperature, names_to = "model", values_to = "fitted")
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.2) +
  geom_line(data = curve_fits,
            aes(y = fitted, colour = model), linewidth = 1) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "Model")
```

Plot the natural spline residuals against temperature to check for any remaining shape.

```{r seoul-spline-residuals}
seoul |>
  mutate(residual = residuals(seoul_ns)) |>
  ggplot(aes(mean_temperature, residual)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Mean temperature (°C)", y = "Natural spline residual")
```

Discuss: Where does a line miss the shape in the points? How does the LOESS span change the curve? Compare the B-spline and natural spline near the observed temperature limits. A natural spline becomes linear beyond its boundary knots, but does that make extrapolation safe? Would a more flexible fit to these observations by itself justify choosing that model?

**Worked interpretation:** The low-span LOESS moves more with local fluctuations. A line misses the broad curved temperature relationship. The two spline fits can differ near the extremes, where data are thin. Natural spline tails are constrained to be linear outside the boundary knots; that constraint does not supply new information outside the observed range. The residual plot is a check for remaining shape, and in-sample fit alone is not a measure of future prediction quality.

## Can we separate trend from seasonality?

### Seoul: fit a trend, then seasonal pairs

Plot rentals against the actual date. Fit a linear `day_index` trend; then add one annual sine and cosine pair. The pair can express a seasonal wave at any phase, rather than fixing its peak to the start of the calendar. Add a second pair for a sharper within-year shape. Both terms in each pair have to enter together. Use `anova()` for a joint, two-degree-of-freedom comparison of the first- and second-pair models.

```{r seoul-fourier}
seoul |>
  ggplot(aes(date, rented_bike_count)) +
  geom_line() + labs(x = "Date", y = "Daily rentals")
seoul_time <- seoul |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )
seoul_trend <- lm(rented_bike_count ~ day_index, data = seoul_time)
seoul_first <- lm(rented_bike_count ~ day_index + S1 + C1, data = seoul_time)
seoul_second <- lm(rented_bike_count ~ day_index + S1 + C1 + S2 + C2,
                   data = seoul_time)
seoul_time |>
  mutate(Trend = fitted(seoul_trend), `First pair` = fitted(seoul_first),
         `Second pair` = fitted(seoul_second)) |>
  pivot_longer(c(Trend, `First pair`, `Second pair`),
               names_to = "model", values_to = "fitted_mean") |>
  ggplot(aes(date, fitted_mean, colour = model)) +
  geom_line() +
  labs(x = "Date", y = "Daily rentals", colour = "Model")
anova(seoul_first, seoul_second)
```

Discuss: What does the trend alone miss? Does the second pair change the fitted seasonal shape, and what does the partial F test say about its extra terms? We have only one year of Seoul data: what can we say about a pattern recurring next year?

**Worked interpretation:** The trend is a straight temporal change and cannot follow the observed within-year rise and fall. A complete sine/cosine pair allows the phase to be estimated. The second pair adds another frequency. For these 352 days, the partial comparison gives $F(2,346) = 21.79$, $p < 0.001$ under the usual independent-error assumptions; the two extra terms improve the fit within this sample. We will check those residual assumptions next. We have not observed whether the seasonal pattern recurs next year.

A factor for calendar season or month gives each category its own offset. Sine and cosine give a smooth within-year pattern. The terms answer different questions; neither makes a one-year record into evidence about future years.

## What is left in the residuals?

### Seoul: check the sequence before choosing an error model

Keep the second-pair mean model from above. Plot its residuals against actual dates and check for long stretches on one side of zero. A missing part of the mean can also create a pattern in the residuals, so inspect the mean model before assuming that dependence alone is the problem.

The ordinary ACF and sequence tests assume equally spaced observations. For those diagnostics use the uninterrupted 131-day run from 1 December 2017 to 10 April 2018. Check the dates before treating lag 1 as one day. `DescTools::RunsTest()` splits a numerical sequence at its median; `lmtest::dwtest()` tests first-order correlation in regression residuals; `Box.test(..., type = "Ljung-Box")` tests several lags together. Use the time plot and ACF to interpret their small P-values rather than treating the tests as a diagnosis of the cause.

```{r seoul-dependence}
seoul_residuals <- seoul_time |>
  mutate(residual = residuals(seoul_second))
seoul_residuals |>
  ggplot(aes(date, residual)) +
  geom_line() + geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Actual date", y = "Second-pair residual")
seoul_run <- seoul_residuals |>
  filter(date >= as.Date("2017-12-01"), date <= as.Date("2018-04-10"))
stopifnot(nrow(seoul_run) == 131,
          all(as.integer(diff(seoul_run$date)) == 1))
DescTools::RunsTest(seoul_run$residual)
acf(seoul_run$residual, main = "Consecutive Seoul residuals")
Box.test(seoul_run$residual, lag = 14, type = "Ljung-Box")
seoul_run_lm <- lm(rented_bike_count ~ day_index + S1 + C1 + S2 + C2,
                   data = seoul_run)
lmtest::dwtest(seoul_run_lm, alternative = "greater")
```

Discuss: What do the time plot, median runs and lag-1 ACF show? A small Durbin-Watson or Ljung-Box P-value indicates what kind of departure? Could a missed mean pattern produce some of it? Why must we not feed the gappy 352-day residual vector into an ordinary daily ACF?

**Worked interpretation:** Long same-sign stretches and prominent ACF bars indicate dependence in the consecutive run. The runs test checks sequencing around the median, Durbin-Watson focuses on lag one and Ljung-Box combines several lags. A small P-value is evidence against the respective null, not evidence that AR(1) is necessarily the right repair. Missing seasonal or weather effects can also leave structure. The 13 missing service days make ordinary lag-1 indexing wrong for the full retained vector.

## Does AR(1) change the tourism analysis?

### Victoria tourism: work through the full comparison

Each row in `motel.csv` is a month from January 1980 to December 1994. `RoomNights` counts occupied room nights. `Time` increases one step per month and `Month` is a calendar-month factor. We can compare an indicator model for the seasonal mean with an AR(1) model for dependence left in its errors. In an AR(1) model, $\varepsilon_t = \phi\varepsilon_{t-1} + \eta_t$ with $|\phi| < 1$, and errors $k$ months apart have correlation $\phi^k$.

```{r tourism-data, message=FALSE}
tourism <- read_csv("../data/motel.csv", show_col_types = FALSE) |>
  mutate(
    Date = as.Date(Date),
    Month = factor(lubridate::month(Date, label = TRUE), ordered = FALSE)
  ) |>
  arrange(Date) |>
  mutate(Time = row_number() - 1L)
stopifnot(identical(tourism$Date,
                    seq.Date(min(tourism$Date), max(tourism$Date), by = "month")))
```

Plot room nights, fit `RoomNights ~ Time + Month` by OLS, plot the fitted mean and the residuals over time, and inspect the residual ACF. Then fit the same mean by GLS with `corAR1(form = ~ Time)`. Compare the `Time` estimates and their standard errors, extract $\phi$ and check the ACF of the *normalised* GLS residuals. Normalising accounts for the fitted variance and correlation structure before checking what remains.

```{r tourism-models}
ggplot(tourism, aes(Date, RoomNights)) +
  geom_line() + labs(x = "Month", y = "Room nights")
tourism_lm <- lm(RoomNights ~ Time + Month, data = tourism)
summary(tourism_lm)
tourism |>
  mutate(fitted_mean = fitted(tourism_lm), residual = residuals(tourism_lm)) |>
  ggplot(aes(Date, RoomNights)) +
  geom_line(colour = "grey60") +
  geom_line(aes(y = fitted_mean), colour = "blue") +
  labs(x = "Month", y = "Room nights")
tourism |>
  mutate(residual = residuals(tourism_lm)) |>
  ggplot(aes(Date, residual)) +
  geom_line() + geom_hline(yintercept = 0, linetype = 2)
acf(residuals(tourism_lm), main = "Tourism OLS residuals")
```

```{r tourism-gls}
tourism_gls <- gls(RoomNights ~ Time + Month,
                   correlation = corAR1(form = ~ Time), data = tourism)
summary(tourism_gls)
bind_rows(
  broom::tidy(tourism_lm) |> filter(term == "Time") |> mutate(model = "OLS"),
  broom::tidy(tourism_gls) |> filter(term == "Time") |> mutate(model = "GLS AR(1)")
) |>
  select(model, estimate, std.error)
coef(tourism_gls$modelStruct$corStruct, unconstrained = FALSE)
acf(residuals(tourism_gls, type = "normalized"),
    main = "Tourism normalised GLS residuals")
```

Discuss: Which observed patterns belong to the mean model and which remain in the errors? What does the estimated $\phi$ imply about nearby months? Did GLS change the `Time` estimate, its standard error, or both? Does the normalised residual ACF look quieter? Correlated errors make ordinary OLS standard errors unreliable, but do not by themselves force a correctly specified OLS slope to be biased.

**Worked interpretation:** The month factor describes a repeating calendar-month mean while the residual ACF asks whether nearby months still move together after that adjustment. Read the `Time` rows of the printed comparison in room nights per month, then compare their standard errors. The printed correlation parameter is $\hat\phi$; its sign and magnitude describe the modelled residual persistence. Fewer conspicuous bars in the normalised ACF are encouraging, but a remaining pattern could point to an incomplete mean or error model.

## Does the same workflow hold for Perth?

### Quarterly house prices: make your own comparison

Each row of `hp.csv` is one quarter. `Perth` is a relative house-price index and `WtAve` is the weighted index across eight Australian capitals. The indices use Darwin in the second quarter of 1986 as a reference value of 100. The published `WtAve` includes Perth itself, so describe an association with that average rather than an independent prediction from the other seven cities. `lubridate::my()` parses the month-year text as the quarter's first month.

```{r perth-data, message=FALSE}
hp <- read_csv("../data/hp.csv", show_col_types = FALSE) |>
  mutate(Quarter = as.Date(lubridate::my(Quarter))) |>
  arrange(Quarter) |>
  mutate(Time = row_number() - 1L)
stopifnot(identical(hp$Quarter,
                    seq.Date(min(hp$Quarter), max(hp$Quarter), by = "3 months")))
hp |> select(Quarter, Time) |> slice(c(1, n()))
```

Plot `Perth` against `Quarter`. Fit `Perth ~ WtAve + Time` with `lm()`, plot its residuals against the quarter and inspect their ACF. Fit that same mean with `gls()` and `corAR1(form = ~ Time)`, compare all coefficient estimates and standard errors, and inspect the normalised residual time plot and ACF. Use `intervals()` to inspect the interval for $\phi$. All operations have appeared in the Seoul or tourism blocks.

```{r perth-analysis}
ggplot(hp, aes(Quarter, Perth)) +
  geom_line() + labs(x = "Quarter", y = "Perth relative price index")
hp_lm <- lm(Perth ~ WtAve + Time, data = hp)
summary(hp_lm)
hp |>
  mutate(residual = residuals(hp_lm)) |>
  ggplot(aes(Quarter, residual)) +
  geom_line() + geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Quarter", y = "OLS residual")
acf(residuals(hp_lm), main = "Perth OLS residuals")
hp_gls <- gls(Perth ~ WtAve + Time,
              correlation = corAR1(form = ~ Time), data = hp)
summary(hp_gls)
bind_rows(
  broom::tidy(hp_lm) |> mutate(model = "OLS"),
  broom::tidy(hp_gls) |> mutate(model = "GLS AR(1)")
) |>
  select(model, term, estimate, std.error)
intervals(hp_gls)
hp_normalised <- residuals(hp_gls, type = "normalized")
hp |>
  mutate(normalised_residual = hp_normalised) |>
  ggplot(aes(Quarter, normalised_residual)) +
  geom_line() + geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Quarter", y = "Normalised GLS residual")
acf(hp_normalised, main = "Perth normalised GLS residuals")
```

Discuss: Where does Perth's index move with `WtAve`, and where does time add information? Is there evidence of serial structure after OLS and after GLS? What changed in the estimated coefficients and their uncertainty? Explain why the overlap between `WtAve` and Perth limits the interpretation.

We have now checked the shape of the mean and the dependence left after fitting it. If a clear pattern survives in the normalised residuals, what would you revisit first: the mean model, the AR(1) error model, or both? Give a reason from the plots rather than another P-value alone.

**Worked interpretation:** In the 45-quarter OLS fit, the `WtAve` coefficient is about 1.01 index units per unit of the published weighted average (SE 0.055), conditional on the time trend. The estimated time coefficient is about -0.066 index units per quarter (SE 0.074). Successive OLS residuals have a sample correlation of about 0.83, so the normalised GLS time plot and ACF are an important recheck. Compare the printed GLS estimates and standard errors with those OLS values. The association with `WtAve` includes Perth's contribution to that average. If a pattern survives GLS, re-examine the specified mean as well as the error correlation.
