---
title: "Lab 11 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
library(car)
library(glmnet)
```

# Computer Laboratory 11: Search, Shrinkage, and Weighting

Last week we built and compared several Auto MPG models by hand. That worked because the candidate set was still manageable.

This week we continue with the same data, but we now let the predictor set become larger. We will first automate part of the model search, then look at what happens when several predictors carry almost the same information. In the second part of the lab, the problem changes: we will use weighted regression because the observations themselves do not all carry the same amount of information.

# Part A: Auto MPG

```{r import-auto, message=FALSE}
auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin))

levels(auto$origin)
```

Europe is the reference level for `origin`. As in Lab 10, we will model `mpg` and exclude `car_name` from the candidate predictors.

## Search from both ends

The null model contains only an intercept. The full candidate model contains the seven predictors we are willing to consider.

```{r search-models}
null_model <- lm(mpg ~ 1, data = auto)

full_model <- lm(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)
```

Fit the null and full Auto MPG models shown here. Run forward search from the null model and backward search from the full model using the same AIC scope. Print each search path and final formula.

```{r run-searches}
search_scope <- list(
  lower = formula(null_model),
  upper = formula(full_model)
)

forward_search <- step(
  null_model,
  scope = search_scope,
  direction = "forward",
  trace = 0
)

backward_search <- step(
  full_model,
  scope = search_scope,
  direction = "backward",
  trace = 0
)

```

```{r search-paths}
forward_search$anova
backward_search$anova

formula(forward_search)
formula(backward_search)
```

Discuss: Did the two routes end at the same model? What would agreement or disagreement tell us about this particular search, and what would it *not* establish?

**Solution.** Forward search adds `weight`, `model_year`, `origin`, `displacement`, `horsepower`, and then `cylinders`. Backward search starts from the full model and removes only `acceleration`.

Both searches therefore finish at

```text
mpg ~ cylinders + displacement + horsepower + weight +
      model_year + origin
```

They agree for these data and this candidate set. Stepwise procedures can still be path-dependent: at each stage they make a local decision from the models currently available. Agreement does not establish that this is the uniquely best model, and it does not account for models or terms outside the specified scope.

Compare the AIC values for the selected and full models.

```{r compare-search-aic}
AIC(full_model, forward_search, backward_search)
```

**Solution.** The two searches agree and end about 1.3 AIC units below the full model. That is a small separation, so the lower AIC by itself is a weak reason to prefer the selected formula. The full model keeps `acceleration`; we should consider whether that term matters for the modelling question, and how we would assess prediction on new cars.

## Keep `origin` in every candidate model

Fit a lower-bound model containing `origin`, then repeat backward search with that model as the lower scope limit. Print the selected formula.

**Solution.** Put `origin` in the lower model. It then becomes part of the lower boundary of the search.

```{r forced-origin}
origin_model <- lm(mpg ~ origin, data = auto)

forced_origin_search <- step(
  full_model,
  scope = list(
    lower = formula(origin_model),
    upper = formula(full_model)
  ),
  direction = "backward",
  trace = 0
)

formula(forced_origin_search)
```

Discuss: What decision did we make before the search? Why could an automatic AIC step not remove `origin` this time?

This controls which models the algorithm is allowed to visit. It does not make the other statistical limitations of stepwise search disappear.

## How much do the numerical predictors overlap?

Make a correlation matrix for `cylinders`, `displacement`, `horsepower`, `weight`, `acceleration` and `model_year`. Then compute the VIFs or adjusted GVIFs for the full model.

```{r correlation-matrix}
auto |>
  select(
    cylinders,
    displacement,
    horsepower,
    weight,
    acceleration,
    model_year
  ) |>
  cor() |>
  round(3)
```

**Solution.** `cylinders`, `displacement`, `horsepower`, and `weight` are strongly correlated. For example, the correlations of cylinders with displacement and displacement with weight are about 0.95 and 0.93. These variables are all carrying information about related aspects of vehicle size and power.

Now inspect variance inflation factors for the full model.

```{r vif-full}
car::vif(full_model)
```

Discuss: Which conditional coefficients might be hard to estimate separately? Why would a high VIF alone be an inadequate reason to delete one of them?

**Solution.** For the one-degree-of-freedom numerical terms, the `GVIF` column is the usual VIF. The values are about 10.7 for `cylinders`, 22.9 for `displacement`, 10.0 for `horsepower`, and 11.1 for `weight`. These values match what we saw in the correlation matrix: the model has difficulty separating the conditional contributions of these overlapping predictors.

`origin` has two degrees of freedom, so its raw GVIF is not directly comparable with the one-degree-of-freedom terms. The adjusted `GVIF^(1/(2*Df))` column is provided for that comparison.

## Simple associations and conditional coefficients

Fit separate simple regressions for `cylinders`, `displacement`, `horsepower` and `weight`. Compare each slope with the corresponding coefficient in the full model.

```{r coefficient-comparison}
overlapping_predictors <- c(
  "cylinders",
  "displacement",
  "horsepower",
  "weight"
)

simple_models <- map(
  overlapping_predictors,
  ~ lm(reformulate(.x, response = "mpg"), data = auto)
)

coefficient_comparison <- tibble(
  Predictor = overlapping_predictors,
  `Simple coefficient` = map2_dbl(
    simple_models,
    overlapping_predictors,
    ~ unname(coef(.x)[.y])
  ),
  `Full-model coefficient` = unname(
    coef(full_model)[overlapping_predictors]
  ),
  `Full-model SE` = unname(
    coef(summary(full_model))[overlapping_predictors, "Std. Error"]
  )
) |>
  mutate(across(where(is.numeric), ~ round(.x, 4)))

knitr::kable(coefficient_comparison)
```

Discuss: Which signs or magnitudes changed? Describe the full-model coefficient as an association conditional on the other included predictors; do not interpret a sign reversal as a causal benefit.

**Solution.** Every simple-regression coefficient is negative: cars with more cylinders, greater displacement, more horsepower, or greater weight tend to have lower `mpg`. In the combined model, the coefficient magnitudes change substantially and the displacement coefficient becomes positive.

That sign change does not mean greater displacement causes fuel economy to improve. The full-model displacement coefficient compares cars with the other included predictors held fixed, including weight, cylinders, and horsepower. The data contain relatively little independent variation with which to make that comparison. The fitted model can still describe or predict the response reasonably while its individual conditional coefficients are difficult to estimate precisely.

## Ridge and lasso on the same folds

Stepwise regression makes a hard decision: a term is either in the model or out of it. Penalised regression takes a different approach. We can keep a larger starting model but penalise large coefficients. Ridge usually keeps all predictors and shrinks their coefficients. Lasso can shrink some coefficients all the way to zero.

Build the model matrix shown here. Use one fold assignment for both `cv.glmnet()` fits. Plot both coefficient paths and both CV curves. Extract coefficients at `lambda.min` and `lambda.1se`.

```{r glmnet-data}
X <- model.matrix(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)[, -1]

y <- auto$mpg
```

Last week we introduced cross-validation as a way to evaluate performance on held-out observations. Here `cv.glmnet()` uses the same idea to choose the amount of shrinkage. We use the same folds for ridge and lasso so that their cross-validation results are directly comparable.

```{r cross-validated-penalties}
set.seed(161251)
fold_id <- sample(rep(1:10, length.out = nrow(auto)))

ridge_cv <- cv.glmnet(
  X,
  y,
  alpha = 0,
  foldid = fold_id
)

lasso_cv <- cv.glmnet(
  X,
  y,
  alpha = 1,
  foldid = fold_id
)
```

### 3.1 Coefficient paths

Plot the coefficient paths. What changes as the penalty becomes stronger?

```{r coefficient-paths, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge coefficient paths")
plot(lasso_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso coefficient paths")
par(mfrow = c(1, 1))
```

**Solution.** Both methods shrink coefficients towards zero as the penalty increases. Ridge shrinks the correlated predictors together and does not set them exactly to zero. The lasso paths can reach zero, so lasso can produce a smaller active set of predictors.

The predictors use different units, so the vertical sizes of the returned coefficients should not be read as a direct ranking of predictor importance. `glmnet` standardises internally while fitting, then reports coefficients on the original scales.

### 3.2 Choosing the penalty by cross-validation

```{r cv-curves, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv)
title("Ridge cross-validation")
plot(lasso_cv)
title("Lasso cross-validation")
par(mfrow = c(1, 1))

penalty_choices <- tibble(
  Method = c("Ridge", "Lasso"),
  `lambda.min` = c(ridge_cv$lambda.min, lasso_cv$lambda.min),
  `lambda.1se` = c(ridge_cv$lambda.1se, lasso_cv$lambda.1se)
)

knitr::kable(penalty_choices, digits = 4)
```

Discuss: What trade-off is made by choosing `lambda.1se` rather than `lambda.min`?

**Solution.** `lambda.min` gives the smallest mean cross-validated error among the fitted penalty values. `lambda.1se` chooses the most strongly regularised model whose estimated error is within one standard error of that minimum. It accepts a small amount of uncertainty in predictive performance in exchange for stronger shrinkage and, for lasso, usually a simpler active set.

Cross-validation estimates performance on held-out folds from this dataset. It is more relevant to prediction than training RSS or adjusted R-squared, but it is still an estimate with sampling and fold-to-fold uncertainty.

### 3.3 Which coefficients remain?

```{r penalised-coefficients}
round(as.matrix(coef(ridge_cv, s = "lambda.min")), 4)
round(as.matrix(coef(ridge_cv, s = "lambda.1se")), 4)

round(as.matrix(coef(lasso_cv, s = "lambda.min")), 4)
round(as.matrix(coef(lasso_cv, s = "lambda.1se")), 4)
```

Discuss: Which coefficients are exactly zero under lasso? What trade-off does `lambda.1se` make? Why is a zero dummy coefficient for one `origin` level different from removing the `origin` term as a whole?

**Solution.** Ridge retains every coefficient at both penalty choices, but shrinks them. At `lambda.min`, lasso sets `cylinders`, `displacement`, and `acceleration` to zero. At the more strongly regularised `lambda.1se`, it also sets the Japan-versus-Europe origin contrast to zero.

For this particular fold assignment, lasso retains weight and horsepower from the strongly correlated size-and-power group while dropping cylinders and displacement. This is one way lasso can behave with correlated predictors: it may retain some members of an overlapping group and remove others. It is not a general promise about which member will be chosen, and a different sample or fold assignment could change the active set.

# Part B: Weighted regression for magazine readership

So far this week we have been deciding how to handle predictors in the mean model. Weighted regression is a different problem. Now the rows themselves do not all carry the same amount of information.

The magazine data contain an estimated mean number of magazines read for each age and sex group. Each row summarises `N` people and reports the within-group standard deviation `StDev`.

```{r import-magazines, message=FALSE}
mags <- readr::read_csv("../data/AveMags.csv") |>
  mutate(
    Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
    Older = as.integer(Age >= 50),
    precision_weight = N / StDev^2
  )

glimpse(mags)
```

## Which magazine means are most precise?

```{r magazine-data-plot}
ggplot(mags, aes(x = Age, y = Avemags)) +
  geom_point(aes(size = N, colour = StDev, shape = Sex)) +
  labs(
    x = "Age",
    y = "Mean number of magazines",
    size = "Group size",
    colour = "Within-group SD"
  ) +
  theme_minimal()
```

The fitted pattern needs to allow the age trend to change between teenagers, adults, and older adults. We will use the following mean structure rather than rebuilding it from scratch:

```{r magazine-mean-model}
magazine_unweighted <- lm(
  Avemags ~ Age * Adult + Age * Older + Sex,
  data = mags
)
```

This is a hierarchical formula. It includes `Age`, both group indicators, and both corresponding age interactions. The teenage slope is the coefficient of `Age`; the adult slope adds the `Age:Adult` coefficient; and the older-adult slope also adds `Age:Older`. The `Adult` and `Older` main effects allow level changes at the group boundaries, so this model does not force the fitted pieces to meet exactly at those boundaries. The sex difference is additive and is held constant across ages.

```{r magazine-unweighted-fit}
mags |>
  mutate(fitted = fitted(magazine_unweighted)) |>
  ggplot(aes(x = Age, y = Avemags, colour = Sex)) +
  geom_point() +
  geom_line(aes(y = fitted)) +
  labs(
    x = "Age",
    y = "Mean number of magazines"
  ) +
  theme_minimal()
```

**Solution.** The unweighted fit increases through the teenage ages, is approximately flat through adulthood, and decreases for the older group. Females have a higher fitted mean than males at each age because the model contains an additive sex effect but no age-by-sex interaction.

Compute each group's precision weight as `N/StDev^2` and list the groups from highest to lowest weight.

For a mean based on (N_i) independent people with within-group standard deviation (SD_i), its sampling variance is proportional to

\[
\frac{SD_i^2}{N_i}.
\]

We therefore use inverse-variance weights proportional to

\[
w_i = \frac{N_i}{SD_i^2}.
\]

```{r inspect-magazine-weights}
mags |>
  select(Age, Sex, N, StDev, precision_weight) |>
  arrange(desc(precision_weight))
```

Discuss: Why do both group size and within-group variation enter the weight?

**Solution.** The age-67 groups receive by far the largest weights because they combine large sample sizes with relatively modest standard deviations. The teenage groups generally receive much less weight. The 16-year-old male group has the smallest weight because its estimated mean is based on only 18 people and has a large within-group standard deviation.

## Hold the mean formula fixed and change the weights

Fit the provided age-piece and sex formula once without weights and once with `weights = N/StDev^2`. Compare the fitted values and coefficients. Then plot ordinary residuals for the first fit and $\sqrt{w_i}(y_i-\hat y_i)$ for the weighted fit.

```{r fit-weighted-model}
magazine_weighted <- lm(
  Avemags ~ Age * Adult + Age * Older + Sex,
  data = mags,
  weights = precision_weight
)

coefficient_comparison <- tibble(
  Term = names(coef(magazine_unweighted)),
  Unweighted = unname(coef(magazine_unweighted)),
  Weighted = unname(coef(magazine_weighted))
) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(coefficient_comparison)
```





**Solution.** Weighting changes the fit because the highly weighted adult and older groups now have more influence than the relatively imprecise teenage groups. The fitted female-minus-male difference decreases from about 1.13 magazines in the unweighted model to about 0.72 in the weighted model. The weighted teenage fitted values, especially for females, move down, while several fitted values for older males move up slightly.

The weighting has not changed the response or the mean-model formula. It has changed the criterion being minimised: discrepancies from more precise group means count more heavily.

```{r compare-magazine-fits}
magazine_fits <- bind_rows(
  mags |>
    mutate(
      fitted = fitted(magazine_unweighted),
      fit = "Unweighted"
    ),
  mags |>
    mutate(
      fitted = fitted(magazine_weighted),
      fit = "Weighted"
    )
)

ggplot(magazine_fits, aes(x = Age, colour = Sex)) +
  geom_point(aes(y = Avemags, size = precision_weight), alpha = 0.5) +
  geom_line(aes(y = fitted, linetype = fit), linewidth = 0.9) +
  labs(
    x = "Age",
    y = "Mean number of magazines",
    size = "Precision weight",
    linetype = "Fit"
  ) +
  theme_minimal()
```

### Ordinary and weighted residuals

For the unweighted model, inspect the ordinary residuals. For the weighted model, inspect

\[
\sqrt{w_i}(y_i - \hat{y}_i),
\]

which places the residuals on the scale used by weighted least squares.

```{r compare-magazine-residuals}
residual_comparison <- bind_rows(
  mags |>
    transmute(
      Age,
      Sex,
      residual = resid(magazine_unweighted),
      model = "Unweighted: ordinary residuals"
    ),
  mags |>
    transmute(
      Age,
      Sex,
      residual = resid(magazine_weighted) * sqrt(precision_weight),
      model = "Weighted: weighted residuals"
    )
)

ggplot(residual_comparison, aes(x = Age, y = residual, colour = Sex)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_point() +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  labs(x = "Age", y = "Residual") +
  theme_minimal()
```

Discuss: Which fitted contrasts change? What assumption about the precision of a group mean justifies the weights? Why do the two residual panels have different scales?

**Solution.** The weighted residuals are scaled by the square root of their precision weights, so the panels are on different scales. A small weighted residual means the discrepancy is small relative to the assumed standard deviation of that group mean; it does not by itself show that the group is unimportant.

**Solution.** The ordinary residual plot treats every group mean as equally precise even though the sample sizes and standard deviations differ considerably. The weighted residuals rescale each discrepancy by the information behind that row. A raw discrepancy from a precise group mean matters more; the same discrepancy from a noisy group mean matters less.

We should not compare the vertical magnitudes across the two panels as if they were in identical units. The useful comparison is whether the residual structure is reasonable on the error scale assumed by each fit. Some variation and individual large residuals remain, but the weighted analysis no longer asks the low-information teenage rows to influence the fit as much as the large, more precise older groups.

## Final decision

In one short paragraph, state when you would use term-based search, ridge or lasso, and weighted regression in these two examples. Explain what you would need to assess before claiming the selected Auto MPG model predicts new cars well.

The two parts of this lab solve different problems:

- Stepwise search and penalised regression respond to a large or redundant predictor set.
- VIFs and coefficient comparisons diagnose why individual conditional effects can become unstable when predictors overlap.
- Cross-validation chooses the amount of shrinkage using held-out predictions.
- Weighted regression changes how much each observation contributes when the observations have different precisions.

None of these methods chooses a scientifically meaningful model on its own. The model still depends on the question, the available variables, and the assumptions connecting the fitted calculation to that question.
