---
title: "Computer Laboratory 11: Search, Shrinkage, and Weighting"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
library(car)
library(glmnet)
```

# Computer Laboratory 11: Search, Shrinkage, and Weighting

Last week we built and compared several Auto MPG models by hand. That worked because the candidate set was still manageable.

This week we keep the same data but make the problem harder. We first automate part of the model search, then look at what happens when several predictors carry almost the same information. In the second part, the problem changes: the magazine data contain group means with very different levels of precision.

The lab uses [auto_mpg.csv](https://knowlton.co.nz/161251/data/auto_mpg.csv) and [AveMags.csv](https://knowlton.co.nz/161251/data/AveMags.csv). The downloadable R Markdown file reads both datasets directly from the course site.

# Part A: Auto MPG

## Search from both ends

Let's start with the same response and candidate predictors from Lab 10. The null model contains only an intercept; the full model contains the seven predictors we are willing to consider.

```{r import-auto, message=FALSE}
auto <- read_csv("https://knowlton.co.nz/161251/data/auto_mpg.csv") |>
  mutate(origin = factor(origin))

levels(auto$origin)
```

As in Lab 10, we will model `mpg` and exclude `car_name` from the candidate predictors. `origin` is a factor, with Europe as the reference level.

```{r search-models}
null_model <- lm(mpg ~ 1, data = auto)

full_model <- lm(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)

search_scope <- list(
  lower = formula(null_model),
  upper = formula(full_model)
)
```

### Start with forward search

Here is the forward search from the intercept-only model. Run it and inspect the order in which terms enter.

```{r forward-search}
forward_search <- step(
  null_model,
  scope = search_scope,
  direction = "forward",
  trace = 0
)

forward_search$anova
formula(forward_search)
```

### Repeat the search backwards

Now start from `full_model` and run backward elimination using the same `search_scope`.

Print the search path and the final formula.

```{r backward-search, eval=FALSE}
# Fit backward_search using step().
# Start from full_model.
# Use search_scope.
# Set direction = "backward".
# Then print:
# backward_search$anova
# formula(backward_search)
```

Discuss: Did the forward and backward routes end at the same model? What would agreement or disagreement tell us about this particular search, and what would it *not* establish?

Now compare the AIC values for `full_model`, `forward_search` and `backward_search`.

```{r compare-search-aic, eval=FALSE}
# Compare the three models with AIC().
```

Discuss: How large is the difference between the lowest AIC and the full model's AIC? If the values are close, what else would you consider before choosing a model?

## Keep `origin` in every candidate model

The first search allowed every candidate term to be removed. Now make one modelling decision in advance: `origin` must stay in every model considered by the search.

Fit a lower-bound model containing `origin`, then repeat backward search from `full_model` using that model as the lower scope limit. Print the selected formula.

```{r forced-origin, eval=FALSE}
# 1. Fit origin_model with mpg explained by origin.
# 2. Run backward stepwise search from full_model.
# 3. Use origin_model as the lower model in scope.
# 4. Print the resulting formula.
```

Discuss: What decision did we make before the search? Why can the AIC search no longer remove `origin`?

## How much do the numerical predictors overlap?

Several of the Auto MPG predictors describe related aspects of the same car. Start with their pairwise correlations.

```{r correlation-matrix}
auto |>
  select(
    cylinders,
    displacement,
    horsepower,
    weight,
    acceleration,
    model_year
  ) |>
  cor() |>
  round(3)
```

Identify the pairs with the strongest correlations. Which group of variables appears to be carrying substantially overlapping information?

Now calculate variance inflation factors for `full_model`.

```{r vif-full, eval=FALSE}
# Use car::vif() on full_model.
```

Discuss: Which numerical terms have the largest VIFs? Which conditional coefficients might therefore be hard to estimate separately? Why would a high VIF alone be an inadequate reason to delete a predictor?

## What happens to the coefficients?

The VIFs tell us that some coefficients are difficult to separate. We can see the consequence by comparing a simple regression with the same predictor's coefficient in the full model.

Start with `cylinders`.

```{r cylinders-simple}
cylinders_simple <- lm(mpg ~ cylinders, data = auto)

coef(cylinders_simple)
coef(full_model)["cylinders"]
```

Now repeat this for `displacement`, `horsepower` and `weight`. For each predictor, compare the simple-regression slope with its coefficient in `full_model`. Also inspect the full-model standard error.

```{r coefficient-comparison, eval=FALSE}
# Fit:
# displacement_simple <- lm(...)
# horsepower_simple   <- lm(...)
# weight_simple       <- lm(...)
#
# For each predictor, compare:
# 1. the simple-regression slope,
# 2. the coefficient from full_model,
# 3. the standard error from summary(full_model).
```

Discuss: Which signs or magnitudes changed after the other predictors were included? Describe the full-model coefficients as associations conditional on the other included predictors. Do not interpret a sign reversal as a causal benefit.

## Ridge and lasso on the same folds

We have now seen the problem with trying to separate strongly overlapping predictors using ordinary least squares. Ridge and lasso keep the larger predictor set but control coefficient size with a penalty.

The model matrix and common fold assignment are supplied because the point here is to interpret the penalised fits rather than spend the lab rebuilding the setup.

```{r glmnet-data}
X <- model.matrix(
  mpg ~ cylinders + displacement + horsepower + weight +
    acceleration + model_year + origin,
  data = auto
)[, -1]

y <- auto$mpg

set.seed(161251)
fold_id <- sample(rep(1:10, length.out = nrow(X)))
```

Run ridge and lasso on the same folds.

```{r cross-validated-penalties}
ridge_cv <- cv.glmnet(
  X,
  y,
  alpha = 0,
  foldid = fold_id
)

lasso_cv <- cv.glmnet(
  X,
  y,
  alpha = 1,
  foldid = fold_id
)
```

### Compare the coefficient paths

The code below draws both coefficient paths.

```{r coefficient-paths, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge coefficient paths")
plot(lasso_cv$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso coefficient paths")
par(mfrow = c(1, 1))
```

Examine the two plots.

Discuss: As the penalty becomes stronger, what happens to the ridge coefficients? What happens to the lasso coefficients? Which method can set coefficients exactly to zero?

### Choose the penalty by cross-validation

Now inspect the cross-validation curves.

```{r cv-curves, fig.width=9, fig.height=5}
par(mfrow = c(1, 2))
plot(ridge_cv)
title("Ridge cross-validation")
plot(lasso_cv)
title("Lasso cross-validation")
par(mfrow = c(1, 1))
```

`lambda.min` minimises the estimated cross-validation error. `lambda.1se` uses a stronger penalty while remaining within one standard error of that minimum.

Extract both values for ridge and lasso.

```{r penalty-values, eval=FALSE}
# Print:
# ridge_cv$lambda.min
# ridge_cv$lambda.1se
# lasso_cv$lambda.min
# lasso_cv$lambda.1se
```

Discuss: Which value is larger, `lambda.min` or `lambda.1se`? What trade-off are we making when we choose `lambda.1se`?

### Which coefficients remain?

Here is the ridge coefficient vector at `lambda.min`.

```{r ridge-min-coefficients}
round(as.matrix(coef(ridge_cv, s = "lambda.min")), 4)
```

Now extract the ridge coefficients at `lambda.1se` and the lasso coefficients at both `lambda.min` and `lambda.1se`.

```{r penalised-coefficients, eval=FALSE}
# Use the same coef(..., s = ...) syntax as above to obtain:
# 1. ridge at lambda.1se
# 2. lasso at lambda.min
# 3. lasso at lambda.1se
#
# Round the coefficient matrices to four decimal places.
```

Use those coefficient vectors to answer the following.

Discuss: Does ridge set any coefficients exactly to zero? Which coefficients are exactly zero under lasso at `lambda.min`? Which additional coefficients, if any, become zero at `lambda.1se`? Among `cylinders`, `displacement`, `horsepower` and `weight`, which variables does lasso retain? Why is a zero coefficient for one `origin` dummy variable different from removing the `origin` term as a whole?

# Part B: Weighted regression for magazine readership

So far we have been changing how predictors enter the mean model. Weighted regression solves a different problem. Here each row is already a group mean, and some of those means are estimated much more precisely than others.

Each row in the magazine data is the mean number of magazines read for one age and sex group. The columns `N` and `StDev` give the group size and within-group standard deviation.

## Start with the mean model

Load the data and look at how the group means vary with age.

```{r import-magazines, message=FALSE}
mags <- read_csv("https://knowlton.co.nz/161251/data/AveMags.csv") |>
  mutate(
    Sex = factor(Sex, levels = c(0, 1), labels = c("Male", "Female")),
    Older = as.integer(Age >= 50)
  )

glimpse(mags)
```

```{r magazine-data-plot}
ggplot(mags, aes(x = Age, y = Avemags)) +
  geom_point(aes(size = N, colour = StDev, shape = Sex)) +
  labs(
    x = "Age",
    y = "Mean number of magazines",
    size = "Group size",
    colour = "Within-group SD"
  ) +
  theme_minimal()
```

Before changing the weights, keep the mean structure fixed. The supplied variable `Adult` separates the teenage rows from the adult rows, and `Older` marks ages 50 and above. We will allow the age slope to change across those ranges while keeping the sex difference additive.

The mean model is supplied. The point of this section is to change the precision assumption, not to rediscover the mean structure.

```{r magazine-mean-model}
magazine_unweighted <- lm(
  Avemags ~ Age * Adult + Age * Older + Sex,
  data = mags
)
```

The teenage slope is the coefficient of `Age`. The adult slope adds the `Age:Adult` coefficient, and the older-adult slope also adds `Age:Older`. The main effects for `Adult` and `Older` allow level changes at the boundaries, so the fitted pieces are not forced to meet exactly.

```{r magazine-unweighted-fit}
mags |>
  mutate(fitted = fitted(magazine_unweighted)) |>
  ggplot(aes(x = Age, y = Avemags, colour = Sex)) +
  geom_point() +
  geom_line(aes(y = fitted)) +
  labs(x = "Age", y = "Mean number of magazines") +
  theme_minimal()
```

Discuss: What broad age pattern does this unweighted model capture? What does the absence of an age-by-sex interaction assume about the fitted sex difference?

## Which magazine means are most precise?

For a mean based on $N_i$ independent people with within-group standard deviation $SD_i$, the sampling variance is proportional to

$$
\frac{SD_i^2}{N_i}.
$$

Weighted least squares uses weights proportional to the inverse variance, so here

$$
w_i = \frac{N_i}{SD_i^2}.
$$

A group mean gets more weight when it is based on more people, and less weight when the observations within that group are more variable.

Calculate `precision_weight` for every row, then list the groups from highest to lowest weight.

```{r magazine-weights, eval=FALSE}
# Add precision_weight = N / StDev^2 to mags.
```

```{r inspect-magazine-weights, eval=FALSE}
# Select Age, Sex, N, StDev and precision_weight.
# Arrange the rows from highest to lowest precision_weight.
```

Discuss: Which groups receive the greatest weight? Which receive the least? Explain why both `N` and `StDev` affect the precision of a group mean.

## Hold the mean formula fixed and change the weights

Now fit the same mean formula again, but this time tell `lm()` to use `weights = precision_weight`.

```{r fit-weighted-model, eval=FALSE}
# Fit magazine_weighted using exactly the same formula as magazine_unweighted.
# Add weights = precision_weight.
```

Compare the coefficients from the two models.

```{r compare-magazine-coefficients, eval=FALSE}
# Print the coefficients from magazine_unweighted.
# Print the coefficients from magazine_weighted.
```

Discuss: Which coefficients change the most? Has weighting changed the response or the mean-model formula? What has changed in the criterion used to fit the model?

### Compare the fitted values

Add the fitted values from the unweighted and weighted models to a combined data frame, then plot both fits against age.

```{r compare-magazine-fits, eval=FALSE}
# Create magazine_fits by binding:
# 1. mags with fitted(magazine_unweighted) and fit = "Unweighted"
# 2. mags with fitted(magazine_weighted) and fit = "Weighted"
#
# Plot Avemags against Age.
# Show Sex by colour.
# Show precision_weight by point size.
# Draw the fitted values with different line types for the two fits.
```

Discuss: Where do the weighted and unweighted fitted values differ most? Relate those differences to the precision weights you calculated above.

### Compare ordinary and weighted residuals

For the unweighted model, use the ordinary residuals. For the weighted model, calculate

$$
\sqrt{w_i}(y_i - \hat{y}_i),
$$

which places the residuals on the scale used by weighted least squares.

```{r compare-magazine-residuals, eval=FALSE}
# Create ordinary residuals for magazine_unweighted.
# Create weighted residuals for magazine_weighted by multiplying
# resid(magazine_weighted) by sqrt(precision_weight).
# Plot both sets of residuals against Age in separate panels.
```

Discuss: Why do the two residual panels have different vertical scales? What assumption about the precision of the group means justifies the weighted analysis?

## Final decision

In one short paragraph, state when you would use term-based search, ridge or lasso, and weighted regression in these two examples. Explain what you would need to assess before claiming the selected Auto MPG model predicts new cars well.

Everything in this lab still assumes that we have chosen a sensible shape for the mean. In Lab 12 we relax that too: what do we do when a straight-line effect is not enough?

[View the worked solutions online](https://knowlton.co.nz/161251/labs/lab11-sols.html).
