---
title: "Lab 10"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

Let's work through each idea first with wage data, then repeat the same analysis with Auto MPG.

The wage data contain 3,000 male workers recorded from 2003 to 2009. Each row is one worker. Wage is recorded in thousands of US dollars. The Auto MPG data contain 392 cars from model years 1970 to 1982. Each row is one car.

Both datasets are observational, so the fitted relationships describe associations rather than causal effects.

| Wage variable | Meaning |
|:--|:--|
| `wage` | recorded wage, thousands of US dollars |
| `age` | age, years |
| `education` | education level |
| `year` | year the wage was recorded |
| `jobclass` | industrial or information work |

| Auto MPG variable | Meaning |
|:--|:--|
| `mpg` | fuel economy, miles per gallon |
| `weight` | vehicle weight, pounds |
| `origin` | USA, Europe or Japan |
| `model_year` | model year, 70--82 |
| `horsepower` | engine power, horsepower |

Let's leave `car_name` out of the modelling.

```{r import-data, message=FALSE}
wage <- readr::read_csv("../data/wage.csv") |>
  mutate(education = factor(education))

auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(wage)
glimpse(auto)
```

## What changes across groups?

### Wage example

Let's work through the complete three-model comparison using wage and age.

First plot wage against age, with education shown by colour.

```{r wage-raw-plot}
ggplot(wage, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Discuss: How does wage vary with age? Do the education groups appear to have similar vertical positions and slopes?

Now fit the three models.

```{r wage-fit-lines}
w1 <- lm(wage ~ age, data = wage)
w2 <- lm(wage ~ age + education, data = wage)
w3 <- lm(wage ~ age * education, data = wage)
```

#### One common line

Plot the fitted values from `w1`.

```{r wage-one-line}
wage |>
  mutate(fitted = predict(w1)) |>
  arrange(age) |>
  ggplot(aes(x = age, y = wage)) +
  geom_point(aes(colour = education), alpha = 0.2) +
  geom_line(aes(y = fitted), colour = "black", linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What is this model assuming about the education groups?

#### Parallel lines

Now plot the fitted values from `w2`.

```{r wage-parallel-lines}
wage |>
  mutate(fitted = predict(w2)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What changed from the one-line model? What is still being held the same across education groups?

#### Separate lines

Finally, plot the fitted values from `w3`.

```{r wage-separate-lines}
wage |>
  mutate(fitted = predict(w3)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What changed from the parallel-lines model? Which term in the formula allows that change?

```text
education * age = education + age + education:age
```

#### Read the separate-lines coefficients

Run the usual model summary.

```{r wage-read-interaction}
summary(w3)
```

The `age` coefficient is the age slope for the reference education group. The education coefficients change the intercept, and the age-by-education coefficients change the slope. Age zero is outside these data, so the intercept differences are not useful overall wage comparisons.

### Auto MPG: repeat the complete analysis

Now repeat the whole comparison using fuel economy, weight and origin.

First plot `mpg` against `weight`, with `origin` shown by colour.

```{r auto-raw-plot, eval=FALSE}
# Plot mpg against weight, with origin shown by colour.
```

Discuss: How does fuel economy vary with weight? Do the origin groups appear to have similar vertical positions and slopes?

Fit the three models below.

```text
a1: mpg ~ weight
a2: mpg ~ weight + origin
a3: mpg ~ weight * origin
```

```{r auto-fit-lines, eval=FALSE}
# Fit a1, a2 and a3.
```

#### One common line

Plot the fitted values from `a1`.

```{r auto-one-line, eval=FALSE}
# Add predict(a1) as fitted values and plot the common fitted line.
```

Discuss: What is this model assuming about the origin groups?

#### Parallel lines

Plot the fitted values from `a2`.

```{r auto-parallel-lines, eval=FALSE}
# Add predict(a2) as fitted values and plot one line for each origin.
```

Discuss: What changed from `a1`? What is still being held the same across origins?

#### Separate lines

Plot the fitted values from `a3`.

```{r auto-separate-lines, eval=FALSE}
# Add predict(a3) as fitted values and plot one line for each origin.
```

Discuss: What changed from `a2`? Which term allows the weight slopes to differ?

#### Read the separate-lines coefficients

Run `summary(a3)`.

```{r auto-read-interaction, eval=FALSE}
# Run summary(a3).
```

Discuss: Which coefficient is the USA weight slope? Which coefficients change the intercepts for Europe and Japan? Which coefficients change their slopes? Why are the origin main-effect coefficients not overall origin differences?

## Do the groups improve the model?

A partial F test compares a reduced model with a larger model containing all of the reduced model's terms. It tests the added terms together.

### Wage example

Let's complete the set of nested comparisons before changing datasets.

#### Does education add information after age?

```{r compare-wage-education}
anova(w1, w2)
```

Discuss: How many parameters were added? What does the P-value say about moving from one common line to parallel education lines?

#### Do the age slopes differ by education?

```{r compare-wage-slopes}
anova(w2, w3)
```

Discuss: Which terms were added? What does the P-value say about whether the age slope varies across education groups?

#### What does the full comparison test?

```{r compare-full-wage}
anova(w1, w3)
```

Discuss: How many parameters are added? What does this joint test establish?

#### Does term order matter?

A sequential ANOVA table gives each term credit for variation it explains when it enters the model.

```{r wage-sequential-anova}
anova(lm(wage ~ education + age, data = wage))
anova(lm(wage ~ age + education, data = wage))
```

Discuss: The fitted model is identical in both cases. Why do the sums of squares change?

### Auto MPG: repeat the complete comparison set

Now answer the same four questions using `a1`, `a2` and `a3`.

#### Does origin add information after weight?

```{r compare-auto-origin, eval=FALSE}
# Compare a1 with a2 using anova().
```

Discuss: How many parameters were added? What does the P-value say about moving from one common line to parallel origin lines?

#### Do the weight slopes differ by origin?

```{r compare-auto-slopes, eval=FALSE}
# Compare a2 with a3 using anova().
```

Discuss: Which terms were added? What does the P-value say about whether the weight slope varies across origins?

#### What does the full comparison test?

```{r compare-full-auto, eval=FALSE}
# Compare a1 with a3 using anova().
```

Discuss: How many parameters are added? What does this joint test establish?

#### Does term order matter?

```{r auto-sequential-anova, eval=FALSE}
# Compare:
# anova(lm(mpg ~ origin + weight, data = auto))
# anova(lm(mpg ~ weight + origin, data = auto))
```

Discuss: Why is `anova(a1, a2)` the clearer answer to the question "does origin add information after weight?"

## When does another predictor earn its place?

A smaller RSS is guaranteed when another predictor is added, so RSS alone cannot decide whether a larger model is worth keeping.

Let's use adjusted R-squared, residual standard error and AIC alongside the partial F test. `broom::glance()` puts these model summaries into one tidy row.

### 1. Add a clearly useful predictor

#### Wage example

Let's add year to the wage model.

```{r wage-add-year}
w4 <- lm(wage ~ age * education + year, data = wage)

bind_rows(
  w3 = broom::glance(w3),
  w4 = broom::glance(w4),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w3, w4)
```

Discuss: Do the four pieces of evidence all favour adding year? How large is the improvement?

#### Repeat this for Auto MPG

Now add `model_year` to the Auto MPG model and call it `a4`.

```{r auto-add-model-year, eval=FALSE}
# Fit a4: mpg ~ weight * origin + model_year.
# Use broom::glance() to compare a3 and a4.
# Compare a3 with a4 using anova().
```

Discuss: Does model year earn its place?

### 2. Add a predictor with a smaller contribution

#### Wage example

Let's add `jobclass` to the wage model.

```{r wage-add-jobclass}
w5 <- lm(wage ~ age * education + year + jobclass, data = wage)

bind_rows(
  w4 = broom::glance(w4),
  w5 = broom::glance(w5),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w4, w5)
```

Discuss: The improvement is much smaller than it was for year. Is `jobclass` still worth keeping in this descriptive model?

#### Repeat this for Auto MPG

Now add `horsepower` to `a4` and call the model `a5`.

```{r auto-add-horsepower, eval=FALSE}
# Fit a5: mpg ~ weight * origin + model_year + horsepower.
# Use broom::glance() to compare a4 and a5.
# Compare a4 with a5 using anova().
```

Discuss: Do adjusted R-squared, residual standard error, AIC and the partial F test tell the same story? Which model, `a4` or `a5`, is the better compact descriptive model?

## What if prediction is the goal?

### 1. Separate model fit from prediction

Everything so far has been calculated on the same observations used to fit the models.

Discuss: What could be done to assess prediction for cars that were not used to fit the model? Why are adjusted R-squared, residual standard error and AIC not direct estimates of that prediction error?

Do not implement cross-validation here. That comes next week.

## Does the working model fit the pattern?

### 1. Check the working model

#### Wage example

Let's check the wage model before treating it as finished.

Run the standard diagnostic plots for `w5`.

```{r wage-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w5)
par(mfrow = c(1, 1))
```

Now plot the residuals against age.

```{r wage-residuals-by-age}
wage |>
  mutate(residual = residuals(w5)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: What pattern should be visible if the straight-line age effect is adequate? What pattern is actually present?

### 2. Revise the mean structure

#### Wage example

The curve suggests that the mean structure still needs work. Quadratic regression was introduced earlier in the course, so let's add `age^2` while keeping the linear age term.

```{r wage-revise-mean}
w6 <- lm(
  wage ~ age * education + year + jobclass + I(age^2),
  data = wage
)

bind_rows(
  w5 = broom::glance(w5),
  w6 = broom::glance(w6),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w5, w6)
```

Repeat the diagnostic plots and the residual-versus-age plot for `w6`.

```{r wage-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w6)
par(mfrow = c(1, 1))

wage |>
  mutate(residual = residuals(w6)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: Has the broad age pattern been reduced? What concerns remain?

### 3. Check the Auto MPG working model

#### Repeat the diagnostic check for Auto MPG

Now make the same diagnostic check for `a4`.

Run the standard diagnostic plots, then plot the residuals against `weight`.

```{r auto-diagnostics, eval=FALSE}
# Plot the standard diagnostic panels for a4.
# Then plot residuals(a4) against weight using the wage plot above as the pattern.
```

Discuss: Is there a systematic pattern left in the mean?

### 4. Revise the Auto MPG mean structure

#### Repeat the model-revision step for Auto MPG

Now fit:

```text
a6: mpg ~ weight * origin + I(weight^2) + model_year
```

Compare `a4` with `a6`, then repeat both diagnostic checks.

```{r auto-revise-mean, eval=FALSE}
# Fit a6.
# Compare a4 and a6 using broom::glance() and anova().
# Repeat the standard diagnostics.
# Plot residuals(a6) against weight.
```

Discuss: What improved? What still needs attention?

The standard diagnostic plots do not establish independence. That has to be considered from the way the observations were collected and whether groups of observations might share unmodelled structure.

## Pulling it together

### Final question

Write a short conclusion for Auto MPG covering:

1. what changed between the common-line, parallel-line and separate-line models;
2. what the partial F tests said about origin;
3. whether horsepower was kept after model year, and why;
4. what the diagnostics led to changing;
5. why these in-sample comparisons do not establish prediction performance for new cars.

Keep this to about six sentences.

With a small candidate set, these decisions can still be made by hand. Once the predictor list gets longer, that becomes tedious very quickly. Lab 11 returns to Auto MPG and looks at automated search and shrinkage.
