---
title: "Computer Laboratory 12: Curves, trend and seasonality : solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory 12: Curves, trend and seasonality

This week we relax the shape of the mean model. We will use daily Seoul bike demand to compare LOESS, cubic B-splines and natural cubic splines, then let time itself enter the model through a trend and smooth seasonal terms.

The lab uses [seoul_bike_daily.csv](https://knowlton.co.nz/161251/data/seoul_bike_daily.csv). The downloadable R Markdown file reads the dataset directly from the course site.

## How much curve do we need for temperature?

Each row in `seoul_bike_daily.csv` is one calendar day. `rented_bike_count` is the number of rentals and `mean_temperature` is in °C. The 13 days when the system was not fully functioning should not represent ordinary demand. Keep the original `date` and `day_index` after filtering so that a gap in service remains a gap in time.

```{r seoul-data, message=FALSE}
seoul_all <- read_csv("https://knowlton.co.nz/161251/data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(date = as.Date(date)) |>
  arrange(date)
seoul <- seoul_all |>
  filter(functioning_day_all == "yes")
c(all_days = nrow(seoul_all), outage_days = nrow(seoul_all) - nrow(seoul),
  analysed_days = nrow(seoul))
```

### Seoul: compare the smooths and spline models

Start with the points and two LOESS curves. `span` is the fraction of nearby observations used in each local fit. A small span can follow local changes but can also chase noise; a large span smooths more strongly.

```{r seoul-loess}
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.3,
              se = FALSE, aes(colour = "span 0.3")) +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.9,
              se = FALSE, aes(colour = "span 0.9")) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "LOESS")
```

Now fit a line, a cubic B-spline and a natural cubic spline, using `df = 5` for both splines. Here `df` controls the size of the spline basis, not the degree of a single polynomial. The spline model is linear in its fitted coefficients, even though its curve bends with temperature. Predict on the same temperature grid to compare the curves directly.

```{r seoul-splines}
seoul_line <- lm(rented_bike_count ~ mean_temperature, data = seoul)
seoul_bs <- lm(rented_bike_count ~ splines::bs(mean_temperature, df = 5), data = seoul)
seoul_ns <- lm(rented_bike_count ~ splines::ns(mean_temperature, df = 5), data = seoul)
temperature_grid <- tibble(mean_temperature = seq(
  min(seoul$mean_temperature), max(seoul$mean_temperature), length.out = 250
))
curve_fits <- temperature_grid |>
  mutate(
    Line = predict(seoul_line, newdata = temperature_grid),
    `B-spline` = predict(seoul_bs, newdata = temperature_grid),
    `Natural spline` = predict(seoul_ns, newdata = temperature_grid)
  ) |>
  pivot_longer(-mean_temperature, names_to = "model", values_to = "fitted")
ggplot(seoul, aes(mean_temperature, rented_bike_count)) +
  geom_point(alpha = 0.2) +
  geom_line(data = curve_fits,
            aes(y = fitted, colour = model), linewidth = 1) +
  labs(x = "Mean temperature (°C)", y = "Daily rentals", colour = "Model")
```

Plot the natural spline residuals against temperature to check for any remaining shape.

```{r seoul-spline-residuals}
seoul |>
  mutate(residual = residuals(seoul_ns)) |>
  ggplot(aes(mean_temperature, residual)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "Mean temperature (°C)", y = "Natural spline residual")
```

Discuss: Where does a line miss the shape in the points? How does the LOESS span change the curve? Compare the B-spline and natural spline near the observed temperature limits. A natural spline becomes linear beyond its boundary knots, but does that make extrapolation safe? Would a more flexible fit to these observations by itself justify choosing that model?

**Worked interpretation:** The low-span LOESS moves more with local fluctuations. A line misses the broad curved temperature relationship. The two spline fits can differ near the extremes, where data are thin. Natural spline tails are constrained to be linear outside the boundary knots; that constraint does not supply new information outside the observed range. The residual plot is a check for remaining shape, and in-sample fit alone is not a measure of future prediction quality.

## Can we separate trend from seasonality?

### Seoul: fit a trend, then seasonal pairs

Plot rentals against the actual date. Fit a linear `day_index` trend; then add one annual sine and cosine pair. The pair can express a seasonal wave at any phase, rather than fixing its peak to the start of the calendar. Add a second pair for a sharper within-year shape. Both terms in each pair have to enter together. Use `anova()` for a joint, two-degree-of-freedom comparison of the first- and second-pair models.

```{r seoul-fourier}
seoul |>
  ggplot(aes(date, rented_bike_count)) +
  geom_line() + labs(x = "Date", y = "Daily rentals")
seoul_time <- seoul |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )
seoul_trend <- lm(rented_bike_count ~ day_index, data = seoul_time)
seoul_first <- lm(rented_bike_count ~ day_index + S1 + C1, data = seoul_time)
seoul_second <- lm(rented_bike_count ~ day_index + S1 + C1 + S2 + C2,
                   data = seoul_time)
seoul_time |>
  mutate(Trend = fitted(seoul_trend), `First pair` = fitted(seoul_first),
         `Second pair` = fitted(seoul_second)) |>
  pivot_longer(c(Trend, `First pair`, `Second pair`),
               names_to = "model", values_to = "fitted_mean") |>
  ggplot(aes(date, fitted_mean, colour = model)) +
  geom_line() +
  labs(x = "Date", y = "Daily rentals", colour = "Model")
anova(seoul_first, seoul_second)
```

Discuss: What does the trend alone miss? Does the second pair change the fitted seasonal shape, and what does the partial F test say about its extra terms? We have only one year of Seoul data: what can we say about a pattern recurring next year?

**Worked interpretation:** The trend is a straight temporal change and cannot follow the observed within-year rise and fall. A complete sine/cosine pair allows the phase to be estimated. The second pair adds another frequency. For these 352 days, the partial comparison gives $F(2,346) = 21.79$, $p < 0.001$ under the usual independent-error assumptions; the two extra terms improve the fit within this sample. We will check those residual assumptions next. We have not observed whether the seasonal pattern recurs next year.

A factor for calendar season or month gives each category its own offset. Sine and cosine give a smooth within-year pattern. The terms answer different questions; neither makes a one-year record into evidence about future years.

## Wrapping up

We have changed the mean in two different ways. Splines let the relationship with a numerical predictor bend, while trend and sine/cosine terms let the expected response change systematically through time.

In one short paragraph, explain when you would use LOESS, a regression spline, and a trend plus Fourier terms. Then state what these mean-model choices still do **not** tell us about observations collected close together in time.

**Worked interpretation:** LOESS is useful for exploring nonlinear shape without committing to one global regression equation. A regression spline puts a flexible nonlinear effect directly into the regression model, with its flexibility controlled by the spline basis and degrees of freedom. Trend and Fourier terms describe systematic change through time and smooth periodic structure in the mean. None of these choices establishes that the remaining errors are independent. Observations close together in time may still have related residuals after the mean has been fitted.

That remaining question is the starting point for Lab 13: after fitting a reasonable mean, are the residuals still dependent?
