---
title: "Lab 10 solutions"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

Let's work through each idea first with wage data, then repeat the same analysis with Auto MPG.

The wage data contain 3,000 male workers recorded from 2003 to 2009. Each row is one worker. Wage is recorded in thousands of US dollars. The Auto MPG data contain 392 cars from model years 1970 to 1982. Each row is one car.

Both datasets are observational, so the fitted relationships describe associations rather than causal effects.

| Wage variable | Meaning |
|:--|:--|
| `wage` | recorded wage, thousands of US dollars |
| `age` | age, years |
| `education` | education level |
| `year` | year the wage was recorded |
| `jobclass` | industrial or information work |

| Auto MPG variable | Meaning |
|:--|:--|
| `mpg` | fuel economy, miles per gallon |
| `weight` | vehicle weight, pounds |
| `origin` | USA, Europe or Japan |
| `model_year` | model year, 70--82 |
| `horsepower` | engine power, horsepower |

Let's leave `car_name` out of the modelling.

```{r import-data, message=FALSE}
wage <- readr::read_csv("../data/wage.csv") |>
  mutate(education = factor(education))

auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(wage)
glimpse(auto)
```

## What changes across groups?

Let's start by plotting wage against age, with education shown by colour.

```{r wage-raw-plot}
ggplot(wage, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Discuss: How does wage vary with age? Do the education groups appear to have similar vertical positions and slopes?

Wage generally rises with age before flattening or falling later, and the education groups sit at different wage levels. The raw points also suggest that the age pattern may not be identical across education groups.

Let's fit three models.

```{r wage-fit-lines}
w1 <- lm(wage ~ age, data = wage)
w2 <- lm(wage ~ age + education, data = wage)
w3 <- lm(wage ~ age * education, data = wage)
```

These correspond to one common line, parallel lines, and separate lines.

```text
education * age = education + age + education:age
```

Let's plot the fitted values from each model.

```{r wage-one-line}
wage |>
  mutate(fitted = predict(w1)) |>
  arrange(age) |>
  ggplot(aes(x = age, y = wage)) +
  geom_point(aes(colour = education), alpha = 0.2) +
  geom_line(aes(y = fitted), colour = "black", linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

```{r wage-parallel-lines}
wage |>
  mutate(fitted = predict(w2)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

```{r wage-separate-lines}
wage |>
  mutate(fitted = predict(w3)) |>
  arrange(education, age) |>
  ggplot(aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.2) +
  geom_line(aes(y = fitted, group = education), linewidth = 1) +
  labs(x = "Age (years)", y = "Wage (US$1,000s)")
```

Discuss: What changes when `education` is added? What changes when `age:education` is added?

Adding education shifts the fitted lines vertically while keeping a common age slope. Adding the interaction also allows the age slopes to differ.

Now repeat the same steps for Auto MPG.

Plot `mpg` against `weight`, with `origin` shown by colour.

```{r auto-raw-plot}
ggplot(auto, aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Weight (lb)",
    y = "Fuel economy (miles per gallon)",
    colour = "Origin"
  )
```

Discuss: How does fuel economy vary with weight? Do the origin groups appear to have similar vertical positions and slopes?

Fuel economy falls as weight increases. The origins occupy different parts of the weight range, and the fitted relationship may differ in both position and slope.

Now fit:

```text
a1: mpg ~ weight
a2: mpg ~ weight + origin
a3: mpg ~ weight * origin
```

```{r auto-fit-lines}
a1 <- lm(mpg ~ weight, data = auto)
a2 <- lm(mpg ~ weight + origin, data = auto)
a3 <- lm(mpg ~ weight * origin, data = auto)
```

Make the same three fitted-line plots.

```{r auto-one-line}
auto |>
  mutate(fitted = predict(a1)) |>
  arrange(weight) |>
  ggplot(aes(x = weight, y = mpg)) +
  geom_point(aes(colour = origin), alpha = 0.3) +
  geom_line(aes(y = fitted), colour = "black", linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

```{r auto-parallel-lines}
auto |>
  mutate(fitted = predict(a2)) |>
  arrange(origin, weight) |>
  ggplot(aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fitted, group = origin), linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

```{r auto-separate-lines}
auto |>
  mutate(fitted = predict(a3)) |>
  arrange(origin, weight) |>
  ggplot(aes(x = weight, y = mpg, colour = origin)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(y = fitted, group = origin), linewidth = 1) +
  labs(x = "Weight (lb)", y = "Fuel economy (miles per gallon)")
```

Discuss: Which term moves the origin lines up and down? Which term allows the weight slopes to differ?

The `origin` term shifts the lines vertically. The `weight:origin` interaction allows the slopes to differ.

Let's look at the separate-lines wage model once in the usual coefficient table.

```{r wage-read-interaction}
summary(w3)
```

The `age` coefficient is the age slope for the reference education group. The education coefficients change the intercept, and the age-by-education coefficients change the slope. Age zero is outside these data, so the intercept differences are not useful overall wage comparisons.

Now run `summary(a3)`.

```{r auto-read-interaction}
summary(a3)
```

Discuss: Which coefficient is the USA weight slope? Which coefficients change the intercepts for Europe and Japan? Which coefficients change their slopes? Why are the origin main-effect coefficients not overall origin differences?

USA is the reference origin, so `weight` is the USA slope. `originEurope` and `originJapan` change the intercepts, while `weight:originEurope` and `weight:originJapan` change the slopes. Because the interaction is present, the main-effect origin coefficients describe differences at zero pounds, which is outside the data.

## Do the groups improve the model?

A partial F test compares a reduced model with a larger model containing all of the reduced model's terms. It tests the added terms together.

Let's ask whether education adds information after age.

```{r compare-wage-education}
anova(w1, w2)
```

Discuss: How many parameters were added? What does the P-value say about moving from one common line to parallel education lines?

Four education parameters are added. The P-value is below 0.001, so there is evidence that education improves the model after age is already included.

Now ask the same question for the cars: after accounting for weight, does origin improve the model?

```{r compare-auto-origin}
anova(a1, a2)
```

Discuss: How many parameters were added? What does the P-value say about moving from one common line to parallel origin lines?

Two origin contrasts are added. The P-value is about 0.0024, so there is evidence that origin improves the model after weight is included.

Next, let's ask whether the slopes differ. For the wage data:

```{r compare-wage-slopes}
anova(w2, w3)
```

Discuss: Which terms were added? What does the P-value say about whether the age slope varies across education groups?

Four interaction coefficients are added. The P-value is about 0.013, giving evidence that at least some age slopes differ by education.

Now make the same comparison for Auto MPG.

```{r compare-auto-slopes}
anova(a2, a3)
```

Discuss: Which terms were added? What does the P-value say about whether the weight slope varies across origins?

Two interaction coefficients are added. The P-value is about 0.044, giving modest evidence that the weight slopes differ among origins.

Finally, compare the simplest and fullest models directly.

```{r compare-full-lines}
anova(w1, w3)
anova(a1, a3)
```

Discuss: How many parameters are added in each comparison? What does each joint test establish?

The wage comparison adds eight parameters and strongly favours the separate-lines model. The Auto MPG comparison adds four parameters and also favours separate lines over one common line, with P about 0.0011. These are joint tests and do not say that every added coefficient is individually useful.

There is one complication worth seeing before moving on. A sequential ANOVA table gives each term credit for variation it explains when it enters the model. If predictors overlap, term order matters.

```{r wage-sequential-anova}
anova(lm(wage ~ education + age, data = wage))
anova(lm(wage ~ age + education, data = wage))
```

Discuss: The fitted model is identical in both cases. Why do the sums of squares change?

Education and age share some information. The first term gets credit for the shared part before the second term is added.

Now repeat this with `origin` and `weight`.

```{r auto-sequential-anova}
anova(lm(mpg ~ origin + weight, data = auto))
anova(lm(mpg ~ weight + origin, data = auto))
```

Discuss: Why is `anova(a1, a2)` the clearer answer to the question "does origin add information after weight?"

The direct comparison writes the question into the models: `a1` already contains weight, and `a2` adds origin. The answer therefore does not depend on whichever term happened to appear first in a sequential table.

## When does another predictor earn its place?

A smaller RSS is guaranteed when another predictor is added, so RSS alone cannot decide whether a larger model is worth keeping.

Let's use adjusted R-squared, residual standard error and AIC alongside the partial F test. `broom::glance()` puts these model summaries into one tidy row.

Add year to the wage model.

```{r wage-add-year}
w4 <- lm(wage ~ age * education + year, data = wage)

bind_rows(
  w3 = broom::glance(w3),
  w4 = broom::glance(w4),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w3, w4)
```

Discuss: Do the four pieces of evidence all favour adding year? How large is the improvement?

All four favour `w4`, although the changes in the fit measures are modest. Adjusted R-squared rises from about 0.260 to 0.263, residual standard error falls slightly, AIC falls by about 9.5, and the partial F test gives P about 0.0007.

Now add `model_year` to the Auto MPG model and call it `a4`.

```{r auto-add-model-year}
a4 <- lm(mpg ~ weight * origin + model_year, data = auto)

bind_rows(
  a3 = broom::glance(a3),
  a4 = broom::glance(a4),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a3, a4)
```

Discuss: Does model year earn its place?

Yes. Adjusted R-squared rises from about 0.703 to 0.830, residual standard error falls from about 4.25 to 3.22 MPG, AIC falls by more than 200, and the partial F test gives P below 0.001.

Let's make one more addition to each model.

For wage, add `jobclass`.

```{r wage-add-jobclass}
w5 <- lm(wage ~ age * education + year + jobclass, data = wage)

bind_rows(
  w4 = broom::glance(w4),
  w5 = broom::glance(w5),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w4, w5)
```

Discuss: The improvement is much smaller than it was for year. Is `jobclass` still worth keeping in this descriptive model?

The changes are small but point in the same direction. Adjusted R-squared rises slightly, residual standard error falls slightly, AIC falls by about 8.8, and the partial F test gives P about 0.001. Keeping job class is reasonable within this candidate set.

For Auto MPG, add `horsepower` to `a4` and call the model `a5`.

```{r auto-add-horsepower}
a5 <- lm(mpg ~ weight * origin + model_year + horsepower, data = auto)

bind_rows(
  a4 = broom::glance(a4),
  a5 = broom::glance(a5),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a4, a5)
```

Discuss: Do adjusted R-squared, residual standard error, AIC and the partial F test tell the same story? Which model, `a4` or `a5`, is the better compact descriptive model?

They all point towards `a4`. Adding horsepower slightly reduces adjusted R-squared, barely changes residual standard error, raises AIC, and gives a partial F-test P-value of about 0.38.

## What if prediction is the goal?

Everything so far has been calculated on the same observations used to fit the models.

Discuss: What could be done to assess prediction for cars that were not used to fit the model? Why are adjusted R-squared, residual standard error and AIC not direct estimates of that prediction error?

Use validation data or cross-validation so predictions are assessed on observations that did not fit the model. The in-sample measures above are not direct estimates of prediction error for new cars.

## Does the working model fit the pattern?

Let's check the wage model before treating it as finished.

Run the standard diagnostic plots for `w5`.

```{r wage-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w5)
par(mfrow = c(1, 1))
```

Now plot the residuals against age.

```{r wage-residuals-by-age}
wage |>
  mutate(residual = residuals(w5)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: What pattern should be visible if the straight-line age effect is adequate? What pattern is actually present?

The residuals should have no systematic relationship with age. Instead, there is a broad curve, so the straight-line age term is missing structure in the mean.

Quadratic regression was introduced earlier in the course, so let's add `age^2` while keeping the linear age term.

```{r wage-revise-mean}
w6 <- lm(
  wage ~ age * education + year + jobclass + I(age^2),
  data = wage
)

bind_rows(
  w5 = broom::glance(w5),
  w6 = broom::glance(w6),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(w5, w6)
```

Repeat the diagnostic plots and the residual-versus-age plot for `w6`.

```{r wage-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w6)
par(mfrow = c(1, 1))

wage |>
  mutate(residual = residuals(w6)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

Discuss: Has the broad age pattern been reduced? What concerns remain?

The broad age curve is reduced, although other departures remain. The plots do not establish independence.

Now make the same diagnostic check for `a4`.

Run the standard diagnostic plots, then plot the residuals against `weight`.

```{r auto-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(a4)
par(mfrow = c(1, 1))

auto |>
  mutate(residual = residuals(a4)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

Discuss: Is there a systematic pattern left in the mean?

There is clear curvature in the residuals against weight, so the mean structure still needs work.

Now fit:

```text
a6: mpg ~ weight * origin + I(weight^2) + model_year
```

```{r auto-revise-mean}
a6 <- lm(
  mpg ~ weight * origin + I(weight^2) + model_year,
  data = auto
)

bind_rows(
  a4 = broom::glance(a4),
  a6 = broom::glance(a6),
  .id = "model"
) |>
  select(model, adj.r.squared, sigma, AIC)

anova(a4, a6)

par(mfrow = c(2, 2))
plot(a6)
par(mfrow = c(1, 1))

auto |>
  mutate(residual = residuals(a6)) |>
  ggplot(aes(x = weight, y = residual)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  geom_hline(yintercept = 0) +
  labs(x = "Weight (lb)", y = "Residual (mpg)")
```

Discuss: What improved? What still needs attention?

The broad U-shaped pattern is substantially reduced. Some tail and spread concerns remain. The standard diagnostic plots do not establish independence; that has to be considered from the way the observations were collected and whether groups of observations might share unmodelled structure.

## Pulling it together

A reasonable Auto MPG conclusion is:

Fuel economy decreases as weight increases, and allowing origin to change the fitted lines improves on one common line. There is modest evidence that the weight slopes differ among origins. Adding model year produces a large improvement, while horsepower adds little after weight, origin and model year are already included, so `a4` is a reasonable compact model before diagnostics. The residual plot for `a4` shows clear curvature, and adding `weight^2` improves the mean structure. The revised model still has tail and spread concerns, so the diagnostics do not establish that it is fully adequate. Finally, these are in-sample comparisons; validation or cross-validation is needed to assess prediction for new cars.

With a small candidate set, these decisions can still be made by hand. Once the predictor list gets longer, that becomes tedious very quickly. Lab 11 returns to Auto MPG and looks at automated search and shrinkage.
