---
title: "Lab 10"
subtitle: "161.251 Regression Modelling"
author: "Nick Knowlton nknowlton@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", dev = "png")
library(tidyverse)
```

# Computer Laboratory: Building and Comparing Linear Models

I will demonstrate each modelling step using wage data. When you reach a **Your turn** section, repeat that step using Auto MPG before moving on. The two datasets have different variables and different answers, but the modelling questions are the same.

The wage data contain 3,000 male workers, recorded from 2003 to 2009. Each row is one worker. Wage is recorded in thousands of US dollars. The Auto MPG data contain 392 cars from model years 1970 to 1982. Each row is one car. Both datasets are observational, so the fitted relationships describe associations, not causal effects.

| Wage variable | Meaning |
|:--|:--|
| `wage` | recorded wage, thousands of US dollars |
| `age` | worker's age, in years |
| `education` | education level, from less than high school to advanced degree |
| `year` | year the wage was recorded, 2003--2009 |
| `jobclass` | industrial or information work |

| Auto MPG variable | Meaning |
|:--|:--|
| `mpg` | fuel economy, miles per gallon |
| `weight` | vehicle weight, pounds |
| `origin` | USA, Europe or Japan |
| `model_year` | vehicle model year, 70--82 |
| `horsepower` | engine power, horsepower |

We will not use `car_name` as a predictor.

By the end of the lab, you should be able to read factor-by-covariate models as fitted lines, test nested models, compare a small set of candidate models, and use residual plots to identify a missing feature of the mean structure.

```{r import-data, message=FALSE}
wage <- readr::read_csv("../data/wage.csv") |>
  mutate(education = factor(education))

auto <- readr::read_csv("../data/auto_mpg.csv") |>
  mutate(origin = factor(origin, levels = c("USA", "Europe", "Japan")))

glimpse(wage)
glimpse(auto)
```

## What changes across groups?

### Instructor demonstration: education and age

Start with the wage data. Does the relationship between age and wage look similar at every education level?

```{r wage-raw-plot}
ggplot(wage, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.25) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Fit three models. The first has one line for everyone, the second allows education groups to have different intercepts but a common age slope, and the third allows both intercepts and slopes to differ.

```{r wage-fit-lines}
w1 <- lm(wage ~ age, data = wage)
w2 <- lm(wage ~ age + education, data = wage)
w3 <- lm(wage ~ age * education, data = wage)
```

```{.text}
education * age = education + age + education:age
```

The interaction lets the age slope vary with education. The next plot shows the fitted lines for all three models. Lines for each education group stop at that group's observed age range.

```{r wage-model-geometry}
age_values <- seq(min(wage$age), max(wage$age), length.out = 100)

education_grid <- wage |>
  group_by(education) |>
  summarise(
    age = list(seq(min(age), max(age), length.out = 100)),
    .groups = "drop"
  ) |>
  unnest(age)

common_lines <- tibble(age = age_values) |>
  mutate(
    fitted = predict(w1, newdata = tibble(age = age_values)),
    education = factor(NA_character_, levels = levels(wage$education)),
    model = "One common line"
  )

parallel_lines <- education_grid |>
  mutate(
    fitted = predict(w2, newdata = education_grid),
    model = "Parallel lines"
  )

separate_lines <- education_grid |>
  mutate(
    fitted = predict(w3, newdata = education_grid),
    model = "Separate lines"
  )

model_levels <- c("One common line", "Parallel lines", "Separate lines")
line_data <- bind_rows(common_lines, parallel_lines, separate_lines) |>
  mutate(model = factor(model, levels = model_levels))

point_data <- crossing(
  wage,
  model = factor(model_levels, levels = model_levels)
)

ggplot(point_data, aes(x = age, y = wage, colour = education)) +
  geom_point(alpha = 0.18) +
  geom_line(
    data = filter(line_data, model != "One common line"),
    aes(y = fitted, group = interaction(education, model))
  ) +
  geom_line(
    data = filter(line_data, model == "One common line"),
    aes(x = age, y = fitted),
    inherit.aes = FALSE,
    colour = "black"
  ) +
  facet_wrap(~ model) +
  labs(
    x = "Age (years)",
    y = "Wage (US$1,000s)",
    colour = "Education"
  )
```

Read the coefficient table for the separate-lines model.

```{r wage-read-interaction}
summary(w3)
```

**What to look for:** the `age` coefficient is the slope for the reference education group. The education coefficients shift the intercepts, and the age-by-education coefficients adjust the slopes.

<details>
<summary>Worked interpretation</summary>

The reference group is workers with less than a high-school education. Its age slope is the `age` coefficient. The education coefficients give intercept differences at age zero, and the interaction coefficients give slope differences from the reference group. Age zero is outside the data, so those intercept differences are not useful overall wage comparisons.

</details>

### Your turn: origin and vehicle weight

Now repeat the same steps for Auto MPG.

1. Plot `mpg` against `weight`, using `origin` for colour.
2. Fit `a1`, `a2` and `a3` with the formulas below.
3. Draw the one-line, parallel-lines and separate-lines fits. Use the wage geometry code as your template and keep each origin's fitted lines within its observed weight range.
4. Run `summary(a3)`. Identify the USA weight slope, the Europe and Japan intercept adjustments, and the slope adjustments. Explain why the intercept adjustments are not overall origin differences.

```{.text}
a1: mpg ~ weight
a2: mpg ~ weight + origin
a3: mpg ~ weight * origin
```

```{r auto-plot-and-models}
# Plot mpg against weight, coloured by origin.
# Fit a1, a2 and a3 using the formulas above.
# Use wage-model-geometry as your template for the three fitted-line panels.
```

```{r auto-read-interaction}
# Run summary(a3), then answer questions 4 above.
```

## Do the groups improve the model?

A partial F test compares a reduced model with a larger model that contains all the reduced model's terms. It tests the added terms together, not one coefficient at a time.

### Instructor demonstration: does education add information after age?

The question is whether education improves the model after age is already included. Compare `w1` with `w2`.

```{r compare-wage-education}
anova(w1, w2)
```

The test adds four education parameters. The P-value is below 0.001, so the parallel-lines model improves on one common line. This is evidence that at least one education group differs after age is accounted for.

### Your turn: does origin add information after weight?

Compare `a1` with `a2`. State how many parameters are added, report the P-value, and describe what changes in the fitted lines.

```{r compare-auto-origin}
# Use compare-wage-education as your template.
# Compare a1 (reduced) with a2 (full) using anova().
```

### Instructor demonstration: do the age slopes differ?

Compare `w2` with `w3`. The question is whether the age slope varies across education groups.

```{r compare-wage-slopes}
anova(w2, w3)
```

The test adds four age-by-education interaction coefficients. The P-value is about 0.013, which is evidence that at least some age slopes differ by education.

### Your turn: do the weight slopes differ?

Compare `a2` with `a3`. Report the added terms and the P-value. What does the test say about the weight slopes across origins?

```{r compare-auto-slopes}
# Use compare-wage-slopes as your template.
# Compare a2 (reduced) with a3 (full) using anova().
```

### Compare the simplest and fullest models

The earlier tests changed one part of the model at a time. Now compare `w1` directly with `w3`, then make the same comparison for Auto MPG. State the number of added parameters and interpret the joint test.

```{r compare-full-lines}
# Compare w1 with w3 using anova().
# Then compare a1 with a3.
```

### Why does sequential ANOVA depend on order?

A sequential ANOVA table gives each term credit for the variation it explains when it enters the model. When predictors overlap, changing the order can change the sums of squares.

```{r wage-sequential-anova}
anova(lm(wage ~ education + age, data = wage))
anova(lm(wage ~ age + education, data = wage))
```

Repeat this check for `mpg ~ origin + weight` and `mpg ~ weight + origin`. Why is comparing `a1` with `a2` a clearer answer to “does origin add information after weight?”

```{r auto-sequential-anova}
# Fit the two models in the two term orders, then run anova() for each.
```

## When does another predictor earn its place?

Adjusted R-squared, residual standard error and AIC compare fit while accounting for model complexity in different ways. Larger adjusted R-squared and smaller residual standard error or AIC are preferred when the models use the same response and observations. A partial F test answers a specific nested-model question.

### Instructor demonstration: add year

Add the year the wage was recorded to `w3`, and call the result `w4`.

```{r wage-add-year}
w4 <- lm(wage ~ age * education + year, data = wage)

tibble(
  model = c("w3", "w4"),
  adjusted_R2 = c(summary(w3)$adj.r.squared, summary(w4)$adj.r.squared),
  residual_SE = c(sigma(w3), sigma(w4)),
  AIC = c(AIC(w3), AIC(w4))
)

anova(w3, w4)
```

The P-value is about 0.0007. Check whether the three fit measures also favour `w4`, and explain what year adds after age and education are already in the model.

### Your turn: add model year

Fit `a4` by adding `model_year` to `a3`. Recreate the same comparison table for `a3` and `a4`, then run `anova(a3, a4)`. State which measures improve and what the partial F test says.

```{r auto-add-model-year}
# Use wage-add-year as your template.
# Fit a4: mpg ~ weight * origin + model_year.
# Compare a3 with a4 using the same three measures and anova().
```

### Instructor demonstration: add job class

Add `jobclass` to `w4` and call the result `w5`.

```{r wage-add-jobclass}
w5 <- lm(wage ~ age * education + year + jobclass, data = wage)

tibble(
  model = c("w4", "w5"),
  adjusted_R2 = c(summary(w4)$adj.r.squared, summary(w5)$adj.r.squared),
  residual_SE = c(sigma(w4), sigma(w5)),
  AIC = c(AIC(w4), AIC(w5))
)

anova(w4, w5)
```

The partial F test gives a P-value of about 0.001. Decide whether the small changes in the other measures are enough to keep `jobclass` in this descriptive model.

### Your turn: add horsepower

Fit `a5` by adding `horsepower` to `a4`. Recreate the same comparison table and run `anova(a4, a5)`. Do the four measures tell a consistent story?

```{r auto-add-horsepower}
# Use wage-add-jobclass as your template.
# Fit a5: mpg ~ weight * origin + model_year + horsepower.
# Compare a4 with a5 using adjusted R-squared, residual standard error,
# AIC and anova().
```

Choose between `a4` and `a5` as a compact descriptive model. Support your choice with at least three pieces of evidence. Do not conclude that an excluded predictor has no association in every possible model.

## What if prediction is the goal?

The measures above are calculated on the observations used to fit each model. They are not direct estimates of prediction error for new observations.

For the Auto MPG analysis, what could you do to assess how well a model predicts cars that were not used to fit it? Why would the answer to that question matter if prediction, rather than describing these data, were the goal?

## Does the working model fit the pattern?

Use the standard diagnostic plots for `w5`. For each plot, describe what you see and explain why it matters. Also plot residuals against age: that makes it easier to see whether a straight-line age effect leaves a pattern in the mean.

### Instructor demonstration: diagnose and revise the wage model

```{r wage-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w5)
par(mfrow = c(1, 1))
```

```{r wage-residuals-by-age}
wage |>
  mutate(residual = residuals(w5)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

The residual pattern suggests that the straight-line age effect needs more work. Add a quadratic age term to `w5`, giving `w6`. In an R formula, write the squared term as `I(age^2)`.

```{r wage-revise-mean}
w6 <- lm(
  wage ~ age * education + year + jobclass + I(age^2),
  data = wage
)

anova(w5, w6)

tibble(
  model = c("w5", "w6"),
  adjusted_R2 = c(summary(w5)$adj.r.squared, summary(w6)$adj.r.squared),
  residual_SE = c(sigma(w5), sigma(w6)),
  AIC = c(AIC(w5), AIC(w6))
)
```

Now repeat both diagnostic checks for `w6`.

```{r wage-revised-diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(w6)
par(mfrow = c(1, 1))
```

```{r wage-revised-residuals-by-age}
wage |>
  mutate(residual = residuals(w6)) |>
  ggplot(aes(x = age, y = residual)) +
  geom_point(alpha = 0.25) +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  labs(x = "Age (years)", y = "Residual (US$1,000s)")
```

<details>
<summary>Worked interpretation</summary>

The broad age curve is reduced, though the diagnostic plots still show departures that deserve attention. These plots do not establish independence.

</details>

### Your turn: diagnose and revise the Auto MPG model

Use the same checks on your chosen model, `a4`. Plot the standard four diagnostic panels and plot residuals against `weight`.

```{r auto-diagnostics, fig.width=9, fig.height=7}
# Plot the standard diagnostic panels for a4.
# Use wage-diagnostics as your template.
```

Plot residuals against weight as well. What pattern do you see?

```{r auto-residuals-by-weight}
# Use wage-residuals-by-age as your template, replacing age with weight.
```

Now fit `a6`:

```{.text}
mpg ~ weight * origin + I(weight^2) + model_year
```

Compare `a4` and `a6` with a partial F test and the same three fit measures.

```{r auto-revise-mean}
# Fit a6 using the formula above.
# Compare a4 with a6 using anova(), adjusted R-squared,
# residual standard error and AIC.
```

Repeat the standard diagnostic plots and residual-versus-weight plot for `a6`. State what improved and what still needs attention.

```{r auto-revised-diagnostics, fig.width=9, fig.height=7}
# Repeat auto-diagnostics and auto-residuals-by-weight for a6.
```

We cannot diagnose independence from these plots alone. Consider how the observations were collected and whether groups of observations might share unmodelled structure. Do not treat a diagnostic plot or an influential-point label as an automatic reason to delete a car.

## Independent synthesis

Write a short conclusion for Auto MPG. Include:

1. what the common-line, parallel-lines and separate-lines models allowed to change;
2. what the partial F tests said about origin;
3. whether you kept horsepower after `model_year`, and why;
4. what the diagnostics led you to change;
5. why the in-sample comparisons do not establish prediction performance for new cars.

Keep your conclusion to about six sentences. Lab 11 returns to Auto MPG and considers a larger candidate set.
