---
title: Massey University --- School of Mathematical and Computational Sciences
subtitle: 161.251 Regression Modelling
author: "Staff member responsible for this workshop: Jonathan Marshall j.c.marshall@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
output:
    html_document:
      code_download: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(comment="", dev="png")
library(tidyverse)
library(broom)
```

  

## Computer Laboratory Tutorial 11 B

Today we are looking at summarised data on magazine reading by `Age` group and `Sex` (1 = females, 0 = males).

```{r, message=FALSE}
mags <- read_csv("https://www.massey.ac.nz/~jcmarsha/161251/data/AveMags.csv")
mags
```

1. We'll start by doing a plot of `Avemags` by `Age` and `Sex`. You should notice that there is initially an increasing trend, followed by a flat period and then a decreasing trend.
    
    ```{r}
    mags |> ggplot(aes(x=Age, y=Avemags, col=as_factor(Sex))) +
      geom_point()
    ```

    Eyeballing the chart, we see there is a change in slope around 20 and 50. We have the `Adult` variable available for the change at 20, but we'll need another one for 50+
    
    ```{r}
    mags_change <- mags |> mutate(Older = if_else(Age >= 50, 1, 0))
    ```

2. Now we'll fit a linear regression for `Avemags` by `Age` and `Sex`, allowing the trend with `Age` to differ based on `Adult` or `Older`.

    ```{r}
    mod1 <- lm(Avemags ~ Age + Sex + Age*Adult + Age*Older, data=mags_change)
    summary(mod1)
    ```

    *It seems there is a strong effect of `Age` and `Sex` and evidence for an interaction with of `Adult` with `Age`. In the teenage years magazine reading increases, and then drops and is flat in Adult, and then declines for Older adults.*
    
    *It's a little unclear if the effect of `Older` needs to be considered. It is not significant in the summary, but is in an anova - this is because of the order of considering things and the fact that it's all additive on top of the teenage slope. A comparison of a model without `Older` suggests it may not be needed:*
    
    ```{r}
    mod1b <- lm(Avemags ~ Age + Sex + Age*Adult, data=mags_change)
    anova(mod1b, mod1)
    ```

3. We'll visualise the model in part 2 by using `broom::augment()` to add `.fitted` columns to the data.

    ```{r}
    augment(mod1) |>
      ggplot(mapping = aes(x=Age, y=Avemags, col=as_factor(Sex))) +
      geom_point() +
      geom_line(aes(y=.fitted))
    ```

    Notice that the lines are parallel. This is something that we forced on the data.
    
    *This was forced on the data as we hvae no interaction terms between `Sex` and `Age` - that is, the effect of `Sex` can only be additive on top of whatever trend we use for `Age`.*
    
4. Let's see if you can get an alternate model formulated.
    
    ```{r}
    mod2 <- lm(Avemags ~ Age*Sex + Age*Adult*Sex + Age*Older*Sex, data=mags_change)
    summary(mod2)
    ```
    ```{r}
    anova(mod1, mod2)
    ```
    
    *There is little evidence from the anova test that the second model is better than the first.*
    
    ```{r}
    augment(mod2) |>
      ggplot(mapping = aes(x=Age, y=Avemags, col=as_factor(Sex))) +
      geom_point() +
      geom_line(aes(y=.fitted))
    ```

    - This shows the difference in slopes is most marked in the teenage years, where teenage females appear to increase their readership of magazines at a faster rate than teenage boys.
    
5. Produce diagnostic plots for your model in part 3 (i.e. not the one with the `Sex` interaction).

    ```{r}
    par(mfrow=c(2,2))
    plot(mod1)
    ```
    
    *The diagnostics largely look fine. There is arguably some increasing variance with fitted value (presumably from teenagers) but it's not too concerning and everything else looks fine.*

### Weighted regression

6. Each of the observations in our dataset are *averages* across a bunch of individuals. So each point is not worth the same "weight": They are different both in terms of the number of individuals contributing to each observation AND the precision (standard deviations) of the observations. The following plot visualises the data with number of individuals

    ```{r}
    mags_change |>
      ggplot(mapping = aes(x=Age, y=Avemags, size=N, col=StDev, shape=as_factor(Sex))) +
      geom_point()
    ```
    
    *We have the most evidence of magazine use among females generally (larger triangles?) and among older readers. There is the lowest amount of information from teenagers. In particular, we have a lot of variation from teenage boys around age 16.*

7. Redo `mod1` and `mod2` using `weights = N/StDev^2`, and produce the same plots and comparisons.

    ```{r}
    wt1 <- lm(Avemags ~ Age + Sex + Age*Adult + Age*Older, data=mags_change,
              weights = N/StDev^2)
    wt2 <- lm(Avemags ~ Age*Sex + Age*Adult*Sex + Age*Older*Sex, data=mags_change,
              weights = N/StDev^2)
    anova(wt1, wt2)
    augment(wt2) |>
      ggplot(mapping = aes(x=Age, y=Avemags, col=as_factor(Sex))) +
      geom_point(mapping = aes(size=`(weights)`)) +
      geom_line(aes(y=.fitted))
    ```

    *There is evidence now that the model with interaction between `Sex` and `Age` is more useful than the one without. This is largely driven by differences in teenagers, with boys reading fewer magazines than girls as they get older.*

8. Look in particular at the unweighted residuals from `mod2` and the weighted residuals from the corresponding model in part 7. Plot them versus `Age`, coloured by `Sex`.

    ```{r}
    mod2 |> augment() |>
      ggplot() +
      geom_point(aes(x=Age, y=.resid, col=as_factor(Sex))) +
      labs(y="residuals")
    wt2 |> augment() |>
      ggplot() +
      geom_point(aes(x=Age, y=.resid*sqrt(`(weights)`), col=as_factor(Sex))) +
      labs(y="weighed residuals")
    ```
    
    *We can see that the large spread in residuals for teenagers in `mod2` is greatly reduced in `wt2` as we just don't have much information for those ages compared to older age groups.*

9. What is your conclusion about the magazine reading habits of teenagers, adults, and older adults?

    *Females tend to read more magazine than males by around 0.5 - 1 magazine on average. The difference is most marked in teenagers, where teenage girls read more magazines as they get older, whereas males increase but only slightly. The magazine reading of both sexes reduces dramatically and roughly flat lines during adulthood, tailing off as they get older again.*
    