---
title: Massey University --- School of Mathematical and Computational Sciences
subtitle: 161.251 Regression Modelling
author: "Staff member responsible for this workshop: Jonathan Marshall j.c.marshall@massey.ac.nz"
date: "last updated `r format(lubridate::today(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment="", dev="png")
library(tidyverse)
```

# Computer Laboratory  Exercise 10A

The purpose of this exercise is to gain some familiarity with the `step()` function for stepwise regression. You might first like to look back at Lab 5A to where we did stepwise regression by hand.

## Automatic Variable Selection for Swiss Fertility Data

```{r}
data(swiss)
head(swiss)
```

1. In lab 5A you fitted the model 
    
    ```{r}
    swiss.lm.full= lm(Fertility ~ .,  data=swiss)
    summary(swiss.lm.full)
    ```
    
    Now we'll use the `step()` function to perform backwards model selection using the AIC criterion. We should get the same result as in Lab 5A.
    
    ```{r}
    swiss.lm.1 = lm(Fertility~1, data=swiss)
    swiss.step.back = step(swiss.lm.full, direction="backward")
    summary(swiss.step.back)
    ```
    
    Note that here we didn't bother specifying the `scope` argument, as we're going backwards only from a full model: the lowest model to consider will be the intercept only model.
    
    - Check back to your Lab 5A that you end up with the same model.
    - Is this what you expect? Could it have been different?

2.  It is of interest to know if we end up with the same model going forward. This time the `scope` argument should be the largest model (e.g. use `scope = formula(swiss.lm.full)`). Try this yourself.

    - In which order do the variables enter the model?
    - Do you get the same model as in backward selection?

3.  Suppose that we start with the variable `Examination` in the model, and consider all models up to the full model. We can do that with:

    ```{r}
    swiss.lm.e = lm(Fertility ~ Examination, data=swiss)
    summary(swiss.lm.e)
    
    scope <- list(lower=formula(swiss.lm.e), upper=formula(swiss.lm.full))
    swiss.step.for2 = step( swiss.lm.e, scope=scope, direction="both")
    summary(swiss.step.for2)
    ```

    - You should notice that Examination now stays in the model. Why is that?

4. Redo the above stepwise regression starting with `swiss.lm.e` but this time allow the scope to include an intercept only model. What happens?

5.  Suppose we felt the data on `Education` was unreliable and we exclude it from the model. Setup a step-wise regression that considers all models from the intercept only model up to the largest model excluding `Education`.

    - Adjust the `scope` argument so that `Education` isn't considered.
    - Do all the other variables enter in?

