---
title: "Lecture 17 - Orthogonal Polynomials"
author: "Sergio Sclovich"
format: 
  revealjs:
    embed-resources: true
    slide-number: true
    incremental: false
    highlight-style: zenburn
    highlight-lines: true
    ratio: "4:3"
    progress: true
    slideNumberFormat: "%current%/%total%"
    theme: moon
    css: "../format.css"
execute:
  output: true
  eval: true
  code-fold: true   # enable code folding
editor: 
  markdown: 
    wrap: 72
---

```{r setup, echo=FALSE}
library(broom)
library(knitr)
library(ggplot2); theme_set(theme_bw(base_size=15))
library(patchwork)
library(tidyverse)
#library(car)

# Ensure the plot options are set correctly
#opts_chunk$set(dev.args=list(bg='transparent'), comment="", warning=FALSE, echo=FALSE)
opts_chunk$set(comment="", warning=FALSE, echo=FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.dim=c(8,5), out.width="70%", fig.retina=2)

```


## learning outcomes

-- Understand the concept of Orthogonality.

-- Implement Orthogonal Polynomial models in r.


##  Dependencies between estimates of individual regression coefficients

$$
\mathrm{Var}(\hat{\boldsymbol{\beta}}) =
\begin{bmatrix}
\mathrm{Var}(\hat{\beta}_0) & \mathrm{Cov}(\hat{\beta}_0, \hat{\beta}_1) & \cdots & \mathrm{Cov}(\hat{\beta}_0, \hat{\beta}_p) \\
\mathrm{Cov}(\hat{\beta}_1, \hat{\beta}_0) & \mathrm{Var}(\hat{\beta}_1) & \cdots & \mathrm{Cov}(\hat{\beta}_1, \hat{\beta}_p) \\
\vdots & \vdots & \ddots & \vdots \\
\mathrm{Cov}(\hat{\beta}_p, \hat{\beta}_0) & \mathrm{Cov}(\hat{\beta}_p, \hat{\beta}_1) & \cdots & \mathrm{Var}(\hat{\beta}_p)
\end{bmatrix}
= \sigma^2 (X^\top X)^{-1}
$$



##  Dependencies between estimates of individual regression coefficients

$$
\mathrm{Var}(\hat{\boldsymbol{\beta}}) =
\sigma^2 (X^\top X)^{-1}
$$
Where $X$ is the design matrix and $\sigma^2$ is the estimated standard error of the residuals.

following the definition of the correlation:
$$cor(X, Y) = \frac{cov(X,Y)}{\sigma_X\sigma_Y}.$$

##  Dependencies between estimates of individual regression coefficients

<!-- $$ -->
<!-- \mathrm{Corr}(\hat{\boldsymbol{\beta}}) = -->
<!-- \begin{bmatrix} -->
<!-- 1 & \mathrm{Corr}(\hat{\beta}_0, \hat{\beta}_1) & \cdots & \mathrm{Corr}(\hat{\beta}_0, \hat{\beta}_p) \\ -->
<!-- \mathrm{Corr}(\hat{\beta}_1, \hat{\beta}_0) & 1 & \cdots & \mathrm{Corr}(\hat{\beta}_1, \hat{\beta}_p) \\ -->
<!-- \vdots & \vdots & \ddots & \vdots \\ -->
<!-- \mathrm{Corr}(\hat{\beta}_p, \hat{\beta}_0) & \mathrm{Corr}(\hat{\beta}_p, \hat{\beta}_1) & \cdots & 1 -->
<!-- \end{bmatrix} -->
<!-- $$ -->
$$
\small
\mathrm{Corr}(\hat{\boldsymbol{\beta}}) =
\begin{bmatrix}
1 & \mathrm{Corr}(\hat{\beta}_0, \hat{\beta}_1) & \cdots & \mathrm{Corr}(\hat{\beta}_0, \hat{\beta}_p) \\
\mathrm{Corr}(\hat{\beta}_1, \hat{\beta}_0) & 1 & \cdots & \mathrm{Corr}(\hat{\beta}_1, \hat{\beta}_p) \\
\vdots & \vdots & \ddots & \vdots \\
\mathrm{Corr}(\hat{\beta}_p, \hat{\beta}_0) & \mathrm{Corr}(\hat{\beta}_p, \hat{\beta}_1) & \cdots & 1
\end{bmatrix}
$$




## Crowds for the Brisbane Lions example


This Data has the number of members and average crowds sizes for the
    Brisbane Lions AFL club from 1993 to 2003.


    
|  Variables | Description                    |
| :- | :---------------------- |
|  Year | Calendar year       |
| Members | Number of Lions members |
|  Crowd | Average home crowd size |
    

The aim of this Example is to regress `Crowd` on `Year`.

## Crowds for the Brisbane Lions example

```{r getLionsData, echo=0, eval=-2}
Lions <- read.csv(file="lions.csv", header=TRUE)
#str(Lions)
#summary(Lions) |> kable()
plot(Crowd~Year, data=Lions)
```


## Crowds for the Brisbane Lions example

```{r Lions.lm1}
Lions.lm1 <- lm(Crowd~Year, data=Lions) 
summary(Lions.lm1)
```

As expected from the scatter plot, there is evidence the
Lion's home crowds are increasing through time.

## Crowds for the Brisbane Lions example

What does the intercept mean?

```{r vcovLions.lm1}
vcov(Lions.lm1)
cov2cor(vcov(Lions.lm1)
)
```

Dramatic negative correlation between $\hat{\beta_0}$ and $\hat{\beta_1}$.

If we change the slope of the regression line even a little bit, this will have a big impact on the value of the intercept (especially since the values of the explanatory variables are so far from 0).


## Collinearity

The rows of the design matrix (a column of ones, and a column of the `Year`) are almost perfectly `collinear`.

`Collinearity` means that one of the columns is (approximately) equal to a linear combination of the other columns.

Hence the regression coefficients tend to be highly variable when collinearity occurs.

What do we do?

## Introducing Orthogonal Polynomials

Let $P_r(x_i)$ denote a polynomial of order $r$ in $x_i$,
    where $i = 1, \ldots, n$; $$P_r(x_i) = a_0 + a_1x_i + a_2x_i^2 + \ldots + a_r x_i^r$$

The polynomials $P_r$ and $P_s$ are `orthogonal` if

$$\sum_{i=1}^n P_r(x_i) P_s(x_i) = 0~~~~~~(r \ne s)$$

## Introducing Orthogonal Polynomials


We can fit a polynomial of order `p` to the data using orthogonal polynomials:

For example: $E[Y_i] = 2+3x_i+4x_i^2$ can be re-expressed as

$$E[Y_i] = 1+3(\frac{1}{3}x_i+\frac{1}{3})+ 4(x_i^2+\frac{1}{2}x_i).$$

## Introducing Orthogonal Polynomials

$$E[Y_i] = \alpha_0 + \alpha_1 P_1(x_i) + \alpha_2 P_2(x_i) + \ldots + \alpha_p P_p(x_i)$$

This will be completely equivalent to fitting (both produce the same fitted values)

$$E[Y_i] = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \ldots + \beta_p x_i^p$$
<!-- with the $\beta$'s being a simple transformation of the $\alpha$'s. -->

The resulting estimated regression coefficients are  now uncorrelated.

## Introducing Orthogonal Polynomials

We can create a design matrix as follows:

$$
\small
X = \left [ \begin{array}{ll}
    1 & x_1 - \bar x \\
    1 & x_2 - \bar x \\
    \vdots & \vdots \\
    1 & x_n - \bar x
    \end{array}
    \right ]
$$

Note: Columns of $X$ are orthogonal because they can be expressed
    as two orthogonal polynomials.

Column 1 is $P_0 = 1$.

Column 2 is $P_1(x_i) = x_i - \bar{x}$.

## Introducing Orthogonal Polynomials

We can show that $\sum_{i=1}^n {P_0(x_i) P_1(x_i)} = \sum_{i=1}^n {(x_i - \bar{x})} = 0$.

When we have an orthogonal matrix $X_{orth}$, we have the following implications: 

$\rightarrow X_{orth}^T X_{orth}$ is a diagonal matrix

$\rightarrow \sigma^2 (X_{orth}^T X_{orth})^{-1}$ is also a diagonal matrix

$\rightarrow$ parameter estimates based on a model with this design matrix will be uncorrelated.


## Return to Brisbane Lions Example

We will define `Yr` to be a `centred` version of `Year`.

```{r Lions.lm2}
Lions$Yr <- Lions$Year - mean(Lions$Year)
Lions.lm2 <- lm(Crowd~Yr, data=Lions)
```

## Return to Brisbane Lions Example

```{r}
coef(Lions.lm1)
coef(Lions.lm2)
```

The models `Lions.lm1` and `Lions.lm2` have the same estimated slope parameters (as expected).

```{r cov2cor.vcovLions.lm2}
cov2cor(vcov(Lions.lm2))
```

## Revisiting the FEV Data Example

In a previous example we fitted a polynomial regression of `FEV` (a measure of lung function) on `Age`.

The estimated regression coefficients for this model are highly correlated.


```{r, echo=TRUE}
Fev <- read.csv(file="./fev.csv", header=TRUE)
Fev.lm6 <- lm(FEV~poly(Age,degree=6, raw=TRUE), data=Fev)
```

## Revisiting the FEV Data Example

```{r, echo=FALSE}
head(cov2cor(vcov(Fev.lm6)))[,1:3]
```



## Revisiting the FEV Data Example

```{r Fev.lm.poly}
Fev.lm.poly <- lm(FEV~poly(Age,degree=6),data=Fev)
head(round(cov2cor(vcov(Fev.lm.poly)),digits=6))[,1:3]
```

## Revisiting the FEV Data Example

```{r}
summary(Fev.lm.poly)
```

## Obtaining the design matrix

The `model.matrix()` command extracts the `X` matrix used in model fitting. It is quite large for the `FEV` data, so the following example compares the matrices used for the Lions example above.



::::{.columns}


:::{.column width="50%"}


```{r getModelMats}
model.matrix(Lions.lm1)
```


:::


:::{.column width=50%"}

```{r getModelMats2}
model.matrix(Lions.lm2)
```

:::


::::
