---
title: "161.251 Regression Modelling — Compiled Lectures"
author: "Jonathan Godfrey and Nick Knowlton"
date: "updated on `r format(Sys.Date(), '%d %B %Y')`"
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Lecture notes for 161.251 Regression Modelling at Massey University."
---

# Preface {-}

These notes are the compiled lecture material for the course 161.251 Regression
Modelling taught at Massey University by Jonathan Godfrey (Weeks 1--8 and
appendices) and Nick Knowlton (Weeks 9--12).

We use R in almost every lecture; you should be able to run the code on your own.
The material makes use of `tidyverse` packages. Graphs are made using a mixture
of base R graphics and ggplot functions, with the choice being made on the basis
of convenience.

Please ask for help if you don't understand the R code being demonstrated.

## Table of Contents {-}

```{r preface-toc, echo=FALSE, results='asis'}
lectures <- read.csv("../course/lectures.csv", stringsAsFactors = FALSE)
lectures <- lectures[lectures$IncludeInBook == "yes", ]
lectures <- lectures[order(lectures$LectureNo), ]

# Build filename from slug (matches bookdown's chapter ID-based naming)
make_link <- function(title, slug) {
  sprintf("%s.html", slug)
}

current_week <- -1
for (i in seq_len(nrow(lectures))) {
  wk <- lectures$Week[i]
  if (wk != current_week) {
    if (wk == 0) {
      cat("\n### Appendices {-}\n\n")
    } else {
      cat(sprintf("\n### Week %d {-}\n\n", wk))
    }
    current_week <- wk
  }
  link <- make_link(lectures$LectureTitle[i], lectures$Slug[i])
  presenter <- if (lectures$Presenter[i] == "Nick Knowlton") " (NK)" else ""
  cat(sprintf(
    "- [%d. %s%s](%s)\n",
    lectures$LectureNo[i], lectures$LectureTitle[i], presenter, link
  ))
}
```

```{r setup-book, include=FALSE}
source("../shared/setup.R")

knitr::opts_chunk$set(
  fig.path = "_main_files/figure-html/",
  cache.path = "../build/cache/book/"
)

# Write package bibliography
if (!file.exists("packages.bib")) {
  knitr::write_bib(file = "packages.bib")
}
```