---
title: "161.251 Assignment 2, Semester 1 2024"
author: "YOUR NAME HERE"
date: "Due Friday 25 October 2024"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
#knitr::opts_chunk$set(echo = TRUE)

pv <- read_csv("../data/ModelingPVCapacity.csv")
pv_final <- pv |> mutate(MonthEnd = dmy(MonthEnd)) |>
  select(-Row, -Month, -Year, -Time)

nsw <- read_csv("../data/NSWHospitals.csv")
```

## Exercise 1

The data frame `pv` has data on the amount of installed photovoltaic cells (solar panels) in New Zealand.  We wish to model the `Capacity` (in MWh) as a function of time.

1. The following code constructs a `Time` variable from the `MonthEnd` variable.

    ```{r}
    pv_final <- pv_final |>
      mutate(Time = time_length(MonthEnd - min(MonthEnd), unit="day"))
    ```

    - Why is this important for a time series dataset like this?

2. Using `pv_final`, create separate models for `Capacity` in terms of `Time` with either a linear trend or a quadratic trend.

    - Show summary output of each model.

3. Utilise the `broom::augment()` function (or an equivalent) to visualise the result of *BOTH* models on the *SAME* chart against the data.

    - Ensure you show the data on the chart.
    - Consider how you might show the fitted values from both models on the chart.
    - Consider if you should show uncertainty in the model fits.

4. Perform an appropriate test, or provide evidence, to decide which model is best for modelling `Capacity` in terms of `Time`.

5. In November 2014, Contact Energy reduced the amount it would pay when customers sold PV-generated power back to the national grid, a move that was immediately followed by other companies. The following code adds an indicator variable for this.
    
    ```{r}
    pv_change <- pv_final |>
      mutate(PostNov2014 = if_else(MonthEnd >= '2014-11-30', 1, 0))
    ```

    Utilising the `pv_change` dataset, refit the quadratic model for capacity incorporating the `PostNov2014` variable as an additive effect.
    
    - What is your conclusion about the effect of the Buy back amount reducing from this model? Include evidence.

6. Adapt your model from part 5 to include an interaction between your quadratic in time and the `PostNov2014` variable.

    - What is the purpose of including the interaction in this way?
    - What is your conclusion about the rate of PV uptake in New Zealand before and after November 2014?

7. Generate diagnostic plots from your model in part 6. Are there any problems with the linear model assumptions?


## Exercise 2 

The data frame `nsw` contains information about the costs and treatment of various hospitals in different areas ("SLA"s) of New South Wales, Australia.

The response variable we will focus on is called `StdCostRatio`,  which is a standardised measure of how expensive various hospitals are, per patient (average=100). The remaining variables are described in an appendix. You don't need to know the definitions.

1. Fit a  linear regression of `StdCostRatio` on all variables excluding the first 4 columns (i.e.  from `SupplyBeds10000` to `Nocar`. Show the model output.

2. Calculate variable inflation factors for the preceding regression.  Which variables have the highest VIFs?   Is multicollinearity a problem?

3. Use Backwards elimination to find a linear regression of `StdCostRatio` based on the columns from `SupplyBeds10000` to `Nocar` but with fewer variables.

    - Show a summary of the final model.
    - Which variable is the strongest predictor?

4. Update your model from 3 to include the last variable removed in the stepwise regression, and compare the adjusted $R^2$ of these two models.

    - Which model would you prefer, and why?

5. The `StdCostRatio` is a summarised statistic computed based on different population sizes (available in `Population`). Re-fit the regression model chosen by the `step()` routine in part 3 with `weights` set to `Population`.

    - Give the model summary output and a brief explanation of the difference with the unweighted model.

6. Plot the weighted residuals against fitted values for your weighted regression from 5.

    - You should find an observation that is somewhat unusual.
    - Which area is this observation from?

## Exercise 3

We return to the model for `pv` capacity from Exercise 1.

(a)	Using the Durbin-Watson test, test the residuals from the  model in Q1(c)(i) for autocorrelation.   What do you conclude?   *Hint you will need to install the library lmtest.

```{r}

```

*Comment:  *

(b)(i).  	Calculate   variables

SinTime = Sin(Time* 2  * 3.1415927/3.6525)  and 
CosTime = Cos(Time* 2 * 3.1415927/3.6525) 

Fit a linear model of  Capacity on  Time   I(Time^2)   Changeslope   Jump   SinTime   and   CosTime .   Show  a summary of the regression output. If a variable is non-significant then eliminate it and re-fit the regression.   

```{r}

```

 (ii).	Is this a better model than before?  Quote evidence.

*Comment:   *

 (iii).	What time of year is peak for installation of new PV capacity?   (Hint check the data for when the sinusoidal part of the model $\approx$ 1.  When does this occur? Check against the column “MonthEnd”)

```{r}

```

*Comment:    *          

(c) Has the autocorrelation been removed?  Check using dwtest and show an acf plot of the residuals from the most recent model in (b). What do you conclude? 

```{r}

```

*Comment:  *

(d) Use the gls function to fit a first-order autoregressive model

library(nlme)

Cap.gls <- gls(Capacity ~ Time + I(Time^2) + Changeslope + Jump + SinTime, correlation = corAR1() )

Estimate the first-order autocorrelation coefficient. 

```{r}

```

*Comment:   *