---
title: "161.251 Assignment 2, Semester 1 2024"
author: "YOUR NAME HERE"
date: "Due Friday 25 October 2024"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
#knitr::opts_chunk$set(echo = TRUE)

pv <- read_csv("../data/ModelingPVCapacity.csv")
pv_final <- pv |> mutate(MonthEnd = dmy(MonthEnd)) |>
  select(-Row, -Month, -Year, -Time)

nsw <- read_csv("../data/NSWHospitals.csv")
```

## Exercise 1

The data frame `pv` has data on the amount of installed photovoltaic cells (solar panels) in New Zealand.  We wish to model the `Capacity` (in MWh) as a function of time.

1. The following code constructs a `Time` variable from the `MonthEnd` variable.

    ```{r}
    pv_final <- pv_final |>
      mutate(Time = time_length(MonthEnd - min(MonthEnd), unit="month"))
    ```

    - Why is this important for a time series dataset like this?

    A time variable is important so that we can adequately model trend
    through time via a continuously increasing variable.

2. Using `pv_final`, create separate models for `Capacity` in terms of `Time` with either a linear trend or a quadratic trend.

```{r}
pv.1 <- lm(Capacity ~ Time, data=pv_final)
pv.2 <- lm(Capacity ~ poly(Time, 2), data=pv_final)
summary(pv.1)
summary(pv.2)
```

    - Show summary output of each model.

3. Utilise the `broom::augment()` function (or an equivalent) to visualise the result of *BOTH* models on the *SAME* chart against the data.

    - Ensure you show the data on the chart.
    - Consider how you might show the fitted values from both models on the chart.
    - You need not represent the uncertainty of the fitted values.

```{r}
library(broom)
lst(pv.1, pv.2) |> map_dfr(augment, newdata=pv_final, .id='model') |>
  ggplot() +
  geom_point(aes(x=Time, y=Capacity)) +
  geom_line(aes(x=Time, y=.fitted, col=model))
```

4. Perform an appropriate test, or provide evidence, to decide which model is best for modelling `Capacity` in terms of `Time`.

```{r}
anova(pv.1, pv.2)
```
The polynomial model is clearly superior.

5. In November 2014, Contact Energy reduced the amount it would pay when customers sold PV-generated power back to the national grid, a move that was immediately followed by other companies. The following code adds an indicator variable for this.
    
    ```{r}
    pv_change <- pv_final |>
      mutate(PostNov2014 = if_else(MonthEnd >= '2014-11-30', 1, 0))
    ```

    Utilising the `pv_change` dataset, refit the quadratic model for capacity incorporating the `PostNov2014` variable as an additive effect.
    
    - What is your conclusion about the effect of the Buy back amount reducing from this model? Include evidence.

```{r}
pv.3 <- lm(Capacity ~ poly(Time, 2) + PostNov2014, data=pv_change)
summary(pv.3)
```

This suggests that uptake of PV *increased* after november 2014, though it is only moderately significant. This doesn't seem to make much sense.

6. Adapt your model from part 5 to include an interaction between your quadratic in time and the `PostNov2014` variable.

    - What is the purpose of including the interaction in this way?
    - What is your conclusion about the rate of PV uptake in New Zealand before and after November 2014?

```{r}
pv.4 <- lm(Capacity ~ poly(Time, 2) * PostNov2014, data=pv_change)
summary(pv.4)
```

The interaction allows all the terms in the polynomial trend to adjust after November 2014, rather than having just a "jump".

This makes a bunch more sense: both the intercept term and the polynomial "slope" terms are decreased post 2014.

7. Generate diagnostic plots from your model in part 6. Are there any problems with the linear model assumptions?

```{r}
par(mfrow=c(2,2))
plot(pv.4)
```

There's clearly some patterns on the fitted vs residuals and scale-location plots. Normality is probably fine. The patterns are likely due to autocorrelation.


## Exercise 2 

The data frame `nsw` contains information about the costs and treatment of various hospitals in different areas ("SLA"s) of New South Wales, Australia.

The response variable we will focus on is called `StdCostRatio`,  which is a standardised measure of how expensive various hospitals are, per patient (average=100). The remaining variables are described in an appendix. You don't need to know the definitions.

1. Fit a  linear regression of `StdCostRatio` on all variables excluding the first 4 columns (i.e.  from `SupplyBeds10000` to `Nocar`. Show the model output.

```{r}
nsw_reduc <- nsw |> select(-c(1:4))
nsw.1 <- lm(StdCostRatio ~ ., data=nsw_reduc)
summary(nsw.1)
```

2. Calculate variable inflation factors for the preceding regression.  Which variables have the highest VIFs?   Is multicollinearity a problem?

```{r}
library(car)
vif(nsw.1)
```

Yes, lots of the VIF are very large.

3. Use Backwards elimination to find a linear regression of `StdCostRatio` based on the columns from `SupplyBeds10000` to `Nocar` but with fewer variables.

    - Show a summary of the final model.
    - Which variable is the strongest predictor?

```{r}
nsw.null <- lm(StdCostRatio ~ 1, data=nsw_reduc)
nsw.full <- lm(StdCostRatio ~ ., data=nsw_reduc)
scope <- list(lower=formula(nsw.null), upper=formula(nsw.full))
nsw.step <- step(nsw.full, scope=scope, direction="backward")
summary(nsw.step)
```

Strongest predictor appears to be ARIAscore as that has the lowest P-value.

4. Update your model from 3 to include the last variable removed in the stepwise regression, and compare the adjusted $R^2$ of these two models.

    - Which model would you prefer, and why?

The last variable to be removed was `Overcrowded`. Let's add that in.

```{r}
nsw.step2 <- update(nsw.step,  ~ . + Overcrowded)
summary(nsw.step)
summary(nsw.step2)
```

The model with `Overcrowded` has a very slightly increased adjusted R^2. Very slightly. There's not a strong argument for preffering one over the other.

5. The `StdCostRatio` is a summarised statistic computed based on different population sizes (available in `Population`). Re-fit the regression model chosen by the `step()` routine in part 3 with `weights` set to `Population`.

    - Give the model summary output and a brief explanation of the difference with the unweighted model.

```{r}
nsw.step.wt <- lm(formula(nsw.step), weight=Population, data=nsw_reduc)
summary(nsw.step.wt)
summary(nsw.step)
```

6. Plot the weighted residuals against fitted values for your weighted regression from 5.

    - You should find an observation that is somewhat unusual.
    - Which area is this observation from?

```{r}
nsw_resid <- augment(nsw.step.wt, nsw) |>
  mutate(wtresid = .resid * sqrt(Population))

nsw_resid |>
  ggplot() +
  geom_point(aes(x=.fitted, y=wtresid))

nsw_resid |> filter(wtresid > 6000)
```

## Exercise 3

We return to the model for `pv` capacity from Exercise 1.

1. The following code adds a variable for `Month` of the year:

    ```{r}
    pv_month <- pv_change |>
      mutate(Month = factor(month(MonthEnd, label=TRUE),
                            ordered=FALSE))
    ```

    Try adding `Month` to your model in Exercise 1 part 6.

    - Is there evidence of seasonality?
    - If so, what is the peak month for installation?

```{r}
pv.5 <- lm(Capacity ~ poly(Time, 2)*PostNov2014+Month, data=pv_month)
anova(pv.5)
summary(pv.5)
```

2. Assess the residuals of the model fit in part 1 for autocorrelation.

    - What do you conclude?
    
```{r}
acf(resid(pv.5))
```

Lots of autocorrelation here.

3. Use the `gls` function to fit a first-order autoregressive model based on the model in part 1.

    - What is your estimate of the first-order autocorrelation coefficient?
    - Has modelling with an AR1 process removed the autocorrelation in the residuals?

```{r}
library(nlme)
pv.6 <- gls(Capacity ~ poly(Time, 2)+PostNov2014+Month, correlation=corAR1(), data=pv_month)
summary(pv.6)
anova(pv.6)
plot(pv.6)
acf(resid(pv.6))
```

The acf plot is different, but we haven't really solved the underlying problem with just the AR1() process. It may be we need a stronger auto-correlation process for these data, as the coefficient was essentially 1.
