## Regularized Regression: Ridge, Lasso, and Cross-Validation

In the previous lecture we made regression models more flexible by adding spline basis terms. In the Seoul bike data, that immediately raises a new question: what happens when we combine flexible weather effects with many calendar and seasonal predictors?

By the end of this lecture, students should be able to:

1. Explain why large or highly correlated predictor sets can destabilise ordinary least squares.
2. Write the penalised least-squares objectives for ridge and lasso regression.
3. Distinguish L2 shrinkage from L1 shrinkage.
4. Use cross-validation to choose the tuning parameter `lambda`.
5. Recognise regularisation as a general tool for controlling model complexity.

```{r regularized-setup, include=FALSE}
library(glmnet)
```

`r xfun::embed_file("../Data/seoul_bike_hourly.csv")`

```{r read seoul hourly, eval=-1, echo=-2, message=FALSE}
seoul_hourly = read_csv("../Data/seoul_bike_hourly.csv")
seoul_hourly
```

## Why Regularise?

Ordinary least squares can struggle when:

- predictors overlap strongly,
- the design matrix becomes large,
- flexible terms are added aggressively,
- or the analyst wants prediction stability more than a sparse symbolic model.

Regularisation keeps the regression framework, but adds a penalty that discourages overly complex coefficient vectors.

## The Core Equations

**Ordinary least squares** chooses $\beta$ to minimise

$$\sum_{i=1}^n (y_i - \hat{y}_i)^2.$$

**Ridge regression** adds an $L_2$ penalty:

$$\hat{\beta}^{\text{ridge}} = \arg\min_\beta \left\{ \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^p \beta_j^2 \right\}.$$

**Lasso regression** adds an $L_1$ penalty:

$$\hat{\beta}^{\text{lasso}} = \arg\min_\beta \left\{ \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^p |\beta_j| \right\}.$$

The tuning parameter $\lambda$ controls how strongly we penalise coefficient size.

## Ridge vs Lasso with the Seoul Bike Data

This is a natural continuation of the variable-selection and multicollinearity discussions from Lectures 27 and 29. We have:

- weather predictors,
- hour and season effects,
- holiday information,
- and now spline terms for temperature as well.

```{r seoul candidate columns}
seoul_hourly |>
  select(
    rented_bike_count,
    hour,
    season,
    holiday,
    functioning_day,
    temperature,
    humidity,
    wind_speed,
    visibility,
    dew_point_temperature,
    solar_radiation,
    rainfall,
    snowfall
  )
```

### Overlapping Predictors

```{r ridge-lasso-overlap, warning=FALSE, message=FALSE}
seoul_hourly |>
  ggplot(aes(x = temperature, y = dew_point_temperature)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Temperature",
    y = "Dew point temperature",
    title = "Some predictors carry overlapping information"
  )
```

## Fitting Ridge and Lasso

```{r ridge-lasso-fit, warning=FALSE, message=FALSE}
X <- model.matrix(
  rented_bike_count ~ hour + season + holiday + functioning_day +
    splines::ns(temperature, df = 5) +
    humidity + wind_speed + visibility +
    dew_point_temperature + solar_radiation + rainfall + snowfall,
  data = seoul_hourly
)[, -1]
y <- seoul_hourly$rented_bike_count

set.seed(42)

ridge_fit <- cv.glmnet(X, y, alpha = 0, standardize = TRUE)
lasso_fit <- cv.glmnet(X, y, alpha = 1, standardize = TRUE)

par(mfrow = c(1, 2))
plot(ridge_fit, main = "Ridge Regression")
plot(lasso_fit, main = "Lasso")
```

Cross-validation chooses `lambda` from the data. The two reference values are usually:

- `lambda.min`: the value with smallest estimated prediction error,
- `lambda.1se`: a more conservative choice that often gives a simpler model.

This is the same design problem we saw earlier in the course, but richer:

- Lecture 27 asked which predictors belong in the model.
- Lecture 29 showed that overlapping predictors can destabilise coefficients.
- Here we allow the full design matrix and control its complexity by penalising coefficient size.

## Coefficient Comparison

```{r ridge-lasso-coefs, warning=FALSE, message=FALSE}
ridge_coef_matrix <- coef(ridge_fit, s = "lambda.min") |>
  as.matrix()
ridge_coef <- tibble(
  term = rownames(ridge_coef_matrix),
  estimate = as.numeric(ridge_coef_matrix[, 1])
) |>
  filter(term != "(Intercept)") |>
  mutate(model = "Ridge")

lasso_coef_matrix <- coef(lasso_fit, s = "lambda.min") |>
  as.matrix()
lasso_coef <- tibble(
  term = rownames(lasso_coef_matrix),
  estimate = as.numeric(lasso_coef_matrix[, 1])
) |>
  filter(term != "(Intercept)") |>
  mutate(model = "Lasso")

bind_rows(ridge_coef, lasso_coef) |>
  mutate(term = fct_reorder(term, abs(estimate))) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_point() +
  facet_wrap(~ model, scales = "free_x") +
  labs(
    x = "Coefficient estimate",
    y = NULL,
    title = "Ridge shrinks coefficients; lasso can remove terms"
  )
```

```{r ridge-lasso-counts, warning=FALSE, message=FALSE}
cat("Ridge: number of non-zero coefficients:",
    sum(coef(ridge_fit, s = "lambda.min") != 0), "\n")
cat("Lasso: number of non-zero coefficients:",
    sum(coef(lasso_fit, s = "lambda.min") != 0), "\n")
```

## Ridge vs Lasso Summary

| Feature | Ridge | Lasso |
|---------|-------|-------|
| Penalty | $L_2$ (squared) | $L_1$ (absolute) |
| Coefficient shrinkage | Toward zero | Some to exactly zero |
| Variable selection | No | Yes |
| Correlated predictors | Shares coefficient weight | Often keeps a smaller subset |
| When to use | Many small effects | Many candidates, only some likely to matter |

## Bridge to Time-Indexed Regression

Regularisation controls coefficient complexity. It does **not** remove time structure from the mean or make dependent residuals independent.

```{r seoul daily preview, warning=FALSE, message=FALSE}
seoul_daily <- read_csv("../Data/seoul_bike_daily.csv", show_col_types = FALSE)

seoul_daily |>
  select(
    day_index,
    rented_bike_count,
    mean_temperature,
    total_solar_radiation,
    season
  ) |>
  slice(1:8)
```

In the next lecture we will switch to the daily aggregate and ask a different question: how do we model trend and seasonality in the mean response over time?
