
```{r seasonality-setup, include=FALSE}
library(tidyverse)
library(tsibble)
library(fable)
library(feasts)
library(lubridate)
```

## Time Series Data: A Regression Perspective

Time series data differs from cross-sectional data in that observations are collected over time, creating potential dependencies between successive measurements. This violates the standard regression assumption of independent errors.

In this lecture and the next, we'll address two key questions:

1. **This lecture**: How do we model systematic patterns like trend and seasonality using regression predictors?
2. **Next lecture**: What happens when residuals remain correlated after fitting our model?

For now, we assume the independence assumption holds and focus on capturing temporal structure through well-chosen predictors.

## Independent vs non-independent errors

Assumption A2 of   linear regression 
$$ y_i = \beta_0 + \beta_1  x_{i1} + \beta_2 x_{2i}  + ... \beta_p  x_{pi} + \varepsilon_i$$

is that the errors $\varepsilon_i$ are independent. That is, there is no carry-over of information from one data row $i$  to another, apart from what is expressed through differences in the covariates (or factors, if relevant).

In the final topics of this course we will consider how to detect whether this assumption is broken, and how to model 
data with correlated errors 
$cor(\varepsilon_{i-1},\varepsilon_i) \ne 0$.

For the moment we suppose the independence assumption A2 remains correct, and consider some simple ways of handling data that have a regular periodic component, such as patterns that depend on seasons of the year, or time of the day. 

## Indicator Variables for Seasonality

Often time series show some form of seasonal variation.  One way to handle this in a regression equation is to use indicator variables which pick out the most obvious seasonal influences (e.g. summer/ winter, or the Christmas effect on retail figures. )    

We consider data on Ski equipment sales versus PDI (personal disposable income).  In the following data, both Ski Sales and PDI increased as time went by, but there appears to be a seasonal aspect to sales (which makes sense). 

`r xfun::embed_file("../../data/SkiSales.csv")`

```{r Ski Sales, eval=-1, echo=-2}
SkiSales = read.csv("SkiSales.csv",header=TRUE)
SkiSales = read.csv("../../data/SkiSales.csv",header=TRUE)
SkiSales |> head() |> kable()
```

```{r PlotSkiSales}
SkiSales |> mutate(Quarter =  as.factor(Quarter)) |> ggplot(aes(y=Sales, x= PDI)) +
geom_point(aes(pch=Quarter, col= Quarter))

SkiSales.lm1 = lm(Sales ~ PDI, data=SkiSales)
Residuals =  SkiSales.lm1$residuals
SkiSales |> mutate(Quarter = as.factor(Quarter)) |> ggplot(aes(x=PDI, y=Residuals)) +
geom_point(aes(pch=Quarter, col= Quarter)) +
geom_abline(h=0)
```

The residual plot shows that residuals tended to be positive in the first and fourth quarter of the year
(plotted using o,x) and negative in second and third quarter (plotted using triangle,+).    It therefore
seems reasonable to create an indicator variable for winter
 (first and   fourth quarter). 
 
```{r winterIndicator}
SkiSales |> mutate(Winter =as.numeric( (Quarter ==1)+ (Quarter ==4))) -> SkiSales
SkiSales.lm2 = lm(Sales ~ PDI + Winter, data=SkiSales)
summary(SkiSales.lm2)
plot(SkiSales.lm2$residuals ~ PDI,  pch=Quarter, col= Quarter, data=SkiSales)
```

The regression shows ski sales are significantly higher in winter, as expected, and the residuals do not show any particular pattern with Quarter.   

However, for completeness, we will just check whether there is any tendency for ski sales to be higher or lower in the first half of winter (Christmas??) or summer. 

```{r Q4Q2Indicator}
SkiSales |> mutate(Q4=as.numeric(  Quarter ==4 ),
 Q2=as.numeric(  Quarter ==2 )) -> SkiSales
SkiSales.lm3 = lm(Sales ~ PDI + Winter+ Q4 + Q2, data=SkiSales)
summary(SkiSales.lm3)
anova(SkiSales.lm2, SkiSales.lm3)
```

We can conclude that a simple winter indicator is sufficient to express the effect of season on sales. 

## Effect of month

We consider some data on a householder’s monthly electricity bill from May 1994 to May 1998. During this period the householder sometimes let paying guests rent part of the house, and we may expect the electricity bill was  higher at those times.  The householder's problem was to estimate the extra cost, so it could be legitimately claimed as a business expense for tax purposes.  

One might expect the electricity bill to be higher in winter months, and in later years because of price rises, and also for periods when there were paying guests renting part of the house. It was also supposed that there might be an effect for the number of days included in the bill, and for whether the electricity bill was an estimate or the meter had actually been read. 

The focus *for us* is how to adjust the analysis for the effect of season (month)  on the bill.  To examine this, we consider the residuals of Bill after adjusting for the non-weather-related variables.

`r xfun::embed_file("../../data/Electricity.csv")`

```{r Electricity vs weather, eval=-1, echo=-2}
Electricity= read.csv("Electricity.csv",header=TRUE)
Electricity = read.csv("../../data/Electricity.csv",header=TRUE) 
Electricity |> head() |> kable()
Electricity.lm1 =lm(Bill~Days + Estimate+ Guestwks + Time , data=Electricity)
resids =Electricity.lm1$residuals
plot(resids~ Time, data=Electricity)
lines(resids~ Time, lty=1,col=4, data=Electricity)
abline(h=0, lty=2)
```


Although the effect of season is not clear, there are portions of the graph with several positive residuals in sequence, and several negative residuals. Time starts here at month=5 (May).

### Indicator variables

One way of modelling seasonality would be to estimate an effect for each Month.  The dataset here is really too short for such an approach, but we can try it for purposes of illustration.

```{r factorMonth}
Electricity.lm2= lm(resids ~ factor(Month), data=Electricity)
summary(Electricity.lm2)
```

We see from the F test that the mean(resids) varies significantly with Month.   
F-statistic: 4.633 on 11 and 37 DF,  p-value: 0.0002 

The t-tests indicate that, compared to the baseline month January,  the resids are unusually high in July through to October.  This makes sense because more electricity would be used for lighting and heating during winter months. 

Usually, however, we would like a more parsimonious model. So we consider one next. 

## Sinusoidal models

It would seem reasonable to model electricity use by temperature, and temperature by a simple Sine curve, with peak in winter and lowest electricity use in summer.
Perhaps an initial guess is to set the '0' for the sine curve to  April since the power bill in that month includes the autumn equinox when temperatures should be ‘average’.

We can compute variables

SineTime=sin(  (Time-4)$\times 2\times$
 3.1415927 /12) ,    and

CosTime=cos(  (Time-4)$\times 2\times$
3.1415927 /12).

The rationale for also calculating a Cosine curve is that sometimes we don’t  know for sure where the ‘0’  should be.  If we include a Cosine curve in the model, its effect is to allow the ‘0’ to shift earlier or later.  

How it works:  suppose we want to fit a curve of the form   $A\sin( x+\phi)$  where $A$ is an amplitude and 
$\phi$ is an unknown shift parameter.  Then  trigonometry tells us that
$$A.\sin(x+\phi)  
    =A\sin(x)\cos(\phi) + A\cos(x)\sin(\phi) \\
    = A_1 \sin(x) + A_2\cos(x)   $$
    
So a  regression of  $Y$ on the two columns with numbers 
$\sin(x)$ and $\cos(x)$  will automatically calculate the slope parameters $A_1$ and $A_2$   and their relative size will automatically determine the optional choice of starting point $\phi$.  Easy as!

If we find out later the extra cos() term is  not needed, then we can stick with our guess of April. 

```{r sinusoidal first guess}
Electricity |> mutate(SinTime = sin((Time-4)*2*3.1415927 /12),
CosTime = cos((Time-4)*2*3.1415927 /12)) -> Electricity
Electricity.lm3 = lm(resids ~ SinTime  + CosTime, data=Electricity)
summary(Electricity.lm3)
```

The regression shows that the coefficient of CosTime is definitely non-zero.  If it had been zero then this would mean  $\sin(\phi)=0$  and hence $\phi=0$  or in other words that there was no need for any further adjustment from my guess of April as the “average” month.       

Since the CosTime is highly significant it looks as if my guess was wrong.  Let’s plot the fitted values on the same graph as my working ‘response’ variable.The plot below shows that the fitted values equal zero close to the first data point, at Time =5. 

```{r plotSinusoidal}
plot(resids~ Time, data=Electricity)
lines(resids~ Time, lty=1,col=4, data=Electricity)
lines(Electricity.lm3$fitted.values ~ Time, lty=2,col=2,lwd=2, data=Electricity)
abline(h=0, lty=2)
```

 
So we will try with 5 as the adjustment. The working below shows success. 

```{r second sinusoidal}
Electricity |> mutate(SinTime = sin((Time-5)*2*3.1415927 /12),
CosTime = cos((Time-5)*2*3.1415927 /12)) -> Electricity
Electricity.lm4 = lm(resids ~ SinTime  + CosTime, data=Electricity)
summary(Electricity.lm4)
```
 
### Adding more sin and cos terms

Now considering again the plot, I am not convinced  that the
relationship between resids and month is  a simple
sine curve. In particular resids rise more steeply in winter.
 
The next step is to add a component with 
double the period of the sin curve, i.e.

Sin2Time=sin(  (Time-5)$\times 4\times$
 3.1415927 /12) ,    and

Cos2Time=cos( (Time-5)$\times 4\times$ 3.1415927 /12).

Adding terms in this way gives a more flexible set of shapes.

```{r Augmented sin curve}
Electricity |> mutate(Sin2Time = sin((Time-5)*4*3.1415927 /12),
Cos2Time = cos((Time-5)*4*3.1415927 /12)) -> Electricity
Electricity.lm5 = lm(resids ~ SinTime  + CosTime+ Sin2Time + Cos2Time, data=Electricity)
summary(Electricity.lm5)
anova(Electricity.lm4,Electricity.lm5)
```

```{r plot augmented sinusoidal}
plot(resids~ Time, data=Electricity)
lines(resids~ Time, lty=1,col=4, data=Electricity)
lines(Electricity.lm5$fitted.values ~ Time, lty=2,col=2,lwd=2, data=Electricity)
abline(h=0, lty=2)
```
 
The regression output indicates that the extra sinusoidal terms have improved the fit significantly. 

The graph suggests:  first, that  the peak in power costs in winter is steeper than previously shown, with estimated maximum in August;  estimated minimum power use is in December;  and there was a fairly flat period of  electricity use for the four months between February and May inclusive.  

(The graph actually shows a tiny peak in March but it is not clear whether this is real or an “artifact”    of the model  i.e. something forced by the fact we are fitting the data using simple wavy curves and so one is going to get wiggles whether real or not.) 

The only way to check would be to add in even more terms such as sines and cosines with 3 times the frequency.  However we do not have a lot of data so we will stop here. 

## Fourier Terms: A Systematic Approach

The sine and cosine terms we've been adding are examples of **Fourier terms**. For monthly data with period $m = 12$, we can create pairs of sine and cosine terms at different frequencies:

$$S_k(t) = \sin\left(\frac{2\pi k t}{m}\right), \quad C_k(t) = \cos\left(\frac{2\pi k t}{m}\right)$$

where $k = 1, 2, \ldots, K$ and $K \leq m/2$. Each pair $(S_k, C_k)$ captures oscillations at frequency $k/m$ cycles per time unit.

- $k = 1$: fundamental frequency (1 cycle per 12 months)
- $k = 2$: second harmonic (2 cycles per 12 months, i.e., 6-month period)
- $k = 3$: third harmonic (3 cycles per 12 months, i.e., 4-month period)

Adding more harmonics increases flexibility, similar to adding more knots in a spline. The key advantage is that Fourier terms provide a **systematic, orthogonal basis** for seasonal patterns.

```{r fourier-terms, tidy=FALSE}
# Create Fourier terms systematically
Electricity |> 
  mutate(
    S1 = sin(2 * pi * Time / 12),
    C1 = cos(2 * pi * Time / 12),
    S2 = sin(4 * pi * Time / 12),
    C2 = cos(4 * pi * Time / 12)
  ) -> Electricity_fourier

# Fit model with Fourier terms
Electricity.lm_fourier <- lm(resids ~ S1 + C1 + S2 + C2, data = Electricity_fourier)
summary(Electricity.lm_fourier)
```

This is equivalent to what we did manually with `SinTime`, `CosTime`, `Sin2Time`, `Cos2Time`, but the Fourier approach is more systematic and extends naturally to higher harmonics or different seasonal periods.

## STL Decomposition: A Visual Diagnostic

**STL** (Seasonal and Trend decomposition using LOESS) provides a visual way to separate a time series into three components:

1. **Trend**: long-term movement
2. **Seasonal**: repeating periodic patterns
3. **Remainder**: what's left after removing trend and seasonality

STL is not a regression model — it's a **descriptive diagnostic** that helps us understand what patterns exist in the data before we build our regression model.

```{r stl-decomposition, warning=FALSE, message=FALSE}
# Create a monthly time series object for STL decomposition
Electricity_ts <- ts(Electricity$Bill, start = c(1994, 5), frequency = 12)

# STL decomposition using base R
Electricity_stl <- stl(Electricity_ts, s.window = 7, t.window = 7)

# Plot the decomposition
plot(Electricity_stl, main = "STL Decomposition of Electricity Bills")
```

The decomposition shows:
- **Trend component**: gradual increase in bills over time
- **Seasonal component**: regular winter peaks
- **Remainder**: irregular fluctuations

This visual diagnostic confirms what our regression models are trying to capture: a trend (via `Time`) and seasonality (via sine/cosine terms or indicators). The remainder should look like random noise if our model is adequate.

### Putting it together

Finally one should put all the variables together into the same model: 
 
 
```{r everything}
 Electricity.lm6 =lm(Bill~Days + Estimate+ Guestwks + Time+SinTime  + CosTime+ Sin2Time + Cos2Time, data=Electricity)
 summary(Electricity.lm6)
plot(Bill~ Time, data=Electricity)
lines(Bill~ Time, lty=1,col=4, data=Electricity)
lines(Electricity.lm6$fitted.values ~ Time, lty=2,col=2,lwd=2, data=Electricity)
```

The model is not perfect, but is now picking up many of the features of the data. 
 

