## Beyond Parametric Models

So far we have focused on parametric regression models — models where the form of the relationship between $Y$ and $X$ is specified by a small number of parameters (e.g. $\beta_0 + \beta_1 x$, or $\beta_0 + \beta_1 x + \beta_2 x^2$).

In this lecture we introduce three approaches that relax the strict parametric form:

- **Splines**: piecewise polynomials joined at knots
- **LOESS**: locally estimated scatterplot smoothing
- **Regularized models**: ridge regression and lasso

## Why Go Beyond Parametric Models?

- Parametric models impose a global structure on the data
- Real relationships may be more complex than a straight line or simple polynomial
- We may want to let the data "speak for themselves" without imposing a rigid functional form
- Regularization helps when we have many predictors relative to observations

## Motivation: The Flexibility Problem

A straight line can underfit. A high-degree polynomial can overfit. Flexible regression needs control.

```{r motivation, warning=FALSE, message=FALSE}
library(tidyverse)
library(splines)
library(glmnet)

auto_mpg <- read_csv("../../data/auto_mpg.csv", show_col_types = FALSE)

auto_mpg |>
  ggplot(aes(x = horsepower, y = mpg)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Horsepower",
    y = "Miles per gallon",
    title = "Fuel economy falls nonlinearly as horsepower increases"
  )
```

Today's methods all control flexibility in different ways:

- **Splines** control where and how curves bend
- **LOESS** controls locality through a smoothing span
- **Ridge and lasso** control coefficient size through penalties

---

## Example 1: LOESS — Exploratory Local Smoothing

### The LOESS Idea

Instead of fitting one global model, LOESS fits many small local regressions:

1. For each point $x_0$ where we want a prediction, select a neighbourhood of nearby points
2. Fit a low-degree polynomial (usually linear or quadratic) using weighted least squares, where points closer to $x_0$ get more weight
3. The predicted value at $x_0$ is the fitted value from this local model

### LOESS with Auto MPG

```{r loess-auto, warning=FALSE, message=FALSE}
auto_mpg |>
  ggplot(aes(x = horsepower, y = mpg)) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    span = 0.3,
    se = FALSE,
    linewidth = 1
  ) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    span = 0.8,
    se = FALSE,
    linewidth = 1
  ) +
  labs(
    x = "Horsepower",
    y = "Miles per gallon",
    title = "LOESS span controls the smoothness of the fitted curve"
  )
```

### The Span Parameter

- `span` controls the fraction of data used in each local neighbourhood
- Small span (0.3) → more local, more flexible, noisier
- Large span (0.8) → smoother, less flexible, more stable
- Typical values: 0.2 to 0.8
- Can be chosen by cross-validation or AICc

### LOESS as a Visual Tool

LOESS is excellent for exploratory graphics. It is less convenient when the model needs to be written down, interpreted through coefficients, or combined into a larger regression design.

```{r loess-spline-compare, warning=FALSE, message=FALSE}
# Fit a natural spline for comparison
spline_hp <- lm(mpg ~ ns(horsepower, df = 4), data = auto_mpg)

hp_grid <- tibble(horsepower = seq(
  min(auto_mpg$horsepower),
  max(auto_mpg$horsepower),
  length.out = 200
))

hp_predictions <- hp_grid |>
  mutate(
    natural_spline = predict(spline_hp, newdata = hp_grid)
  ) |>
  pivot_longer(
    cols = c(natural_spline),
    names_to = "model",
    values_to = "mpg_hat"
  )

auto_mpg |>
  ggplot(aes(x = horsepower, y = mpg)) +
  geom_point(alpha = 0.4) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    span = 0.5,
    se = FALSE,
    linewidth = 1,
    colour = "#E69F00"
  ) +
  geom_line(
    data = hp_predictions,
    aes(x = horsepower, y = mpg_hat, linetype = model),
    linewidth = 1,
    colour = "#0072B2"
  ) +
  labs(
    x = "Horsepower",
    y = "Miles per gallon",
    linetype = "Model",
    title = "LOESS shows the shape; splines give a regression model"
  )
```

The LOESS curve is useful for seeing the empirical shape of the data. The natural spline captures a similar nonlinear pattern but remains a formal regression model — easier to combine with other predictors, compare models, and make predictions.

### Slide Summary: LOESS

| Aspect | Description |
|--------|-------------|
| **Purpose** | Explore the shape of a relationship |
| **Mechanism** | Fit simple regressions in local neighbourhoods |
| **Control** | Span |
| **Strength** | Very good visual diagnostic |
| **Limitation** | Less convenient as a formal regression model |

---

## Example 2: Splines — Flexible Regression

### What is a Spline?

A spline is a piecewise polynomial function that is smooth at the join points (knots).

- We already saw piecewise linear models in Lecture 18
- Splines generalise this to higher-degree polynomials with smoothness constraints
- Common types: linear splines, cubic splines, natural cubic splines, B-splines

### The Key Conceptual Point

A spline model is still a linear model in its coefficients:

$$y = \beta_0 + \beta_1 b_1(x) + \beta_2 b_2(x) + \cdots + \beta_k b_k(x) + \varepsilon$$

The $b_j(x)$ terms are transformed versions of $x$. The model is nonlinear in $x$ because $x$ has been expanded into spline basis functions — but it is still linear in the $\beta$ coefficients.

### Splines with Wage Data

The relationship between `age` and `wage` is nonlinear and interpretable. Students can reason about the biology and social mechanisms without needing domain-specific knowledge.

```{r spline-wage-data, warning=FALSE, message=FALSE}
wage <- read_csv("../../data/wage.csv", show_col_types = FALSE)

wage |>
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE, colour = "#D55E00") +
  labs(
    x = "Age",
    y = "Wage",
    title = "A straight line is too rigid for this relationship"
  )
```

### Fitting Linear, B-Spline, and Natural Spline Models

```{r spline-wage-models, warning=FALSE, message=FALSE}
# Linear model
lm_linear <- lm(wage ~ age, data = wage)

# B-spline with 5 degrees of freedom
lm_bs <- lm(wage ~ bs(age, df = 5), data = wage)

# Natural spline with 5 degrees of freedom
lm_ns <- lm(wage ~ ns(age, df = 5), data = wage)

# Predict on a grid
age_grid <- tibble(age = seq(min(wage$age), max(wage$age), length.out = 200))

predictions <- age_grid |>
  mutate(
    linear = predict(lm_linear, newdata = age_grid),
    bs = predict(lm_bs, newdata = age_grid),
    natural_spline = predict(lm_ns, newdata = age_grid)
  ) |>
  pivot_longer(
    cols = c(linear, bs, natural_spline),
    names_to = "model",
    values_to = "fitted_wage"
  )

wage |>
  ggplot(aes(x = age, y = wage)) +
  geom_point(alpha = 0.12) +
  geom_line(
    data = predictions,
    aes(x = age, y = fitted_wage, linetype = model),
    linewidth = 1
  ) +
  labs(
    x = "Age",
    y = "Wage",
    linetype = "Model",
    title = "Splines allow flexible but controlled nonlinear regression"
  )
```

### What the Models Do

- The **ordinary linear model** has one slope across all ages. It cannot capture the rise and plateau pattern.
- The **B-spline model** allows the slope and curvature to change across age ranges. It follows the data more closely.
- The **natural spline** behaves like a spline in the middle of the data but becomes linear near the boundaries. That reduces unstable edge behaviour.

### Natural Splines vs Ordinary B-Splines

The key difference is **boundary behaviour**:

```{r spline-boundary, warning=FALSE, message=FALSE}
# Focus on the boundary region
predictions |>
  filter(model %in% c("bs", "natural_spline")) |>
  ggplot(aes(x = age, y = fitted_wage, colour = model)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Age",
    y = "Fitted wage",
    colour = "Model",
    title = "Natural splines constrain the curve to be linear at the boundaries"
  )
```

Natural splines are often a better default for regression because we rarely trust highly curved behaviour at the edges, where there are fewer observations.

### Slide Summary: Regression Splines

| Aspect | Description |
|--------|-------------|
| **Purpose** | Model nonlinear relationships inside a regression framework |
| **Mechanism** | Replace $x$ with spline basis functions |
| **Control** | Degrees of freedom or knots |
| **Strength** | Works naturally with `lm()`, prediction, and model comparison |
| **Limitation** | Boundary behaviour can be unstable |

### Slide Summary: Natural Splines

| Aspect | Description |
|--------|-------------|
| **Purpose** | A more stable spline model near the edges of the data |
| **Mechanism** | Constrain the fitted curve to be linear in the tails |
| **Control** | Degrees of freedom or knots |
| **Strength** | Often a sensible default for regression |
| **Limitation** | Still requires choosing the amount of flexibility |

---

## Example 3: Ridge and Lasso — Penalised Regression

### The Problem with Many Predictors

When we have many predictors (possibly more than observations), ordinary least squares:

- May have high variance
- May not have a unique solution (if $p > n$)
- Tends to overfit

### The Core Equations

**Ordinary least squares** chooses $\beta$ to minimise:

$$\sum_{i=1}^n (y_i - \hat{y}_i)^2$$

**Ridge regression** adds an $L_2$ penalty:

$$\hat{\beta}^{\text{ridge}} = \arg\min_\beta \left\{ \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^p \beta_j^2 \right\}$$

**Lasso** adds an $L_1$ penalty:

$$\hat{\beta}^{\text{lasso}} = \arg\min_\beta \left\{ \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{j=1}^p |\beta_j| \right\}$$

### Ridge vs Lasso with Bike Sharing Data

The bike sharing daily dataset is ideal because the predictors have intuitive redundancy: `temp` and `feels_like_temp` overlap strongly, `month` and `season` overlap, `weekday` and `workingday` partly overlap. This gives the penalty something real to do.

```{r ridge-lasso-bike-data, warning=FALSE, message=FALSE}
bike <- read_csv("../../data/bike_daily.csv", show_col_types = FALSE)

bike |>
  select(
    count,
    season,
    year,
    month,
    holiday,
    weekday,
    workingday,
    weather,
    temp,
    feels_like_temp,
    humidity,
    windspeed
  )
```

### Overlapping Predictors

```{r ridge-lasso-overlap, warning=FALSE, message=FALSE}
bike |>
  ggplot(aes(x = temp, y = feels_like_temp)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Temperature",
    y = "Feels-like temperature",
    title = "Some predictors carry overlapping information"
  )
```

### Fitting Ridge and Lasso

```{r ridge-lasso-fit, warning=FALSE, message=FALSE}
# Create model matrix
X <- model.matrix(count ~ . - 1, data = bike)
y <- bike$count

set.seed(42)

# Ridge (alpha = 0)
ridge_fit <- cv.glmnet(X, y, alpha = 0)

# Lasso (alpha = 1)
lasso_fit <- cv.glmnet(X, y, alpha = 1)

# Plot cross-validation curves
par(mfrow = c(1, 2))
plot(ridge_fit, main = "Ridge Regression")
plot(lasso_fit, main = "Lasso")
```

The plots show cross-validation error versus log(lambda). The vertical dashed lines indicate $\lambda_{\min}$ (optimal lambda) and $\lambda_{1\text{se}}$ (more conservative choice). Cross-validation chooses the penalty strength from the data, not by hand.

### Coefficient Comparison

```{r ridge-lasso-coefs, warning=FALSE, message=FALSE}
# Extract coefficients at lambda.min
ridge_coef_matrix <- coef(ridge_fit, s = "lambda.min") |>
  as.matrix()
ridge_coef <- tibble(
  term = rownames(ridge_coef_matrix),
  estimate = as.numeric(ridge_coef_matrix[, 1])
) |>
  filter(term != "(Intercept)") |>
  mutate(model = "Ridge")

lasso_coef_matrix <- coef(lasso_fit, s = "lambda.min") |>
  as.matrix()
lasso_coef <- tibble(
  term = rownames(lasso_coef_matrix),
  estimate = as.numeric(lasso_coef_matrix[, 1])
) |>
  filter(term != "(Intercept)") |>
  mutate(model = "Lasso")

bind_rows(ridge_coef, lasso_coef) |>
  mutate(term = fct_reorder(term, abs(estimate))) |>
  ggplot(aes(x = estimate, y = term)) +
  geom_point() +
  facet_wrap(~ model, scales = "free_x") +
  labs(
    x = "Coefficient estimate",
    y = NULL,
    title = "Ridge shrinks coefficients; lasso can remove terms"
  )
```

Notice the key difference:

- **Ridge** shrinks coefficients towards zero but keeps all predictors in the model
- **Lasso** shrinks some coefficients exactly to zero, performing automatic variable selection

```{r ridge-lasso-counts, warning=FALSE, message=FALSE}
cat("Ridge: number of non-zero coefficients:", 
    sum(coef(ridge_fit, s = "lambda.min") != 0), "\n")
cat("Lasso: number of non-zero coefficients:", 
    sum(coef(lasso_fit, s = "lambda.min") != 0), "\n")
```

### Ridge vs Lasso Summary

| Feature | Ridge | Lasso |
|---------|-------|-------|
| Penalty | $L_2$ (squared) | $L_1$ (absolute) |
| Coefficient shrinkage | Toward zero | Some to exactly zero |
| Variable selection | No | Yes |
| Correlated predictors | Shares coefficient | Picks one |
| When to use | Many small effects | Sparse true model |

### Slide Summary: Ridge

| Aspect | Description |
|--------|-------------|
| **Penalty** | $L_2$, $\sum \beta_j^2$ |
| **Main effect** | Shrinks coefficients towards zero |
| **Best use** | Many correlated predictors, each possibly carrying some information |
| **Interpretation** | Keeps predictors but reduces instability |

### Slide Summary: Lasso

| Aspect | Description |
|--------|-------------|
| **Penalty** | $L_1$, $\sum |\beta_j|$ |
| **Main effect** | Shrinks some coefficients exactly to zero |
| **Best use** | Many candidate predictors, only some likely to matter |
| **Interpretation** | Regularisation plus variable selection |

---

## Synthesis: All Methods Control Flexibility

| Method | Controls flexibility by... |
|--------|---------------------------|
| **LOESS** | Adjusting the span — how local the fit is |
| **Splines** | Changing the basis used for $x$ (degrees of freedom, knots) |
| **Ridge** | Penalising the $L_2$ size of coefficients ($\lambda$) |
| **Lasso** | Penalising the $L_1$ size of coefficients ($\lambda$) |

Cross-validation gives a practical way to choose how much flexibility to allow for all these methods.

### Summary

- **LOESS** provides flexible, assumption-light smoothing through local weighted regression — excellent for exploratory visualisation
- **Splines** extend polynomial regression with piecewise fits and smoothness constraints, keeping the benefits of a formal regression model
- **Ridge regression** shrinks coefficients to handle multicollinearity and high-dimensional data
- **Lasso** performs automatic variable selection by shrinking some coefficients to exactly zero

These methods bridge the gap between rigid parametric models and the need for flexible, data-driven modelling.
