## Regression Models for Trend and Seasonality

In the previous lecture, a simple seasonal indicator was enough for the ski-sales data. In other situations, the pattern across the year is smoother and more complicated. We want a model that is more flexible than a single winter indicator, but more parsimonious than fitting a separate effect for every month.

By the end of this lecture, students should be able to:

1. Use factor indicators to represent seasonality.
2. Build sine and cosine predictors for periodic behaviour.
3. Explain why sine-plus-cosine allows an unknown phase shift.
4. Recognise Fourier terms as a systematic version of the same idea.

## Electricity Bills: Month Effects and Seasonal Shape

We consider monthly electricity bills for a household from May 1994 to May 1998. During some months there were paying guests staying in part of the house, so the householder wanted to estimate the extra business-related electricity cost.

After allowing for the number of days in the bill, whether the bill was estimated, the guest weeks, and a time trend, we then examine what seasonal structure remains.

`r xfun::embed_file("../../data/Electricity.csv")`

```{r read-electricity}
Electricity <- read.csv("../../data/Electricity.csv", header = TRUE)
Electricity |> head() |> kable()
```

### Residual Seasonality After Basic Adjustment

```{r electricity-initial-model}
Electricity_lm1 <- lm(Bill ~ Days + Estimate + Guestwks + Time, data = Electricity)
resids <- resid(Electricity_lm1)

plot(resids ~ Time, data = Electricity, ylab = "Residuals")
lines(resids ~ Time, data = Electricity, col = 4)
abline(h = 0, lty = 2)
```

The plot suggests that some months tend to lie above zero and others below zero, so there is still seasonal structure after adjusting for the non-weather-related variables.

### A Separate Effect for Each Month

```{r electricity-month-factor}
Electricity_lm2 <- lm(resids ~ factor(Month), data = Electricity)
summary(Electricity_lm2)
```

This shows that the average residual differs by month. That is useful diagnostically, but using eleven indicator variables for a short series is not very parsimonious.

## Sinusoidal Terms

A smoother alternative is to represent seasonality by sine and cosine waves. Suppose we try
$$\sin\left(\frac{2\pi (t - c)}{12}\right)$$
for monthly data, where the shift constant $c$ says where the cycle starts.

An initial guess is to place the zero point near April, because that is close to autumn and might correspond to an "average" month for electricity use.

So we create
$$\mathrm{SinTime} = \sin\left(\frac{2\pi (Time - 4)}{12}\right), \qquad
\mathrm{CosTime} = \cos\left(\frac{2\pi (Time - 4)}{12}\right).$$

The cosine term matters because we may not know the best phase shift in advance. If the true seasonal pattern is
$$A\sin(x + \phi),$$
then trigonometry gives
$$A\sin(x + \phi) = A\sin(x)\cos(\phi) + A\cos(x)\sin(\phi)
= A_1 \sin(x) + A_2 \cos(x).$$

So a regression on both sine and cosine can estimate the amplitude and the phase shift automatically.

### First Guess: Shift by Four Months

```{r electricity-first-sinusoid}
Electricity <- Electricity |>
  mutate(
    SinTime = sin((Time - 4) * 2 * pi / 12),
    CosTime = cos((Time - 4) * 2 * pi / 12)
  )

Electricity_lm3 <- lm(resids ~ SinTime + CosTime, data = Electricity)
summary(Electricity_lm3)
```

```{r electricity-first-sinusoid-plot}
plot(resids ~ Time, data = Electricity, ylab = "Residuals")
lines(resids ~ Time, data = Electricity, col = 4)
lines(Electricity_lm3$fitted.values ~ Time, data = Electricity, col = 2, lwd = 2, lty = 2)
abline(h = 0, lty = 2)
```

The coefficient of `CosTime` is clearly not zero, which tells us our guessed phase shift is not quite right.

### Second Guess: Shift by Five Months

The fitted curve crosses zero closer to `Time = 5`, so we try that shift instead.

```{r electricity-second-sinusoid}
Electricity <- Electricity |>
  mutate(
    SinTime = sin((Time - 5) * 2 * pi / 12),
    CosTime = cos((Time - 5) * 2 * pi / 12)
  )

Electricity_lm4 <- lm(resids ~ SinTime + CosTime, data = Electricity)
summary(Electricity_lm4)
```

Now the cosine term is no longer doing much work, which suggests the revised phase shift is much more sensible.

### A Sharper Seasonal Shape: Add Second Harmonics

The fitted curve still looks a little too smooth. In particular, the rise toward winter is steeper than a simple sine wave suggests, so we add terms with twice the frequency:
$$\sin\left(\frac{4\pi (Time - 5)}{12}\right), \qquad
\cos\left(\frac{4\pi (Time - 5)}{12}\right).$$

```{r electricity-second-harmonic}
Electricity <- Electricity |>
  mutate(
    Sin2Time = sin((Time - 5) * 4 * pi / 12),
    Cos2Time = cos((Time - 5) * 4 * pi / 12)
  )

Electricity_lm5 <- lm(resids ~ SinTime + CosTime + Sin2Time + Cos2Time, data = Electricity)
summary(Electricity_lm5)
anova(Electricity_lm4, Electricity_lm5)
```

```{r electricity-second-harmonic-plot}
plot(resids ~ Time, data = Electricity, ylab = "Residuals")
lines(resids ~ Time, data = Electricity, col = 4)
lines(Electricity_lm5$fitted.values ~ Time, data = Electricity, col = 2, lwd = 2, lty = 2)
abline(h = 0, lty = 2)
```

The extra sine and cosine pair gives a noticeably more flexible shape. It captures a steeper winter peak and a flatter period through part of the year without requiring a separate coefficient for each month.

### Putting the Model Together

```{r electricity-full-model}
Electricity_lm6 <- lm(
  Bill ~ Days + Estimate + Guestwks + Time + SinTime + CosTime + Sin2Time + Cos2Time,
  data = Electricity
)
summary(Electricity_lm6)

plot(Bill ~ Time, data = Electricity)
lines(Bill ~ Time, data = Electricity, col = 4)
lines(Electricity_lm6$fitted.values ~ Time, data = Electricity, col = 2, lwd = 2, lty = 2)
```

This final model is still a regression model. The main difference is that the seasonal predictors are now smooth waves rather than month indicators.

## Fourier Terms in a Modern Example

The Seoul bike data use the same idea, but on a daily time scale. The sine and cosine terms are usually written more systematically as **Fourier terms**.

`r xfun::embed_file("../../data/seoul_bike_daily.csv")`

```{r read-seoul-daily-trend-seasonality, message=FALSE}
seoul_daily <- read_csv("../../data/seoul_bike_daily.csv", show_col_types = FALSE) |>
  mutate(
    S1 = sin(2 * pi * day_index / 365),
    C1 = cos(2 * pi * day_index / 365),
    S2 = sin(4 * pi * day_index / 365),
    C2 = cos(4 * pi * day_index / 365)
  )

seoul_daily |> head() |> kable()
```

```{r seoul-fourier-model}
seoul_fourier <- lm(
  rented_bike_count ~ day_index + S1 + C1 + S2 + C2 +
    mean_temperature + total_rainfall + total_solar_radiation,
  data = seoul_daily
)
summary(seoul_fourier)
```

```{r seoul-fourier-plot}
seoul_daily |>
  mutate(fitted = fitted(seoul_fourier)) |>
  ggplot(aes(x = day_index, y = rented_bike_count)) +
  geom_line(alpha = 0.6) +
  geom_line(aes(y = fitted), colour = "red", linewidth = 1) +
  labs(
    x = "Day index",
    y = "Daily rented bike count",
    title = "Fourier terms give a smooth seasonal fit for bike demand"
  )
```

The electricity example introduced the ideas in a small, interpretable setting. The Seoul bike model shows the same method on a larger dataset, where the sine and cosine pairs become a compact way to represent annual seasonality.

## Looking Ahead

So far we have assumed that once the mean structure is well chosen, the residuals behave independently. The next lecture asks whether that is actually true.
